### 2014-08-23 standing-variation-paper.tex
initializing repro

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{green}(#1)} }
\newcommand{\plr}[1]{{\it\color{blue}(#1)}}

\newcommand{\var}{\mathop{\mbox{Var}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

\bibliographystyle{plain}

\begin{document}

\section{Introduction}

In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift. In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel. This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
Recall questions of origin of adaptations

References on parallel adapation, standing variation, etc.:
  coop,
  pennings and hermisson.
Note pennings and hermisson found standing deleterious variation was very important.

recall peromyscus examples with patchy environments

Summarize previous paper

Question assumptions of no standing variation and homogenous selection

\section{General model}

outline here assumptions common to both models.

Stick to constant-speed waves.
Note monty's empircal speeds paper.

\section{Standing variation}

\subsection{The set-up}

% selection pressure changes at $t=0$ from $1-s_d$ to $1+s_b$ -- homogeneous

We assume that the species range
is a homogenous, one- or two-dimensional region. 
There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
We assume that different mutations are distinguishable --
either as selectively equivalent mutations, or by linked variation.
For a long time in the past, the mutated type has been at a selective disadvantage (long enough to be at selection-mutation equilibrium),
but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
After fixation, alleles are either descended
from families of mutants present as standing variation when the selective regime changed,
or from new mutants arising since that time.

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the haploid fitness
\gc{heterozygous fintess, with no reference to the fitness of the
  homozygote for the mutant alleles}.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

% Branching-process-sized point mass approx to local groups present at $t=0$
% -- depends on local population density large enough relative to migration (to be "branching")
% -- and migration not too large (since we approximate as a point mass).
% Find mean density of clusters of successful standing mutants. 
% Is greater-than-one size of clusters likely to make a difference?

\paragraph{Rates of origination of standing and new mutations}
We make use of the commonly used approximation that neglects competition between relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a branching process.
After $t=0$, the offspring of a single mutant form (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
As in \cite{ralphcoop2010}, each new offspring has a very small probability of being a mutant and establishing locally,
so the collections of times and locations at which mutants appear and establish locally 
is well--approximated by a Poisson process in space and time.
This the rate of this Poisson process, 
the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type
is approximately $\lambda = 2 \mu \rho s_b$.

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendents of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendents are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}

\plr{XXX More discussion on whether this is a good approximation?  
Compute typical size and radius of a standing cluster?}
\gc{Point out connection to standard mutation selection balance formula}

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}.

If the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance) then the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$.

\plr{recall PPP computations; draw cones of opportunity for new mutations }

\paragraph{}
So we can put these ingredients together for a simple model of the
geographic spread of alleles. A cartoon example of this process is
shown in Figure \ref{fig:cartoon} Initially, when the selection
pressures change at $t=0$, a set of standing variants can start to
spread having escaped loss through drift. These variants are shown as
little lightning bolts, and occur at a density $\lambda_0$ across
space. 
They spread at velocity $v$ carving out cones in space-time. As these
alleles proceed in their geographic spread other new alleles can arise
and become established in parallel, these are shown as little
stars. These new mutations arise and become established at rate $\lambda$.

The wave of advance works in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). As such
our framework is applicable here.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{spreading_alleles_trimmed}
  \end{center}
  \caption{
Cartoon of the geographic spread of standing and new mutations. 
}
  \label{fig:cartoon}

\end{figure}

\subsection{G6PD}

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Sharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd ,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd, Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd
  Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their prescence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to various compounds, notably
those present in fava beans.

% G6PD frequency across world http://www.google.com/imgres?imgurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%3Adoi%2F10.1371%2Fjournal.pmed.1001339.g002%2Flargerimage&imgrefurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%253Adoi%252F10.1371%252Fjournal.pmed.1001339&h=2938&w=2381&tbnid=oD_tX3ea8IbLfM%3A&zoom=1&docid=k37axqAZDS7ljM&ei=5rdOU5TVDMqj8QGntICYBg&tbm=isch&ved=0CF8QMygLMAs&iact=rc&uact=3&dur=1447&page=1&start=0&ndsp=15

% Spatial distribution of G6PD alleles:  http://www.malariajournal.com/content/12/1/418

%Hedrick sex-specific review http://www.eeslmu.de/eeswiki/images/Hedrick_Review.pdf
% Fry: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3654548/
% kidwell http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1213615/
%Rice: http://www.jstor.org/discover/10.2307/2408385?uid=3739560&uid=2129&uid=2&uid=70&uid=4&uid=3739256&sid=21103992458511

%IDEA: could we look at all freq. reports of g6pd and see if male and
%female freq. diffs support model of sexual-conflict?

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants}. So the average width of
an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km.

G6pd is 515 codons long, with 140 variants reported in it REF. Assuming a
mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
$\mu \approx 1 \times 10^{-6}$ a generation. 
The dispersal and demographic parameters of humans is unclear,
particularly as we are concerned with the effective population density.
We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
We use $\sigma=10,~50,$ and $100$ as our dispersal parameter.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ %
{\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
  the frequency of G6pd-deficiency alleles. Size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
  \url{http://www.malariajournal.com/content/12/1/418}. Countries with
  endemic malaria are colored yellow. Figure taken from \citep{Howes-g6pd-variants}.

}
\end{center}
\end{figure}

% \begin{figure}[ht]
% \begin{center}
%   \includegraphics[width=1.0\textwidth]{G6PD_charlengths}
% \caption{ %
% {\bf Characteristic length,} as a function of dispersal distance $\sigma$ and selective disadvantage $s_d$.  XXX at what parameters. XXX
% } \label{Fig-G6PD-charlength}
% \end{center}
% \end{figure}

%$A-$ ?? risk lower of severe malaria for both female heterozygotes and male
 %hemizygotes estimates of selection pressure \citep {Ruwende-g6pd}
%Mahidol487A \citep{Louicharoen-g6pd}
%Show's effect in het and homozy females

%Selection coefficient (s) 	A- 0.044 (0.019–0.048)  Med	0.034
%(0.014–0.049)  ``The mean age of the A− allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years. For the Med alleles, the mean age was 3330 years with a 95\% credibility interval of 1600 to 6640 years"
%\citet{tishkoff-g6pd}
%s 0.24 and t= 40 generations \citep{Slatkin-age-est}
%The estimated age is ~100 generations with an upper bound of <150
%generations A- 0.1 < s < 0.2 \citep{Saunders-g6pd}
%Mahidol487A mutation started to increase at ~1500 YBP, with a
%selection intensity of ~0.23 \citep{Louicharoen-g6pd}
%Show's effect in het and homozy females

\subsection{Characteristic parameters}

% a characteristic length of regions and a proportion of haplotypes (or of space?) that are from standing variation

In \citet{ralphcoop2010}, studying the model without standing variation,
we defined a {\em characteristic length} which gave the spatial scale across with mutants with distinct origins would establish.
This was proportional to the mean distance between neighboring established mutants,
but had the advantage of being easier to calculate.
Furthermore, the time scale over which adaptation occurred could be found by dividing the characteristic length 
by the speed at which the mutants spread.
We will define the characteristic length for this new model,
as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
%(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)

Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
As it spreads, it will cover a distance $L$ in time $L/v$.
The number of other mutations appearing in the circle it has covered up until this time is Poisson with mean
$\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
(and $\lambda_0 2 L + \lambda \pi L^2 /v$ in one dimension).
% $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
Therefore, if we define $\chi$ to be the unique positive solution to
\[
    \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
\]
then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
before it is expected that one other successful mutation would have arisen in the area covered so far.
The explicit formula for $\chi$ is cumbersome; here we omit it.
(This is in two dimensions; in one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
This can be rewritten 
\begin{equation} \label{eqn:defines_chi}
%    \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^2 + \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^3 
%        = \frac{ - \log(1-s_d)^3 }{ 4 s_b^2 \sigma^2 \pi \rho \mu } .
   \frac{\sqrt{2s_b} }{-\log(1-s_d) } \chi^2 + \frac{1}{\sigma} \chi^3 = \frac{1}{\sqrt{2s_b} \rho\mu\pi}
\end{equation}
From this we see that $\chi$ decreases with $\rho$ and $\mu$.
Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
$\chi = \sqrt( -\log(s_d) / 2 s_b \rho \mu \pi ) + O(1/\sigma)$.
On the other hand, if the mutant allele is highly deleterious before $t=0$,
then the characteristic length is close to the value from \citet{ralphcoop2010}:
$\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d))$.

XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi/\sqrt{\pi \lambda_0})^2$ then
$f = 1-\gamma f^{3/2}$, where
$\gamma = (4 \pi^{5/2} / \sqrt{2}) \mu^2 s_b^{3/2} \rho^2 / \sigma \log(1/(1-s_d))$.
Then if $\gamma \ll 1$, $f \approx 1$, while if $\gamma \gg 1$, then $f \approx 0$.

By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
and occur within the first $\chi/v$ generations.
Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
we expect to find one mutational origin;
we denote the probability that this mutation is from standing variation by $\pi_0$.
Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
and it is an elementary consequence of the Poisson process that
\begin{align} \label{eqn:pizero}
    \pi_0 = \lambda_0 \pi \chi^2 .
\end{align}
\marginnote{put in value of $\chi$ computed above.}
(In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)

In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
calculated for our G6pd parameters.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
} \label{Fig-G6PD-charlength}
\end{center}
\end{figure}

% \begin{figure}[ht]
% \begin{center}
%   \includegraphics[width=1.0\textwidth]{charlen-by-sd-sigma}
% \caption{ %
% {\bf Characteristic length,} as a function of dispersal distance $\sigma$ and selective disadvantage $s_d$.  XXX at what parameters. XXX
% }
% \end{center}
% \end{figure}

\subsection{Standing Variation -- Results}

The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

In Figure \ref{G6PD_chartimes}

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_chartimes}   %{charlen-by-sd-limit}
\caption{ %
{\bf Mean adaptation times,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
} \label{G6PD_chartimes}
\end{center}
\end{figure}

% proportion of haplotypes (and of space) that are from standing variation

We have defined $\lambda_0$ to be the mean density of standing variants that reach local fixation.
Define $\nu_+$ to be the mean density of new mutants whose offspring fix locally.
Since the probability that a mutant arising at $(x,t)$ is lucky enough to be born in a location not already occupied by mutants
is $\P\{ \tau > t \}$,
we can see  $\nu_+ = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
$\nu_+ = \lambda \E[\tau] $.
Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
\begin{equation}
    \nu_+ = (\mbox{mean density of new patches}) = \frac{1}{1-\log(1-s_d) \E[\tau]} .
\end{equation}
Since the mean proportion of patches that come from standing variation is $\lambda_0 / (\lambda_0 + \nu_+)$,
this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
There are $\lambda_0 + \nu_+$ patches per unit area, so
the typical patch (i.e.\ with distribution given by the Palm measure) occupies area $1/(\lambda_0 + \nu_+)$.

We can also find the mean proportion of space covered by standing variants.
If by time $t$ a geographic location has not yet been reached by the mutation,
the probability it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$, 
and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
Therefore, as is standard for competing exponentials,
the probability a given location is reached first by a standing variant,
and therefore the mean proportion of space covered  by standing variants,
is
\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt .
\end{equation}
\marginnote{put in this value in terms of Gaussian CDF?}
To evaluate this integral, again see Appendix \ref{apx:integrals}.

Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
i.e.\ $a_0 = z_0 / \lambda_0$.
We can compare this to the corresponding quantity $a_+$ for a new variant through the formula
$a_0 / a_+ = z_0 / (1-z_0)$.

In Figure \ref{G6PD_standing_var_proportion}

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
\caption{ %
{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the sickle-cell example.  XXX say more.
} \label{G6PD_standing_var_proportion}
\end{center}
\end{figure}

\marginnote{do mean (and variance?) of distance between haplotypes?}

influence of migration on things -- plot these versus migration

{\tt Simulations:} compare mean density against $s_d$ at different values of $\sigma$; omit variability if needed for readability.
Compare to panmictic results.

\subsubsection{Multiple variant types}

One interesting problem that we can seek to address with this work is
the extent to which

Lets imagine for the moment that all classes of beneficial allele have
the same beneficial selection coefficient. 
Each class of mutations could have its own mutation rate $\mu_j$ and
have a disadvantage $s_{d,j}$ prior to the switch. As they have the
same selection coefficient all of the waves travel outward at a rate
$v$.

\begin{align}
    \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } ; ~~      \lambda_{j} = \mu_j \rho s_b;
\end{align}

\begin{equation}
    z_0 = \int_0^\infty {\lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3} \; \exp \left( - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
\end{equation}

\section{Local establishment and comparison to panmixia}

In the above (and in \citet{ralphcoop2010}) we have assumed that once the mutation appears,
conditional on eventual fixation, it begins to spread spatially at speed $v$ instantly,
effectively neglecting the time it must first spend fixing locally.
In \citet{ralphcoop2010} we addressed this by noting that there would be no change at all in our results 
if all mutations had to wait the same amount of time before fixing locally,
and that this time was short relative to the time it took the wave to spread across the characteristic length;
we then showed via simulation that this was reasonable in certain situations.
In this section we examine this assumption in more detail, although mostly through heuristic arguments,
and also compare our results above to those of \citet{softsweeps}.

We are assuming that shortly after a new mutation appears, 
it can be approximated by a branching process growing at rate $s$
until the point that it grows large enough to ``feel'' spatial structure,
at which point it begins to spread as a more--or--less deterministic wave.
Although we are not aware of good analysis of this transition, 
the relevant size when spatial structure becomes important
should be something close to $\sigma^2 \rho$ 
(i.e.\ Wright's ``local effective population size'' \citep{XXX}).
Let $Z_t$ be a continuous-time branching process with $Z_0=1$ and $\E[Z_t] = e^{s_b t}$.
Then we know that there exists a random variable $W$ such that $\lim_{t\to\infty} e^{-s_b t} Z_t = W$ almost surely,
so that if $\tau$ is the time $Z_t$ reaches size $\sigma^2 \rho$, i.e.\ $\tau = \inf\{ t \ge 0: Z_t > \sigma^2 \rho$,
then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$.
From this we know that $\tau \approx (1/s_b) (\log (\sigma^2 \rho) - \log W)$;
although more detailed information is available (e.g.\ a CLT for $\tau$, \citet{XXX}),
we will stick to the loose interpretation.
\marginnote{check jaegers.  isn't there a CLT for $\tau$?}
As $\sigma^2 \rho$ increases, the mean of $\tau$ increases, but the variance approaches a constant
($\var[\tau] \to (1/s_b) \var[\log W]$).

So, roughly speaking, we need to evaluate the importance of a delay of about $T = (1/s_b) \log (\sigma^2 \rho)$.
New mutations will appear and fix during this time if $4N\mu s_b \ge 1/T$,
i.e.\ if $4 N \mu \ge 1/\log (\sigma^2 \rho)$.
Our model will still be a good approximation, however, 
as long as $T$ is short relative to the time a wave takes to spread between nearby mutational origins.
This can be worked out,
but it is simpler to note that if the converse is true, 
that the process is largely unaffected by spatial structure,
and so the panmictic model is a good approximation for the true process.
Since both our spatial model and the panmictic model underestimate the true degree of parallel adaptation,
it suffices to compare results from the two models to avoid applying the wrong model.

% When would we expect more mutations in this time?
% Roughly, if populations are larger than $C/\mu$ for some $C \approx 100$.
% Compare to panmixia: increases degree of parallel mutation if... ?

\citet{softsweepsII}, it is shown that under a panmictic model with certain assumptions on the parameters,
the number of independent origins due to both standing variation and new mutation seen in a sample of size $n$
has approximately the Ewens distribution with parameters $n$ and $\theta = 4 N \mu$.
As $n$ increases, the total number of types seen grows as $\log n$, 
so the ``mean density of types'' corresponding to our $\lambda_0 + \nu_+$ in their model is infinite.
\marginnote{Graph: sigma against mean number of origins; different curves for spatial, panmictic (flat), and simulation;
different values of theta as different line types?  but how to do this?
Mention Monty's empirical speeds paper.}
We could compare the probability of parallel adaptation under the two models (see eq'n 18 in \citet{softsweepsII}),
but this seems less useful than an assessment of the degree of parallel adaptation.

% \section{Applications}

% The FRIGIDA (FRI) in {\it  Arabidopsis thaliana} is a major
% determinant of genetic variation in flowering time. Recessive loss of function
% alleles at FRI result in early flowering, due to the almost complete
% loss the of the requirement of vernalization (a prolonged chill).  
% This is thought to potentially confer an advantage due to allowing a
% more rapid cycling life history.

% Sweeps of FRI loss of function alleles \citep{Hagenblad-FRIsweep,Toomajian-FRIsweep}

% 20 independent nonfunctional FRI haplotypes \citep{Shindo-FRI}

% open reading frame (ORF) of 609 amino acids

\section{Discussion}

Thoughts
species coherence.
The Biological Reality of Species: Gene Flow, Selection, and Collective Evolution

\bibliography{standing_patches_refs}

\appendix

%%%%%%%%%%%%
\section{Integrals appearing in the text}
    \label{apx:integrals}

In the text at several points appear integrals of the form
\begin{align}
  \int_0^\infty t^c \exp \left( - \alpha t^d - \beta t^{d+1} \right) dt 
\end{align}
where $c$ is a positive integer, $d$ is the dimension, and $\alpha$ and $\beta$ are positive real numbers.
This could be evaluated through standard numerical methods; below we describe a power series expansion.
Changing variables to $u = \beta t^{d+1}$, this becomes
\begin{align}
    \left( (d+1)^{-1} \beta^{ (1-c)/(d+1) } \right) \int_0^\infty u^{(c+1-d)/(d+1)} \exp\left( - \alpha \beta^{-d/(d+1)} u^{d/(d+1)} - u \right) du ,
\end{align}
so it suffices to evaluate the function
\begin{equation}
    G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt ,
\end{equation}
in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic paramters.
Since $a$ and $b$ only depend on the dimension and the quantity being computed,
we are interested in $G$ as a function of $x$.
At least in the case $b=d/(d+1)$ it is possible to express $G$ as a finite sum of gamma functions,
but we proceed with a simpler method.
Note that $\partial_x G(a,b,x) = -G(a+b,b,x)$,
and that at $x=0$, the function $G$ is the gamma function $G(a,b,0) = \Gamma(a+1)$.
Therefore, a Taylor series for $G$ would be
\[
    G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
\]
It is easy to check using Stirling's formula that $\limsup_{n \to \infty} ( x^n \Gamma(a+nb+1)/n! )^{1/n} = 0$
if $b<1$, so the sum converges.

\end{document}

### 2014-08-23 g6pd_figures.R
adding more files

@@ -1,0 +1,106 @@
+
+source('~/Dropbox/postdocs/Peter/Spatial_adaptation/Spatial_adaptation/standing-variation-fns.R')
+#install.packages("wesanderson")
+library("wesanderson")
+
+my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
+
+ mu=1e-6; rhos=c(0.2,2); sb=0.05; sdvals=seq(0.0001,0.3,length=800); 
+
+sigmas=c(100,50,10)
+
+layout(t(1:2))
+par(mar=c(4,4.1,2,0.3))
+for(j in 1:2){
+	rho=rhos[j]
+	if(j == 1) plot(sdvals,sdvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab=expression(chi),cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
+	if(j == 2) plot(sdvals,sdvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab="",cex.lab=1.5,cex.axis=1.5,main=expression(rho==2))
+	
+	for(i in 1:3){
+	
+	sigma=sigmas[i]; charlength.sd<-sapply(sdvals, function (sd) { charLength(mu,rho,sb,sd,sigma)$value } )
+	 lines(sdvals,charlength.sd,lwd=2,col=my.col[i])
+	 abline(h= newCharLength(mu, rho, sb, sd=NA, sigma)$value,col=my.col[i],lwd=2,lty=2)  ##new mut. only
+	}
+	
+	 ##contribution of standing var only, i.e. old variation
+	 charlength.standing.only<-sapply(sdvals, function (sd) {  oldCharLength(mu, rho, sb, sd, sigma)$value})
+	 lines(sdvals,charlength.standing.only ,lwd=2,col="black",lty=3)
+	  if(j == 2) legend("topright",col=c("grey","grey","black",my.col[1:3]),lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
+	 	lapply(sigmas , function(x){  {substitute( sigma==myS, list(myS=x) )} }))))
+ 	}
+ 	
+dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlengths.pdf")
+ 
[…]

### 2014-08-23 standing-variation-fns.R
adding more files

@@ -1,0 +1,139 @@
+#
+# Implement theory from standing variation parts
+
+# mu is mutation rate
+# rho is population density
+# sigma is mean squared dispersal distance
+# sb >0 is advantage of mutation for t>0
+# sd >0 is disadvantage of mutation for t<0
+# 
+# ... and
+# lambda <- 2 * mu * rho * sb    # is rate of new successful mutations
+# lambda0 <- lambda / log(1/(1-sd))         # is approximate rate of standing successful mutations
+# v <- sigma * sqrt(2*sb)        # speed of wave of advance
+
+everything <- function (mu, rho, sb, sd, sigma) {
+    # bundle up everything
+        return( c( 
+            c( z0 = standingProportionArea(mu,rho,sb,sd,sigma)$value,
+                chi = charLength(mu,rho,sb,sd,sigma)$value,
+                Etau = meanTime(mu,rho,sb,sd,sigma)$value),
+            c( mu=mu, rho=rho, sb=sb, sd=sd, sigma=sigma )
+        ) )
+}
+
+meanTime <- function (mu, rho, sb, sd, sigma,include.new=TRUE) {
+    # mean time til adaptation of a point;
+    # E[\tau] is int_0^\infty exp(-lambda0 pi v^2 t^2 - lambda pi v^2 t^3) dt
+    if (sd==0) {
+        return( list(value=0) )
+    } else if (sd==1) {  ## no standing var.
+        f <- function (t) {
+        	stopifnot(include.new)
+            exp( - 4 * mu * rho * sb^2 * sigma^2 * t^3 / 3) 
+        }
+    } else { 
+        f <- function (t) {
+            exp( - 4 * mu * rho * sb^2 * sigma^2 * t^2 * (1/log(1/(1-sd)) + include.new*t/3) )
+        }   
+    }
+
+    xx <- 2^(-30:30)
+    yy <- sapply(xx, function(x) { f(x) } )
+    scale <- xx[ which.min( abs(10-yy*xx) ) ]
+    return( integrate(function(x) { f(x*scale)*scale }, 0, Inf) )
+}
+
+
+
+
+meanNewDensity <- function (mu, rho, sb, sd, sigma) {
+    # mean density of patches arising from new mutations
+    #  is lambda * E[\tau]
+    x <- meanTime(mu, rho, sb, sd, sigma)
+    x$value <- x$value * 2 * mu * rho * sb 
[…]

### 2014-08-25 standing-variation-paper.tex
adding more files
wrote part of discussion

@@
 
-References on parallel adapation, standing variation, etc.:
+References on parallel adaptation, standing variation, etc.:
   coop,
@@
 Stick to constant-speed waves.
-Note monty's empircal speeds paper.
+Note monty's empirical speeds paper.
 
@@
 important early dynamics are determined by the haploid fitness
-\gc{heterozygous fintess, with no reference to the fitness of the
+\gc{heterozygous fitness, with no reference to the fitness of the
   homozygote for the mutant alleles}.
@@
 
-% Branching-process-sized point mass approx to local groups present at $t=0$
+% Branching-process-sized point mass approx. to local groups present at $t=0$
 % -- depends on local population density large enough relative to migration (to be "branching")
@@
 Before $t=0$, on the other hand, the mutation is deleterious, so the
-genetic descendents of each new mutation are (with high probability)
+genetic descendants of each new mutation are (with high probability)
 doomed (doomed I say!) to extinction,
@@
 Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
-the locations of all mutant families extant at $t=0$ whose descendents are destined to fix locally is also,
+the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
 by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
@@
 
-The wave of advance works in the case of heterozygote advantage
-\citep{aronson1975nonlinear}, with the selected allele spreading
-locally to the equilibrium frequency (rather than fixation). As such
-our framework is applicable here.
-
 \begin{figure}[ht]
@@
 studied: the $A-$ found in
-much of sub-Sharan Africa; the Med allele found in the Mediterranean and middle
+much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
 east; and the Mahidol allele found Myanmar and Thailand.
@@
 Given such a strong pressure on these alleles they should have risen
-quickly to fixation, so their prescence at intermediate frequency,
+quickly to fixation, so their presence at intermediate frequency,
 over a broad geographic area, makes it a good candidate for a recently
@@
 Hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to various compounds, notably
+leading to hemolytic anemia when exposed to a variety of compounds, notably
 those present in fava beans.
+
+The wave of advance works in the case of heterozygote advantage
+\citep{aronson1975nonlinear}, with the selected allele spreading
+locally to the equilibrium frequency (rather than fixation). As such
+our framework is applicable here.
 
@@
 {\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
-  the frequency of G6pd-deficiency alleles. Size of the pie chart
+  the frequency of G6pd-deficiency alleles. The size of the pie chart
   indicates the number of G6pd-deficient individuals sampled.
-  \url{http://www.malariajournal.com/content/12/1/418}. Countries with
-  endemic malaria are colored yellow. Figure taken from \citep{Howes-g6pd-variants}.
+. Countries with
+  endemic malaria are colored yellow. Figure taken from
+  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.
 
@@
 
-Thoughts
-species coherence.
-The Biological Reality of Species: Gene Flow, Selection, and Collective Evolution
+Confusing convergence for local adaptation. Mention skin pigmentation?
+
+Our results are predicated on the idea that adaptive variants are
+initially rare within populations, i.e. they are reasonably
+deleterious before the environment switches. In constrast, even distant populations could have
+a shared basis of adaptation if they adapt via common variation,
+e.g. previously neutral (or nearly neutral) variation, shared among populations. 
+Our ability to detect such shared convergent events is at the moment
+quite limited. 
+One way that selection pressures might be weak on alleles, is if they
+are at only one of many loci contributing to variation in a trait. However, if that
+case populations separated by a reasonable amount of drift may still
+adapt via different genetic routes, as the constellation of alleles
+best positioned to initially respond to positive selection, i.e. those at
+intermediate frequency, will be somewhat different.
+
+\paragraph{}
+Our results touch on an old debate on the evolutionary coherence of
+species. 
+Mayr and others have argued that species are coherent \citep{} 
+as they are united by shared gene flow. However, this arguement has
+questioned by a number of authors based on relatively high levels of 
+differentiation in many species \citep{}.
+\citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
+may not be high enough to prevent neutral differentiation, species
+should be viewed as being united as levels of gene flow were still
+high enough for selected alleles to spread across the entire species.
+Our work on geographic convergent adaptation
+\citep[see also][]{ralph2010parallel,RalphCoop:14} brings these conclusions partially into
+question. Our results suggest that wide-spread species should often
+repeatedly adapt in parallel rather than waiting for a single allele
+to migrate across the range.
+
+In support of this idea recent selective sweeps are often
+geographically restricted \citep{}, but see XXX. This is likely in part due to the
+relatively low occurance of species-wide selection pressures \citep{},
+but as noted above even when we know of wide-spread selection
+pressures (e.g. Malaria) the response is usually convergent across large enough
+spatial scales. This view is potentially contradicted by the
+introgression of adaptive alleles across species and sub-species
+boundaries, suggesting that selected alleles do sometimes spread
+despite low migration rates \citep{}. However, at least some of
+these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
+alleles.
+
+% \gc{Needed?:} This view is potentially contradicted by the
+% introgression of adaptive alleles across species and sub-species
+% boundaries, suggesting that selected alleles do sometimes spread
+% despute low migration rates \citep{}. However, at least some of
+% these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
+% alleles. Other cases of selective introgression
+% are concurrent with populations spreading into new environments,
+% and coming in contact with preadapted populations, potentially
+% lessening the chance of convergent adaptation \citep{}.
+
+If gene flow rarely rapidly spread selected alleles across
+species range, how then do selected alleles ever become fixed within species? 
+Repeated bouts of adaptation in particular genomic regions may act to slowly
+push a subset of previously selected alleles to fixation across the
+species range, through the spread of the genetic backgrounds on which they arise.  
+However, it seems likely that this is a slow process compared to the
+initial rapid spread of selected alleles.
+
+One potential hypothesis is that many selected alleles only achieve
+fixation, not through their own species-wide spread, but rather
+through subsequent large-scale
+changes in geographic range size induced by extirpation of the species
+over part of its range. Such drops in range size may fix alleles, or radically change the
+species-wide frequency, of alleles previously restricted to small portion of a species range. 
+Range sizes are in constant in flux, with
+
+Similarly the range fragmentation accompanying speciation may act to fix variants within newly formed
+species that were polymorphic across ancestral species ranges.
+
+Note that this does not mean that:
+Any change in neutral fixation rate.
+Any change in
+
+Many DMIs are locally restricted \citep{Cutter:12}.
+
+species range evolution
+splitting of species.
+\citep{Rosenblum:12}
+
+Futuyma has argued that much of the differentiation within species is \citep{Futuyma:10,FUTUYMA:87}
 

### 2014-08-28 standing-variation-paper.tex
latemlified
added more to discussion, and corrected typo in multiple variant section

@@
 to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
-hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd ,
+hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd,
 Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
 the $A-$, Med, and Mahidol suggest that they have spread in past few
-thousand years \citet{tishkoff-g6pd, Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
+thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
 consistent with the age of other malaria resistance alleles. 
 These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
-selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd
-  Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
+selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
 the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
@@
   \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
-\caption{ %
-{\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
+\caption{ 
+Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
   the frequency of G6pd-deficiency alleles. The size of the pie chart
@@
   endemic malaria are colored yellow. Figure taken from
-  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.
-
-}
+  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.}
 \end{center}
@@
 \begin{equation}
-    z_0 = \int_0^\infty {\lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3} \; \exp \left( - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
+    z_0 = \int_0^\infty {2 \lambda_{0,j} \pi v^2 t - \lambda_j \pi v^2 t^3 } \; \exp \left(\sum_j - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
 \end{equation}
@@
 
-\paragraph{}
+%%Mayr: [t]he steady and high genetic input caused by gene flow is the main factor responsible for genetic cohesion among the populations of a species.
+
+\paragraph{Are species held together by wide-spread selective sweeps?}
 Our results touch on an old debate on the evolutionary coherence of
-species. 
-Mayr and others have argued that species are coherent \citep{} 
-as they are united by shared gene flow. However, this arguement has
+species. Mayr and many others have argued that species are coherent
+evolutionary units as they are united by shared gene flow \citep[pages
+521–522 ][]{Mayr:SpeciesEvol}. However, this argument has
 questioned by a number of authors based on relatively high levels of 
-differentiation in many species \citep{}.
+differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
 \citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
@@
 should be viewed as being united as levels of gene flow were still
-high enough for selected alleles to spread across the entire species.
-Our work on geographic convergent adaptation
-\citep[see also][]{ralph2010parallel,RalphCoop:14} brings these conclusions partially into
-question. Our results suggest that wide-spread species should often
-repeatedly adapt in parallel rather than waiting for a single allele
-to migrate across the range.
+high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
+Indeed large scale genotyping and sequencing, along with more
+sophisticated methods, are highlighting ever more signals of gene flow
+between populations and species \citep{}.
+However, our work on geographic convergent adaptation
+\citep[see also][]{ralph2010parallel,RalphCoop:14} brings the
+conclusion that this gene flow will lead to the rapid spread of
+selected alleles partially into
+question. Our results suggest that species should often adapt to
+wide-spread selection pressures through convergent evolution rather
+than waiting for a single allele to migrate across the range over
+length-scales determined by our characteristic length.
 
@@
 geographically restricted \citep{}, but see XXX. This is likely in part due to the
-relatively low occurance of species-wide selection pressures \citep{},
+relatively low incidence of wide-spread selection pressures \citep{},
 but as noted above even when we know of wide-spread selection
@@
 boundaries, suggesting that selected alleles do sometimes spread
-despite low migration rates \citep{}. However, at least some of
-these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
+despite low migration rates \citep{hedrick2013adaptive}. However, at least some of
+these cases likely reflect the introgression of haplotype complexes
+consisting of multiple tightly linked beneficial
 alleles.
@@
 
-If gene flow rarely rapidly spread selected alleles across
-species range, how then do selected alleles ever become fixed within species? 
-Repeated bouts of adaptation in particular genomic regions may act to slowly
+\paragraph{How then do substitutions occur?}
+It is rare for gene flow to rapidly spread selected alleles across
+species range, how then do selected alleles ever become fixed within
+species? Repeated bouts of adaptation in particular genomic regions may act 
 push a subset of previously selected alleles to fixation across the
@@
 
-One potential hypothesis is that many selected alleles only achieve
+\paragraph{Speciation and extinction as phases of molecular evolution?}
+One potential resolution is that many selected alleles achieve
 fixation, not through their own species-wide spread, but rather
-through subsequent large-scale
-changes in geographic range size induced by extirpation of the species
-over part of its range. Such drops in range size may fix alleles, or radically change the
+through subsequent large-scale changes in geographic range size
+induced by extirpation of the species over parts of its range. Such drops in range size may fix, or radically change the
 species-wide frequency, of alleles previously restricted to small portion of a species range. 
-Range sizes are in constant in flux, with
+Also many modes of speciation are proposed to occur through a
+geographically-limited subset of populations forming the basis of new species, e.g. the splitting off
+of part of the range through a vicariance event or dispersal of a
+subset of individuals. As such specation will lead to the geographic assorting of
+polymorphic ancestral variation, again acting to fix variants within
+newly formed species that were previously polymorphic across ancestral species ranges.
 
-Similarly the range fragmentation accompanying speciation may act to fix variants within newly formed
-species that were polymorphic across ancestral species ranges.
+Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
+e.g. adaptation to local conditions, within species maybe ephemeral and
+subject to loss due to local extinction and the mixing following the collapse of population
+structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
+response to punctuated equilibrium \citep{eldredgegould72}, arguing
+that the observation of stasis and rapid anagensis associated with
+speciation were consistent with micro-evolution . 
+Futuyma argued that despite rapid adaptation over short time-scales, we
+may observe morphological stasis in the
+fossil record as
+much of this adaptation is lost \citep[see
+also][]{Eldredge:05}. Furthermore, he suggests that speciation
+may act as rachet to prevent the loss of differentiation, acting to
+maintain adaptive changes among populations, and prevent their loss by
+interbreeding. At face value the rate of species formation seems too low to
+contribute to this process. However, as \citep{Rosenblum:12}, and many
+others, have argued that in fact the rate of speciation may well be very high,
+but that the majority of incipient species do not persist long due to
+reabsorption or extinction. Repeated bouts of extinction and
+speciation will send waves of alleles
+to fixation along particular lineages.
 
-Note that this does not mean that:
-Any change in neutral fixation rate.
-Any change in
+This link between speciation and substitution does not imply that
+substitutions should be thought of as being clustered at splits in
+inferred phylogenies \citep[see ][for a recent exhange on
+this]{Venditti:10, Pennell:14,Venditti:14,Pennell:14B}. 
+(Note that neutral substitutions will be to be unaffected by this process, as they
+will accumulate in a clock like manner along lineages, as dictated by
+the mutation rate, regardless of the geographic details of their
+polymorphic stage.)
+Firstly, it it likely that splits in
+phylogenies are likely only a tiny proportion of all incipient
+speciation events because extinction rates may be high
+\citep{Rosenblum:12}, such that every lineage has likely past through
+many ``speciation events'' in addition to the observed ones.
+Secondly, under this view, the origination process of alleles and their fixation process
+are uncoupled \citep[in the sense of ][ page XXX]{Gillespiebook} by the geographic structure.
+Such that spatial polymorphisms may accumulate gradually, in
+geographically restricted populations, between large-scale 
+biogeographic events. 
+%Thus selected alleles that eventually become a
+%substitutions may have arisen long before they reach fixation.
 
+Nor does this view imply that adaptive evolution or speciation is driven by the
+shifting balance or genetic revolutions \citep{Wright:32,
+  Mayr-genetic-revol:1954}, whereby extreme genetic drift allows
+populations to cross fitness valleys and substitute novel epistatic
+combinations. While geographic lineage sorting via speciation and
+extinction can be thought of as very large-scale genetic drift events,
+the initial spread of our alleles is due to selection they are only later sorted
+
+This view suggests that the alleles that substitute along a lineage
+may not be those but rather are in part those lucky enough to have
+spread across areas not
+
+contributing to
+
+In reality the evolution of range size is not independent of
+adaptation. 
 Many DMIs are locally restricted \citep{Cutter:12}.
@@
 splitting of species.
-\citep{Rosenblum:12}
-
-Futuyma has argued that much of the differentiation within species is \citep{Futuyma:10,FUTUYMA:87}
 

### 2014-08-28 standing-variation-fns.R

@@ -137,3 +137,27 @@ paramString <- function(ps=c("mu","rho","sb","sd","sigma"), pos=-1) {
     paste( sapply( 1:length(ps), function (k) { sprintf( "%s=%.2g", names(ps)[k], ps[k] ) } ), collapse=", " )
 }
 
+
+MultipleTypes <- function (mu, rho, sb, sd, sigma) {
+    # proportion of space arising from standing variation
+    # is int_0^\infty 2 * lambda0 * pi * v^2 * t * exp( - lambda0 * pi * v^2 * t^2 - lambda * pi * v^2 * t^3 /3 ) dt
+    if (any(sd==1)) {
+        return( list(value=0) )
+    } else if (sd==0) {
+        return( list(value=1) )
+    } else { 
+        lambda.1 <- 2*mu[1]*rho[1]*sb[1]
+        lambdaoh.1 <- lambda.1/ log(1/(1-sd[1]))
+        lambda.2 <- 2*mu[2]*rho[2]*sb[2]]
+        lambdaoh.2 <- lambda.2/ log(1/(1-sd[2]))
+ 
+        v <- sigma * sqrt(2*sb)
+        f <- function (t) {
+            2 * lambdaoh * pi * v^2 * t * exp( - lambdaoh*pi*v^2*t^2 - lambda*pi*v^2*t^3 /3 )
+        }
+        xx <- 2^(-30:30)
+        yy <- sapply(xx, function(x) { f(x) } )
+        scale <- xx[ which.min( abs(10-yy*xx) ) ]
+        return( integrate(function(x) { f(x*scale)*scale }, 0, Inf) )
+    }
+}

### 2014-09-02 standing-variation-paper.tex
changes throughout

@@
 \newcommand{\R}{\mathbb{R}}
+
+%\iflatexml  % no margin notes in latexml
+% \newcommand{\marginnote}[1]{}
+%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
+%\fi
 
@@
 Note pennings and hermisson found standing deleterious variation was very important.
-
-recall peromyscus examples with patchy environments
-
-Summarize previous paper
-
-Question assumptions of no standing variation and homogenous selection
-
-\section{General model}
-
-outline here assumptions common to both models.
-
-Stick to constant-speed waves.
-Note monty's empirical speeds paper.
-
-\section{Standing variation}
 
@@
 We assume that diploid fitness is additive, or at least that the
-important early dynamics are determined by the haploid fitness
-\gc{heterozygous fitness, with no reference to the fitness of the
-  homozygote for the mutant alleles}.
+important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
+  homozygote for the mutant alleles.
 The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
@@
 \marginnote{XXX or do diploid? below is haploid.}
+
+We assume that dispersal occurs by local movement of
+individuals. Specifically we will assume that individuals dispersal a
+normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
+(i.e.\ $\sigma$ is the dispersal distance)
 
@@
 \end{align}
+Note that this is closely related to the standard frequency of a
+allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}
 
@@
 Compute typical size and radius of a standing cluster?}
-\gc{Point out connection to standard mutation selection balance formula}
 
@@
 We assume that the allele, having become established, quickly settles
-down to spread as a traveling wave of constant speed. The behavior of
+down to spread spatially as a traveling wave of constant speed. The behavior of
 this wave of advance of a beneficial allele was first described by 
-\citet{fisher1937wave} and \citet*{KPP1937}.
+\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
+conditions the speed of advance
+of this wave will be $v = \sigma \sqrt{2s}$. See
+\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
+applies to this scenerio. Note that the speed of the wave can be
+changed
 
-If the mean squared distance between parent and child is $\sigma^2$
-(i.e.\ $\sigma$ is the dispersal distance) then the speed of advance
-of this wave will be $v = \sigma \sqrt{2s}$.
-
-\plr{recall PPP computations; draw cones of opportunity for new mutations }
-
-\paragraph{}
+\paragraph{Putting it together.}
 So we can put these ingredients together for a simple model of the
@@
 little lightning bolts, and occur at a density $\lambda_0$ across
-space. 
-They spread at velocity $v$ carving out cones in space-time. As these
+space. They spread at velocity $v$ carving out cones in space-time. As these
 alleles proceed in their geographic spread other new alleles can arise
@@
 stars. These new mutations arise and become established at rate $\lambda$.
+
+As we previously outlined in \citet{ralph2010parallel} this model of
+geographic convergent evolution, when $\lambda_0=0$, is analogous to the \citet{kolmogorov-crystallization}
+model of crystallization. In \citet{kolmogorov-crystallization} model nucleation points form at randomly at
+a constant rate in time and space 
+and initiate the radial growth of new crystals. The crystals at the
+end point of this process form a random tessellation covering space
+the properties of which have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
+The generalized version of this process, for non-constant wave speeds
+and imhomogeneous Poisson processes is termed the 
+Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
+Our combined process with both standing variation and new mutation is
+a specially case of the KJMA tessellation, where the
+spatial-temporally homogeneous
+Poisson origination process of new mutations, is supplemented by a
+single pulse of origination points at time zero with spatial density $\lambda_0$.
+
+If we only allow standing variation then everything about the process
+is relatively simple. All of the
+points in space closer to a particular origination point will be
+reached by that allele before any other
+allele has a chance to spread there. Formally then the spatial patterning,
+given a set of origination points, would be specified by the Voronoi
+tessellation (i.e. the cells formed by assigning points to their
+nearest origin). All of the properties of the patterning of space by alleles is
+determined by the spatial locations of the initiation points, as
+specified by the spatially homogeneous Poisson process, independent of
+the rate of spatial spread. So that the properties
+of the spatial tessellation with standing variation depends only on
+$\lambda_0$, and this
+model is commonly called the Poisson-Voronoi tesselation.
+
+In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
+speeds, as can arise due to fat-tailed dispersal kernels. Here we
+restict ourselves to constant wave speeds for simplicity but similar
+calculations could be done by expansion of the results of the previous paper.
 
@@
 
-\subsection{G6PD}
+\subsection{G6PD example}
+We will describe a number of properties of our process, but we first
+want to introduce a motivating example to provide some concrete
+numbers to use for illustrative purposes.
 
@@
 order of $\sim 10$ million $km^2$. In that area there are at least $15$
-common variants \citep{Howes-g6pd-variants}. So the average width of
+common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). So the average width of
 an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
-
-G6pd is 515 codons long, with 140 variants reported in it REF. Assuming a
+The coding region of G6pd is 515 codons long, with $\sim$140
+functional, distinct variants that cause reduced G6PD activity. Assuming a
 mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
 $\mu \approx 1 \times 10^{-6}$ a generation. 
-The dispersal and demographic parameters of humans is unclear,
-particularly as we are concerned with the effective population density.
+The dispersal and demographic parameters of humans in the past few
+thousans years is unclear,
+particularly as we are concerned with the effective population
+density. 
 We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
 We use $\sigma=10,~50,$ and $100$ as our dispersal parameter. 
+We note that obviously human migration has been
+shaped both by local dispersal and larger-scale expansions \citep[see
+][for a recent discussion]{}, so these parameters are for
+illustrative purposes only.
 
@@
   endemic malaria are colored yellow. Figure taken from
-  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.}
+  \citet{Howes-g6pd-variants}
+  \url{http://www.malariajournal.com/content/12/1/418}. \gc{Not sure
+    whether to include this.}
+} \label{fig-G6PD-map}
 \end{center}
@@
 
-In \citet{ralphcoop2010}, studying the model without standing variation,
+In \citet{ralph2010parallel}, studying the model without standing variation,
 we defined a {\em characteristic length} which gave the spatial scale across with mutants with distinct origins would establish.
@@
 \subsection{Standing Variation -- Results} 
-
 The quantity that is perhaps easiest to compute is the mean time until adaptation.
@@
 One interesting problem that we can seek to address with this work is
-the extent to which
+the extent to which the degree of pleiotropy biases the 
+adaptation towards the repeated use of particular subset of loci
+(i.e. convergence at genetic level). 
+While many alleles may result in the benefical phenotype, many of
+these may be unlikely to contribute to adaptation due to their
+negative pleiotropic consequences. There are at least two ways that
+negative pleiotropy can contribute to high rates of
+convergence when adapting when a single change is sufficient for
+adaptation. 
+First, negative pleiotropic effects can reduce the overall beneficial
+selection coefficient of an allele in the new environment making them
+unlikely to become established and slow to spread (and in the worst
+case making them deleterious). This effect has been studied by a
+number of authors \citep{}. A second contribution of negative
+pleiotropy to shaping genetic convergence, is that alleles that have
+less negative pleiotropy are more likely to segregate before the
+environmental shift, and so get a head start in contributing to
+adaptation. This effect is seemingly less well studied, perhaps in
+part because many modeling approaches tend to use the strong-selection
+weak mutation limit and so ignore standing variation. \gc{Is this fair?}
 
+Here we focus primarily on the second effect. 
 Lets imagine for the moment that all classes of beneficial allele have
@@
 $v$. 
-
+The the rate of type $k$ standing variants per unit area and the rate of de
+novo variants per unit area per generation are
 \begin{align}
@@
 \end{align}
+Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
+the proportion of the species range covered by alleles of type $k$ is
+\begin{equation} \label{eqn-prop-space-allele-k}
+    p_k = \int_0^\infty  \big( 2 \lambda_{0,j} \pi v^2 t + \lambda_j
+    \pi v^2 t^3 \big)  \; \exp \left(\sum_k - \lambda_{0,k} \pi v^2 t^2 - \lambda_k \pi v^2 t^3 / 3 \right) dt .
+\end{equation}
+If we only allow standing variation this collapses to 
+\begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
+    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) }
+\end{equation}
+while if we only allow new variation, i.e. if all variants are highly
+deleterious before the environment switches, then $p_k =\mu_j / (\sum_k  \mu_k)$
 
-\begin{equation}
-    z_0 = \int_0^\infty {2 \lambda_{0,j} \pi v^2 t - \lambda_j \pi v^2 t^3 } \; \exp \left(\sum_j - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
-\end{equation}
+To illustrate some of the properties of this model lets imagine the
+somewhat extreme scenerio that there is a single base-pair where
+a mutation change is relatively free of negative pleiotropy (class $1$), and a
+larger mutational target where changes have more serious pleiotropic
+consequences in the ancestral environment (class $2$). We set $s_{d,1}
+\leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, and think of the second
+class mutational target being ten, a hundred, or one thousand base
+pairs. We set  . We show the expected proportion of space covered by
+the rarer class $1$ mutations in Figure \ref{fig-pleiotropy_calc},
+assuming that both classes share a beneficial selection coefficient
+$s_b=0.05$ once the environment switches. 
+Intuitively the contribution of the rarer mutation decreases as the
+mutational target of the other class becomes larger, and as the
+difference in the negative pleiotropoc consequences of the two classes
+of alleles decreases. The case with standing variation only is the best
+case scenerio for the rarer mutation as we force them to have the same
+advantage in the new environment, so its rate of introduction is
+necessarily lower. However the standing variation only case does seems
+to provide a reasonable rule of thumb. Especially for parameter combinations, higher population densities and
+high dispersal distances, that increase the contribution of standing
+variation (and similarly for high $s_b$).
+
+\begin{figure}[ht]
+\begin{center}
+  \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
+\caption{ %
+{\bf Proportion of space covered by rarer but less pleiotropic
+  mutation}. Empty circles give the result for standing variation only,
+  eqn. \eqref{eqn-prop-space-allele-k-standing-room-only}. Lines give
+  the result allowing both standing and de novo mutations  eqn. \eqref{eqn-prop-space-allele-k}}   \label{fig-pleiotropy_calc}
+\end{center}
+\end{figure}
 

### 2014-09-02 g6pd_figures.R

@@ -90,10 +90,39 @@ for(j in 1:2){
 dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_standing_var_proportion.pdf")
  
  
+ ####effect of pleiotropy on standing variation
+  more.cols<-wes.palette(n=3,"FantasticFox")
+  sd.1<-c(,10^(-seq(log10(0.05),5,length=100)));
+  plot(x=range(sd.1),y=c(0,1),type="n",log="x",xlab=expression(s[d1]),ylab=expression(p[1]),cex.lab=1.5,cex.axis=1.5)
+ mu.2s<-c(1e-7,1e-6,1e-5)
  
-# as.expression( c(lapply(rate , function(x) substitute(2*N*nu==myS, list(myS=x) ) )
- as.expression( c(lapply(rate , function(x) substitute(2*N*nu==myS, list(myS=x) ) ),"Top-hat","Step","Theory"))
- ##areas with Malaria
+ 
+ for(i in 1:length(mu.2s)){
+  	prob<-sapply(sd.1,function(x){MultipleTypes(mu=c(1e-8,mu.2s[i]), rho=2, sb=c(0.05,0.05), sd=c(x,0.05), sigma=10)$value})
+ 	lines(sd.1,prob,col=more.cols[i],lty=1)
+  	prob<-sapply(sd.1,function(x){MultipleTypes(mu=c(1e-8,mu.2s[i]), rho=2, sb=c(0.05,0.05), sd=c(x,0.05), sigma=100)$value})
+ 	lines(sd.1,prob,col=more.cols[i],lty=2)
+
+  	prob<-sapply(sd.1,function(x){MultipleTypes(mu=c(1e-8,mu.2s[i]), rho=0.2, sb=c(0.05,0.05), sd=c(x,0.05), sigma=10)$value})
+ 	lines(sd.1,prob,col=more.cols[i],lty=3)
+  	prob<-sapply(sd.1,function(x){MultipleTypes(mu=c(1e-8,mu.2s[i]), rho=0.2, sb=c(0.05,0.05), sd=c(x,0.05), sigma=100)$value})
+ 	lines(sd.1,prob,col=more.cols[i],lty=4)
+ 
+	prob<-sapply(sd.1,function(x){MultipleTypes(mu=c(1e-8,mu.2s[i]), rho=2, sb=c(0.05,0.05), sd=c(x,0.05), 	sigma=20,standing.var.calc=TRUE)$value})
[…]

### 2014-09-02 standing-variation-fns.R

@@ -138,22 +138,29 @@ paramString <- function(ps=c("mu","rho","sb","sd","sigma"), pos=-1) {
 }
 
 
-MultipleTypes <- function (mu, rho, sb, sd, sigma) {
+MultipleTypes <- function (mu, rho, sb, sd, sigma,include.new=TRUE,standing.var.calc=FALSE) {
     # proportion of space arising from standing variation
     # is int_0^\infty 2 * lambda0 * pi * v^2 * t * exp( - lambda0 * pi * v^2 * t^2 - lambda * pi * v^2 * t^3 /3 ) dt
     if (any(sd==1)) {
         return( list(value=0) )
-    } else if (sd==0) {
-        return( list(value=1) )
+    } else if (any(sd==0)) {
+        return( list(value="wrong") )
     } else { 
-        lambda.1 <- 2*mu[1]*rho[1]*sb[1]
+        lambda.1 <- 2*mu[1]*rho*sb[1]
         lambdaoh.1 <- lambda.1/ log(1/(1-sd[1]))
-        lambda.2 <- 2*mu[2]*rho[2]*sb[2]]
+        lambda.2 <- 2*mu[2]*rho*sb[2]
         lambdaoh.2 <- lambda.2/ log(1/(1-sd[2]))
  
-        v <- sigma * sqrt(2*sb)
+ 		if(standing.var.calc){
+ 			return(list(value=lambdaoh.1/(lambdaoh.1+lambdaoh.2)))
+ 		} 
+ 		
+        v.1 <- sigma * sqrt(2*sb[1])
+        v.2 <- sigma * sqrt(2*sb[2])
+        stopifnot(v.1==v.2)
+    #    recover()
         f <- function (t) {
-            2 * lambdaoh * pi * v^2 * t * exp( - lambdaoh*pi*v^2*t^2 - lambda*pi*v^2*t^3 /3 )
+            (2 * lambdaoh.1 * pi * v.1^2 * t +  include.new*lambda.1 * pi * v.1^2 * t^2 )* exp( - (lambdaoh.1*v.1^2+lambdaoh.2*v.1^2)*pi*t^2 - include.new*(lambda.1*v.1^2+lambda.2*v.2^2)*pi*t^3 /3 )
         }
         xx <- 2^(-30:30)
         yy <- sapply(xx, function(x) { f(x) } )

### 2014-09-02 standing-variation-paper.tex
added char length as function of sb to R script, and edits to man text

@@
 \usepackage{amsmath,amssymb}
-\usepackage{natbib}
+\usepackage[authoryear]{natbib}
 \usepackage{marginnote}
@@
 %\fi
+\bibliographystyle{amnat}
+%\bibliographystyle{plain}
 
-\bibliographystyle{plain}
+\title{The role of standing variation in parallel adaptation}
+\author{Peter L. Ralph$^1$ \\ email: pralph@usc.edu  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }
 
 \begin{document}
+
+\maketitle
+\date{}
+
+\begin{center}
+
+$^1$ Computational Biology and Bioinformatics \\ 
+University of Southern California, Los Angeles, CA \\
+$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
+University of California -- Davis, Davis, CA, 95616
+\end{center}
+
+\section*{Abstract}
 
@@
 
-We assume that dispersal occurs by local movement of
-individuals. Specifically we will assume that individuals dispersal a
-normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
-(i.e.\ $\sigma$ is the dispersal distance)
+%We assume that dispersal occurs by local movement of
+%individuals. Specifically we will assume that individuals dispersal a
+%normally distributed distance, with the mean squared distance between parent and child is $\sigma^2$
+%(i.e.\ $\sigma$ is the dispersal distance)
 
@@
 \citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
-applies to this scenerio. Note that the speed of the wave can be
-changed
+applies to this scenerio. Note that the speed of the wave will vary
+with details of the space that individuals migrate across
+\citep[e.g. see ][ for comparisons to migration on discrete
+grids]{Slatkin-speed:76,SlatkinCharlesworth:78}, however, the velocity
+will usually be primarily determined by the product $\sigma^2$ and $s$.
 
@@
 
+\paragraph{}
 Our results are predicated on the idea that adaptive variants are
@@
 Our ability to detect such shared convergent events is at the moment
-quite limited. 
-One way that selection pressures might be weak on alleles, is if they
-are at only one of many loci contributing to variation in a trait. However, if that
+quite limited. If many loci contributing to variation in a trait, then
+selection on any one allele may be weak allowing them to segregate
+across populations. However, if that
 case populations separated by a reasonable amount of drift may still
@@
 best positioned to initially respond to positive selection, i.e. those at
-intermediate frequency, will be somewhat different.
+intermediate frequency, will be somewhat different. Therefore, it
+seems unlikely that polygenic traits are more suspectible to a shared
+genetic basis to adaptation across populations than simple traits.
 
@@
 \paragraph{How then do substitutions occur?}
-It is rare for gene flow to rapidly spread selected alleles across
+If it is rare for gene flow to rapidly spread selected alleles across
 species range, how then do selected alleles ever become fixed within
@@
 that the observation of stasis and rapid anagensis associated with
-speciation were consistent with micro-evolution . 
+speciation were consistent with micro-evolution. 
 Futuyma argued that despite rapid adaptation over short time-scales, we
@@
 
+%Venditti:10, 
 This link between speciation and substitution does not imply that
@@
 inferred phylogenies \citep[see ][for a recent exhange on
-this]{Venditti:10, Pennell:14,Venditti:14,Pennell:14B}. 
+this]{Pennell:14,Venditti:14,Pennell:14B}. 
 (Note that neutral substitutions will be to be unaffected by this process, as they
@@
 polymorphic stage.)
-Firstly, it it likely that splits in
+First, it it likely that splits in
 phylogenies are likely only a tiny proportion of all incipient
@@
 many ``speciation events'' in addition to the observed ones.
-Secondly, under this view, the origination process of alleles and their fixation process
-are uncoupled \citep[in the sense of ][ page XXX]{Gillespiebook} by the geographic structure.
-Such that spatial polymorphisms may accumulate gradually, in
+Second, under this view, spatial polymorphisms may accumulate gradually, in
 geographically restricted populations, between large-scale 
-biogeographic events. 
+biogeographic events leading to their fixation (or indeed loss). 
+Such that the origination process of alleles and their fixation process
+are reasonably uncoupled \citep[in the sense of ][]{Gillespie:94} by the geographic structure.
+
 %Thus selected alleles that eventually become a
@@
 shifting balance or genetic revolutions \citep{Wright:32,
-  Mayr-genetic-revol:1954}, whereby extreme genetic drift allows
+  Mayr-genetic-revol:1954}, whereby genetic drift allows
 populations to cross fitness valleys and substitute novel epistatic
@@
 the initial spread of our alleles is due to selection they are only later sorted 
+into substitutions via these large scale drift events \citep[see also
+][for discussion]{Futuyma:89}.
 
 This view suggests that the alleles that substitute along a lineage
-may not be those but rather are in part those lucky enough to have
-spread across areas not
-
-contributing to
+may rarely be capable, or need, to spread species-wide. This is not
+inconsistent with genome-wide evidence that a reasonable fraction of 
+substitutions are fixed by positive selection in a number of species
+\citep[most notably Drosophila][]{}. Selection has played a strong
+role in the establishment of these alleles, and their frequency may be
+high enough within populations that they do not contribute
+substantially to polymorphism.
 
 In reality the evolution of range size is not independent of
-adaptation. 
-Many DMIs are locally restricted \citep{Cutter:12}.
-
-species range evolution
-splitting of species.
+adaptation. Many DMIs are locally restricted \citep{Cutter:12}.
 

### 2014-09-02 g6pd_figures.R

@@ -5,6 +5,38 @@ library("wesanderson")
 
 my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
 
+
+###################################################
+###################Char length as function of sd
+###################################################
+
+ mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.001,0.3,length=800); 
+
+sigmas=c(100,50,10)
+
+layout(t(1:2))
+par(mar=c(4,4.1,2,0.3))
+for(j in 1:2){
+	rho=rhos[j]
+	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(chi),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
+	if(j == 2) plot(sbvals,sbvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.4,main=expression(rho==2))
+	for(i in 1:3){
+	sigma=sigmas[i]; charlength.sb<-sapply(sbvals, function (sb) { charLength(mu,rho,sb,sd=sd,sigma)$value } )
+	 lines(sdvals,charlength.sb,lwd=2,col=my.col[i]); 
+	 
+	 charlength.new.only<-sapply(sbvals, function (sb) {  newCharLength(mu, rho, sb, sd=NA, sigma)$value})
+	 lines(sdvals,charlength.new.only,col=my.col[i],lwd=2,lty=2)  ##new mut. only
+	}
+	 ##contribution of standing var only, i.e. old variation
+	 charlength.standing.only<-sapply(sbvals, function (sb) {  oldCharLength(mu, rho, sb, sd=sd, sigma)$value})
+	 lines(sdvals,charlength.standing.only ,lwd=2,col="black",lty=3)
+	  if(j == 2) legend("topright",col=c("grey","grey","black",my.col[1:3]),lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
+	 	lapply(sigmas , function(x){  {substitute( sigma==myS, list(myS=x) )} }))))
+ 	}
+
+###################################################
+###################Char length as function of sd
[…]

### 2014-09-03 standing-variation-paper.tex
wrote intro

@@
 
+There is increasing number of examples of different populations within species often
+adapt to shared environments via independent genetic changes. 
+In some cases this convergent evolution leads to differing phenotypes
+in response to shared selection pressures
+\citep{} in other cases the degree of convergence is precise down the
+same nucleutide change. Such convergent evoltuion within populations
+has been seen for many
+carefully studied phenotypes in a range of species, including drug resistance, pathogen resistance, insecticideresistance, and pigmentation. In many of these examples we 
+need to be careful about whether they changes are truely convergent or
+whether selection pressures differ subtly. 
+We note that we treat the phrases convergent and parallel evolution
+synonymously, as we are concerned with adaptation within a single
+species that can occur via a different or shared genetic routes \citep{Arendt:08}.
+
+The issue of convergent adaptation within species touches on a number
+of important questions in evolutionary biology. These include the extent
+to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
+whether adaptation is mutation-limited \citep{karasov2010}, and whether
+species should be regarded to cohesive units. Convergence within
+populations also effects our
+ability to detect adaptation from population genomic data as no one
+allele sweeps fixation over the entire area affected by a selection
+pressure \citep{softsweepsIII}.
+
+Convergent evolution can occur even within a well mixed population
+subject to a constant selection pressure, either through selection on
+multiple mutations present as standing variation within population
+before selection pressures switch \citep{Orr:01,softsweepsI} or due to
+multiple adaptive alleles arising after selection pressures switch \citet{softsweepsII}. 
+This previous work has shown that 
+a primary determinant of the probability that multiple alleles contribute to adaptation
+is the population scaled mutation rate \citep[see ][ for a review]{MesserPetrov}.
+
+Geographically limited dispersal also increases the chance of
+convergent evolution. For example when selection pressures are
+geographically patchy this can increasing the probability of
+convergent evolution as alleles are unable to spread through intervening populations \citep{RalphCoop:14}.
 In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
@@
 and that a single mutational change was sufficient to adapt populations to the novel environment. 
-We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift.
-In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel.
-This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
-Recall questions of origin of adaptations
+We assumed that there was no standing variation for the adaptive
+allele, such that parallel mutation must be due to multiple mutations
+after the environmental shift. In that setting we found that there was
+a characteristic geographic scale, over which we would expect multiple
+instances of our adaptive allele to have arisen in parallel. This
+characteristic length could be expressed in terms of a simple compound
+parameter determined by our parameters of interest.
 
-References on parallel adaptation, standing variation, etc.:
-  coop,
-  pennings and hermisson.
-Note pennings and hermisson found standing deleterious variation was very important.
+Here we extend our spatial model from \citep{RalphCoop} to include standing
+variation present at mutation-selection balance before selection
+pressures switch.
 
@@
 In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
+as a function 
 calculated for our G6pd parameters.
+
+In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
+calculated for our G6pd parameters.
+
+\begin{figure}[ht]
+\begin{center}
+  \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
+\caption{ %
+{\bf Characteristic length,} as a function of selective advantage,
+holding the prior disadvantage of the allele fixed $s_d=0.05$. 
+We compare this to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+} \label{Fig-G6PD-charlength-sb}
+\end{center}
+\end{figure}
 
@@
 The quantity that is perhaps easiest to compute is the mean time until adaptation.
-Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
+If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
 Then, as shown in Figure XX,
@@
 
-In Figure \ref{G6PD_standing_var_proportion}
+In Figure \ref{G6PD_standing_var_proportion} we show the proportion of
+alleles that spread from standing variation, and the proportion of
+geographic space covered by standing variants.
 
@@
 \caption{ %
-{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the sickle-cell example.  XXX say more.
+{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the G6PD example.  XXX say more.
 } \label{G6PD_standing_var_proportion}
@@
 
-\marginnote{do mean (and variance?) of distance between haplotypes?}
-
-influence of migration on things -- plot these versus migration
-
-{\tt Simulations:} compare mean density against $s_d$ at different values of $\sigma$; omit variability if needed for readability.
-Compare to panmictic results.
+{\tt Simulations:} compare mean density against $s_d$ at different
+values of $\sigma$; omit variability if needed for readability. Compare to panmictic results.
 

### 2014-09-03 g6pd_figures.R

@@ -34,6 +34,8 @@ for(j in 1:2){
 	 	lapply(sigmas , function(x){  {substitute( sigma==myS, list(myS=x) )} }))))
  	}
 
+dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlengths_sb.pdf")
+
 ###################################################
 ###################Char length as function of sd
 ###################################################

### 2014-09-12 standing-variation-paper.tex
Graham's changes

@@
 to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
-whether adaptation is mutation-limited \citep{karasov2010}, and whether
-species should be regarded to cohesive units. Convergence within
-populations also effects our
-ability to detect adaptation from population genomic data as no one
+whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, and whether
+species should be regarded to cohesive units. Convergence evolution within
+populations also effects our ability to detect adaptation from population genomic data as no one
 allele sweeps fixation over the entire area affected by a selection
@@
 geographically patchy this can increasing the probability of
-convergent evolution as alleles are unable to spread through intervening populations \citep{RalphCoop:14}.
-In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
+convergent evolution as alleles are unable to spread through
+intervening populations \citep{RalphCoop:14}. In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
 where the selection pressure was constant across the entire environment, 
@@
 characteristic length could be expressed in terms of a simple compound
-parameter determined by our parameters of interest.
-
-Here we extend our spatial model from \citep{RalphCoop} to include standing
+parameter determined by our parameters of interest. Here we extend
+this spatial model include standing
 variation present at mutation-selection balance before selection
@@
 
-Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
+Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation,
 e.g. adaptation to local conditions, within species maybe ephemeral and
@@
 structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
-response to punctuated equilibrium \citep{eldredgegould72}, arguing
+response to the pattern of punctuated equilibrium \citep{eldredgegould72}, arguing
 that the observation of stasis and rapid anagensis associated with
-speciation were consistent with micro-evolution. 
+speciation were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
 Futuyma argued that despite rapid adaptation over short time-scales, we
@@
 much of this adaptation is lost \citep[see
-also][]{Eldredge:05}. Furthermore, he suggests that speciation
+also][]{lieberman:96,Eldredge:05}. Furthermore, he suggests that speciation
 may act as rachet to prevent the loss of differentiation, acting to
@@
 
+While Futuyma and others have advanced these ideas mainly in the
+setting of adaptation to local conditions \citep[particularly local
+communities ][]{} . Much of the recent adaptation that we see as population
+geneticists is likely adaptation to local conditions, but as we have
+shown even wide-spread selection pressures will likely result in
+alleles only spreading locally. The adaptive substitution
+we see along lineages are the alleles that were lucky enough to
+have spread in populations that escaped local extinction, and to have
+avoided being assorted into species that eventually went extinct. 
+Range sizes change rapidly enough that it seems likely that range size
+changes should change substantially over the soujourn time of most
+alleles that will eventually become fixed in the species, as such it
+seems as though .
+
 %Venditti:10, 
 This link between speciation and substitution does not imply that
-substitutions should be thought of as being clustered at splits in
+substitutions should necessarily be thought of as being clustered at splits in
 inferred phylogenies \citep[see ][for a recent exhange on
 this]{Pennell:14,Venditti:14,Pennell:14B}. 
-(Note that neutral substitutions will be to be unaffected by this process, as they
+We first note that neutral substitutions will be to be unaffected by this process, because they
 will accumulate in a clock like manner along lineages, as dictated by
 the mutation rate, regardless of the geographic details of their
-polymorphic stage.)
-First, it it likely that splits in
+polymorphic stage. Turning to the accumulation of adaptive
+substitutions, 
+it likely that splits in
 phylogenies are likely only a tiny proportion of all incipient
@@
 many ``speciation events'' in addition to the observed ones.
-Second, under this view, spatial polymorphisms may accumulate gradually, in
+Futhermore, under this view, spatial polymorphisms may accumulate gradually, in
 geographically restricted populations, between large-scale 

### 2014-09-13 standing-variation-paper.tex
minor edits

@@
 %\fi
-\bibliographystyle{amnat}
-%\bibliographystyle{plain}
+%% NEED amnat.bst IN REPO
+% \bibliographystyle{amnat}
+\bibliographystyle{plainnat}
 
@@
 
-There is increasing number of examples of different populations within species often
-adapt to shared environments via independent genetic changes. 
+There are an increasing number of examples in which
+different populations within species have
+adapted to similar environments via independent genetic changes. 
 In some cases this convergent evolution leads to differing phenotypes
@@
 \citep{} in other cases the degree of convergence is precise down the
-same nucleutide change. Such convergent evoltuion within populations
-has been seen for many
-carefully studied phenotypes in a range of species, including drug resistance, pathogen resistance, insecticideresistance, and pigmentation. In many of these examples we 
-need to be careful about whether they changes are truely convergent or
+same nucleutide change. 
+Such convergent evoltuion within populations has been seen for many
+carefully studied phenotypes in a range of species, 
+including drug resistance, pathogen resistance, insecticide resistance, and pigmentation. 
+In many of these examples we 
+need to be careful about whether the changes are truly convergent or
 whether selection pressures differ subtly. 
-We note that we treat the phrases convergent and parallel evolution
+We note that we treat the phrases ``convergent'' and ``parallel'' evolution
 synonymously, as we are concerned with adaptation within a single
-species that can occur via a different or shared genetic routes \citep{Arendt:08}.
+species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.
 
 The issue of convergent adaptation within species touches on a number
-of important questions in evolutionary biology. These include the extent
-to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
-whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, and whether
-species should be regarded to cohesive units. Convergence evolution within
-populations also effects our ability to detect adaptation from population genomic data as no one
-allele sweeps fixation over the entire area affected by a selection
-pressure \citep{softsweepsIII}.
+of important questions in evolutionary biology. 
+These include the extent to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
+whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
+and to what degree should species be regarded to cohesive units. 
+Convergent evolution within populations 
+also affects our ability to detect adaptation from population genomic data,
+since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.
 
 Convergent evolution can occur even within a well mixed population
-subject to a constant selection pressure, either through selection on
-multiple mutations present as standing variation within population
-before selection pressures switch \citep{Orr:01,softsweepsI} or due to
-multiple adaptive alleles arising after selection pressures switch \citet{softsweepsII}. 
-This previous work has shown that 
+subject to a constant selection pressure, 
+either through selection on multiple mutations present as standing variation within population
+before selection pressures switch \citep{Orr:01,softsweepsI},
+or due to multiple adaptive alleles that arise after selection pressures switch \citet{softsweepsII}. 
+The previous work has shown that 
 a primary determinant of the probability that multiple alleles contribute to adaptation
@@
 
-Geographically limited dispersal also increases the chance of
-convergent evolution. For example when selection pressures are
-geographically patchy this can increasing the probability of
-convergent evolution as alleles are unable to spread through
-intervening populations \citep{RalphCoop:14}. In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
+\plr{should stick to either ``parallel'' or ``convergent'' perhaps}
+
+Geographically limited dispersal also increases the chance of convergent evolution. 
+For example, 
+patchy geographically selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
+since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
+In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
 where the selection pressure was constant across the entire environment, 
 and that a single mutational change was sufficient to adapt populations to the novel environment. 
-We assumed that there was no standing variation for the adaptive
-allele, such that parallel mutation must be due to multiple mutations
-after the environmental shift. In that setting we found that there was
-a characteristic geographic scale, over which we would expect multiple
-instances of our adaptive allele to have arisen in parallel. This
-characteristic length could be expressed in terms of a simple compound
-parameter determined by our parameters of interest. Here we extend
-this spatial model include standing
-variation present at mutation-selection balance before selection
-pressures switch.
+We assumed that there was no standing variation for the adaptive allele, 
+so that parallel mutation must be due to multiple mutations after the environmental shift. 
+In that setting, 
+we found that there was a characteristic geographic scale over which we would expect 
+multiple instances of our adaptive allele to have arisen in parallel. 
+This characteristic length could be expressed in terms of a simple compound
+parameter determined by our parameters of interest. 
+Here we extend this spatial model to include standing variation 
+present at mutation-selection balance before the selection pressures switch.
 
+%%%%%%%%% %%%%%%%%%%%% %%%%%%%
 \subsection{The set-up}
@@
 We assume that the species range
-is a homogenous, one- or two-dimensional region. 
+is a large, homogenous, one- or two-dimensional region. 
 There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
-We assume that different mutations are distinguishable --
+We assume that separately arising mutations are distinguishable --
 either as selectively equivalent mutations, or by linked variation.
-For a long time in the past, the mutated type has been at a selective disadvantage (long enough to be at selection-mutation equilibrium),
+We also suppose that 
+the mutated type has been at a selective disadvantage 
+for a sufficiently long enough time in the past to be at selection-mutation equilibrium,
 but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
@@
 The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
-(If the mutation is recessive, XXX more complicated things happen.)
+(We leave aside the case where mutations are recessive, however.)
 As for the other parameters,
 suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
-and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
+and that the mean squared distance between parent and child is $\sigma^2$.
 The species occupies an area with mean density
@@
 \paragraph{Rates of origination of standing and new mutations}
-We make use of the commonly used approximation that neglects competition between relatives,
+We make use of the commonly used approximation that neglects competition between close relatives,
 treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
@@
 is well--approximated by a Poisson process in space and time.
-This the rate of this Poisson process, 
-the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type
+The rate of this Poisson process,
+i.e.\ the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
 is approximately $\lambda = 2 \mu \rho s_b$.
 
-Before $t=0$, on the other hand, the mutation is deleterious, so the
-genetic descendants of each new mutation are (with high probability)
-doomed (doomed I say!) to extinction,
+\plr{the following needs better explanation, maybe a figure?}
+Before $t=0$, on the other hand, the mutation is deleterious.
+We have assumed that $s_d$ is not too small
+so that the genetic descendants of each new mutation are (with high probability) doomed to extinction,
 but may persist for some time.
-Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
+Since the times and locations of mutants before $t=0$ also well--approximated by a Poisson process with rate $\mu \rho$,
 the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
-by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
+by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
+% (In principle, more than one member of a mutant family may fix,
+% but we assume these are close together enough that we may treat them as beginning from a single point.)
 If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
@@
 equivalent to the mutants arising later.
-(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
-and the spatial distribution of the family is small relative to the spread between them.)
+(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
+and the spatial distribution of each is small relative to the spread between the families.)
 To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
@@
 \end{align}
-Note that this is closely related to the standard frequency of a
+Note that this is closely related to the equilibrium frequency of a
 allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}
@@
 \paragraph{Geographic spread of alleles} 
-Once an allele has become local established it can begin to spatially spread. 
-We assume that the allele, having become established, quickly settles
-down to spread spatially as a traveling wave of constant speed. The behavior of
-this wave of advance of a beneficial allele was first described by 
-\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
-conditions the speed of advance
-of this wave will be $v = \sigma \sqrt{2s}$. See
-\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
-applies to this scenerio. Note that the speed of the wave will vary
+Once an allele has become locally established it can begin to spread across space. 
+We assume that the allele, once established, 
+quickly settles down to spread spatially as a traveling wave of constant speed. 
+The behavior of this wave of advance of a beneficial allele was first described by 
+\citet{fisher1937wave} and \citet*{KPP1937}. 
+Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
+See \citet{ralph2010parallel} for a fuller review of these travelling waves.
+Note that the speed of the wave will vary
 with details of the space that individuals migrate across
-\citep[e.g. see ][ for comparisons to migration on discrete
-grids]{Slatkin-speed:76,SlatkinCharlesworth:78}, however, the velocity
-will usually be primarily determined by the product $\sigma^2$ and $s$.
+\citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.
+% However, the velocity will usually be primarily determined by the product of $\sigma^2$ and $s$.
 
 \paragraph{Putting it together.}
-So we can put these ingredients together for a simple model of the
-geographic spread of alleles. A cartoon example of this process is
-shown in Figure \ref{fig:cartoon} Initially, when the selection
-pressures change at $t=0$, a set of standing variants can start to
-spread having escaped loss through drift. These variants are shown as
-little lightning bolts, and occur at a density $\lambda_0$ across
-space. They spread at velocity $v$ carving out cones in space-time. As these
-alleles proceed in their geographic spread other new alleles can arise
-and become established in parallel, these are shown as little
-stars. These new mutations arise and become established at rate $\lambda$.
+Now, we can put these ingredients together for a simple model of the geographic spread of alleles. 
+A cartoon example of this process is shown in Figure \ref{fig:cartoon}. 
+Initially, when the selection pressures change at $t=0$, 
+a set of standing variants can start to spread having escaped loss through drift. 
+These variants are shown as little lightning bolts, and occur at a density $\lambda_0$ across space. 
+They spread at velocity $v$, carving out cones in space-time. 
+As these alleles proceed in their geographic spread, other new alleles can arise and become established in parallel, 
+these are shown as little stars. 
+These new mutations arise and become established at rate $\lambda$.
 
 As we previously outlined in \citet{ralph2010parallel} this model of
-geographic convergent evolution, when $\lambda_0=0$, is analogous to the \citet{kolmogorov-crystallization}
-model of crystallization. In \citet{kolmogorov-crystallization} model nucleation points form at randomly at
-a constant rate in time and space 
-and initiate the radial growth of new crystals. The crystals at the
-end point of this process form a random tessellation covering space
-the properties of which have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
+geographic convergent evolution, when $\lambda_0=0$, is analogous to the a model of crystallization due to Kolmogorov \citet{kolmogorov-crystallization}.
+In this model, nucleation points form at random at a constant rate in time and space 
+and initiate the radial growth of new crystals. 
+Eventually, the different orientations of crystals form a random tessellation of space,
+hose properties have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
 The generalized version of this process, for non-constant wave speeds
-and imhomogeneous Poisson processes is termed the 
+and inhomogeneous Poisson processes is termed the 
 Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
 Our combined process with both standing variation and new mutation is
-a specially case of the KJMA tessellation, where the
-spatial-temporally homogeneous
-Poisson origination process of new mutations, is supplemented by a
-single pulse of origination points at time zero with spatial density $\lambda_0$.
+a special case of the KJMA tessellation, 
+where the spatial-temporally homogeneous Poisson origination process of new mutations, 
+is supplemented by a single pulse of origination points at time zero with spatial density $\lambda_0$.
+(For the purposes of analogy, we could imagine that before time $t=0$
+the temperature is high enough that nucleation sites do not persist.)
 
-If we only allow standing variation then everything about the process
-is relatively simple. All of the
-points in space closer to a particular origination point will be
-reached by that allele before any other
-allele has a chance to spread there. Formally then the spatial patterning,
-given a set of origination points, would be specified by the Voronoi
-tessellation (i.e. the cells formed by assigning points to their
-nearest origin). All of the properties of the patterning of space by alleles is
-determined by the spatial locations of the initiation points, as
-specified by the spatially homogeneous Poisson process, independent of
-the rate of spatial spread. So that the properties
-of the spatial tessellation with standing variation depends only on
-$\lambda_0$, and this
-model is commonly called the Poisson-Voronoi tesselation.
+If we only allow standing variation, then everything about the process is relatively simple. 
+All of the points in space closer to a particular origination point will be
+reached by that allele before any other allele has a chance to spread there. 
+Such a tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
+(i.e.\ the cells formed by assigning points to the nearest point in a Poisson process).
+The properties of the patterning of space by alleles is
+determined by the spatial locations of the initiation points, 
+as specified by the spatially homogeneous Poisson process, 
+independent of the rate of spatial spread. 
+Therefore, the properties of the spatial tessellation with standing variation depends only on $\lambda_0$.
 
-In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
-speeds, as can arise due to fat-tailed dispersal kernels. Here we
-restict ourselves to constant wave speeds for simplicity but similar
-calculations could be done by expansion of the results of the previous paper.
+% In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
+% speeds, as can arise due to fat-tailed dispersal kernels. 
+% Here we restict ourselves to constant wave speeds for simplicity but similar
+% calculations could be done by expansion of the results of the previous paper.
 
@@
 
+%%%%%%%%% %%%%%%%%%%%
+
 \subsection{G6PD example}
-We will describe a number of properties of our process, but we first
-want to introduce a motivating example to provide some concrete
+
+Below we describe a number of properties of our process, but we will first introduce 
+a motivating example to provide some concrete
 numbers to use for illustrative purposes.
 
-Over the past ten thousand years malaria resistance alleles have arisen at a number
-of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
-A number of these alleles appear to represent convergence, with
-multiple changes in the same gene. 
-For example, there are multiple
-changes at the $\beta$-globin gene that confers malaria resistance,
-and the sickle cell allele may plausibly have arisen by up to five distinct
-changes at the same base pair within Africa
+Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
+and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
+A number of these alleles appear to be examples of convergent adaptation, 
+as different derived mutations in the same gene are seen in different individuals.
+For example, 
+a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
+and the sickle cell allele may plausibly have arisen by up to five independent occurrences
+of the same base pair mutatoin at different locations within Africa
 \citep{Flint:98,ralph2010parallel}.
 Another particularly impressive case of convergent evolution in response to
-selection pressures imposed by malaria are the numerous changes throughout the
-X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
-described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
-found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
-exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
-resistance alleles at G6PD are particularly common and relatively well
-studied: the $A-$ found in
-much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
-east; and the Mahidol allele found Myanmar and Thailand.
+selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
+with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
+that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
+These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
+rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
+Three G6PD deficiency alleles that also confer resistance to malaria 
+are particularly common and relatively well studied: 
+the $A-$ allele found in much of sub-Saharan Africa; 
+the Med allele found in the Mediterranean and Middle East; 
+and the Mahidol allele found Myanmar and Thailand.
 The $A-$ and Mahidol alleles have been shown
 to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
-hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd,
-Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
-the $A-$, Med, and Mahidol suggest that they have spread in past few
-thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
-consistent with the age of other malaria resistance alleles. 
-These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
-selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
-the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
-$A-$ allele. 
-Given such a strong pressure on these alleles they should have risen
-quickly to fixation, so their presence at intermediate frequency,
-over a broad geographic area, makes it a good candidate for a recently
-balanced polymorphism. 
+hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
+Haplotype-based analysis of genetic diversity surrounding
+the $A-$, Med, and Mahidol suggest that they have spread over the past 
+few thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
+consistent with the age of other known malaria resistance alleles. 
+Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
+selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
+consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
+\plr{what are Ruwende's results? feels a bit vague here}
+Given such a strong pressure on these alleles they should have risen quickly to fixation, 
+so their presence at intermediate frequency over a broad geographic area
+makes them good candidates for recently balanced polymorphisms. 
 Hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to a variety of compounds, notably
-those present in fava beans.
+leading to hemolytic anemia when exposed to a variety of compounds, 
+notably those present in fava beans.
 
-The wave of advance works in the case of heterozygote advantage
-\citep{aronson1975nonlinear}, with the selected allele spreading
-locally to the equilibrium frequency (rather than fixation). As such
-our framework is applicable here.
+Note that although originally the ``wave of advance'' was presented as the spread of an unconditionally favorable allele
+\citep{fisher1936,KPP},
+the same model holds in the case of heterozygote advantage \citep{aronson1975nonlinear}, 
+with the selected allele spreading locally to the equilibrium frequency (rather than fixation),
+with speed determined by the advantage of heterozygotes when rare.
 
@@
 order of $\sim 10$ million $km^2$. In that area there are at least $15$
-common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). So the average width of
-an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
-The coding region of G6pd is 515 codons long, with $\sim$140
-functional, distinct variants that cause reduced G6PD activity. Assuming a
-mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
-$\mu \approx 1 \times 10^{-6}$ a generation. 
-The dispersal and demographic parameters of humans in the past few
-thousans years is unclear,
-particularly as we are concerned with the effective population
-density. 
-We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
-We use $\sigma=10,~50,$ and $100$ as our dispersal parameter. 
-We note that obviously human migration has been
-shaped both by local dispersal and larger-scale expansions \citep[see
-][for a recent discussion]{}, so these parameters are for
-illustrative purposes only.
+common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
+Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
+The coding region of G6pd is 515 codons long, 
+and around 140 distinct deficiency alleles have been observed. 
+Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
+we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
+The dispersal and demographic parameters of humans in the past few thousans years is unclear,
+particularly as we are concerned with the ``effective'' population density
+(i.e.\ population density divided by variance in offspring number).
+We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
+and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
+Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
+\citep[see ][for a recent discussion]{}, 
+so these parameters only provide a rough view of the process.
 
@@
 \caption{ 
-Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
-  the frequency of G6pd-deficiency alleles. The size of the pie chart
-  indicates the number of G6pd-deficient individuals sampled.
-. Countries with
-  endemic malaria are colored yellow. Figure taken from
-  \citet{Howes-g6pd-variants}
-  \url{http://www.malariajournal.com/content/12/1/418}. \gc{Not sure
-    whether to include this.}
+Map of G6pd-deficiency allele frequencies across Asia. 
+The pie chart shows the frequency of G6pd-deficiency alleles. 
+The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
+Countries with endemic malaria are colored yellow. 
+Figure taken from \citet{Howes-g6pd-variants}
+\url{http://www.malariajournal.com/content/12/1/418}. 
+\gc{Not sure whether to include this.}
+\plr{I like it.}
 } \label{fig-G6PD-map}
@@
 %Selection coefficient (s) 	A- 0.044 (0.019–0.048)  Med	0.034
-%(0.014–0.049)  ``The mean age of the A− allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years.
+%(0.014–0.049)  ``The mean age of the A- allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years.
 For the Med alleles, the mean age was 3330 years with a 95\% credibility interval of 1600 to 6640 years"
@@
 %selection intensity of ~0.23 \citep{Louicharoen-g6pd}
-%Show's effect in het and homozy females
+%Shows effect in het and homozy females
+
+%%%%%%%%%%%% %%%%%%%%%%%%%%%
 
@@
 % $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
-Therefore, if we define $\chi$ to be the unique positive solution to
+Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
 \[
@@
 The explicit formula for $\chi$ is cumbersome; here we omit it.
-(This is in two dimensions; in one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
-This can be rewritten 
+(In one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
+Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
+we can rewrite this as
 \begin{equation} \label{eqn:defines_chi}
@@
 Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
-$\chi = \sqrt( -\log(s_d) / 2 s_b \rho \mu \pi ) + O(1/\sigma)$.
+$\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma)$.
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
@@
 
-XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi/\sqrt{\pi \lambda_0})^2$ then
+XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
 $f = 1-\gamma f^{3/2}$, where
@@
 Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
-and it is an elementary consequence of the Poisson process that
+and it is an property of competing Poisson processes that
 \begin{align} \label{eqn:pizero}
@@
 \end{align}
-\marginnote{put in value of $\chi$ computed above.}
 (In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)
 
-In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
-as a function 
-calculated for our G6pd parameters.
-
-In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
-calculated for our G6pd parameters.
+In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths as a function 
+of the G6pd parameters.
 
@@
 
-\subsection{Standing Variation -- Results} 
+%%%%%%%%%%% %%%%%%%%%%%%% %%%%%%%%%
+\subsection{Time to adaptation}
+
 The quantity that is perhaps easiest to compute is the mean time until adaptation.
 If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
-Then, as shown in Figure XX,
+Then, as can be seen from Figure~\ref{fig:cartoon},
 $\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
@@
 \]
-and so we have
+and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
 \begin{align}
@@
 For means of evaluting this integral, see Appendix \ref{apx:integrals}.
-
-In Figure \ref{G6PD_chartimes}
+In Figure \ref{G6PD_chartimes} we show the mean time until adaptation for various values of the parameters.
 
@@
 Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
-then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
+then $a_0$ is given by the proportion of the range occupied by standing variants divided by the mean density of unique standing variants,
 i.e.\ $a_0 = z_0 / \lambda_0$.
-We can compare this to the corresponding quantity $a_+$ for a new variant through the formula
-$a_0 / a_+ = z_0 / (1-z_0)$.
+We can solve for $a_+$, the corresponding mean area occupied by a given new variant, 
+using the formula $a_0 / a_+ = z_0 / (1-z_0)$.
 

### 2014-09-14 standing-variation-paper.tex
light pass through

@@
 %\fi
-%% NEED amnat.bst IN REPO
-% \bibliographystyle{amnat}
-\bibliographystyle{plainnat}
+
+\bibliographystyle{amnat}
 
@@
 
-XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
+XXX could do calculations as in "panmixia" here, e.g.\ if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
 $f = 1-\gamma f^{3/2}$, where
@@
 
-{\tt Simulations:} compare mean density against $s_d$ at different
-values of $\sigma$; omit variability if needed for readability. Compare to panmictic results.
+{\tt Simulations:} 
+compare mean density against $s_d$ at different values of $\sigma$; 
+omit variability if needed for readability. 
+Compare to panmictic results.
 
@@
 One interesting problem that we can seek to address with this work is
-the extent to which the degree of pleiotropy biases the 
+the extent to which the degree of pleiotropy biases 
 adaptation towards the repeated use of particular subset of loci
-(i.e. convergence at genetic level). 
-While many alleles may result in the benefical phenotype, many of
-these may be unlikely to contribute to adaptation due to their
-negative pleiotropic consequences. There are at least two ways that
-negative pleiotropy can contribute to high rates of
-convergence when adapting when a single change is sufficient for
-adaptation. 
-First, negative pleiotropic effects can reduce the overall beneficial
-selection coefficient of an allele in the new environment making them
-unlikely to become established and slow to spread (and in the worst
-case making them deleterious). This effect has been studied by a
-number of authors \citep{}. A second contribution of negative
-pleiotropy to shaping genetic convergence, is that alleles that have
-less negative pleiotropy are more likely to segregate before the
-environmental shift, and so get a head start in contributing to
-adaptation. This effect is seemingly less well studied, perhaps in
-part because many modeling approaches tend to use the strong-selection
-weak mutation limit and so ignore standing variation. \gc{Is this fair?}
+(i.e.\ convergence at genetic level). 
+While many alleles may confer the benefical phenotype, 
+not all will contribute to adaptation if they have negative pleiotropic consequences.
+There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
+when adapting if a single change is sufficient for adaptation. 
+First, negative pleiotropic effects can reduce 
+the overall beneficial selection coefficient of an allele in the new environment,
+making them unlikely to become established and slow to spread 
+(and in the worst case making them deleterious). 
+This effect has been studied by a number of authors \citep{}. 
+A second contribution is that alleles that have
+less negative pleiotropy are more likely to be present as standing variation 
+before the environmental shift.
+This effect seems less well studied, 
+perhaps because many modeling approaches use the strong selection, weak mutation limit 
+and so ignore standing variation. \gc{Is this fair?}
+\plr{well, even in the strong-selection weak-mutation limit you'll have standing variation}
 
 Here we focus primarily on the second effect. 
-Lets imagine for the moment that all classes of beneficial allele have
-the same beneficial selection coefficient. 
+Let's imagine for the moment that all classes of beneficial allele 
+have the same beneficial selection coefficient. 
 Each class of mutations could have its own mutation rate $\mu_j$ and
-have a disadvantage $s_{d,j}$ prior to the switch. As they have the
-same selection coefficient all of the waves travel outward at a rate
-$v$. 
-The the rate of type $k$ standing variants per unit area and the rate of de
-novo variants per unit area per generation are
+have a disadvantage $s_{d,j}$ prior to the switch. 
+As they have the same beneficial selection coefficient after the switch, 
+all of the waves travel outward at a rate $v$. 
+Then the density of type $k$ standing variants per unit area 
+and the input rate of \textit{de novo} variants per unit area per generation are, respectively,
 \begin{align}
-    \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } ; ~~      \lambda_{j} = \mu_j \rho s_b;
+  \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } \qquad \text{and} \qquad    \lambda_{j} = \mu_j \rho s_b .
 \end{align}
@@
 \begin{equation} \label{eqn-prop-space-allele-k}
-    p_k = \int_0^\infty  \big( 2 \lambda_{0,j} \pi v^2 t + \lambda_j
-    \pi v^2 t^3 \big)  \; \exp \left(\sum_k - \lambda_{0,k} \pi v^2 t^2 - \lambda_k \pi v^2 t^3 / 3 \right) dt .
+    p_k = \int_0^\infty  \left( 2 \lambda_{0,k} \pi v^2 t + \lambda_k \pi v^2 t^3 \right)  
+    \; \exp \left( - \sum_j ( \lambda_{0,j} \pi v^2 t^2 + \lambda_j \pi v^2 t^3 / 3 ) \right) dt .
 \end{equation}
@@
 \begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
-    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) }
+    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) } ,
 \end{equation}
-while if we only allow new variation, i.e. if all variants are highly
-deleterious before the environment switches, then $p_k =\mu_j / (\sum_k  \mu_k)$
+while if we only allow new variation, 
+i.e.\ if all variants are highly deleterious before the environment switches, 
+then $p_k =\mu_j / (\sum_k  \mu_k)$.
 
-To illustrate some of the properties of this model lets imagine the
-somewhat extreme scenerio that there is a single base-pair where
-a mutation change is relatively free of negative pleiotropy (class $1$), and a
-larger mutational target where changes have more serious pleiotropic
-consequences in the ancestral environment (class $2$). We set $s_{d,1}
-\leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, and think of the second
-class mutational target being ten, a hundred, or one thousand base
-pairs. We set  . We show the expected proportion of space covered by
-the rarer class $1$ mutations in Figure \ref{fig-pleiotropy_calc},
-assuming that both classes share a beneficial selection coefficient
-$s_b=0.05$ once the environment switches. 
-Intuitively the contribution of the rarer mutation decreases as the
+To illustrate some of the properties of this model, 
+let's imagine the somewhat extreme scenerio that there is a single base-pair 
+at which a possible mutation is relatively free of negative pleiotropy (call this class $1$), 
+and a larger mutational target where changes 
+have more serious pleiotropic consequences in the ancestral environment (class $2$). 
+We set $s_{d,1} \leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, 
+assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
+and think of the second class mutational target being ten, a hundred, or one thousand base pairs. 
+\plr{What is $s_{d,2}$ in figure 7?}
+We show the expected proportion of space covered by
+the rarer, class $1$ mutations in Figure \ref{fig-pleiotropy_calc}.
+Intuitively, the contribution of the rarer mutation decreases as the
 mutational target of the other class becomes larger, and as the
-difference in the negative pleiotropoc consequences of the two classes
-of alleles decreases. The case with standing variation only is the best
-case scenerio for the rarer mutation as we force them to have the same
-advantage in the new environment, so its rate of introduction is
-necessarily lower. However the standing variation only case does seems
-to provide a reasonable rule of thumb. Especially for parameter combinations, higher population densities and
-high dispersal distances, that increase the contribution of standing
-variation (and similarly for high $s_b$).
+difference in the negative pleiotropic consequences of the two classes of alleles decreases. 
+The case with standing variation only is the best case scenario 
+for the rarer mutation as we force them to have the same advantage in the new environment, 
+so its rate of introduction is necessarily lower. 
+However, the standing variation only case does seem to provide a reasonable rule of thumb. 
+Especially for parameter combinations, 
+higher population densities and high dispersal distances, 
+that increase the contribution of standing variation (and similarly for high $s_b$).
 
@@
 conditional on eventual fixation, it begins to spread spatially at speed $v$ instantly,
-effectively neglecting the time it must first spend fixing locally.
+effectively neglecting the time it must first spend escaping demographic stochasticity.
 In \citet{ralphcoop2010} we addressed this by noting that there would be no change at all in our results 
@@
 In this section we examine this assumption in more detail, although mostly through heuristic arguments,
-and also compare our results above to those of \citet{softsweeps}.
+and also compare our results above to the results without geographic structure of \citet{softsweeps}.
 
@@
 Then we know that there exists a random variable $W$ such that $\lim_{t\to\infty} e^{-s_b t} Z_t = W$ almost surely,
-so that if $\tau$ is the time $Z_t$ reaches size $\sigma^2 \rho$, i.e.\ $\tau = \inf\{ t \ge 0: Z_t > \sigma^2 \rho$,
+so that if $\tau$ is the time $Z_t$ reaches size $\sigma^2 \rho$, % i.e.\ $\tau = \inf\{ t \ge 0: Z_t > \sigma^2 \rho\}$,
 then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$.
@@
 but it is simpler to note that if the converse is true, 
-that the process is largely unaffected by spatial structure,
+the process is largely unaffected by spatial structure,
 and so the panmictic model is a good approximation for the true process.
@@
 
-\citet{softsweepsII}, it is shown that under a panmictic model with certain assumptions on the parameters,
+\citet{softsweepsII} show that under a panmictic model with certain assumptions on the parameters,
 the number of independent origins due to both standing variation and new mutation seen in a sample of size $n$
 has approximately the Ewens distribution with parameters $n$ and $\theta = 4 N \mu$.
-As $n$ increases, the total number of types seen grows as $\log n$, 
-so the ``mean density of types'' corresponding to our $\lambda_0 + \nu_+$ in their model is infinite.
+As $n$ increases, the total number of types seen grows as $\log n$.
+In our model, we can increase total population size by increasing either the density $\rho$ or the total amount of area.
+In either case, the predicted number of distinct types grows linearly with $n$,
+much faster than under panmixia.
+\plr{note: this had the wrong conclusion here previously}
 \marginnote{Graph: sigma against mean number of origins; different curves for spatial, panmictic (flat), and simulation;
@@
 Mention Monty's empirical speeds paper.}
-We could compare the probability of parallel adaptation under the two models (see eq'n 18 in \citet{softsweepsII}),
-but this seems less useful than an assessment of the degree of parallel adaptation.
+% We could compare the probability of parallel adaptation under the two models (see eq'n 18 in \citet{softsweepsII}),
+% but this seems less useful than an assessment of the degree of parallel adaptation.
 
@@
 Our results are predicated on the idea that adaptive variants are
-initially rare within populations, i.e. they are reasonably
+initially rare within populations, i.e.\ they are reasonably
 deleterious before the environment switches. In constrast, even distant populations could have
 a shared basis of adaptation if they adapt via common variation,
-e.g. previously neutral (or nearly neutral) variation, shared among populations. 
-Our ability to detect such shared convergent events is at the moment
-quite limited. If many loci contributing to variation in a trait, then
-selection on any one allele may be weak allowing them to segregate
-across populations. However, if that
-case populations separated by a reasonable amount of drift may still
+e.g.\ previously neutral (or nearly neutral) variation, shared among populations. 
+Our ability to detect such shared convergent events is at the moment quite limited. 
+If many loci contribute to variation in a trait, then
+selection on any one allele may be weak, allowing adaptation to use different alleels in different populations.
+However, in that case populations separated by a reasonable amount of drift may still
 adapt via different genetic routes, as the constellation of alleles
-best positioned to initially respond to positive selection, i.e. those at
-intermediate frequency, will be somewhat different. Therefore, it
-seems unlikely that polygenic traits are more suspectible to a shared
+best positioned to initially respond to positive selection, 
+i.e.\ those at intermediate frequency, will be somewhat different. 
+\plr{Not sure I follow that.  Ones at intermediate frequency will end up affecting the population mean less
+than ones at low frequency (but above the demographic stochasticity barrier)?}
+Therefore, it seems unlikely that polygenic traits are more suspectible to a shared
 genetic basis to adaptation across populations than simple traits.
@@
 \paragraph{Are species held together by wide-spread selective sweeps?}
-Our results touch on an old debate on the evolutionary coherence of
-species. Mayr and many others have argued that species are coherent
-evolutionary units as they are united by shared gene flow \citep[pages
-521–522 ][]{Mayr:SpeciesEvol}. However, this argument has
-questioned by a number of authors based on relatively high levels of 
-differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
-\citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
-may not be high enough to prevent neutral differentiation, species
-should be viewed as being united as levels of gene flow were still
-high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
-Indeed large scale genotyping and sequencing, along with more
-sophisticated methods, are highlighting ever more signals of gene flow
-between populations and species \citep{}.
+Our results touch on an old debate on the evolutionary coherence of species. 
+Mayr and many others have argued that species are coherent
+evolutionary units because they are united by shared gene flow \citep[pages 521–522 ][]{Mayr:SpeciesEvol}. 
+However, this argument has been questioned by a number of authors 
+based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
+\citet{Rieseberg2001} and \citet{MorjanRieseberg:04} argued that 
+even if gene flow is not be high enough to prevent neutral differentiation, 
+species should be viewed as being united 
+if levels of gene flow were still high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
+\plr{not clear which of three papers in the last sentence are being cited for what?}
+At present, large scale genotyping and sequencing projects, 
+along with more sophisticated methods, 
+are highlighting ever more signals of gene flow between populations and species \citep{}.
 However, our work on geographic convergent adaptation
-\citep[see also][]{ralph2010parallel,RalphCoop:14} brings the
-conclusion that this gene flow will lead to the rapid spread of
-selected alleles partially into
-question. Our results suggest that species should often adapt to
+\citep[see also][]{ralph2010parallel,RalphCoop:14} 
+concludes that this gene flow will lead to the rapid spread of
+selected alleles partially into question. 
+Our results suggest that species should often adapt to
 wide-spread selection pressures through convergent evolution rather
-than waiting for a single allele to migrate across the range over
-length-scales determined by our characteristic length.
+than waiting for a single allele to migrate across the range.
+% over length-scales determined by our characteristic length.
 
-In support of this idea recent selective sweeps are often
-geographically restricted \citep{}, but see XXX. This is likely in part due to the
-relatively low incidence of wide-spread selection pressures \citep{},
+In support of this idea, recent selective sweeps are often
+geographically restricted \citep{}, but see XXX. 
+This is likely in part due to the relatively low incidence of wide-spread selection pressures \citep{},
 but as noted above even when we know of wide-spread selection
-pressures (e.g. Malaria) the response is usually convergent across large enough
-spatial scales. This view is potentially contradicted by the
-introgression of adaptive alleles across species and sub-species
-boundaries, suggesting that selected alleles do sometimes spread
-despite low migration rates \citep{hedrick2013adaptive}. However, at least some of
-these cases likely reflect the introgression of haplotype complexes
-consisting of multiple tightly linked beneficial
-alleles.
+pressures (e.g.\ malaria) the response is usually convergent, not shared, across large spatial scales. 
+On the other hand, 
+introgression of adaptive alleles across species and sub-species boundaries, 
+suggests that selected alleles do sometimes spread despite low migration rates \citep{hedrick2013adaptive}.
+However, at least some of these cases 
+are likely caused by introgression of haplotype complexes
+consisting of many, tightly linked, beneficial alleles.  
+\plr{is the following making sense?}
+The important missing piece to these puzzles seems to be an understanding
+of when phenotypes are accessible to mutation to a given species,
+and why they may be accessible to one subspecies and not another.
 
-% \gc{Needed?:} This view is potentially contradicted by the
-% introgression of adaptive alleles across species and sub-species
-% boundaries, suggesting that selected alleles do sometimes spread
-% despute low migration rates \citep{}. However, at least some of
-% these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
-% alleles. Other cases of selective introgression
+% Other cases of selective introgression
 % are concurrent with populations spreading into new environments,
@@
 \paragraph{How then do substitutions occur?}
-If it is rare for gene flow to rapidly spread selected alleles across
-species range, how then do selected alleles ever become fixed within
-species? Repeated bouts of adaptation in particular genomic regions may act 
-push a subset of previously selected alleles to fixation across the
+If it is rare for gene flow to rapidly spread selected alleles across a species range, 
+how then do selected alleles ever become fixed within species? 
+Repeated bouts of adaptation in particular genomic regions may act 
+to push a subset of previously selected alleles to fixation across the
 species range, through the spread of the genetic backgrounds on which they arise.  
@@
 initial rapid spread of selected alleles.
+\plr{or, by drift?}
 
@@
 through subsequent large-scale changes in geographic range size
-induced by extirpation of the species over parts of its range. Such drops in range size may fix, or radically change the
-species-wide frequency, of alleles previously restricted to small portion of a species range. 
-Also many modes of speciation are proposed to occur through a
-geographically-limited subset of populations forming the basis of new species, e.g. the splitting off
-of part of the range through a vicariance event or dispersal of a
-subset of individuals. As such specation will lead to the geographic assorting of
+induced by extirpation of the species over parts of its range. 
+\plr{a.k.a. bottlenecks}
+Such drops in range size may fix, or radically change the species-wide frequency, 
+of alleles previously restricted to small portion of a species range. 
+Furthermore, many modes of speciation are proposed to occur through a
+geographically-limited subset of populations forming the basis of new species, 
+e.g.\ the splitting off of part of the range through a vicariance event 
+or dispersal of a subset of individuals. 
+Therefore, specation will cause geographic assortment of
 polymorphic ancestral variation, again acting to fix variants within
@@
 
-Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation,
-e.g. adaptation to local conditions, within species maybe ephemeral and
-subject to loss due to local extinction and the mixing following the collapse of population
-structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
-response to the pattern of punctuated equilibrium \citep{eldredgegould72}, arguing
-that the observation of stasis and rapid anagensis associated with
-speciation were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
+\plr{Seems like we should bring in the work on what happens if these locally fixed alleles cause an incompatibility. (and/or speculate about it)}
+
+Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation within species,
+e.g.\ adaptation to local conditions, 
+may be ephemeral and subject to loss due to local extinction and the mixing 
+following the collapse of population structure \citep{Futuyma:10,FUTUYMA:87}. 
+\plr{"to the pattern of"?}
+In part, this was a response to the pattern of punctuated equilibrium \citep{eldredgegould72}, 
+arguing that the observation of stasis and rapid anagensis associated with speciation 
+were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
 Futuyma argued that despite rapid adaptation over short time-scales, we
-may observe morphological stasis in the
-fossil record as
-much of this adaptation is lost \citep[see
-also][]{lieberman:96,Eldredge:05}. Furthermore, he suggests that speciation
+may observe morphological stasis in the fossil record as much of this adaptation is lost 
+\citep[see also][]{lieberman:96,Eldredge:05}. 
+Furthermore, he suggests that speciation
 may act as rachet to prevent the loss of differentiation, acting to
-maintain adaptive changes among populations, and prevent their loss by
-interbreeding. At face value the rate of species formation seems too low to
-contribute to this process. However, as \citep{Rosenblum:12}, and many
-others, have argued that in fact the rate of speciation may well be very high,
-but that the majority of incipient species do not persist long due to
-reabsorption or extinction. Repeated bouts of extinction and
-speciation will send waves of alleles
+maintain adaptive changes among populations, and prevent their loss by interbreeding. 
+At face value the rate of species formation seems too low to contribute to this process. 
+However, \citep{Rosenblum:12}, and many others, 
+have argued that the rate of speciation may well be quite high,
+but that the majority of incipient species do not persist long due to reabsorption or extinction. 
+Repeated bouts of extinction and speciation will send waves of alleles
 to fixation along particular lineages.
 
-While Futuyma and others have advanced these ideas mainly in the
-setting of adaptation to local conditions \citep[particularly local
-communities ][]{} . Much of the recent adaptation that we see as population
-geneticists is likely adaptation to local conditions, but as we have
-shown even wide-spread selection pressures will likely result in
-alleles only spreading locally. The adaptive substitution
-we see along lineages are the alleles that were lucky enough to
-have spread in populations that escaped local extinction, and to have
-avoided being assorted into species that eventually went extinct. 
-Range sizes change rapidly enough that it seems likely that range size
-changes should change substantially over the soujourn time of most
-alleles that will eventually become fixed in the species, as such it
-seems as though .
+Futuyma and others have advanced these ideas mainly in the
+setting of adaptation to local conditions 
+\citep[particularly local communities ][]{} . 
+Much of the recent adaptation seen by
+population genetics is likely adaptation to local conditions, 
+but as we have shown, even wide-spread selection pressures will likely result in 
+geographically restricted spread of alleles.
+The adaptive substitutions we see along lineages are those alleles 
+lucky enough to have spread in populations that escaped local extinction, 
+and to have avoided being assorted into species that eventually went extinct. 
+\plr{those last two things sound like the same thing}
+Ranges change sufficiently rapidly that
+range size could change substantially over the soujourn time of most alleles 
+that will eventually become fixed in the species, as such it seems as though .
 
@@
 This link between speciation and substitution does not imply that
-substitutions should necessarily be thought of as being clustered at splits in
-inferred phylogenies \citep[see ][for a recent exhange on
-this]{Pennell:14,Venditti:14,Pennell:14B}. 
-We first note that neutral substitutions will be to be unaffected by this process, because they
-will accumulate in a clock like manner along lineages, as dictated by
-the mutation rate, regardless of the geographic details of their
-polymorphic stage. Turning to the accumulation of adaptive
-substitutions, 
-it likely that splits in
-phylogenies are likely only a tiny proportion of all incipient
+substitutions should necessarily be thought of as being clustered at splits in inferred phylogenies 
+\citep[see ][for a recent exhange on this]{Pennell:14,Venditti:14,Pennell:14B}. 
+We first note that neutral substitutions are unaffected by this process, 
+because they accumulate in a clocklike manner along lineages, 
+as dictated by the mutation rate, 
+regardless of the geographic details of their polymorphic stage. 
+Turning to the accumulation of adaptive substitutions, 
+it likely that splits in phylogenies are likely only a tiny proportion of all incipient
 speciation events because extinction rates may be high
-\citep{Rosenblum:12}, such that every lineage has likely past through
+\citep{Rosenblum:12}, such that every lineage has likely passed through
 many ``speciation events'' in addition to the observed ones.
-Futhermore, under this view, spatial polymorphisms may accumulate gradually, in
-geographically restricted populations, between large-scale 
+Futhermore, under this view, 
+spatial polymorphisms may accumulate gradually in geographically restricted populations
+between large-scale 
 biogeographic events leading to their fixation (or indeed loss). 
@@
 Nor does this view imply that adaptive evolution or speciation is driven by the
-shifting balance or genetic revolutions \citep{Wright:32,
-  Mayr-genetic-revol:1954}, whereby genetic drift allows
+shifting balance or genetic revolutions 
+\citep{Wright:32, Mayr-genetic-revol:1954}, 
+whereby genetic drift allows
 populations to cross fitness valleys and substitute novel epistatic

### 2014-09-14 standing-variation-paper.tex
a bit more to discussion

@@
 
-The geographic area of Central and Eastern Asia with malaria is on the
-order of $\sim 10$ million $km^2$. In that area there are at least $15$
-common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
+The geographic area of Central and Eastern Asia with malaria is 
+on the order of ten million square kilomters.
+In that area there are at least $15$ common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
@@
 perhaps because many modeling approaches use the strong selection, weak mutation limit 
-and so ignore standing variation. \gc{Is this fair?}
-\plr{well, even in the strong-selection weak-mutation limit you'll have standing variation}
+and so ignore standing variation. 
+\gc{Is this fair?}
+\plr{Not sure what papers you are thinking of.}
 
@@
 To illustrate some of the properties of this model, 
-let's imagine the somewhat extreme scenerio that there is a single base-pair 
+let's imagine the somewhat extreme scenario that there is a single base-pair 
 at which a possible mutation is relatively free of negative pleiotropy (call this class $1$), 
@@
 
-\paragraph{Are species held together by wide-spread selective sweeps?}
+\paragraph{Are species held together by widespread selective sweeps?}
 Our results touch on an old debate on the evolutionary coherence of species. 
 Mayr and many others have argued that species are coherent
-evolutionary units because they are united by shared gene flow \citep[pages 521–522 ][]{Mayr:SpeciesEvol}. 
+evolutionary units because they are united by shared gene flow \citep[pages 521--522 ][]{Mayr:SpeciesEvol}. 
 However, this argument has been questioned by a number of authors 
@@
 % lessening the chance of convergent adaptation \citep{}.
+
+\plr{added:}
+Even if selective sweeps only bring alleles to fixation locally,
+they are still potentially a stronger homogenizing force
+than neutral mixing through migration.
+We can see this by comparing mean coalescent times.
+Under neutral mixing, the mean number of generations back to the most recent common ancestor
+is on order of the total population size.
+This quantity has not been worked out for a model with simultaneous local sweeps,
+but is close to the ``spatial $\Lambda$-Fleming--Viot'' models of \citet{barton2013modelling},
+in which local sweeps occur independently across the range.
+In these models, the mean number of generations back to the most recent common ancestor for distant samples
+scales with XXX,
+which can be substantially shorter.
+\plr{Maybe I'm forgetting something in their papers -- did they do this? --
+but I think it would be a little bit of work to figure out what parameters to put into their models.}
+Intuitively, we can obtain this as follows:
+suppose that coalescence only occurs during local sweeps,
+which occur at rate $\lambda$
+on a spatial scale such that there are on average $M$ independent origins of the sweep during each bout of adaptation.
+The motion of lineages backwards is not dissimilar to a stepping-stone model with $M$ sites
+that coalesce when they are in the same deme
+with probability dependent on the recombination distance to the sweeping, selected site.
+The mean hitting time of two independent random walks on a graph with $M$ nodes is of order $M$;
+to get the mean number of generations until coalescence, 
+we must divide this by the probability of remaining linked to the selected site as it sweeps
+and by the frequency of selected sites.
+Since $M$ scales proportionally to population density $\rho$,
+coalescent times do as well,
+but possibly with a much smaller coefficient
+(due effectively to an increase in reproductive variance).
 
@@
 spatial polymorphisms may accumulate gradually in geographically restricted populations
-between large-scale 
-biogeographic events leading to their fixation (or indeed loss). 
+between large-scale biogeographic events leading to their fixation (or indeed loss). 
 Such that the origination process of alleles and their fixation process

### 2014-09-14 standing-variation-paper.tex
edits to g6pd

--- 2014-09-13 standing-variation-paper.tex
@@
 %\fi
-%% NEED amnat.bst IN REPO
-% \bibliographystyle{amnat}
-\bibliographystyle{plainnat}
+\bibliographystyle{amnat}
+%\bibliographystyle{plain}
 
@@
 
-There are an increasing number of examples in which
-different populations within species have
-adapted to similar environments via independent genetic changes. 
+There is increasing number of examples of different populations within species often
+adapt to shared environments via independent genetic changes. 
 In some cases this convergent evolution leads to differing phenotypes
@@
 \citep{} in other cases the degree of convergence is precise down the
-same nucleutide change. 
-Such convergent evoltuion within populations has been seen for many
-carefully studied phenotypes in a range of species, 
-including drug resistance, pathogen resistance, insecticide resistance, and pigmentation. 
-In many of these examples we 
-need to be careful about whether the changes are truly convergent or
+same nucleutide change. Such convergent evoltuion within populations
+has been seen for many
+carefully studied phenotypes in a range of species, including drug resistance, pathogen resistance, insecticideresistance, and pigmentation. In many of these examples we 
+need to be careful about whether they changes are truely convergent or
 whether selection pressures differ subtly. 
-We note that we treat the phrases ``convergent'' and ``parallel'' evolution
+We note that we treat the phrases convergent and parallel evolution
 synonymously, as we are concerned with adaptation within a single
-species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.
+species that can occur via a different or shared genetic routes \citep{Arendt:08}.
 
 The issue of convergent adaptation within species touches on a number
-of important questions in evolutionary biology. 
-These include the extent to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
-whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
-and to what degree should species be regarded to cohesive units. 
-Convergent evolution within populations 
-also affects our ability to detect adaptation from population genomic data,
-since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.
+of important questions in evolutionary biology. These include the extent
+to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
+whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, and whether
+species should be regarded to cohesive units. Convergence evolution within
+populations also effects our ability to detect adaptation from population genomic data as no one
+allele sweeps fixation over the entire area affected by a selection
+pressure \citep{softsweepsIII}.
 
 Convergent evolution can occur even within a well mixed population
-subject to a constant selection pressure, 
-either through selection on multiple mutations present as standing variation within population
-before selection pressures switch \citep{Orr:01,softsweepsI},
-or due to multiple adaptive alleles that arise after selection pressures switch \citet{softsweepsII}. 
-The previous work has shown that 
+subject to a constant selection pressure, either through selection on
+multiple mutations present as standing variation within population
+before selection pressures switch \citep{Orr:01,softsweepsI} or due to
+multiple adaptive alleles arising after selection pressures switch \citet{softsweepsII}. 
+This previous work has shown that 
 a primary determinant of the probability that multiple alleles contribute to adaptation
@@
 
-\plr{should stick to either ``parallel'' or ``convergent'' perhaps}
-
-Geographically limited dispersal also increases the chance of convergent evolution. 
-For example, 
-patchy geographically selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
-since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
-In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
+Geographically limited dispersal also increases the chance of
+convergent evolution. For example when selection pressures are
+geographically patchy this can increasing the probability of
+convergent evolution as alleles are unable to spread through
+intervening populations \citep{RalphCoop:14}. In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
 where the selection pressure was constant across the entire environment, 
 and that a single mutational change was sufficient to adapt populations to the novel environment. 
-We assumed that there was no standing variation for the adaptive allele, 
-so that parallel mutation must be due to multiple mutations after the environmental shift. 
-In that setting, 
-we found that there was a characteristic geographic scale over which we would expect 
-multiple instances of our adaptive allele to have arisen in parallel. 
-This characteristic length could be expressed in terms of a simple compound
-parameter determined by our parameters of interest. 
-Here we extend this spatial model to include standing variation 
-present at mutation-selection balance before the selection pressures switch.
+We assumed that there was no standing variation for the adaptive
+allele, such that parallel mutation must be due to multiple mutations
+after the environmental shift. In that setting we found that there was
+a characteristic geographic scale, over which we would expect multiple
+instances of our adaptive allele to have arisen in parallel. This
+characteristic length could be expressed in terms of a simple compound
+parameter determined by our parameters of interest. Here we extend
+this spatial model include standing
+variation present at mutation-selection balance before selection
+pressures switch.
 
-%%%%%%%%% %%%%%%%%%%%% %%%%%%%
 \subsection{The set-up}
@@
 We assume that the species range
-is a large, homogenous, one- or two-dimensional region. 
+is a homogenous, one- or two-dimensional region. 
 There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
-We assume that separately arising mutations are distinguishable --
+We assume that different mutations are distinguishable --
 either as selectively equivalent mutations, or by linked variation.
-We also suppose that 
-the mutated type has been at a selective disadvantage 
-for a sufficiently long enough time in the past to be at selection-mutation equilibrium,
+For a long time in the past, the mutated type has been at a selective disadvantage (long enough to be at selection-mutation equilibrium),
 but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
@@
 The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
-(We leave aside the case where mutations are recessive, however.)
+(If the mutation is recessive, XXX more complicated things happen.)
 As for the other parameters,
 suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
-and that the mean squared distance between parent and child is $\sigma^2$.
+and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
 The species occupies an area with mean density
@@
 \paragraph{Rates of origination of standing and new mutations}
-We make use of the commonly used approximation that neglects competition between close relatives,
+We make use of the commonly used approximation that neglects competition between relatives,
 treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
@@
 is well--approximated by a Poisson process in space and time.
-The rate of this Poisson process,
-i.e.\ the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
+This the rate of this Poisson process, 
+the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type
 is approximately $\lambda = 2 \mu \rho s_b$.
 
-\plr{the following needs better explanation, maybe a figure?}
-Before $t=0$, on the other hand, the mutation is deleterious.
-We have assumed that $s_d$ is not too small
-so that the genetic descendants of each new mutation are (with high probability) doomed to extinction,
+Before $t=0$, on the other hand, the mutation is deleterious, so the
+genetic descendants of each new mutation are (with high probability)
+doomed (doomed I say!) to extinction,
 but may persist for some time.
-Since the times and locations of mutants before $t=0$ also well--approximated by a Poisson process with rate $\mu \rho$,
+Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
 the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
-by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
-% (In principle, more than one member of a mutant family may fix,
-% but we assume these are close together enough that we may treat them as beginning from a single point.)
+by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
 If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
@@
 equivalent to the mutants arising later.
-(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
-and the spatial distribution of each is small relative to the spread between the families.)
+(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
+and the spatial distribution of the family is small relative to the spread between them.)
 To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
@@
 \end{align}
-Note that this is closely related to the equilibrium frequency of a
+Note that this is closely related to the standard frequency of a
 allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}
@@
 \paragraph{Geographic spread of alleles} 
-Once an allele has become locally established it can begin to spread across space. 
-We assume that the allele, once established, 
-quickly settles down to spread spatially as a traveling wave of constant speed. 
-The behavior of this wave of advance of a beneficial allele was first described by 
-\citet{fisher1937wave} and \citet*{KPP1937}. 
-Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
-See \citet{ralph2010parallel} for a fuller review of these travelling waves.
-Note that the speed of the wave will vary
+Once an allele has become local established it can begin to spatially spread. 
+We assume that the allele, having become established, quickly settles
+down to spread spatially as a traveling wave of constant speed. The behavior of
+this wave of advance of a beneficial allele was first described by 
+\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
+conditions the speed of advance
+of this wave will be $v = \sigma \sqrt{2s}$. See
+\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
+applies to this scenerio. Note that the speed of the wave will vary
 with details of the space that individuals migrate across
-\citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.
-% However, the velocity will usually be primarily determined by the product of $\sigma^2$ and $s$.
+\citep[e.g. see ][ for comparisons to migration on discrete
+grids]{Slatkin-speed:76,SlatkinCharlesworth:78}, however, the velocity
+will usually be primarily determined by the product $\sigma^2$ and $s$.
 
 \paragraph{Putting it together.}
-Now, we can put these ingredients together for a simple model of the geographic spread of alleles. 
-A cartoon example of this process is shown in Figure \ref{fig:cartoon}. 
-Initially, when the selection pressures change at $t=0$, 
-a set of standing variants can start to spread having escaped loss through drift. 
-These variants are shown as little lightning bolts, and occur at a density $\lambda_0$ across space. 
-They spread at velocity $v$, carving out cones in space-time. 
-As these alleles proceed in their geographic spread, other new alleles can arise and become established in parallel, 
-these are shown as little stars. 
-These new mutations arise and become established at rate $\lambda$.
+So we can put these ingredients together for a simple model of the
+geographic spread of alleles. A cartoon example of this process is
+shown in Figure \ref{fig:cartoon} Initially, when the selection
+pressures change at $t=0$, a set of standing variants can start to
+spread having escaped loss through drift. These variants are shown as
+little lightning bolts, and occur at a density $\lambda_0$ across
+space. They spread at velocity $v$ carving out cones in space-time. As these
+alleles proceed in their geographic spread other new alleles can arise
+and become established in parallel, these are shown as little
+stars. These new mutations arise and become established at rate $\lambda$.
 
 As we previously outlined in \citet{ralph2010parallel} this model of
-geographic convergent evolution, when $\lambda_0=0$, is analogous to the a model of crystallization due to Kolmogorov \citet{kolmogorov-crystallization}.
-In this model, nucleation points form at random at a constant rate in time and space 
-and initiate the radial growth of new crystals. 
-Eventually, the different orientations of crystals form a random tessellation of space,
-hose properties have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
+geographic convergent evolution, when $\lambda_0=0$, is analogous to the \citet{kolmogorov-crystallization}
+model of crystallization. In \citet{kolmogorov-crystallization} model nucleation points form at randomly at
+a constant rate in time and space 
+and initiate the radial growth of new crystals. The crystals at the
+end point of this process form a random tessellation covering space
+the properties of which have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
 The generalized version of this process, for non-constant wave speeds
-and inhomogeneous Poisson processes is termed the 
+and imhomogeneous Poisson processes is termed the 
 Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
 Our combined process with both standing variation and new mutation is
-a special case of the KJMA tessellation, 
-where the spatial-temporally homogeneous Poisson origination process of new mutations, 
-is supplemented by a single pulse of origination points at time zero with spatial density $\lambda_0$.
-(For the purposes of analogy, we could imagine that before time $t=0$
-the temperature is high enough that nucleation sites do not persist.)
+a specially case of the KJMA tessellation, where the
+spatial-temporally homogeneous
+Poisson origination process of new mutations, is supplemented by a
+single pulse of origination points at time zero with spatial density $\lambda_0$.
 
-If we only allow standing variation, then everything about the process is relatively simple. 
-All of the points in space closer to a particular origination point will be
-reached by that allele before any other allele has a chance to spread there. 
-Such a tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
-(i.e.\ the cells formed by assigning points to the nearest point in a Poisson process).
-The properties of the patterning of space by alleles is
-determined by the spatial locations of the initiation points, 
-as specified by the spatially homogeneous Poisson process, 
-independent of the rate of spatial spread. 
-Therefore, the properties of the spatial tessellation with standing variation depends only on $\lambda_0$.
+If we only allow standing variation then everything about the process
+is relatively simple. All of the
+points in space closer to a particular origination point will be
+reached by that allele before any other
+allele has a chance to spread there. Formally then the spatial patterning,
+given a set of origination points, would be specified by the Voronoi
+tessellation (i.e. the cells formed by assigning points to their
+nearest origin). All of the properties of the patterning of space by alleles is
+determined by the spatial locations of the initiation points, as
+specified by the spatially homogeneous Poisson process, independent of
+the rate of spatial spread. So that the properties
+of the spatial tessellation with standing variation depends only on
+$\lambda_0$, and this
+model is commonly called the Poisson-Voronoi tesselation.
 
-% In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
-% speeds, as can arise due to fat-tailed dispersal kernels. 
-% Here we restict ourselves to constant wave speeds for simplicity but similar
-% calculations could be done by expansion of the results of the previous paper.
+In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
+speeds, as can arise due to fat-tailed dispersal kernels. Here we
+restict ourselves to constant wave speeds for simplicity but similar
+calculations could be done by expansion of the results of the previous paper.
 
@@
 
-%%%%%%%%% %%%%%%%%%%%
-
 \subsection{G6PD example}
-
-Below we describe a number of properties of our process, but we will first introduce 
-a motivating example to provide some concrete
+We will describe a number of properties of our process, but we first
+want to introduce a motivating example to provide some concrete
 numbers to use for illustrative purposes.
 
-Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
-and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
-A number of these alleles appear to be examples of convergent adaptation, 
-as different derived mutations in the same gene are seen in different individuals.
-For example, 
-a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
-and the sickle cell allele may plausibly have arisen by up to five independent occurrences
-of the same base pair mutatoin at different locations within Africa
+Over the past ten thousand years malaria resistance alleles have arisen at a number
+of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
+A number of these alleles appear to represent convergence, with
+multiple changes in the same gene. 
+For example, there are multiple
+changes at the $\beta$-globin gene that confers malaria resistance,
+and the sickle cell allele may plausibly have arisen by up to five distinct
+changes at the same base pair within Africa
 \citep{Flint:98,ralph2010parallel}.
 Another particularly impressive case of convergent evolution in response to
-selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
-with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
-that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
-These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
-rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
-Three G6PD deficiency alleles that also confer resistance to malaria 
-are particularly common and relatively well studied: 
-the $A-$ allele found in much of sub-Saharan Africa; 
-the Med allele found in the Mediterranean and Middle East; 
-and the Mahidol allele found Myanmar and Thailand.
+selection pressures imposed by malaria are the numerous changes throughout the
+X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
+described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
+found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
+exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
+resistance alleles at G6PD are particularly common and relatively well
+studied: the $A-$ found in
+much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
+east; and the Mahidol allele found Myanmar and Thailand.
 The $A-$ and Mahidol alleles have been shown
 to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
-hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
-Haplotype-based analysis of genetic diversity surrounding
-the $A-$, Med, and Mahidol suggest that they have spread over the past 
-few thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
-consistent with the age of other known malaria resistance alleles. 
-Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
-selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
-consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
-\plr{what are Ruwende's results? feels a bit vague here}
-Given such a strong pressure on these alleles they should have risen quickly to fixation, 
-so their presence at intermediate frequency over a broad geographic area
-makes them good candidates for recently balanced polymorphisms. 
+hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd,
+Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
+the $A-$, Med, and Mahidol suggest that they have spread in past few
+thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
+consistent with the age of other malaria resistance alleles. 
+These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
+selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
+the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
+$A-$ allele. Our $s_b$ reflects the selection coefficient averaged
+across heterozygote females and hemizygote males. \\
+
+Given such a strong pressure on these alleles they should have risen
+quickly to fixation, so their presence at intermediate frequency,
+over a broad geographic area, makes it a good candidate for a recently
+balanced polymorphism. 
 Hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to a variety of compounds, 
-notably those present in fava beans.
-
-Note that although originally the ``wave of advance'' was presented as the spread of an unconditionally favorable allele
-\citep{fisher1936,KPP},
-the same model holds in the case of heterozygote advantage \citep{aronson1975nonlinear}, 
-with the selected allele spreading locally to the equilibrium frequency (rather than fixation),
-with speed determined by the advantage of heterozygotes when rare.
+leading to hemolytic anemia when exposed to a variety of compounds, notably
+those present in fava beans (although, It is unclear if this alone is enough of a
+selection pressure to maintain the allele in the face of its
+considerable advantage). 
+The wave of advance works in the case of heterozygote advantage
+\citep{aronson1975nonlinear}, with the selected allele spreading
+locally to the equilibrium frequency (rather than fixation). As such
+our framework is applicable to the spread of G6PD (again assuming that
+$s_b$ is selection coefficient in heterozygotes/hemizygotes).
+We assume that before malaria became prevalent, the G6PD deficiency alleles suffered a
+decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and
+hemizygote male. Assuming that the underlying causes of this drop in fitness have
+not changed, we estimate that $s_d$ has to be roughly
+XXX to XXX, in order to have resulted in the equilibrium frequency
+seen today.
 
@@
 order of $\sim 10$ million $km^2$. In that area there are at least $15$
-common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
-Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
-The coding region of G6pd is 515 codons long, 
-and around 140 distinct deficiency alleles have been observed. 
-Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
-we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
-The dispersal and demographic parameters of humans in the past few thousans years is unclear,
-particularly as we are concerned with the ``effective'' population density
-(i.e.\ population density divided by variance in offspring number).
-We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
-and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
-Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
-\citep[see ][for a recent discussion]{}, 
-so these parameters only provide a rough view of the process.
+common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). So the average width of
+an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
+The coding region of G6pd is 515 codons long, with $\sim$140
+functional, distinct variants that cause reduced G6PD activity. Assuming a
+mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
+$\mu \approx 1 \times 10^{-6}$ a generation. 
+The dispersal and demographic parameters of humans in the past few
+thousans years is unclear,
+particularly as we are concerned with the effective population
+density. 
+We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
+We use $\sigma=10,~50,$ and $100$ as our dispersal parameter. 
+We note that obviously human migration has been
+shaped both by local dispersal and larger-scale expansions \citep[see
+][for a recent discussion]{}, so these parameters are for
+illustrative purposes only.
 
@@
 \caption{ 
-Map of G6pd-deficiency allele frequencies across Asia. 
-The pie chart shows the frequency of G6pd-deficiency alleles. 
-The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
-Countries with endemic malaria are colored yellow. 
-Figure taken from \citet{Howes-g6pd-variants}
-\url{http://www.malariajournal.com/content/12/1/418}. 
-\gc{Not sure whether to include this.}
-\plr{I like it.}
+Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
+  the frequency of G6pd-deficiency alleles. The size of the pie chart
+  indicates the number of G6pd-deficient individuals sampled.
+. Countries with
+  endemic malaria are colored yellow. Figure taken from
+  \citet{Howes-g6pd-variants}
+  \url{http://www.malariajournal.com/content/12/1/418}. \gc{Not sure
+    whether to include this.}
 } \label{fig-G6PD-map}
@@
 %Selection coefficient (s) 	A- 0.044 (0.019–0.048)  Med	0.034
-%(0.014–0.049)  ``The mean age of the A- allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years.
+%(0.014–0.049)  ``The mean age of the A− allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years.
 For the Med alleles, the mean age was 3330 years with a 95\% credibility interval of 1600 to 6640 years"
@@
 %selection intensity of ~0.23 \citep{Louicharoen-g6pd}
-%Shows effect in het and homozy females
-
-%%%%%%%%%%%% %%%%%%%%%%%%%%%
+%Show's effect in het and homozy females
 
@@
 % $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
-Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
+Therefore, if we define $\chi$ to be the unique positive solution to
 \[
@@
 The explicit formula for $\chi$ is cumbersome; here we omit it.
-(In one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
-Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
-we can rewrite this as
+(This is in two dimensions; in one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
+This can be rewritten 
 \begin{equation} \label{eqn:defines_chi}
@@
 Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
-$\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma)$.
+$\chi = \sqrt( -\log(s_d) / 2 s_b \rho \mu \pi ) + O(1/\sigma)$.
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
@@
 
-XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
+XXX could do calculations as in "panmixia" here, e.g. if $f = (\chi/\sqrt{\pi \lambda_0})^2$ then
 $f = 1-\gamma f^{3/2}$, where
@@
 Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
-and it is an property of competing Poisson processes that
+and it is an elementary consequence of the Poisson process that
 \begin{align} \label{eqn:pizero}
@@
 \end{align}
+\marginnote{put in value of $\chi$ computed above.}
 (In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)
 
-In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths as a function 
-of the G6pd parameters.
+In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
+as a function 
+calculated for our G6pd parameters.
+
+In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
+calculated for our G6pd parameters.
 
@@
 
-%%%%%%%%%%% %%%%%%%%%%%%% %%%%%%%%%
-\subsection{Time to adaptation}
-
+\subsection{Standing Variation -- Results} 
 The quantity that is perhaps easiest to compute is the mean time until adaptation.
 If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
-Then, as can be seen from Figure~\ref{fig:cartoon},
+Then, as shown in Figure XX,
 $\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
@@
 \]
-and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
+and so we have
 \begin{align}
@@
 For means of evaluting this integral, see Appendix \ref{apx:integrals}.
-In Figure \ref{G6PD_chartimes} we show the mean time until adaptation for various values of the parameters.
+
+In Figure \ref{G6PD_chartimes}
 
@@
 Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
-then $a_0$ is given by the proportion of the range occupied by standing variants divided by the mean density of unique standing variants,
+then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
 i.e.\ $a_0 = z_0 / \lambda_0$.
-We can solve for $a_+$, the corresponding mean area occupied by a given new variant, 
-using the formula $a_0 / a_+ = z_0 / (1-z_0)$.
+We can compare this to the corresponding quantity $a_+$ for a new variant through the formula
+$a_0 / a_+ = z_0 / (1-z_0)$.
 

### 2014-09-17 standing-variation-paper.tex

--- 2014-09-14 standing-variation-paper.tex
@@
 consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
-\plr{what are Ruwende's results? feels a bit vague here}
-Given such a strong pressure on these alleles they should have risen quickly to fixation, 
-so their presence at intermediate frequency over a broad geographic area
-makes them good candidates for recently balanced polymorphisms. 
+\plr{what are Ruwende's results? feels a bit vague here} \\
+
+Given such a strong pressure on these alleles they should have risen
+quickly to fixation, so their presence at intermediate frequency,
+over a broad geographic area, makes it a good candidate for a recently
+balanced polymorphism. 
 Hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to a variety of compounds, 
-notably those present in fava beans.
-
-Note that although originally the ``wave of advance'' was presented as the spread of an unconditionally favorable allele
-\citep{fisher1936,KPP},
-the same model holds in the case of heterozygote advantage \citep{aronson1975nonlinear}, 
-with the selected allele spreading locally to the equilibrium frequency (rather than fixation),
-with speed determined by the advantage of heterozygotes when rare.
+leading to hemolytic anemia when exposed to a variety of compounds, notably
+those present in fava beans (although, It is unclear if this alone is enough of a
+selection pressure to maintain the allele in the face of its
+considerable advantage). 
+The wave of advance works in the case of heterozygote advantage
+\citep{aronson1975nonlinear}, with the selected allele spreading
+locally to the equilibrium frequency (rather than fixation). As such
+our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
+We assume that before malaria became prevalent, the G6PD deficiency alleles suffered a
+decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and
+hemizygote male. Assuming that the underlying causes of this drop in fitness have
+not changed, we estimate that $s_d$ has to be roughly
+XXX to XXX, in order to have resulted in the equilibrium frequency
+seen today.
 

### 2014-09-22 standing-variation-paper.tex
changes throughout, added G6PD_chartimes_sd_sb.pdf

@@
 present at mutation-selection balance before the selection pressures switch.
+
+We show that convergent adaptation within wide-spread species is likely to be
+common, as has already been seen for a number of traits. 
+On the basis of this we argue, in our discussion, that the genetics of
+adaptation may often be
+geographically local, and that we should expect to see relatively few
+wide-spread sweeps even when selection pressures are wide-spread. 
+We discuss some implications, and history, of this view for the evolutionary
+coherence of species and molecular evolution.
 
@@
 
-\plr{the following needs better explanation, maybe a figure?}
-Before $t=0$, on the other hand, the mutation is deleterious.
-We have assumed that $s_d$ is not too small
-so that the genetic descendants of each new mutation are (with high probability) doomed to extinction,
-but may persist for some time.
-Since the times and locations of mutants before $t=0$ also well--approximated by a Poisson process with rate $\mu \rho$,
+Before $t=0$, on the other hand, the allele is deleterious.
+The genetic descendants of each new mutation are (with high probability) doomed to extinction,
+but may persist for some time (note that we have assumed that $s_d$ is not too small).
+Since the times and locations of mutants before $t=0$ also
+well--approximated by a 
+Poisson process with rate $\mu \rho$,
 the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
@@
 
-\plr{XXX More discussion on whether this is a good approximation?  
-Compute typical size and radius of a standing cluster?}
-
 \paragraph{Geographic spread of alleles} 
@@
 \citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.
-% However, the velocity will usually be primarily determined by the product of $\sigma^2$ and $s$.
 
@@
 Therefore, the properties of the spatial tessellation with standing variation depends only on $\lambda_0$.
-
-% In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
-% speeds, as can arise due to fat-tailed dispersal kernels. 
-% Here we restict ourselves to constant wave speeds for simplicity but similar
-% calculations could be done by expansion of the results of the previous paper.
 
@@
 Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
-selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
-consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
-\plr{what are Ruwende's results? feels a bit vague here} \\
+selection coefficient of $0.05-0.3$
+\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
+This is in agreement with the selection coefficients calculated by
+\citet{Ruwende-g6pd} on the basis of the present day levels of
+resistance to malaria due to the $A-$ allele. \\
 
@@
 \url{http://www.malariajournal.com/content/12/1/418}. 
-\gc{Not sure whether to include this.}
-\plr{I like it.}
 } \label{fig-G6PD-map}
@@
 Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
-\[
+\begin{equation}
     \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
-\]
+\end{equation}
 then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
@@
 Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
-$\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma)$.
+\begin{equation} \label{eqn:chi_standing}
+\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma).
+\end{equation}
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
 then the characteristic length is close to the value from \citet{ralphcoop2010}:
-$\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d))$.
-
-XXX could do calculations as in "panmixia" here, e.g.\ if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
-$f = 1-\gamma f^{3/2}$, where
-$\gamma = (4 \pi^{5/2} / \sqrt{2}) \mu^2 s_b^{3/2} \rho^2 / \sigma \log(1/(1-s_d))$.
-Then if $\gamma \ll 1$, $f \approx 1$, while if $\gamma \gg 1$, then $f \approx 0$.
-
+\begin{equation} \label{eqn:chi_new}
+\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d)).
+\end{equation}
+Theses two end points help build our intuition for the interaction of
+parameters in shaping the geographic scale of convergent evolution.
 By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
@@
  Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
-we expect to find one mutational origin;
-we denote the probability that this mutation is from standing variation by $\pi_0$.
-Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
-and it is an property of competing Poisson processes that
-\begin{align} \label{eqn:pizero}
-    \pi_0 = \lambda_0 \pi \chi^2 .
-\end{align}
-(In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)
-
-In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths as a function 
-of the G6pd parameters.
+ we expect to find one mutational origin.
+
+%XXX could do calculations as in "panmixia" here, e.g.\ if $f = (\chi \sqrt{\pi \lambda_0})^2$ then
+%$f = 1-\gamma f^{3/2}$, where
+%$\gamma = (4 \pi^{5/2} / \sqrt{2}) \mu^2 s_b^{3/2} \rho^2 / \sigma \log(1/(1-s_d))$.
+%Then if $\gamma \ll 1$, $f \approx 1$, while if $\gamma \gg 1$, then $f \approx 0$.
+
+%%%GRAHAM: I commented this out. We return to do proportion standing
+%%%below, so perhaps nicer to restrict that to one place. 
+% By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
+% and occur within the first $\chi/v$ generations.
+% Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
+% we expect to find one mutational origin;
+% we denote the probability that this mutation is from standing variation by $\pi_0$.
+% Therefore, $\pi_0$ is roughly the proportion of establishing mutations that come from standing variation,
+% and it is an property of competing Poisson processes that
+% \begin{align} \label{eqn:pizero}
+%     \pi_0 = \lambda_0 \pi \chi^2 .
+% \end{align}
+% (In one dimension, $\pi_0 = 2 \lambda_0 \chi$.)
+
+In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb} we show the range of characteristic lengths as a function 
+of various parameters chosen to match the evolution of malaria
+resistance at G6pd. 
+These curves match our intuition that population densities result in
+smaller characteristic lengths (as would higher mutation rates). 
+Unsurprisingly, allowing standing variation increases the mutational
+input and so decreases the characteristic length below that predicted
+by only new mutations (i.e. eqn. \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
+As the prior deleterious effect of the allele ($s_d$), the
+characteristic length increases untill it reaches that predicted by
+new mutation alone. Higher dispersal distances lead to larger
+characteristic lengths, as alleles spread geographically more rapidly blocking others
+from arising. A larger selective advantage acts in two conflicting
+ways: aiding the rapid geographic spread of established alleles, and also help more independent copies
+escape drift and become established. The effect on helping establish
+alleles wins out in this as increasing the selective $s_b$ decreases
+the characteristic length. This effect strongest when only standing
+variation contributes ( $s_b^{-1/2}$, eqn \eqref{eqn:chi_standing}), 
+as in that case the speed of spread does not matter, the dependence
+when only new mutations contribute is much weaker ( $s_b^{-1/6}$, eqn \eqref{eqn:chi_new})
+Overall, the range of characteristic lengths observed are reasonably
+consistent with the average diameter of a G6PD variant in Eurasia of
+$800$, especially for the lower population density.
+
+\begin{figure}[ht]
+\begin{center}
+  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
+\caption{ %
+{\bf Characteristic length,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+} \label{Fig-G6PD-charlength}
+\end{center}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
-\caption{ %
-{\bf Characteristic length,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
-} \label{Fig-G6PD-charlength}
-\end{center}
-\end{figure}
-
 % \begin{figure}[ht]
@@
 The quantity that is perhaps easiest to compute is the mean time until adaptation.
-If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
+If fix ourselves at some geographic location, and let $\tau \ge 0$ be the time at which the point is reached by the advantageous mutations.
 Then, as can be seen from Figure~\ref{fig:cartoon},
@@
 Since we assume these are a Poisson process, 
-\[
+\begin{equation}
     \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
-\]
+\end{equation}
 and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
@@
 For means of evaluting this integral, see Appendix \ref{apx:integrals}.
-In Figure \ref{G6PD_chartimes} we show the mean time until adaptation for various values of the parameters.
+
+In Figure \ref{G6PD_chartimes} we show the mean time until adaptation
+for various values of the parameters chosen to match the case of
+adaptation at G6PD. Increasing $\sigma$ and decreasing $s_d$ lower 
+the time to adaptation, as alleles spread geographically more quickly
+and are present as standing variation more often
+respectively. Increasing $s_b$ strongly decreases the time to
+adaptation, as it both causes more alleles to escape drift and to
+rapidly spread. Given that the G6PD alleles likely spread over a few
+thousand years, i.e. less than a few hundred generations. This
+time-scale seems quite plausible given Figure \ref{G6PD_chartimes},
+except perhaps for the lowest dispersal distances.
 
@@
 \begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_chartimes}   %{charlen-by-sd-limit}
+  \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
 \caption{ %
-{\bf Mean adaptation times,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+{\bf Mean adaptation times}, as a function of selective disadvantage
+and , compared to the corresponding quantity without standing variation, and to the quantity only considering standing variants.
 } \label{G6PD_chartimes}
@@
 
-We have defined $\lambda_0$ to be the mean density of standing variants that reach local fixation.
-Define $\nu_+$ to be the mean density of new mutants whose offspring fix locally.
+\subsection{The contribution of standing variation.}
+We can use our framework to address how much adaptation arises from
+standing mutation. We have defined $\lambda_0$ to be the mean density
+of standing variants that are present and escape loss when the
+environment shifts.
+We will define $\nu_+$ to be the mean density of newly arising alleles
+that spread --having arisen in an area free of other adaptive alleles.
 Since the probability that a mutant arising at $(x,t)$ is lucky enough to be born in a location not already occupied by mutants
@@
 \end{equation}
-Since the mean proportion of patches that come from standing variation is $\lambda_0 / (\lambda_0 + \nu_+)$,
+Therefore, the mean proportion of patches that come from standing variation
+is 
+\begin{equation} \label{prop_patches_standing}
+\lambda_0 / (\lambda_0 + \nu_+),
+\end{equation}
 this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
@@
 is
-\begin{equation}
+\begin{equation} \label{prop_space_standing}
     z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt .
 \end{equation}
-\marginnote{put in this value in terms of Gaussian CDF?}
 To evaluate this integral, again see Appendix \ref{apx:integrals}.
+%\marginnote{put in this value in terms of Gaussian CDF?}
 
@@
 \caption{ %
-{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the G6PD example.  XXX say more.
+{\bf Mean proportion of patches arising from standing variation.} The
+parameters given are for the G6PD example.
+The left and right panels show the expected proportion of patches that
+arise from standing variation and the expected proportion of
+geographic space that is covered by adaptation from standing variation
+(eqn. \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
 } \label{G6PD_standing_var_proportion}
@@
 
-{\tt Simulations:} 
-compare mean density against $s_d$ at different values of $\sigma$; 
-omit variability if needed for readability. 
-Compare to panmictic results.
+%{\tt Simulations:} 
+%compare mean density against $s_d$ at different values of $\sigma$; 
+%omit variability if needed for readability. 
+%Compare to panmictic results.
 
 \subsubsection{Multiple variant types}
-
-One interesting problem that we can seek to address with this work is
-the extent to which the degree of pleiotropy biases 
+Another problem that we can seek to address with this work is
+the extent to which pleiotropy biases 
 adaptation towards the repeated use of particular subset of loci
@@
 (and in the worst case making them deleterious). 
-This effect has been studied by a number of authors \citep{}. 
+This effect has been studied by a number of authors \citep{REFS}. 
 A second contribution is that alleles that have
@@
 before the environmental shift.
-This effect seems less well studied, 
-perhaps because many modeling approaches use the strong selection, weak mutation limit 
-and so ignore standing variation. 
-\gc{Is this fair?}
-\plr{Not sure what papers you are thinking of.}
 
@@
 Let's imagine for the moment that all classes of beneficial allele 
-have the same beneficial selection coefficient. 
+have the same beneficial selection coefficient. Or simply beneficial
+selection coefficients are similar enough that our selective exclusion
+approximation holds over the time-scales over which we examine the process.
 Each class of mutations could have its own mutation rate $\mu_j$ and
@@
 have more serious pleiotropic consequences in the ancestral environment (class $2$). 
-We set $s_{d,1} \leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, 
+We set $s_{d,1} \leq s_{d,2}=0.05$ and $\mu_1=1 \times 10^{-8}$, 
 assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
 and think of the second class mutational target being ten, a hundred, or one thousand base pairs. 
-\plr{What is $s_{d,2}$ in figure 7?}
 We show the expected proportion of space covered by
@@
 The case with standing variation only is the best case scenario 
-for the rarer mutation as we force them to have the same advantage in the new environment, 
-so its rate of introduction is necessarily lower. 
-However, the standing variation only case does seem to provide a reasonable rule of thumb. 
-Especially for parameter combinations, 
+for the rarer mutation, so its rate of introduction is necessarily lower. 
+However, the standing-variation-only case does seem to provide a
+reasonable rule of thumb, especially for parameter combinations, 
 higher population densities and high dispersal distances, 
@@
 we will stick to the loose interpretation.
-\marginnote{check jaegers.  isn't there a CLT for $\tau$?}
 As $\sigma^2 \rho$ increases, the mean of $\tau$ increases, but the variance approaches a constant
-($\var[\tau] \to (1/s_b) \var[\log W]$).
+($\var[\tau] \to (1/s_b) \var[\log W]$). \plr{check jaegers.  isn't there a CLT for $\tau$?}
 
@@
 
-% \section{Applications}
-
-% The FRIGIDA (FRI) in {\it  Arabidopsis thaliana} is a major
-% determinant of genetic variation in flowering time. Recessive loss of function
-% alleles at FRI result in early flowering, due to the almost complete
-% loss the of the requirement of vernalization (a prolonged chill).  
-% This is thought to potentially confer an advantage due to allowing a
-% more rapid cycling life history.
-
-% Sweeps of FRI loss of function alleles \citep{Hagenblad-FRIsweep,Toomajian-FRIsweep}
-
-% 20 independent nonfunctional FRI haplotypes \citep{Shindo-FRI}
-
-% open reading frame (ORF) of 609 amino acids
-
 \section{Discussion}
 
-Confusing convergence for local adaptation. Mention skin pigmentation?
+Overall our results suggest that convergent evolution among
+populations may be common, especially when standing variation is not
+so deleterious that it is present before environmental shifts.
+
+Allowing standing variation in our model unsurprisingly increases the
+extent of convergence among population, and importantly may greatly
+lower the time untill the species becomes adapted across the
+geographic range of the selection pressure. Allowing standing
+variation also biases the type of variation towards that with fewer
+pleiotropic effects as these are more common as standing variation 
+before the environmental shift. This bias can in some cases easily
+overcome quite significant differences in mutational target sizes
+among loci allowing the same locus to be repeatedly the source of
+adaptation even if there are seemingly many different routes to adaptation.
+
+As we have argued in \citet{ralph2010parallel} the ease with which
+geographic convergent adaptation occurs means that we should
+incorporating it more widely into our thinking about the genetic basis
+of adaptation. For example, the absence of European skin pigmentation alleles in
+ancient DNA from European recovered from several thousand years ago
+has led to the suggestion that these individuals had dark
+pigmentation. However, given the results here and the partial convergent
+adaption of skin pigmentation between Europeans and East Asians
+\citep{} it seems just as plausible that these ancient individuals
+simply had a different complement of ``light-skin'' pigmentation
+alleles. Such convergence may considerably complicate the 
+exploration of adaptation among populations using variants mapped in a
+limited set of populations \citep{BergCoop:14}.
+
+More generally, if geographic convergence is common we should often expect to see
+selected alleles which are strongly geographically restricted as they
+have simply not had time to neutral mixing across the random boundaries
+established in their initial spread. This pattern may be very hard to
+distinguish from local adaptation using population genomic approaches
+alone. This is especially problematic as boundaries between convergent alleles may often
+occur where gene flow rates are low, i.e. ecological breaks, even if
+the alleles concerned have no bearing on the ecological differences
+across these breaks.
+
+Each of these local sweeps will be associated with the haplotype on
+which the particular allele arose. Under our model our standing
+variants are still very young, and so we do not expect a strongly
+reduced hitchhiking effect. As such following the initial period of
+adaptation, we should expect the population to be partitioned into a
+set of geographically resticted long haplotypes. Given sufficient time
+these haplotypes will mix together leading to a signal of a sweep with
+multiple mutations if our selected allele occur at the same locus, or
+multiple partial sweeeps if our loci are scattered across the genomes.
 
@@
 If many loci contribute to variation in a trait, then
-selection on any one allele may be weak, allowing adaptation to use different alleels in different populations.
+selection on any one allele may be weak, allowing adaptation to use different alleles in different populations.
 However, in that case populations separated by a reasonable amount of drift may still
@@
 i.e.\ those at intermediate frequency, will be somewhat different. 
+So it may be the case that polygenic traits are even less suspectible to a shared
+genetic basis to adaptation across populations than simple traits.   
+However, we currently lack models of this, and our ability to infer
+such shared events means that empirically an answer for this question
+may be limited for the moment.
+
 \plr{Not sure I follow that.  Ones at intermediate frequency will end up affecting the population mean less
 than ones at low frequency (but above the demographic stochasticity barrier)?}
-Therefore, it seems unlikely that polygenic traits are more suspectible to a shared
-genetic basis to adaptation across populations than simple traits.
+\gc{Yes but their initial freq. change depends on their
+  heterozygosity, so that if we only have to move the mean a short
+  distance to a new optimum we will not use rare variants. }
 
@@
 species should be viewed as being united 
-if levels of gene flow were still high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
-\plr{not clear which of three papers in the last sentence are being cited for what?}
+if levels of gene flow were still high enough for selected alleles to
+spread across the entire species \citep[see also][]{Ellstrand2014s}.
 At present, large scale genotyping and sequencing projects, 
@@
 
-In support of this idea, recent selective sweeps are often
-geographically restricted \citep{}, but see XXX. 
-This is likely in part due to the relatively low incidence of wide-spread selection pressures \citep{},
+In support of this idea, recent selective sweeps seem to often be 
+geographically restricted \citep{Coop}, but see XXX. 
+This is likely in part due to the relatively low incidence of wide-spread selection pressures,
 but as noted above even when we know of wide-spread selection
-pressures (e.g.\ malaria) the response is usually convergent, not shared, across large spatial scales. 
+pressures (e.g.\ malaria) the response is usually convergent, not
+shared, across large spatial scales. 
 On the other hand, 
@@
 are likely caused by introgression of haplotype complexes
-consisting of many, tightly linked, beneficial alleles.  
-\plr{is the following making sense?}
-The important missing piece to these puzzles seems to be an understanding
-of when phenotypes are accessible to mutation to a given species,
-and why they may be accessible to one subspecies and not another.
+consisting of many, tightly linked, beneficial alleles (that perhaps
+are inaccessible by mutation over reasonable time-scales for a population in a new
+environment).  
+We are currently also limited to scanning genome-wide for 
+species-wide sweeps in a small set of organisms where genome-wide
+sequence data for a number of populations are available. 
+This will obviously change dramatically over the coming years,
+allowing us to form an improved picture of the relationship between
+the level of neutral population structure, and the age and geographic
+extent of spread of selected alleles.
+
+%%Graham: I feel that this moved things in an interesting but
+%%tangential direction. Also I think that introgression may reflect
+%%earlier environmental exposure of populations. I t
+
+%% Interesting point but felt like we were 
+%The important missing piece to these puzzles seems to be an understanding
+%of when phenotypes are accessible to mutation to a given species,
+%and why they may be accessible to one subspecies and not another.
 
@@
 
-\plr{added:}
 Even if selective sweeps only bring alleles to fixation locally,
@@
 than neutral mixing through migration.
-We can see this by comparing mean coalescent times.
 Under neutral mixing, the mean number of generations back to the most recent common ancestor
@@
 This quantity has not been worked out for a model with simultaneous local sweeps,
-but is close to the ``spatial $\Lambda$-Fleming--Viot'' models of \citet{barton2013modelling},
+but will be somewhat analogous to the ``spatial $\Lambda$-Fleming--Viot'' models of \citet{barton2013modelling},
 in which local sweeps occur independently across the range.
-In these models, the mean number of generations back to the most recent common ancestor for distant samples
-scales with XXX,
-which can be substantially shorter.
-\plr{Maybe I'm forgetting something in their papers -- did they do this? --
-but I think it would be a little bit of work to figure out what parameters to put into their models.}
-Intuitively, we can obtain this as follows:
-suppose that coalescence only occurs during local sweeps,
-which occur at rate $\lambda$
-on a spatial scale such that there are on average $M$ independent origins of the sweep during each bout of adaptation.
-The motion of lineages backwards is not dissimilar to a stepping-stone model with $M$ sites
-that coalesce when they are in the same deme
-with probability dependent on the recombination distance to the sweeping, selected site.
-The mean hitting time of two independent random walks on a graph with $M$ nodes is of order $M$;
-to get the mean number of generations until coalescence, 
-we must divide this by the probability of remaining linked to the selected site as it sweeps
-and by the frequency of selected sites.
-Since $M$ scales proportionally to population density $\rho$,
-coalescent times do as well,
-but possibly with a much smaller coefficient
-(due effectively to an increase in reproductive variance).
+Lineages that are close to the sweep ($\sim v/\chi$ Morgans) will be moved
+towards the center of the sweep (a displacement $O(\chi)$), and pairs
+of lineages within an area swept and close to the sweep could be
+forced to coalesce  of the sweep will be forced to
+coalesce \citep[see ][for work on geographic hitchhiking]{barton2013genetic}. The overall rate of coalesce and
+mixing time of lineages will depend on the rate of sweeps, their
+geographic scale, and the rate of recombination and could be calculated by combining the result presented here with
+\citet{barton2013modelling} and  \citep{barton2013genetic}. 
+However, if geographical sweeps are common
+then this may substantially speed up the rate of mixing compared to
+neutral drift and migration.
+
+% In these models, the mean number of generations back to the most recent common ancestor for distant samples
+% scales with XXX,
+% which can be substantially shorter.
+% \plr{Maybe I'm forgetting something in their papers -- did they do this? --
+% but I think it would be a little bit of work to figure out what parameters to put into their models.}
+% Intuitively, we can obtain this as follows:
+% suppose that coalescence only occurs during local sweeps,
+% which occur at rate $\lambda$
+% on a spatial scale such that there are on average $M$ independent origins of the sweep during each bout of adaptation.
+% The motion of lineages backwards is not dissimilar to a stepping-stone model with $M$ sites
+% that coalesce when they are in the same deme
+% with probability dependent on the recombination distance to the sweeping, selected site.
+% The mean hitting time of two independent random walks on a graph with $M$ nodes is of order $M$;
+% to get the mean number of generations until coalescence, 
+% we must divide this by the probability of remaining linked to the selected site as it sweeps
+% and by the frequency of selected sites.
+% Since $M$ scales proportionally to population density $\rho$,
+% coalescent times do as well,
+% but possibly with a much smaller coefficient
+% (due effectively to an increase in reproductive variance).
 
@@
 how then do selected alleles ever become fixed within species? 
-Repeated bouts of adaptation in particular genomic regions may act 
+Drift alone will slowly sort variants within species. While repeated bouts of adaptation in particular genomic regions may act 
 to push a subset of previously selected alleles to fixation across the
@@
 initial rapid spread of selected alleles.
-\plr{or, by drift?}
 
@@
 through subsequent large-scale changes in geographic range size
-induced by extirpation of the species over parts of its range. 
-\plr{a.k.a. bottlenecks}
+induced by extirpation of the species over parts of its range
+\citep[see ][ for how such a model could be constructed]{barton2013modelling}. 
 Such drops in range size may fix, or radically change the species-wide frequency, 
@@
 
-\plr{Seems like we should bring in the work on what happens if these locally fixed alleles cause an incompatibility. (and/or speculate about it)}
-
+Such ideas are not completely new and represent a perhaps logical consequence of
+an allopatric, or at least parapatric, view of the geography of speciation.
+However, it is worth revisiting this idea as population genomic
+sampling becomes geographically broad allowing us to return to themes
+in biogeography. 
 Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation within species,
@@
 following the collapse of population structure \citep{Futuyma:10,FUTUYMA:87}. 
-\plr{"to the pattern of"?}
-In part, this was a response to the pattern of punctuated equilibrium \citep{eldredgegould72}, 
-arguing that the observation of stasis and rapid anagensis associated with speciation 
+Futuyma offered this as an explanation of the pattern of punctuated equilibrium \citep{eldredgegould72}, 
+and argued that the observation of stasis and rapid anagensis associated with speciation 
 were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
@@
 Much of the recent adaptation seen by
-population genetics is likely adaptation to local conditions, 
+population genetics may be adaptation to local conditions, 
 but as we have shown, even wide-spread selection pressures will likely result in 
 geographically restricted spread of alleles.
-The adaptive substitutions we see along lineages are those alleles 
-lucky enough to have spread in populations that escaped local extinction, 
-and to have avoided being assorted into species that eventually went extinct. 
-\plr{those last two things sound like the same thing}
 Ranges change sufficiently rapidly that
 range size could change substantially over the soujourn time of most alleles 
-that will eventually become fixed in the species, as such it seems as though .
+that will eventually become fixed in the species, making it likely that. 
+The adaptive substitutions we see along lineages may be those alleles 
+simply lucky enough to have spread in local populations or incipient species that escaped extinction.
 
 %Venditti:10, 
-This link between speciation and substitution does not imply that
+Such a link between speciation and substitution would not imply that
 substitutions should necessarily be thought of as being clustered at splits in inferred phylogenies 
@@
 
-Nor does this view imply that adaptive evolution or speciation is driven by the
+This view would not imply that adaptive evolution or speciation is driven by the
 shifting balance or genetic revolutions 
@@
 substitutions are fixed by positive selection in a number of species
-\citep[most notably Drosophila][]{}. Selection has played a strong
-role in the establishment of these alleles, and their frequency may be
-high enough within populations that they do not contribute
-substantially to polymorphism.
-
-In reality the evolution of range size is not independent of
-adaptation. Many DMIs are locally restricted \citep{Cutter:12}.
+\citep[most notably Drosophila][]{}, as under this view selection has played a strong
+role in the establishment of these alleles and they not contribute to
+polymorphism within geographically narrow samples. It will be
+interesting to see as we get more broadly geographic population
+genomics sampling for a range of species whether the class of alleles that contribute 
+to local differentiation are the similar as those underlying species
+divergence, and the extent to which the answer to this depends on the
+age and type of population structure within species.
+
+Finally, we close by noting that range expansion, and speciation, are obviously not seperate from
+adaptive differentiation. The invasion of novel geographic spaces may
+lead to burst of adaptive differentiation, at least at a
+subset of genes, and speciation may be associated with rapidly
+adaption to novel environments. Conversely, the relatively slow geographic spread of adaptive
+alleles if they only offer a local advantage, or the selective exclusion
+described here, may allow Dobzhansky-Muller incompatibilities (DMIs) to arise
+within species effectively offering a mechanism for 
+hybrid incompatibilites to evolve in parapatry effectively fracturing
+the range \citep{}. Indeed the alleles that act as components in many
+of the DMIs studied to
+date are geographically restricted \citep[see][]{Cutter:12}. As such
+it seems possible that populations within species are always on the
+verge of speciation, and that as outlined here that this may drive
+some proportion of molecular divergence.
 

### 2014-09-22 g6pd_figures.R

@@ -7,7 +7,7 @@ my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
 
 
 ###################################################
-###################Char length as function of sd
+###################Char length as function of sb
 ###################################################
 
  mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.001,0.3,length=800); 
@@ -65,9 +65,9 @@ for(j in 1:2){
  	}
  	
 dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlengths.pdf")
- 
+
 ###############################################################
-####Mean time standing
+####Mean time standing as func. of sd
 ###############################################################
  layout(t(1:2))
 par(mar=c(4,4.1,2,0.3))
@@ -95,6 +95,54 @@ for(j in 1:2){
  	}
  	
 dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_chartimes.pdf")
+
+
+###############################################################
+####Mean time standing as func. of sd & sb
+###############################################################
+ layout(t(1:2))
+par(mar=c(4,4.1,2,0.3))
+ mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.001,0.3,length=800); 
+
+for(j in 1:2){
+	rho=rhos[j]
+	if(j == 1) {	plot(sdvals,sdvals,type="n",ylim=c(0,250),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab="Generations",cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
+		}
+	if(j == 2){ plot(sdvals,sdvals,type="n",ylim=c(0,250),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
+}	
+
+	for(i in 1:3){
+		if(j == 1){ 
+			sigma=sigmas[i]; 
+			charlength.sd<-sapply(sdvals, function (sd) {  meanTime(mu, rho, sb, sd, sigma)$value } )
+			 lines(sdvals,charlength.sd,lwd=2,col=my.col[i])
[…]

### 2014-09-22 standing-variation-paper.tex
.

@@
 common, as has already been seen for a number of traits. 
-On the basis of this we argue, in our discussion, that the genetics of
-adaptation may often be
-geographically local, and that we should expect to see relatively few
-wide-spread sweeps even when selection pressures are wide-spread. 
+On the basis of this we argue, in our discussion, 
+that the genetics of adaptation may often be geographically local, 
+% and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
+and that we should only expect to see widespread selective sweeps
+when adaptation is highly constrained (e.g.\ by small mutation rate or having to cross a fitness valley).
 We discuss some implications, and history, of this view for the evolutionary
@@
 adapt via different genetic routes, as the constellation of alleles
-best positioned to initially respond to positive selection, 
-i.e.\ those at intermediate frequency, will be somewhat different. 
+at intermediate frequency will be somewhat different,
+and these are the alleles that selection moves most quickly to change the trait value.
+% best positioned to initially respond to positive selection, 
+% i.e.\ those at intermediate frequency, will be somewhat different. 
 So it may be the case that polygenic traits are even less suspectible to a shared

### 2014-09-22 standing-variation-paper.tex
edited through.

@@
 \title{The role of standing variation in parallel adaptation}
-\author{Peter L. Ralph$^1$ \\ email: pralph@usc.edu  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }
+\author{
+Graham Coop$^{*,1}$ \\ email: gmcoop@ucdavis.edu 
+\and 
+Peter L. Ralph$^{*,2}$ \\ email: pralph@usc.edu  
+}
 
@@
 
-$^1$ Computational Biology and Bioinformatics \\ 
+    $^*$ These authors contributed equally to this work. \\
+$^1$ Center for Population Biology \& Department of Evolution and Ecology \\ 
+University of California -- Davis, Davis, CA, 95616\\
+$^2$ Computational Biology and Bioinformatics \\ 
 University of Southern California, Los Angeles, CA \\
-$^2$Center for Population Biology \& Department of Evolution and Ecology \\ 
-University of California -- Davis, Davis, CA, 95616
 \end{center}
@@
 
+%%%%%%%% %%%%%%%%%%%%
 \section{Introduction}
 
-There are an increasing number of examples in which
-different populations within species have
-adapted to similar environments via independent genetic changes. 
-In some cases this convergent evolution leads to differing phenotypes
-in response to shared selection pressures
-\citep{} in other cases the degree of convergence is precise down the
-same nucleutide change. 
-Such convergent evoltuion within populations has been seen for many
-carefully studied phenotypes in a range of species, 
-including drug resistance, pathogen resistance, insecticide resistance, and pigmentation. 
+There are an increasing number of examples known in which
+different populations of a single species have
+adapted to similar environments by means of independent genetic changes. 
+In some cases this convergent evolution has produced different phenotypes
+despite shared selection pressures \citep{}; 
+in other cases independent adaptations are identical down the same nucleotide change. 
+Such convergent evolution within populations has been seen for many
+carefully studied phenotypes across a range of species, 
+including drug resistance in pathogens,
+resistance to pathogens or pesticides,
+and the molecular basis of pigmentation changes.
+\plr{The next sentence should be more specific (give example?) or maybe be deleted, I think?}
 In many of these examples we 
@@
 whether selection pressures differ subtly. 
-We note that we treat the phrases ``convergent'' and ``parallel'' evolution
-synonymously, as we are concerned with adaptation within a single
+The phrase ``parallel evolution'' is also used to refer to such convergent evolution;
+here we use these synonymously, 
+as we are concerned with adaptation within a single
 species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.
@@
 and to what degree should species be regarded to cohesive units. 
-Convergent evolution within populations 
-also affects our ability to detect adaptation from population genomic data,
+Convergent evolution also affects our ability to detect adaptation from population genomic data,
 since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.
@@
 subject to a constant selection pressure, 
-either through selection on multiple mutations present as standing variation within population
+either through selection on multiple mutations present as standing variation within the population
 before selection pressures switch \citep{Orr:01,softsweepsI},
-or due to multiple adaptive alleles that arise after selection pressures switch \citet{softsweepsII}. 
+or due to multiple adaptive alleles that arise after selection pressures switch \citep{softsweepsII}. 
 The previous work has shown that 
 a primary determinant of the probability that multiple alleles contribute to adaptation
-is the population scaled mutation rate \citep[see ][ for a review]{MesserPetrov}.
-
-\plr{should stick to either ``parallel'' or ``convergent'' perhaps}
-
-Geographically limited dispersal also increases the chance of convergent evolution. 
+is the product of the population size and the mutation rate \citep[see ][ for a review]{MesserPetrov}.
+
+Spatial population structure,
+as caused for example by geographically limited dispersal, 
+also increases the chance of convergent evolution. 
 For example, 
@@
 since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
-In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
+In \citep{ralphcoop2010} we formulated a simple model of parallel adaptation in a geographic setting
 where the selection pressure was constant across the entire environment, 
-and that a single mutational change was sufficient to adapt populations to the novel environment. 
+and a single mutational change was sufficient to adapt after the change in environment.
 We assumed that there was no standing variation for the adaptive allele, 
-so that parallel mutation must be due to multiple mutations after the environmental shift. 
+so that parallel mutation must be due to multiple new mutations occurring after the environmental shift. 
 In that setting, 
-we found that there was a characteristic geographic scale over which we would expect 
-multiple instances of our adaptive allele to have arisen in parallel. 
+we found the characteristic geographic scale over which 
+multiple instances of the adaptive allele are expected to arise in parallel. 
 This characteristic length could be expressed in terms of a simple compound
 parameter determined by our parameters of interest. 
-Here we extend this spatial model to include standing variation 
+In this paper, we extend this spatial model to include standing variation 
 present at mutation-selection balance before the selection pressures switch.
 
-We show that convergent adaptation within wide-spread species is likely to be
+Below,
+we show that convergent adaptation within a wide-spread species is likely to be
 common, as has already been seen for a number of traits. 
-On the basis of this we argue, in our discussion, 
+On the basis of this we argue, in the discussion, 
 that the genetics of adaptation may often be geographically local, 
 % and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
-and that we should only expect to see widespread selective sweeps
+and that most widespread selective sweeps should occur
 when adaptation is highly constrained (e.g.\ by small mutation rate or having to cross a fitness valley).
@@
 %%%%%%%%% %%%%%%%%%%%% %%%%%%%
-\subsection{The set-up}
+\subsection{Model description}
 
@@
 We assume that the species range
-is a large, homogenous, one- or two-dimensional region. 
+is a large, homogeneous, one-- or two--dimensional region. 
 There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
@@
 homozygote for the mutant alleles.
-The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
+The number of offspring has finite variance, 
+which would divide the probability of establishment of advantageous alleles below,
+but to keep the formulas simpler we assume to be equal to one.
 (We leave aside the case where mutations are recessive, however.)
@@
 suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
-and that the mean squared distance between parent and child is $\sigma^2$.
+and that the mean squared geographic distance between parent and child is $\sigma^2$.
 The species occupies an area with mean density
-$\rho$ haploid individuals (or chromosomes) per unit area.
-\marginnote{XXX or do diploid? below is haploid.}
+$\rho$ of haploid individuals (or chromosomes) per unit area.
 
@@
 as a branching process.
-After $t=0$, the offspring of a single mutant form (approximately) a branching process with growth rate $s_b$,
+After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
 so each new mutant establishes locally with probability $p_s \approx 2s_b$.
@@
 The rate of this Poisson process,
-i.e.\ the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
+i.e.\ the mean number of nut mutants per unit area and per generation that
+appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
 is approximately $\lambda = 2 \mu \rho s_b$.
@@
 If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
-and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
-equivalent to the mutants arising later.
-(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
+and that these progenitors are near to each other in space, 
+we can then treat each such family as equivalent to a single new mutation,
+but with somewhat larger probability of local establishment.
+(This approximation will be good 
+if the logarithm of the size of each extant family is small relative to the establishment time,
 and the spatial distribution of each is small relative to the spread between the families.)
 To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
-At time $t=0$, there are $Z_T$ individuals present with the mutation,
-and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
-Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
+At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
+and each has probability $p_s$ of establishing, approximately independently.
+Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
 so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
-the rate is
+the mean number of clusters of standing variants destined to fix locally, per unit area, is
 \begin{align*}
@@
 \end{align*}
-Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
+For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
 we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
@@
 Note that this is closely related to the equilibrium frequency of a
-allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}
+allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}.
 
@@
 Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
-See \citet{ralph2010parallel} for a fuller review of these travelling waves.
+(See \citet{ralphcoop2010} for a more thorough review of these travelling waves.)
 Note that the speed of the wave will vary
@@
 \paragraph{Putting it together.}
-Now, we can put these ingredients together for a simple model of the geographic spread of alleles. 
-A cartoon example of this process is shown in Figure \ref{fig:cartoon}. 
+Now, we can put these ingredients together for a simple model of the geographic spread of alleles, 
+a cartoon example of which is shown in Figure \ref{fig:cartoon}. 
 Initially, when the selection pressures change at $t=0$, 
 a set of standing variants can start to spread having escaped loss through drift. 
-These variants are shown as little lightning bolts, and occur at a density $\lambda_0$ across space. 
+The originating mutations of these variants are depicted by lightning bolts, and occur at a density $\lambda_0$ across space. 
 They spread at velocity $v$, carving out cones in space-time. 
 As these alleles proceed in their geographic spread, other new alleles can arise and become established in parallel, 
-these are shown as little stars. 
+whose origins are indicated by stars.
 These new mutations arise and become established at rate $\lambda$.
 
-As we previously outlined in \citet{ralph2010parallel} this model of
-geographic convergent evolution, when $\lambda_0=0$, is analogous to the a model of crystallization due to Kolmogorov \citet{kolmogorov-crystallization}.
-In this model, nucleation points form at random at a constant rate in time and space 
+As we outlined in \citet{ralphcoop2010} this model of
+geographic convergent evolution, when $\lambda_0=0$, is analogous to a model of crystallization due to \citet{kolmogorov-crystallization}.
+In this model, nucleation sites form at random at a constant rate in time and space 
 and initiate the radial growth of new crystals. 
 Eventually, the different orientations of crystals form a random tessellation of space,
-hose properties have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
+whose properties have been studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
 The generalized version of this process, for non-constant wave speeds
-and inhomogeneous Poisson processes is termed the 
-Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
+and inhomogeneous Poisson processes is known as the 
+Kolmogorov--Johnson--Mehl--Avrami tessellation \citep{fanfoni-tomellini}.
 Our combined process with both standing variation and new mutation is
@@
 (For the purposes of analogy, we could imagine that before time $t=0$
-the temperature is high enough that nucleation sites do not persist.)
-
-If we only allow standing variation, then everything about the process is relatively simple. 
-All of the points in space closer to a particular origination point will be
-reached by that allele before any other allele has a chance to spread there. 
-Such a tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
+the temperature is high enough that nucleation sites appear but do not persist long.)
+
+If we ignore the effects of new mutation,
+then everything about the process is relatively simple:
+each point in space will be first reached by the wave whose origination point lies closest to it.
+Such a random tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
 (i.e.\ the cells formed by assigning points to the nearest point in a Poisson process).
-The properties of the patterning of space by alleles is
+The properties of this tesselation by alleles is
 determined by the spatial locations of the initiation points, 
-as specified by the spatially homogeneous Poisson process, 
+which are sampled from a spatially homogeneous Poisson process, 
 independent of the rate of spatial spread. 
-Therefore, the properties of the spatial tessellation with standing variation depends only on $\lambda_0$.
+Introducing new mutations cause some qualitative changes
+beyond dependence on new parameters:
+the cells formed by a Voronoi tesselation have straight sides,
+but the introduction of new mutations cause these to curve.
 
@@
 Cartoon of the geographic spread of standing and new mutations. 
+The mean spatial density of successful standing variants at $t=0$ (lightning bolts) is $\lambda_0$,
+and the mean density in space $\times$ time of successful new mutations (stars) is $\lambda$.
+\plr{Add $t=0$ and direction of time to the figure.}
 }
   \label{fig:cartoon}
-
 \end{figure}
@@
 
-Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
+Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
 and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
@@
 For example, 
-a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
+a number of changes that confer malaria resistance have been observed in the $\beta$-globin gene;
 and the sickle cell allele may plausibly have arisen by up to five independent occurrences
-of the same base pair mutatoin at different locations within Africa
-\citep{Flint:98,ralph2010parallel}.
+of the same base pair mutation at different locations within Africa
+\citep{Flint:98,ralphcoop2010}.
 Another particularly impressive case of convergent evolution in response to
 selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
-with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
+with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
 that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
@@
 rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
-Three G6PD deficiency alleles that also confer resistance to malaria 
+Whether these all confer resistance to malaria is unknown,
+but malaria is thought to be the primary driver of these polymorphisms.
+Three G6PD deficiency alleles known to confer resistance to malaria 
 are particularly common and relatively well studied: 
@@
 The $A-$ and Mahidol alleles have been shown
-to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
+to be protective against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
 hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
 Haplotype-based analysis of genetic diversity surrounding
-the $A-$, Med, and Mahidol suggest that they have spread over the past 
-few thousand years  \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
+$A-$, Med, and Mahidol suggest that they have spread over the past 
+few thousand years \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
 consistent with the age of other known malaria resistance alleles. 
@@
 considerable advantage). 
-The wave of advance works in the case of heterozygote advantage
+The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
+applies as well in the case of heterozygote advantage
 \citep{aronson1975nonlinear}, with the selected allele spreading
-locally to the equilibrium frequency (rather than fixation). As such
-our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
-We assume that before malaria became prevalent, the G6PD deficiency alleles suffered a
-decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and
-hemizygote male. Assuming that the underlying causes of this drop in fitness have
-not changed, we estimate that $s_d$ has to be roughly
-XXX to XXX, in order to have resulted in the equilibrium frequency
-seen today.
+locally to the equilibrium frequency (rather than fixation). 
+Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
+We assume that before malaria became prevalent, 
+G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
+Assuming that the underlying causes of this drop in fitness have not changed, 
+we estimate that $s_d$ has to be roughly XXX to XXX, 
+in order to have resulted in the equilibrium frequency seen today.
 
@@
 The geographic area of Central and Eastern Asia with malaria is 
-on the order of ten million square kilomters.
+on the order of ten million square kilometers.
 In that area there are at least $15$ common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
-The coding region of G6pd is 515 codons long, 
+The coding region of G6PD is 515 codons long, 
 and around 140 distinct deficiency alleles have been observed. 
-Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
-we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
-The dispersal and demographic parameters of humans in the past few thousans years is unclear,
+Assuming a mutation rate of $\approx 10^{-8}$ per base pair per generation, 
+we can take as an order-of-magnitude estimate $\mu \approx 1 \times 10^{-6}$ per generation. 
+The dispersal and demographic parameters of humans in the past few thousand years is unclear,
 particularly as we are concerned with the ``effective'' population density
 (i.e.\ population density divided by variance in offspring number).
-We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
-and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
-Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
+We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per km$^2$,
+and three values for the dispersal distance: $\sigma=10$, $50$ and $100$ kilometers per generation.
+Clearly, human migration has been shaped both by local dispersal and larger-scale expansions 
 \citep[see ][for a recent discussion]{}, 
@@
 \caption{ 
-Map of G6pd-deficiency allele frequencies across Asia. 
-The pie chart shows the frequency of G6pd-deficiency alleles. 
-The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
+Map of G6PD-deficiency allele frequencies across Asia. 
+The pie chart shows the frequency of G6PD-deficiency alleles. 
+The size of the pie chart indicates the number of G6PD-deficient individuals sampled.
 Countries with endemic malaria are colored yellow. 
@@
 
-\subsection{Characteristic parameters}
+\subsection{The geographic resolution of adaptation from new and standing variation}
 
@@
 
-In \citet{ralph2010parallel}, studying the model without standing variation,
-we defined a {\em characteristic length} which gave the spatial scale across with mutants with distinct origins would establish.
+In \citet{ralphcoop2010}, studying the model without standing variation,
+we defined a {\em characteristic length} which gave the spatial scale across which mutants with distinct origins would establish.
 This was proportional to the mean distance between neighboring established mutants,
@@
 Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
-As it spreads, it will cover a distance $L$ in time $L/v$.
+If it does not encounter other locally fixed, beneficial mutants,
+it will cover a distance $L$ in time $L/v$.
 The number of other mutations appearing in the circle it has covered up until this time is Poisson with mean
 $\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
-(and $\lambda_0 2 L + \lambda \pi L^2 /v$ in one dimension).
+(and $2 \lambda_0 L + 2 \lambda \pi L^2 /v$ in one dimension).
 % $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
@@
 The explicit formula for $\chi$ is cumbersome; here we omit it.
-(In one dimension the characteristic length is $\chi = \frac{ \sqrt{ 4 \lambda_0^2 + 2\lambda v } - 2 \lambda_0 }{ 4 \lambda v }$.)
+(In one dimension the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.)
 Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
@@
 \end{equation}
-Theses two end points help build our intuition for the interaction of
+These two end points help build our intuition for the interaction of
 parameters in shaping the geographic scale of convergent evolution.
@@
 of various parameters chosen to match the evolution of malaria
-resistance at G6pd. 
+resistance at G6PD. 
 These curves match our intuition that population densities result in
@@
 input and so decreases the characteristic length below that predicted
-by only new mutations (i.e. eqn. \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
+by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
 As the prior deleterious effect of the allele ($s_d$), the
-characteristic length increases untill it reaches that predicted by
-new mutation alone. Higher dispersal distances lead to larger
-characteristic lengths, as alleles spread geographically more rapidly blocking others
-from arising. A larger selective advantage acts in two conflicting
-ways: aiding the rapid geographic spread of established alleles, and also help more independent copies
-escape drift and become established. The effect on helping establish
-alleles wins out in this as increasing the selective $s_b$ decreases
-the characteristic length. This effect strongest when only standing
-variation contributes ( $s_b^{-1/2}$, eqn \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter, the dependence
-when only new mutations contribute is much weaker ( $s_b^{-1/6}$, eqn \eqref{eqn:chi_new})
+characteristic length increases until it reaches that predicted by new mutation alone. 
+Higher dispersal distances lead to larger characteristic lengths, 
+since more rapid geographic spread blocks others from arising. 
+A larger selective advantage acts in two conflicting
+ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
+escape drift and become established. 
+The effect of helping establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
+\plr{perhaps we should note: (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
+showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1--4}$.) }
+This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
+as in that case the speed of spread does not matter, 
+and the dependence on $s_b$ when only new mutations contribute is much weaker ($s_b^{-1/6}$, equation \eqref{eqn:chi_new})
 Overall, the range of characteristic lengths observed are reasonably
 consistent with the average diameter of a G6PD variant in Eurasia of
-$800$, especially for the lower population density.
+800km, especially for the lower population density.
 
@@
 \caption{ %
-{\bf Characteristic length,} as a function of selective disadvantage, compared to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+{\bf Characteristic length,} as a function of selective disadvantage, 
+compared to the corresponding quantity without standing variation (from \cite{ralphcoop2010}) 
+and to the quantity only considering standing variants (i.e.\ $\sqrt{\lambda_0}$).
+The other parameters are as estimated from G6PD,
+$\mu = 10^{-8}$ and $s_b = XXX$???.
 } \label{Fig-G6PD-charlength}
@@
 {\bf Characteristic length,} as a function of selective advantage,
-holding the prior disadvantage of the allele fixed $s_d=0.05$. 
-We compare this to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-8}$ (right???).  
+We compare this to the corresponding quantity without standing variation (from \cite{ralphcoop2010}) 
+and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
 } \label{Fig-G6PD-charlength-sb}
@@
 
-The quantity that is perhaps easiest to compute is the mean time until adaptation.
-If fix ourselves at some geographic location, and let $\tau \ge 0$ be the time at which the point is reached by the advantageous mutations.
+It is also straightforward to compute the mean time until adaptation.
+Imagine ourselves at some geographic location, and let $\tau \ge 0$ be the time at which we are first reached by some advantageous mutation.
 Then, as can be seen from Figure~\ref{fig:cartoon},
@@
 Since we assume these are a Poisson process, 
+in two dimensions
 \begin{equation}
@@
 \end{equation}
-and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
+and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \} dt$,
 \begin{align}
@@
 \end{align}
-For means of evaluting this integral, see Appendix \ref{apx:integrals}.
+For means of evaluating this integral, see Appendix \ref{apx:integrals}.
 
@@
 rapidly spread. Given that the G6PD alleles likely spread over a few
-thousand years, i.e. less than a few hundred generations. This
-time-scale seems quite plausible given Figure \ref{G6PD_chartimes},
-except perhaps for the lowest dispersal distances.
+thousand years, i.e.\ less than a few hundred generations, 
+this time scale seems quite plausible, except perhaps for the lowest dispersal distances.
 
@@
 \subsection{The contribution of standing variation.}
-We can use our framework to address how much adaptation arises from
-standing mutation. We have defined $\lambda_0$ to be the mean density
+We can use our framework to address what proportion of new adaptive variants arise from standing variation.
+We have defined $\lambda_0$ to be the mean density
 of standing variants that are present and escape loss when the
@@
 We will define $\nu_+$ to be the mean density of newly arising alleles
-that spread --having arisen in an area free of other adaptive alleles.
-Since the probability that a mutant arising at $(x,t)$ is lucky enough to be born in a location not already occupied by mutants
+that spread having arisen in an area free of other adaptive alleles.
+Since the probability that a mutant arising at location $x$ and time $t$ 
+is lucky enough to be born in a location not already occupied by mutants
 is $\P\{ \tau > t \}$,
@@
 \begin{equation} \label{prop_patches_standing}
-\lambda_0 / (\lambda_0 + \nu_+),
+\lambda_0 / (\lambda_0 + \nu_+) .
 \end{equation}
-this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
+% this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
 There are $\lambda_0 + \nu_+$ patches per unit area, so
@@
 geographic space that is covered by adaptation from standing variation
-(eqn. \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
+(equation \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
 } \label{G6PD_standing_var_proportion}
@@
 
-\subsubsection{Multiple variant types}
+\subsection{Multiple variant types}
 Another problem that we can seek to address with this work is
@@
 (i.e.\ convergence at genetic level). 
-While many alleles may confer the benefical phenotype, 
+While many alleles may confer the beneficial phenotype, 
 not all will contribute to adaptation if they have negative pleiotropic consequences.
@@
 Let's imagine for the moment that all classes of beneficial allele 
-have the same beneficial selection coefficient. Or simply beneficial
-selection coefficients are similar enough that our selective exclusion
-approximation holds over the time-scales over which we examine the process.
-Each class of mutations could have its own mutation rate $\mu_j$ and
-have a disadvantage $s_{d,j}$ prior to the switch. 
+have the same beneficial selection coefficient. 
+Or, at least that beneficial selection coefficients are similar enough that our selective exclusion
+approximation holds over the time scale on which we examine the process.
+Each class of mutations $j$ has its own mutation rate $\mu_j$ and selective disadvantage $s_{d,j}$ prior to the environmental switch. 
 As they have the same beneficial selection coefficient after the switch, 
 all of the waves travel outward at a rate $v$. 
-Then the density of type $k$ standing variants per unit area 
+Then, the density of type $j$ standing variants per unit area 
 and the input rate of \textit{de novo} variants per unit area per generation are, respectively,
 \begin{align}
-  \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } \qquad \text{and} \qquad    \lambda_{j} = \mu_j \rho s_b .
+  \lambda_{0,j} = \frac{ 2 \mu_j \rho s_b }{ -\log(1-s_{d,j}) } \qquad \text{and} \qquad    \lambda_{j} = 2 \mu_j \rho s_b .
 \end{align}
+\plr{note: I added a 2 to $\lambda_j$.}
 Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
-the proportion of the species range covered by alleles of type $k$ is
-\begin{equation} \label{eqn-prop-space-allele-k}
-    p_k = \int_0^\infty  \left( 2 \lambda_{0,k} \pi v^2 t + \lambda_k \pi v^2 t^3 \right)  
-    \; \exp \left( - \sum_j ( \lambda_{0,j} \pi v^2 t^2 + \lambda_j \pi v^2 t^3 / 3 ) \right) dt .
+the proportion of the species range covered by alleles of type $j$ is
+\begin{equation} \label{eqn-prop-space-allele-j}
+    p_j = \int_0^\infty  \left( 2 \lambda_{0,j} \pi v^2 t + \lambda_j \pi v^2 t^3 \right)  
+    \; \exp \left( - \sum_k ( \lambda_{0,k} \pi v^2 t^2 + \lambda_k \pi v^2 t^3 / 3 ) \right) dt .
 \end{equation}
 If we only allow standing variation this collapses to 
-\begin{equation}  \label{eqn-prop-space-allele-k-standing-room-only}
-    p_k = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) } ,
+\begin{equation}  \label{eqn-prop-space-allele-j-standing-room-only}
+    p_j = \frac{  \mu_j / \log(1-s_{d,j}) }{\sum_k  \mu_k / \log(1-s_{d,k}) } ,
 \end{equation}
@@
 i.e.\ if all variants are highly deleterious before the environment switches, 
-then $p_k =\mu_j / (\sum_k  \mu_k)$.
+then $p_j =\mu_j / (\sum_k  \mu_k)$.
 
 To illustrate some of the properties of this model, 
-let's imagine the somewhat extreme scenario that there is a single base-pair 
-at which a possible mutation is relatively free of negative pleiotropy (call this class $1$), 
+let's imagine the somewhat extreme scenario that there is a single base pair 
+at which a possible mutation is free of negative pleiotropy (call this class $1$);
 and a larger mutational target where changes 
@@
 assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
-and think of the second class mutational target being ten, a hundred, or one thousand base pairs. 
+and think of the second class as ten, a hundred, or one thousand base pairs. 
 We show the expected proportion of space covered by
-the rarer, class $1$ mutations in Figure \ref{fig-pleiotropy_calc}.
+the rarer, class $1$ mutations in Figure \ref{fig:pleiotropy_calc}.
 Intuitively, the contribution of the rarer mutation decreases as the
@@
 reasonable rule of thumb, especially for parameter combinations, 
-higher population densities and high dispersal distances, 
+like higher population densities and high dispersal distances, 
 that increase the contribution of standing variation (and similarly for high $s_b$).
@@
   mutation}. Empty circles give the result for standing variation only,
-  eqn. \eqref{eqn-prop-space-allele-k-standing-room-only}. Lines give
-  the result allowing both standing and de novo mutations  eqn. \eqref{eqn-prop-space-allele-k}}   \label{fig-pleiotropy_calc}
+  equation \eqref{eqn-prop-space-allele-j-standing-room-only}. Lines give
+  the result allowing both standing and \textit{de novo} mutations (equation \eqref{eqn-prop-space-allele-j}).
+  \plr{add other parameter values}
+  }   \label{fig:pleiotropy_calc}
 \end{center}
@@
 
-\section{Local establishment and comparison to panmixia}
+\subsection{Local establishment and comparison to panmixia}
 
@@
 In this section we examine this assumption in more detail, although mostly through heuristic arguments,
-and also compare our results above to the results without geographic structure of \citet{softsweeps}.
+and also compare the results above to the results without geographic structure of \citet{softsweeps}.
 
@@
 so that if $\tau$ is the time $Z_t$ reaches size $\sigma^2 \rho$, % i.e.\ $\tau = \inf\{ t \ge 0: Z_t > \sigma^2 \rho\}$,
-then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$.
+then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$ \citep{jaegers1975branching}.
 From this we know that $\tau \approx (1/s_b) (\log (\sigma^2 \rho) - \log W)$;
-although more detailed information is available (e.g.\ a CLT for $\tau$, \citet{XXX}),
+although more detailed information is available (e.g.\ a CLT for $\tau$, \citet{nagaev1971limit}),
 we will stick to the loose interpretation.
-As $\sigma^2 \rho$ increases, the mean of $\tau$ increases, but the variance approaches a constant
-($\var[\tau] \to (1/s_b) \var[\log W]$). \plr{check jaegers.  isn't there a CLT for $\tau$?}
 
@@
 much faster than under panmixia.
-\plr{note: this had the wrong conclusion here previously}
-\marginnote{Graph: sigma against mean number of origins; different curves for spatial, panmictic (flat), and simulation;
-different values of theta as different line types?  but how to do this?
-Mention Monty's empirical speeds paper.}
 % We could compare the probability of parallel adaptation under the two models (see eq'n 18 in \citet{softsweepsII}),
@@
 
-Overall our results suggest that convergent evolution among
-populations may be common, especially when standing variation is not
-so deleterious that it is present before environmental shifts.
-
-Allowing standing variation in our model unsurprisingly increases the
-extent of convergence among population, and importantly may greatly
-lower the time untill the species becomes adapted across the
-geographic range of the selection pressure. Allowing standing
-variation also biases the type of variation towards that with fewer
-pleiotropic effects as these are more common as standing variation 
-before the environmental shift. This bias can in some cases easily
-overcome quite significant differences in mutational target sizes
+Our results further suggest that convergent evolution among populations may be quite common, 
+especially when standing variation is not so deleterious that it is present before environmental shifts.
+
+Allowing standing variation in our model, unsurprisingly, 
+increases the extent of convergence among populations.
+Importantly, it may greatly lower the time until the species becomes adapted 
+across the geographic range of the selection pressure. 
+Adaptation through standing variation also biases the type of variation towards those alleles 
+with fewer pleiotropic effects, since these are more common as standing variation before the environmental shift. 
+This bias can in some cases easily overcome quite significant differences in mutational target sizes
 among loci allowing the same locus to be repeatedly the source of
 adaptation even if there are seemingly many different routes to adaptation.
-
-As we have argued in \citet{ralph2010parallel} the ease with which
+(See Figure~\ref{fig:pleiotropy_calc}.)
+
+As we have argued in \citet{ralphcoop2010} the ease with which
 geographic convergent adaptation occurs means that we should
-incorporating it more widely into our thinking about the genetic basis
+incorporate it more widely into our thinking about the genetic basis
 of adaptation. For example, the absence of European skin pigmentation alleles in
@@
 selected alleles which are strongly geographically restricted as they
-have simply not had time to neutral mixing across the random boundaries
-established in their initial spread. This pattern may be very hard to
-distinguish from local adaptation using population genomic approaches
+have simply not had time for neutral mixing to spread them across the landscape.
+This pattern may be very hard to distinguish from local adaptation using population genomic approaches
 alone. This is especially problematic as boundaries between convergent alleles may often
-occur where gene flow rates are low, i.e. ecological breaks, even if
+occur where gene flow rates are low, i.e.\ ecological breaks, even if
 the alleles concerned have no bearing on the ecological differences
@@
 
-Each of these local sweeps will be associated with the haplotype on
-which the particular allele arose. Under our model our standing
-variants are still very young, and so we do not expect a strongly
-reduced hitchhiking effect. As such following the initial period of
+Each of these local sweeps will be associated with the haplotype on which the particular allele arose. 
+Under the parameter regime we study,
+standing variants are still quite young, 
+so we do not expect a strongly reduced hitchhiking effect. 
+As such, following the initial period of
 adaptation, we should expect the population to be partitioned into a
-set of geographically resticted long haplotypes. Given sufficient time
-these haplotypes will mix together leading to a signal of a sweep with
-multiple mutations if our selected allele occur at the same locus, or
-multiple partial sweeeps if our loci are scattered across the genomes.
+set of geographically restricted long haplotypes. 
+Given sufficient time these haplotypes will mix together, 
+leading to a signal of a sweep with multiple mutations if our selected allele occurs at the same locus, 
+or to multiple partial sweeps if the loci are scattered across the genomes.
 
@@
 genetic basis to adaptation across populations than simple traits.   
-However, we currently lack models of this, and our ability to infer
-such shared events means that empirically an answer for this question
-may be limited for the moment.
-
-\plr{Not sure I follow that.  Ones at intermediate frequency will end up affecting the population mean less
-than ones at low frequency (but above the demographic stochasticity barrier)?}
-\gc{Yes but their initial freq. change depends on their
-  heterozygosity, so that if we only have to move the mean a short
-  distance to a new optimum we will not use rare variants. }
+However, we currently lack good methods with which to test this.
+% models of this, and our ability to infer
+% such shared events means that empirically an answer for this question
+% may be limited for the moment.
 
@@
 \citet{Rieseberg2001} and \citet{MorjanRieseberg:04} argued that 
-even if gene flow is not be high enough to prevent neutral differentiation, 
+even if gene flow is not high enough to prevent neutral differentiation, 
 species should be viewed as being united 
@@
 However, our work on geographic convergent adaptation
-\citep[see also][]{ralph2010parallel,RalphCoop:14} 
+\citep[see also][]{ralphcoop2010,RalphCoop:14} 
 concludes that this gene flow will lead to the rapid spread of
@@
 environment).  
-We are currently also limited to scanning genome-wide for 
-species-wide sweeps in a small set of organisms where genome-wide
-sequence data for a number of populations are available. 
-This will obviously change dramatically over the coming years,
-allowing us to form an improved picture of the relationship between
-the level of neutral population structure, and the age and geographic
-extent of spread of selected alleles.
+Furthermore,
+currently we can only scan genomes for species-wide sweeps
+in those few organisms with population-scale sequence data.
+This is rapidly changing,
+and will allow us to form an improved picture of the relationship 
+between the level of neutral population structure, 
+and the age and geographic extent of spread of selected alleles.
 
@@
 Lineages that are close to the sweep ($\sim v/\chi$ Morgans) will be moved
-towards the center of the sweep (a displacement $O(\chi)$), and pairs
-of lineages within an area swept and close to the sweep could be
-forced to coalesce  of the sweep will be forced to
-coalesce \citep[see ][for work on geographic hitchhiking]{barton2013genetic}. The overall rate of coalesce and
-mixing time of lineages will depend on the rate of sweeps, their
-geographic scale, and the rate of recombination and could be calculated by combining the result presented here with
-\citet{barton2013modelling} and  \citep{barton2013genetic}. 
-However, if geographical sweeps are common
+towards the center of the sweep (a displacement $O(\chi)$), 
+and pairs of lineages caught up in the same sweep could be forced to coalesce 
+\citep[see ][for work on geographic hitchhiking]{barton2013genetic}. 
+The overall rate of coalescence
+depends on the rate of sweeps, their geographic scale, and the rate of recombination,
+and could be calculated by combining the result presented here with
+\citet{barton2013modelling} and \citet{barton2013genetic}. 
+However, if geographic sweeps are common
 then this may substantially speed up the rate of mixing compared to
@@
 how then do selected alleles ever become fixed within species? 
-Drift alone will slowly sort variants within species. While repeated bouts of adaptation in particular genomic regions may act 
+Drift alone will slowly sort variants within species. 
+Furthermore, repeated bouts of adaptation in particular genomic regions may act 
 to push a subset of previously selected alleles to fixation across the
@@
 or dispersal of a subset of individuals. 
-Therefore, specation will cause geographic assortment of
+In this case, speciation will cause geographic assortment of
 polymorphic ancestral variation, again acting to fix variants within
@@
 an allopatric, or at least parapatric, view of the geography of speciation.
-However, it is worth revisiting this idea as population genomic
-sampling becomes geographically broad allowing us to return to themes
+However, it is worth revisiting this idea as geographically broad population genomic
+sampling allows us to return to themes
 in biogeography. 
-Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation within species,
+Along similar lines, Futuyma has argued that much of the adaptive differentiation within species,
 e.g.\ adaptation to local conditions, 
@@
 Futuyma offered this as an explanation of the pattern of punctuated equilibrium \citep{eldredgegould72}, 
-and argued that the observation of stasis and rapid anagensis associated with speciation 
+and argued that the observation of stasis and rapid anagenesis associated with speciation 
 were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
@@
 Furthermore, he suggests that speciation
-may act as rachet to prevent the loss of differentiation, acting to
+may act as ratchet to prevent the loss of differentiation, acting to
 maintain adaptive changes among populations, and prevent their loss by interbreeding. 
@@
 have argued that the rate of speciation may well be quite high,
-but that the majority of incipient species do not persist long due to reabsorption or extinction. 
+but that the majority of incipient species do not persist long due to readsorption or extinction. 
 Repeated bouts of extinction and speciation will send waves of alleles
@@
 
+\plr{This paragraph, and some of the following, are a bit redundant with the above.}
 Futuyma and others have advanced these ideas mainly in the
@@
 Ranges change sufficiently rapidly that
-range size could change substantially over the soujourn time of most alleles 
+range size could change substantially over the sojourn time of most alleles 
 that will eventually become fixed in the species, making it likely that. 
@@
 many ``speciation events'' in addition to the observed ones.
-Futhermore, under this view, 
+Furthermore, under this view, 
 spatial polymorphisms may accumulate gradually in geographically restricted populations
-between large-scale biogeographic events leading to their fixation (or indeed loss). 
+between large-scale biogeographic events leading to their fixation or loss. 
+\plr{I don't understand the following (but haven't read gillespie '94).}
 Such that the origination process of alleles and their fixation process
@@
 extinction can be thought of as very large-scale genetic drift events,
-the initial spread of our alleles is due to selection they are only later sorted 
-into substitutions via these large scale drift events \citep[see also
-][for discussion]{Futuyma:89}.
+the initial spread of alleles is due to selection they are only later sorted 
+into substitutions via these large scale drift events \citep[see also ][for discussion]{Futuyma:89}.
 

### 2014-09-26 standing-variation-paper.tex
minor edits through

@@
 
-\title{The role of standing variation in parallel adaptation}
+\title{The role of standing variation in geographic convergent adaptation}
 \author{
-Graham Coop$^{*,1}$ \\ email: gmcoop@ucdavis.edu 
-\and 
-Peter L. Ralph$^{*,2}$ \\ email: pralph@usc.edu  
+Peter L. Ralph$^{*,1}$ \\ email: pralph@usc.edu  
+\\and\\
+Graham Coop$^{*,2}$ \\ email: gmcoop@ucdavis.edu 
 }
@@
     $^*$ These authors contributed equally to this work. \\
-$^1$ Center for Population Biology \& Department of Evolution and Ecology \\ 
+$^1$ Computational Biology and Bioinformatics \\ 
+University of Southern California, Los Angeles, CA \\
+$^2$ Center for Population Biology \& Department of Evolution and Ecology \\ 
 University of California -- Davis, Davis, CA, 95616\\
-$^2$ Computational Biology and Bioinformatics \\ 
-University of Southern California, Los Angeles, CA \\
 \end{center}
@@
 \section*{Abstract}
+The extent to which shared selection pressures led to a share genetic
+response among populations touches on many questions in evolutionary
+biology. For a number of well studied traits and species it appears
+that convergent evolution within species is common. Here we explore
+the role of standing genetic variation in the evolution of convergent
+genetic responses in a geographically spread population. The
+relatively slow geographic spread of selected alleles allows multiple
+alleles to spread from standing and new mutation before any one comes
+to dominate the population. When these alleles meet, their progress is
+substantially slowed--if the alleles are selectively equivalent, and their initial patterning of the range can be
+thought of as a random tessellation.
+We extend our previous work on this topic 
+by incorporating standing variation, and we show that various properties of a simple
+model of geographic adaptation can be well understood by analogy to a Poisson
+process models of crystallization. 
+In this framework we derive the geographic scale over which a typical allele is expected
+to dominate, and the time it takes the species to adapt as a whole. We
+also study the proportion of adaptive alleles that arise from standing variation. Finally we explore how the negative
+pleiotropic effects of alleles before an environment change can bias
+the subset of alleles that get to contribute to a species adaptive response. 
+Our results suggest that much of adaptation may be geographically
+local even when selection pressures are wide-spread. We discuss the
+implications of these results for arguments of species coherence and
+the nature of divergence between species.
 
@@
 
-There are an increasing number of examples known in which
-different populations of a single species have
+There are an increasing number of examples where
+different populations within a species have
 adapted to similar environments by means of independent genetic changes. 
 In some cases this convergent evolution has produced different phenotypes
-despite shared selection pressures \citep{}; 
-in other cases independent adaptations are identical down the same nucleotide change. 
+despite shared selection pressures; 
+in other cases independent adaptations are identical down the same
+nucleotide change  \citep{Jeong20141,stern2013genetic,MartinOrgogozo:13}. 
 Such convergent evolution within populations has been seen for many
@@
 and the molecular basis of pigmentation changes.
-\plr{The next sentence should be more specific (give example?) or maybe be deleted, I think?}
-In many of these examples we 
-need to be careful about whether the changes are truly convergent or
-whether selection pressures differ subtly. 
 The phrase ``parallel evolution'' is also used to refer to such convergent evolution;
@@
 a primary determinant of the probability that multiple alleles contribute to adaptation
-is the product of the population size and the mutation rate \citep[see ][ for a review]{MesserPetrov}.
+is the product of the population size and the mutation rate \citep[see][for a review]{MesserPetrov}.
 
@@
 since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
-In \citep{ralphcoop2010} we formulated a simple model of parallel adaptation in a geographic setting
+In \citet{ralphcoop2010} we formulated a simple model of parallel adaptation in a geographic setting
 where the selection pressure was constant across the entire environment, 
@@
 common, as has already been seen for a number of traits. 
-On the basis of this we argue, in the discussion, 
+On the basis of this we argue 
 that the genetics of adaptation may often be geographically local, 
@@
 and that most widespread selective sweeps should occur
-when adaptation is highly constrained (e.g.\ by small mutation rate or having to cross a fitness valley).
+when adaptation is highly constrained (e.g.\ by small mutation rate or
+the need for a linked combination of alleles).
 We discuss some implications, and history, of this view for the evolutionary
@@
 The rate of this Poisson process,
-i.e.\ the mean number of nut mutants per unit area and per generation that
+i.e.\ the mean number of new mutants per unit area and per generation that
 appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
@@
 \end{align}
-Note that this is closely related to the equilibrium frequency of a
-allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}.
+Note that this is closely related to the equilibrium frequency of an
+allele at mutation selection balance $\mu/s_d$ \citep{Haldane:27,Haldane:37}.
 
@@
 and initiate the radial growth of new crystals. 
-Eventually, the different orientations of crystals form a random tessellation of space,
+After their initial spread, the different orientations of crystals form a random tessellation of space,
 whose properties have been studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
@@
 each point in space will be first reached by the wave whose origination point lies closest to it.
-Such a random tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
-(i.e.\ the cells formed by assigning points to the nearest point in a Poisson process).
-The properties of this tesselation by alleles is
+Such a random tessellation of space is known as a Poisson-Voronoi tessellation \citep{Moller:94}
+(i.e.\ the cells formed by assigning regions of space to the nearest point in a Poisson process).
+The properties of this tessellation by alleles is
 determined by the spatial locations of the initiation points, 
@@
 beyond dependence on new parameters:
-the cells formed by a Voronoi tesselation have straight sides,
+the cells formed by a Voronoi tessellation have straight sides,
 but the introduction of new mutations cause these to curve.
@@
 but malaria is thought to be the primary driver of these polymorphisms.
-Three G6PD deficiency alleles known to confer resistance to malaria 
-are particularly common and relatively well studied: 
+Three G6PD deficiency alleles are particularly common and relatively well studied: 
 the $A-$ allele found in much of sub-Saharan Africa; 
@@
 consistent with the age of other known malaria resistance alleles. 
-Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
+Population genetic analyses suggest that these three variants each have a hemizygote/heterozygote 
 selection coefficient of $0.05-0.3$
 \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
-This is in agreement with the selection coefficients calculated by
+This is in reasonable agreement with the selection coefficients calculated by
 \citet{Ruwende-g6pd} on the basis of the present day levels of
@@
 over a broad geographic area, makes it a good candidate for a recently
-balanced polymorphism. 
-Hemizygous males and homozygous females suffer from G6PD deficiency,
+balanced polymorphism due to heterozygote advantage \citep[note that
+the conditions for a balanced polymorphism are
+complicated by the hemizygosity of males][]{XXX}. 
+Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
 leading to hemolytic anemia when exposed to a variety of compounds, notably
 those present in fava beans (although, It is unclear if this alone is enough of a
-selection pressure to maintain the allele in the face of its
-considerable advantage). 
+selection pressure to maintain the polymorphism in the face of these alleles
+considerable malaria resistance advantage). 
 The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
@@
 Assuming that the underlying causes of this drop in fitness have not changed, 
-we estimate that $s_d$ has to be roughly XXX to XXX, 
+we estimate that $s_d$ has to be roughly \gc{XXX to XXX}, 
 in order to have resulted in the equilibrium frequency seen today.
-
-% G6PD frequency across world http://www.google.com/imgres?imgurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%3Adoi%2F10.1371%2Fjournal.pmed.1001339.g002%2Flargerimage&imgrefurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%253Adoi%252F10.1371%252Fjournal.pmed.1001339&h=2938&w=2381&tbnid=oD_tX3ea8IbLfM%3A&zoom=1&docid=k37axqAZDS7ljM&ei=5rdOU5TVDMqj8QGntICYBg&tbm=isch&ved=0CF8QMygLMAs&iact=rc&uact=3&dur=1447&page=1&start=0&ndsp=15
-
-% Spatial distribution of G6PD alleles:  http://www.malariajournal.com/content/12/1/418
 
@@
 
+% G6PD frequency across world http://www.google.com/imgres?imgurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%3Adoi%2F10.1371%2Fjournal.pmed.1001339.g002%2Flargerimage&imgrefurl=http%3A%2F%2Fwww.plosmedicine.org%2Farticle%2Finfo%253Adoi%252F10.1371%252Fjournal.pmed.1001339&h=2938&w=2381&tbnid=oD_tX3ea8IbLfM%3A&zoom=1&docid=k37axqAZDS7ljM&ei=5rdOU5TVDMqj8QGntICYBg&tbm=isch&ved=0CF8QMygLMAs&iact=rc&uact=3&dur=1447&page=1&start=0&ndsp=15
+
+% Spatial distribution of G6PD alleles:  http://www.malariajournal.com/content/12/1/418
+
 %IDEA: could we look at all freq. reports of g6pd and see if male and
@@
 on the order of ten million square kilometers.
-In that area there are at least $15$ common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
+In that area there are at least $15$ common variants \citep[see Figure \ref{fig-G6PD-map][]{Howes-g6pd-variants}. 
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
@@
 Clearly, human migration has been shaped both by local dispersal and larger-scale expansions 
-\citep[see ][for a recent discussion]{}, 
+\citep[see][for a recent discussion]{Pickrell:14}, 
 so these parameters only provide a rough view of the process.
@@
 by the speed at which the mutants spread.
-We will define the characteristic length for this new model,
-as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
+We first define a similar characteristic length for this new model. %,
+%as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
 %(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)
@@
 before it is expected that one other successful mutation would have arisen in the area covered so far.
-The explicit formula for $\chi$ is cumbersome; here we omit it.
+The explicit formula for $\chi$ in two dimensions is cumbersome; here we omit it.
 (In one dimension the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.)
@@
  Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
- we expect to find one mutational origin.
+ we expect to find roughly one mutational origin.
 
@@
 smaller characteristic lengths (as would higher mutation rates). 
-Unsurprisingly, allowing standing variation increases the mutational
+Allowing standing variation increases the mutational
 input and so decreases the characteristic length below that predicted
 by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
-As the prior deleterious effect of the allele ($s_d$), the
-characteristic length increases until it reaches that predicted by new mutation alone. 
+In turn, increasing the prior deleterious effect of the allele ($s_d$)
+acts to increase the
+characteristic length until it reaches that predicted by new mutation alone. 
 Higher dispersal distances lead to larger characteristic lengths, 
 since more rapid geographic spread blocks others from arising. 
-A larger selective advantage acts in two conflicting
+A larger selective advantage ($s_b$) acts in two conflicting
 ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
@@
 The effect of helping establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
-\plr{perhaps we should note: (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
-showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1--4}$.) }
+ (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
+showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1--4}$.) 
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter, 
-and the dependence on $s_b$ when only new mutations contribute is much weaker ($s_b^{-1/6}$, equation \eqref{eqn:chi_new})
+as in that case the speed of spread does not matter. 
+The dependence of the characteristic length on $s_b$ when only new mutations contribute is much weaker ($s_b^{-1/6}$, equation \eqref{eqn:chi_new})
 Overall, the range of characteristic lengths observed are reasonably
 consistent with the average diameter of a G6PD variant in Eurasia of
-800km, especially for the lower population density.
+800km, especially for the lower population density, as long as the
+fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was not too low.
 
@@
 {\bf Characteristic length,} as a function of selective disadvantage, 
-compared to the corresponding quantity without standing variation (from \cite{ralphcoop2010}) 
-and to the quantity only considering standing variants (i.e.\ $\sqrt{\lambda_0}$).
-The other parameters are as estimated from G6PD,
-$\mu = 10^{-8}$ and $s_b = XXX$???.
+compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
+and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
+The other parameters are chosen to match those of G6PD,
+$\mu = 10^{-5}$ and $s_b = 0.05$.
 } \label{Fig-G6PD-charlength}
@@
 {\bf Characteristic length,} as a function of selective advantage,
-holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-8}$ (right???).  
-We compare this to the corresponding quantity without standing variation (from \cite{ralphcoop2010}) 
-and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
+holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
+We compare this to the corresponding quantity without standing
+variation (equation \eqref{eqn:chi_new}) 
+and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
 } \label{Fig-G6PD-charlength-sb}
@@
 Imagine ourselves at some geographic location, and let $\tau \ge 0$ be the time at which we are first reached by some advantageous mutation.
-Then, as can be seen from Figure~\ref{fig:cartoon},
-$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
+Then, as can be seen ffrom the
+perspective of the grey dot in Figure~\ref{fig:cartoon}, 
+$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$
+extending back to $t=0$ (light grey dashed lines) is empty of successful mutations.
 Since we assume these are a Poisson process, 
@@
 \end{equation}
-and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \} dt$,
+i.e. the combined probability that area of a circle of radius $vt$ surrounding our
+point at $t=0$ was free of successful standing variants, and no
+successful, new mutations arose in the cone (that has radius $r=vt$ at
+$t=0$ and height $h=t$, and hence volume $\pi r^2 h/3$). 
+Since $\E[\tau] = \int_0^\infty \P\{ \tau > t \} dt$,
 \begin{align}
@@
 (and in the worst case making them deleterious). 
-This effect has been studied by a number of authors \citep{REFS}. 
+This first effect has been well studied by a number of authors
+\citep{Orr:00,Otto:04,WelchWaxman:03,Chevin:10} and its role in precise genetic
+convergence examined \citep{Orr:05,Chevin:10,Unckless:09}. 
 A second contribution is that alleles that have
 less negative pleiotropy are more likely to be present as standing variation 
-before the environmental shift.
+before the environmental shift, and so are more able to respond
+immediately.
 
@@
   the result allowing both standing and \textit{de novo} mutations (equation \eqref{eqn-prop-space-allele-j}).
-  \plr{add other parameter values}
+  Here we hold $s_b=s_{d,2}=0.05$ and $\mu_1=1 \times 10^{-8}$.
   }   \label{fig:pleiotropy_calc}
@@
 In this section we examine this assumption in more detail, although mostly through heuristic arguments,
-and also compare the results above to the results without geographic structure of \citet{softsweeps}.
+and also compare the results above to the results without geographic structure of \citet{softsweepsII}.
 
@@
 should be something close to $\sigma^2 \rho$ 
-(i.e.\ Wright's ``local effective population size'' \citep{XXX}).
+(i.e.\ Wright's ``local effective population size'', \citet{Wright:43}).
 Let $Z_t$ be a continuous-time branching process with $Z_0=1$ and $\E[Z_t] = e^{s_b t}$.
@@
 so that if $\tau$ is the time $Z_t$ reaches size $\sigma^2 \rho$, % i.e.\ $\tau = \inf\{ t \ge 0: Z_t > \sigma^2 \rho\}$,
-then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$ \citep{jaegers1975branching}.
+then $\sigma^2 \rho = Z_\tau \approx e^{s_b \tau} W$ \citep{jagers1975branching}.
 From this we know that $\tau \approx (1/s_b) (\log (\sigma^2 \rho) - \log W)$;
@@
 much faster than under panmixia.
+
+\gc{The results of \citet{softsweepsI} and \citet{softsweepsII} suggest that in a
+panmictic population the number of independent alleles (and their frequencies), in a sample,
+is relatively independent of $s_b$ and $s_d$  \citep[although this
+breaks down with fluctuating population size,
+][]{Wilson-softsweep}. In the panmictic model the
+lack of dependence on $s_b$ comes about because while increasing $s_b$
+increases the rate at which independent mutations become established,
+it also accelerates the frequency gain of established alleles decreasing the time period in which new alleles can arise and
+hope to be at significant frequency in the population. These two
+effects cancel each other out leading to no strong effect of $s_b$ on
+the number of independent alleles. 
+Decreasing $s_d$ increases the number of standing variants within a
+population, increasing the number of alleles that manage to establish
+and spread from standing variation
+\citet{softsweepsI,Orr:01}. However, having more established standing alleles
+acts to exclude the spread of new alleles that arise once the environment
+switches. These two opposing effects again cancel out leading to little overall
+effect of $s_d$ on the number of independent alleles.
+In contrast our results show that the
+characteristic length (closely related to the density of independent
+alleles) depends on both $s_b$ and $s_d$ in a geographical spread population. 
+Like the panmictic model our alleles also act to exclude one and
+other, however, the geographic spread of an allele is slow (area grows
+as $s_b$) compared to the initial exponential growth of an allele in a
+panmictic population. That means that the role of selection in helping
+alleles become established can dominate leading to more independent origins, both by
+being  weaker before, and stronger after, the environmental switch.
+
+}
+
 % We could compare the probability of parallel adaptation under the two models (see eq'n 18 in \citet{softsweepsII}),
@@
 especially when standing variation is not so deleterious that it is present before environmental shifts. 
-
-Allowing standing variation in our model, unsurprisingly, 
-increases the extent of convergence among populations.
-Importantly, it may greatly lower the time until the species becomes adapted 
+When the geographical area where a species experiences a 
+selection pressure is greater than our characteristic
+length we should expect multiple independent alleles to arise and
+spread in that area. Allowing standing variation in our model  
+decreases the characteristic length, i.e. it increases the extent of
+convergence among populations.
+While at face value this increase seems unsurprising, as explained above, this
+differs markedly from the case of multiple competing alleles in a panmictic population \citep{softsweepsI,softsweepsII}. 
+Importantly, allowing standing variation may greatly lower the time until the species becomes adapted 
 across the geographic range of the selection pressure. 
@@
 
+\paragraph{The confusing signal of geographic convergent evolution.}
 As we have argued in \citet{ralphcoop2010} the ease with which
@@
 has led to the suggestion that these individuals had dark
-pigmentation. However, given the results here and the partial convergent
-adaption of skin pigmentation between Europeans and East Asians
-\citep{} it seems just as plausible that these ancient individuals
-simply had a different complement of ``light-skin'' pigmentation
+pigmentation \citep{olalde2014derived,lazaridis2013ancient, Wilde2014}. However, given our results and the partially convergent
+basis of skin pigmentation between Europeans and East Asians
+\citep{norton-human-pigmentation,edwards2010pigmentation} it seems just as plausible that these ancient individuals
+simply had adapted to high latitudes via a different complement of ``light-skin'' pigmentation
 alleles. Such convergence may considerably complicate the 
-exploration of adaptation among populations using variants mapped in a
-limited set of populations \citep{BergCoop:14}.
+exploration of phenotypes and adaptation among populations using variants mapped in a
+limited set of populations \citep{Berg:14}.
 
@@
 the alleles concerned have no bearing on the ecological differences
-across these breaks.
+across these breaks \citep[see][ for a wide-ranging discussion of how
+allelic differentiation may build along particular zones]{bierne2011coupling}. 
+\gc{In cases, such as malaria resistance and skin pigmentation, where we
+know something about the genetics, phenotypic distributions, 
+and potential selection agents we can usually avoid mistaking the }
+
+%In cases, such as G6PD, where we have a relatively clear understanding of the genetics and
+%phenotypic consequences we can avoid confusing the convergent response
 
@@
 set of geographically restricted long haplotypes. 
-Given sufficient time these haplotypes will mix together, 
-leading to a signal of a sweep with multiple mutations if our selected allele occurs at the same locus, 
-or to multiple partial sweeps if the loci are scattered across the genomes.
-
-\paragraph{}
+Given sufficient time these haplotypes will mix together through
+migration and drift, 
+potentially leading to a signal of a sweep from multiple independent
+mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
+or to multiple partial sweeps if the loci are scattered across the
+genomes.
+
 Our results are predicated on the idea that adaptive variants are
 initially rare within populations, i.e.\ they are reasonably
-deleterious before the environment switches. In constrast, even distant populations could have
+deleterious before the environment switches. In contrast, even distant populations could have
 a shared basis of adaptation if they adapt via common variation,
@@
 If many loci contribute to variation in a trait, then
-selection on any one allele may be weak, allowing adaptation to use different alleles in different populations.
-However, in that case populations separated by a reasonable amount of drift may still
+selection on any one allele may be weak, which might adaptation to use
+the same alleles in different populations.
+However, populations separated by a reasonable amount of drift may still
 adapt via different genetic routes, as the constellation of alleles
-at intermediate frequency will be somewhat different,
+at intermediate frequency will be somewhat different \citep{Barton:89},
 and these are the alleles that selection moves most quickly to change the trait value.
@@
 % i.e.\ those at intermediate frequency, will be somewhat different. 
-So it may be the case that polygenic traits are even less suspectible to a shared
+So it may be the case that polygenic traits are even less susceptible to a shared
 genetic basis to adaptation across populations than simple traits.   
@@
 based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
-\citet{Rieseberg2001} and \citet{MorjanRieseberg:04} argued that 
-even if gene flow is not high enough to prevent neutral differentiation, 
-species should be viewed as being united 
-if levels of gene flow were still high enough for selected alleles to
-spread across the entire species \citep[see also][]{Ellstrand2014s}.
+Even if gene flow is not high enough to prevent neutral
+differentiation or local adaptation, a number of authors have argued
+that that species are cohesive as gene flow were still high enough for
+globally selected alleles, and hitchhiking haplotypes, to
+spread across entire species \citep[see also][]{Rieseberg2001,MorjanRieseberg:04,Ellstrand2014}.
 At present, large scale genotyping and sequencing projects, 
 along with more sophisticated methods, 
-are highlighting ever more signals of gene flow between populations and species \citep{}.
+are highlighting ever more signals of gene flow between populations and species \citep{Patterson:12,SousaHey:13,Hellenthal:14}.
 However, our work on geographic convergent adaptation
@@
 
-In support of this idea, recent selective sweeps seem to often be 
-geographically restricted \citep{Coop}, but see XXX. 
+In support of this idea, putative, recent selective sweeps seem to often be 
+geographically restricted \citep{Pickrell:09,Coop:09,Granka:14},
+rather than range-wide \citep[but see][ for a potential example]{Clark:07,Long:13}. 
 This is likely in part due to the relatively low incidence of wide-spread selection pressures,
@@
 introgression of adaptive alleles across species and sub-species boundaries, 
-suggests that selected alleles do sometimes spread despite low migration rates \citep{hedrick2013adaptive}.
+suggests that selected alleles do sometimes spread despite low
+migration rates \citep[see ][ for a recent review]{hedrick2013adaptive}.
 However, at least some of these cases 
@@
 This is rapidly changing,
-and will allow us to form an improved picture of the relationship 
+and will allow us to form a much improved picture of the relationship 
 between the level of neutral population structure, 
-and the age and geographic extent of spread of selected alleles.
+and the age and geographic spread of selected alleles across many species.
 
@@
 Under neutral mixing, the mean number of generations back to the most recent common ancestor
-is on order of the total population size.
+is on order of the total effective population size.
 This quantity has not been worked out for a model with simultaneous local sweeps,
@@
 Futuyma argued that despite rapid adaptation over short time-scales, we
-may observe morphological stasis in the fossil record as much of this adaptation is lost 
+may observe morphological stasis in the fossil record as much of this
+adaptation is lost to local extinction and the collapse of population structure
 \citep[see also][]{lieberman:96,Eldredge:05}. 
@@
 have argued that the rate of speciation may well be quite high,
-but that the majority of incipient species do not persist long due to readsorption or extinction. 
+but that the majority of incipient species do not persist long due to reabsorption or extinction. 
 Repeated bouts of extinction and speciation will send waves of alleles
 to fixation along particular lineages. 
-
-\plr{This paragraph, and some of the following, are a bit redundant with the above.}
-Futuyma and others have advanced these ideas mainly in the
-setting of adaptation to local conditions 
-\citep[particularly local communities ][]{} . 
-Much of the recent adaptation seen by
-population genetics may be adaptation to local conditions, 
-but as we have shown, even wide-spread selection pressures will likely result in 
-geographically restricted spread of alleles.
-Ranges change sufficiently rapidly that
-range size could change substantially over the sojourn time of most alleles 
-that will eventually become fixed in the species, making it likely that. 
-The adaptive substitutions we see along lineages may be those alleles 
-simply lucky enough to have spread in local populations or incipient species that escaped extinction.
+%Ranges change sufficiently rapidly that
+%range size could change substantially over the sojourn time of most alleles 
+%that will eventually become fixed in the species, making it likely that.
+
+%\plr{This paragraph, and some of the following, are a bit redundant with the above.}
+%Futuyma and others have advanced these ideas mainly in the
+%setting of adaptation to local conditions 
+%\citep[particularly local communities ][]{} . 
+%Much of the recent adaptation seen by
+%population genetics may be adaptation to local conditions, 
+%but as we have shown, even wide-spread selection pressures will likely result in 
+%geographically restricted spread of alleles.
+%The adaptive substitutions we see along lineages may be those alleles 
+%simply lucky enough to have spread in local populations or incipient species that escaped extinction.
 
@@
 substitutions should necessarily be thought of as being clustered at splits in inferred phylogenies 
-\citep[see ][for a recent exhange on this]{Pennell:14,Venditti:14,Pennell:14B}. 
+\citep[see ][for a recent exchange on this]{Pennell:14,Venditti:14,Pennell:14B}. 
 We first note that neutral substitutions are unaffected by this process, 
@@
 between large-scale biogeographic events leading to their fixation or loss. 
-\plr{I don't understand the following (but haven't read gillespie '94).}
-Such that the origination process of alleles and their fixation process
-are reasonably uncoupled \citep[in the sense of ][]{Gillespie:94} by the geographic structure.
+Such that the origination time of alleles and when they fix could be 
+reasonably uncoupled by the geographic structure \citep[in a manner,
+similar to that argued by ][]{Gillespie:94}.
 
@@
 
-This view suggests that the alleles that substitute along a lineage
-may rarely be capable, or need, to spread species-wide. This is not
-inconsistent with genome-wide evidence that a reasonable fraction of 
+%The adaptive substitutions we see along lineages may be those alleles 
+%simply lucky enough to have spread in local populations or incipient species that escaped extinction.
+
+There is genome-wide evidence that a reasonable fraction of 
 substitutions are fixed by positive selection in a number of species
-\citep[most notably Drosophila][]{}, as under this view selection has played a strong
-role in the establishment of these alleles and they not contribute to
-polymorphism within geographically narrow samples. It will be
-interesting to see as we get more broadly geographic population
-genomics sampling for a range of species whether the class of alleles that contribute 
+\citep[most notably Drosophila, ][]{Sella:09}. Under this geographic view of
+fixation, selection has played a strong
+role in the establishment of these alleles locally. As we get more broadly geographic population
+genomics sampling for a range of species there is an opportunity to
+study whether the class of alleles that contribute 
 to local differentiation are the similar as those underlying species
@@
 
-Finally, we close by noting that range expansion, and speciation, are obviously not seperate from
+Finally, we close by noting that range expansion, and speciation, are obviously not separate from
 adaptive differentiation. The invasion of novel geographic spaces may
@@
 adaption to novel environments. Conversely, the relatively slow geographic spread of adaptive
-alleles if they only offer a local advantage, or the selective exclusion
-described here, may allow Dobzhansky-Muller incompatibilities (DMIs) to arise
+alleles within ranges, e.g. if they only offer a local advantage or
+if they are selective excluded, may allow Dobzhansky-Muller incompatibilities (DMIs) to arise
 within species effectively offering a mechanism for 
-hybrid incompatibilites to evolve in parapatry effectively fracturing
-the range \citep{}. Indeed the alleles that act as components in many
+hybrid incompatibilities to evolve in parapatry effectively fracturing
+the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. Indeed the alleles that act as components in many
 of the DMIs studied to
 date are geographically restricted \citep[see][]{Cutter:12}. As such
-it seems possible that populations within species are always on the
+it seems possible that populations within species may often be on the
 verge of speciation, and that as outlined here that this may drive
 some proportion of molecular divergence.
+
+\subsubsection*{Acknowledgements}
 
@@
 \end{equation}
-in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic paramters.
+in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic parameters.
 Since $a$ and $b$ only depend on the dimension and the quantity being computed,

### 2014-09-27 standing-variation-paper.tex
changes throughout, also updated figs

@@
   \caption{
-Cartoon of the geographic spread of standing and new mutations. 
-The mean spatial density of successful standing variants at $t=0$ (lightning bolts) is $\lambda_0$,
-and the mean density in space $\times$ time of successful new mutations (stars) is $\lambda$.
-\plr{Add $t=0$ and direction of time to the figure.}
+{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. Time
+runs down the page, and a single spatial dimension runs across the
+page. The environment switches at $t=0$, and following that adaptive
+alleles that are successful, at escaping loss by drift, spread from
+standing variants (lightning bolts) and new mutations (stars). 
+The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
+and the mean density in space $\times$ time of successful new mutations is $\lambda$.
+Our successful alleles spread spatially outwards at speed $v$ and hence carve
+space-time cones in previously unadapted areas. When our adaptive
+alleles meet their rapid advance is halted due to them selectively
+excluding one another, and so their boundaries form straight lines on
+short time-scales (overly longer time-scales they will mix into each
+other). The grey dashed line shows a cone radiating backward in space-time,
+originating from a point $(x,t)$, with slope $v$, its radius at $t=0$
+is $v t$. Any successful new allele that arose within this backwards
+pointing cone, and any successful standing variant present on the base of this cone, would have spread to $x$ by time $t$.
 }
@@
 Whether these all confer resistance to malaria is unknown,
-but malaria is thought to be the primary driver of these polymorphisms.
+but malaria is thought to be the primary driver of these polymorphisms
+\citep[see][ for a general review]{hedrick2011population}.
 Three G6PD deficiency alleles are particularly common and relatively well studied: 
@@
 the conditions for a balanced polymorphism are
-complicated by the hemizygosity of males][]{XXX}. 
+complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
 Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
@@
 G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
-Assuming that the underlying causes of this drop in fitness have not changed, 
-we estimate that $s_d$ has to be roughly \gc{XXX to XXX}, 
+Assuming that the underlying causes and strength of this drop in fitness have not changed, 
+we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
+\geq 0.05$),
 in order to have resulted in the equilibrium frequency seen today.
@@
 on the order of ten million square kilometers.
-In that area there are at least $15$ common variants \citep[see Figure \ref{fig-G6PD-map][]{Howes-g6pd-variants}. 
+In that area there are at least $15$ common variants \citep[see Figure \ref{fig-G6PD-map}][]{Howes-g6pd-variants}. 
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
@@
 \caption{ 
-Map of G6PD-deficiency allele frequencies across Asia. 
+{\bf Map of G6PD-deficiency allele frequencies across Asia.} 
 The pie chart shows the frequency of G6PD-deficiency alleles. 
@@
 \caption{ %
-{\bf Characteristic length,} as a function of selective disadvantage, 
+{\bf Characteristic length, as a function of selective disadvantage}, 
 compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
@@
 The other parameters are chosen to match those of G6PD,
-$\mu = 10^{-5}$ and $s_b = 0.05$.
+$\mu = 10^{-5}$ and $s_b = 0.05$. As the characteristic length with
+only standing variation is independent of dispersal distance it is
+plotted as a single black dotted line. 
 } \label{Fig-G6PD-charlength}
@@
 \caption{ %
-{\bf Characteristic length,} as a function of selective advantage,
+{\bf Characteristic length, as a function of selective advantage},
 holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
@@
 Imagine ourselves at some geographic location, and let $\tau \ge 0$ be the time at which we are first reached by some advantageous mutation.
-Then, as can be seen ffrom the
+Then, as can be seen from the
 perspective of the grey dot in Figure~\ref{fig:cartoon}, 
@@
 in two dimensions
-\begin{equation} 
+\begin{equation} \label{eqn-no-mut-t}
     \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
@@
 
-We can also find the mean proportion of space covered by standing variants.
-If by time $t$ a geographic location has not yet been reached by the mutation,
-the probability it will be reached by time $t+dt$ 
+We can also find the mean proportion of space covered by standing
+variants (we will restrict ourselves to two dimensions).
+At time $t$ a geographic location has not yet been reached by the
+mutation with probability given by equation \eqref{eqn-no-mut-t}.
+The probability, given that it has not been reached by $t$, that it will be reached by time $t+dt$ 
 by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
-and the probability it is reached by a new variant is $\lambda \pi v^2 t^2 dt$.
+which is $\lambda_0$ multiplied by the thin slice of extra area in our
+expanded circle at $t=0$, which has gone from a radius $vt$ to
+$v(t+dt)$. The corresponding probability the point is reached by a new
+variant is 
+$\lambda \pi v^2 t^2 dt$, which is $\lambda$ multiplied by the
+sliver of  extra volume in our space-time cone at time $t+dt$ compared to that at time $t$.
 Therefore, as is standard for competing exponentials,
@@
 alleles that spread from standing variation, and the proportion of
-geographic space covered by standing variants.
+geographic space covered by standing variants for parameters chosen to match our G6PD example.
+Even for relatively large deleterious costs prior to the environmental
+switch standing variants still makes up quite large proportion of the
+adaptive alleles, and an even larger proportion of the range (as they
+get to occupy a larger area than new mutations, as they start earlier).
 
@@
 (equation \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
-} \label{G6PD_standing_var_proportion}
+We set $\mu=10^{-5}$ and $s_b=0.05$ to match our G6PD example.} \label{G6PD_standing_var_proportion}
 \end{center}
@@
 While many alleles may confer the beneficial phenotype, 
-not all will contribute to adaptation if they have negative pleiotropic consequences.
+not all will contribute equally to adaptation if they have negative pleiotropic consequences.
 There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
@@
 \end{align}
-\plr{note: I added a 2 to $\lambda_j$.}
-Using these rates it follows that, at the time when every location has been reached by an adaptive allele,
+Using these rates, and an argument analogous to that used to derive
+equation \eqref{prop_space_standing},
+at the time when every location has been reached by an adaptive allele,
 the proportion of the species range covered by alleles of type $j$ is
 \begin{equation} \label{eqn-prop-space-allele-j}
-    p_j = \int_0^\infty  \left( 2 \lambda_{0,j} \pi v^2 t + \lambda_j \pi v^2 t^3 \right)  
-    \; \exp \left( - \sum_k ( \lambda_{0,k} \pi v^2 t^2 + \lambda_k \pi v^2 t^3 / 3 ) \right) dt .
+    p_j = \int_0^\infty  \left( 2 \lambda_{0,j} \pi v^2 t +  \lambda_j \pi v^2 t^2 \right)  
+    \; \exp \left( - \sum_k ( \lambda_{0,k} \pi v^2 t^2 + \lambda_k
+      \pi v^2 t^3 / 3 ) \right) dt .  %%deleted \lambda_j \pi v^2 t^3
+                                %%out of term out front
 \end{equation}
@@
 i.e.\ if all variants are highly deleterious before the environment switches, 
-then $p_j =\mu_j / (\sum_k  \mu_k)$.
+ $p_j =\mu_j / (\sum_k  \mu_k)$.
 
@@
 let's imagine the somewhat extreme scenario that there is a single base pair 
-at which a possible mutation is free of negative pleiotropy (call this class $1$);
+at which a possible mutation is relatively free of negative pleiotropy (call this class $1$);
 and a larger mutational target where changes 
@@
 assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
-and think of the second class as ten, a hundred, or one thousand base pairs. 
+and think of the second class of alleles as arising at one of ten, a hundred, or one thousand base pairs. 
 We show the expected proportion of space covered by
@@
 Intuitively, the contribution of the rarer mutation decreases as the
-mutational target of the other class becomes larger, and as the
+mutational target of the second class becomes larger, and as the
 difference in the negative pleiotropic consequences of the two classes of alleles decreases. 
 The case with standing variation only is the best case scenario 
-for the rarer mutation, so its rate of introduction is necessarily lower. 
+for the rarer mutation, so its rate of introduction after $t=0$ is necessarily lower. 
 However, the standing-variation-only case does seem to provide a
@@
 
+\gc{As discussed above, we could incorporate differences in the beneficial selection
+coefficients of the different classes of alleles, to allow for
+negative pleiotropic effects acting to surpress the advantage of an
+allele once the environment switches (effect one above). One simple way to do
+this is to simply replace $s_b$ with $s_{b,j}$ resulting in a class-specific new allele establishment rates ($\lambda_{j}$) and rates of spread ($v_j$).
+These then could be used in equation
+\eqref{eqn-prop-space-allele-j}. We could imagine our two class model, outlined
+above, with class one having the additional advantage of
+$s_{1,b}>s_{2,b}$. This would further increase the contribution of the rarer class,
+because our rare class would overcome drift more often and spread more
+rapidly. However, a straight-forward application of the logic of used
+to construct equation \eqref{eqn-prop-space-allele-j},
+ignores the fact that classes of alleles with higher $s_b$ is not
+completely selected excluded from spreading into geographic areas occupied by alleles with lower
+$s_b$. Given enough time the most advantageous type (type 1, in this case) would arise and
+spread everywhere, and so if we used different $s_b$s in equation
+\eqref{eqn-prop-space-allele-j}, it would only be a short-term approximation to
+a longer term dynamic. Even after the initial tessellation has formed the allele 1s would have an selective advantage $\delta s
+=s_{b,1}-s_{b,2}$ in areas where allele 2 have already spread, and so
+would arise at rate $2\rho \mu_1 \delta s $ in these areas and spread through them at
+speed $\sigma \sqrt{\delta s}$. An extension of our Poisson process model could incorporate these
+ effects, by thinning the Poisson process of establishing mutations by
+ correctly. 
+Whether allele $2$ persists would
+depend on the linkage arrangements between loci. If the loci of allele
+1 and 2 are unlinked, then allele 1 can spread without disrupting
+allele 2. However, if they are linked the spread of allele $1$ may
+push allele $2$ out of the population. More complicated dynamics,
+including spatial Dobzhansky-Muller incompatibilities \citep{,Kondrashov:03,ralphcoop2010}, could ensue if
+there are epistatic interactions between the alleles. }
+
+% An extension of this Poisson process model could incorporate these
+% effects, thinning out mutations  
+% It could be interesting to investigate pleiotropy's combined
+% effect of adaptation, through standing variation and rate of spread, by coupling together
+% $s_{b,k}$ and $s_{d,k}$ perhaps through Fisher's geometric model.
+% Using Fisher's geometric model we could allow $s_{b}$ and $s_d$ to be coupled as a consequence of being an
+% the fitness in two related environments of some underlying phenotype
+% drawn from a continuous distribution,
+% rather than our $k$ discrete classes \citep{Chevin:10}.
+
+%While it may be of interest to explore this model
+%incorporating both effects of negative ple
+
 \begin{figure}[ht]
@@
 \caption{ %
-{\bf Proportion of space covered by rarer but less pleiotropic
+{\bf Proportion of space covered by rarer but less negatively pleiotropic
   mutation}. Empty circles give the result for standing variation only,
@@
 So, roughly speaking, we need to evaluate the importance of a delay of about $T = (1/s_b) \log (\sigma^2 \rho)$.
-New mutations will appear and fix during this time if $4N\mu s_b \ge 1/T$,
-i.e.\ if $4 N \mu \ge 1/\log (\sigma^2 \rho)$.
+New mutations will appear and become established during this time if
+\gc{$4 \rho \sigma^2 \mu s_b \ge 1/T$,
+i.e.\ if $4 \rho \sigma^2 \mu \ge 1/\log (\sigma^2 \rho)$}.
 Our model will still be a good approximation, however, 
@@
 and spread from standing variation
-\citet{softsweepsI,Orr:01}. However, having more established standing alleles
+\citep{softsweepsI,Orr:01}. However, having more established standing alleles
 acts to exclude the spread of new alleles that arise once the environment
@@
 While at face value this increase seems unsurprising, as explained above, this
-differs markedly from the case of multiple competing alleles in a panmictic population \citep{softsweepsI,softsweepsII}. 
+differs markedly from the case of multiple competing alleles in a
+panmictic population \citep{softsweepsI,softsweepsII}. 
 Importantly, allowing standing variation may greatly lower the time until the species becomes adapted 
@@
 alone. This is especially problematic as boundaries between convergent alleles may often
-occur where gene flow rates are low, i.e.\ ecological breaks, even if
+occur where gene flow rates are low, i.e.\ historical and ecological breaks, even if
 the alleles concerned have no bearing on the ecological differences
@@
 allelic differentiation may build along particular zones]{bierne2011coupling}. 
-\gc{In cases, such as malaria resistance and skin pigmentation, where we
-know something about the genetics, phenotypic distributions, 
-and potential selection agents we can usually avoid mistaking the }
+\gc{We are rarely as fortunate, to know as much about the 
+ genetics, phenotypic distributions, 
+and potential selection agents as we do for malaria resistance. And so we must be wary of mistaking the
+strange spatial distributions of particular alleles for 
+adaptation to some very specific selection pressure (e.g. the distribution of the Mahidol allele in Figure
+\ref{fig-G6PD-map}), when they are simply elements of a larger geographic mosaic of
+alleles responding to a broadly shared selection pressure. }
 
@@
 or to multiple partial sweeps if the loci are scattered across the
-genomes.
+genome \citep{CoopRalph:12}.
 
@@
 adapt via different genetic routes, as the constellation of alleles
-at intermediate frequency will be somewhat different \citep{Barton:89},
-and these are the alleles that selection moves most quickly to change the trait value.
+best able to respond to selection will be somewhat different \citep{Barton:89}.
 % best positioned to initially respond to positive selection, 
@@
 genetic basis to adaptation across populations than simple traits.   
-However, we currently lack good methods with which to test this.
+However, we currently lack good models and methods with which to test this.
 % models of this, and our ability to infer
@@
 Mayr and many others have argued that species are coherent
-evolutionary units because they are united by shared gene flow \citep[pages 521--522 ][]{Mayr:SpeciesEvol}. 
+evolutionary units because they are united by shared gene flow \citep[pages 521--522][]{Mayr:SpeciesEvol}. 
 However, this argument has been questioned by a number of authors 
@@
 \citep[see also][]{ralphcoop2010,RalphCoop:14} 
-concludes that this gene flow will lead to the rapid spread of
+bring the idea that this gene flow will lead to the rapid species-wide spread of
 selected alleles partially into question. 
@@
 environment).  
-Furthermore,
-currently we can only scan genomes for species-wide sweeps
-in those few organisms with population-scale sequence data.
+Currently we can only scan genomes for species-wide sweeps
+in those few organisms with population-scale sequence data, and so we
+are very limited in our view .
 This is rapidly changing,
@@
 and pairs of lineages caught up in the same sweep could be forced to coalesce 
-\citep[see ][for work on geographic hitchhiking]{barton2013genetic}. 
-The overall rate of coalescence
+\citep[see ][for work on geographic
+hitchhiking]{barton2013genetic}. In this case lineages and alleles are literally
+hitchhiking across space.
+The overall rate of lineage movement and coalescence
 depends on the rate of sweeps, their geographic scale, and the rate of recombination,
@@
 how then do selected alleles ever become fixed within species? 
-Drift alone will slowly sort variants within species. 
+Drift alone will act only slowly to sort variants within species into
+divergence among species. 
 Furthermore, repeated bouts of adaptation in particular genomic regions may act 
@@
 initial rapid spread of selected alleles.
-
+\
 \paragraph{Speciation and extinction as phases of molecular evolution?}
@@
 induced by extirpation of the species over parts of its range
-\citep[see ][ for how such a model could be constructed]{barton2013modelling}. 
+\citep[see ][and reference therein for how such a model could be constructed]{barton2013modelling}. 
 Such drops in range size may fix, or radically change the species-wide frequency, 
@@
 and argued that the observation of stasis and rapid anagenesis associated with speciation 
-were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
+were consistent with micro-evolution. 
 Futuyma argued that despite rapid adaptation over short time-scales, we
@@
 alleles within ranges, e.g. if they only offer a local advantage or
-if they are selective excluded, may allow Dobzhansky-Muller incompatibilities (DMIs) to arise
+if they are selective excluded, may allow Dobzhansky-Muller incompatibilities to arise
 within species effectively offering a mechanism for 
 hybrid incompatibilities to evolve in parapatry effectively fracturing
-the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. Indeed the alleles that act as components in many
-of the DMIs studied to
+the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. Indeed the
+alleles that act as components in many of the Dobzhansky-Muller incompatibilities studied to
 date are geographically restricted \citep[see][]{Cutter:12}. As such
@@
 \subsubsection*{Acknowledgements}
+We thank Michael Turelli, Jon Seger and the Coop lab for helpful conversations.
 

### 2014-09-27 g6pd_figures.R

@@ -10,27 +10,27 @@ my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
 ###################Char length as function of sb
 ###################################################
 
- mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.001,0.3,length=800); 
+ mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.00001,0.3,length=1600); 
 
 sigmas=c(100,50,10)
 
 layout(t(1:2))
-par(mar=c(4,4.1,2,0.3))
+par(mar=c(4,4.1,2,0.5))
 for(j in 1:2){
 	rho=rhos[j]
 	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(chi),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
 	if(j == 2) plot(sbvals,sbvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.4,main=expression(rho==2))
 	for(i in 1:3){
 	sigma=sigmas[i]; charlength.sb<-sapply(sbvals, function (sb) { charLength(mu,rho,sb,sd=sd,sigma)$value } )
-	 lines(sdvals,charlength.sb,lwd=2,col=my.col[i]); 
+	 lines(sbvals,charlength.sb,lwd=2,col=my.col[i]); 
 	 
 	 charlength.new.only<-sapply(sbvals, function (sb) {  newCharLength(mu, rho, sb, sd=NA, sigma)$value})
-	 lines(sdvals,charlength.new.only,col=my.col[i],lwd=2,lty=2)  ##new mut. only
+	 lines(sbvals,charlength.new.only,col=my.col[i],lwd=2,lty=2)  ##new mut. only
 	}
 	 ##contribution of standing var only, i.e. old variation
 	 charlength.standing.only<-sapply(sbvals, function (sb) {  oldCharLength(mu, rho, sb, sd=sd, sigma)$value})
-	 lines(sdvals,charlength.standing.only ,lwd=2,col="black",lty=3)
[…]

### 2014-09-29 standing-variation-paper.tex
changes throughout, also updated figs
abstract

@@
 
+% standing deleterious variation
+% higher mutation rate => more importance of standing
+% G6PD a combination of new and standing
+% interpolates betw panmixia and 2010
+% pleiotropy pre-shift biases adaptation
+% establishment phase
+% selective sweeps and species coherence, gene flow
+
 \section*{Abstract}
-The extent to which shared selection pressures led to a share genetic
-response among populations touches on many questions in evolutionary
-biology. For a number of well studied traits and species it appears
-that convergent evolution within species is common. Here we explore
-the role of standing genetic variation in the evolution of convergent
-genetic responses in a geographically spread population. The
-relatively slow geographic spread of selected alleles allows multiple
-alleles to spread from standing and new mutation before any one comes
-to dominate the population. When these alleles meet, their progress is
-substantially slowed--if the alleles are selectively equivalent, and their initial patterning of the range can be
-thought of as a random tessellation.
-We extend our previous work on this topic 
-by incorporating standing variation, and we show that various properties of a simple
-model of geographic adaptation can be well understood by analogy to a Poisson
-process models of crystallization. 
-In this framework we derive the geographic scale over which a typical allele is expected
-to dominate, and the time it takes the species to adapt as a whole. We
-also study the proportion of adaptive alleles that arise from standing variation. Finally we explore how the negative
-pleiotropic effects of alleles before an environment change can bias
-the subset of alleles that get to contribute to a species adaptive response. 
-Our results suggest that much of adaptation may be geographically
-local even when selection pressures are wide-spread. We discuss the
-implications of these results for arguments of species coherence and
-the nature of divergence between species.
+The extent to which populations experiencing shared selective pressures
+adapt through a shared genetic response
+is relevant to many questions in evolutionary biology. 
+In a number of well studied traits and species, 
+it appears that convergent evolution within species is common. 
+In this paper, 
+we explore how standing, deleterious genetic variation contributes to
+convergent genetic responses in a geographically spread population,
+extending our previous work on the topic.
+Geographically limited dispersal slows the spread of each selected allele,
+hence allowing other alleles 
+-- newly arisen mutants or present as standing variation -- 
+to spread before any one comes to dominate the population. 
+When such alleles meet, their progress is substantially slowed 
+-- if the alleles are selectively equivalent, 
+they mix slowly, dividing the species range into a random tessellation,
+which can be well understood by analogy to a Poisson process model of crystallization. 
+In this framework, 
+we derive the geographic scale over which a typical allele is expected to dominate, 
+the time it takes the species to adapt as a whole,
+and the proportion of adaptive alleles that arise from standing variation.
+Finally, we explore how negative pleiotropic effects of alleles before an environment change 
+can bias the subset of alleles that get to contribute to a species adaptive response. 
+We apply the results to the many geographically localized G6PD deficiency alleles thought to confer resistance to malaria,
+whose large mutational target size and deleterious effects make them likely candidates to have been present as deleterious standing variation.
+We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
+which suggest that these arose both from standing variation and new mutations since the advent of malaria.
+Our results suggest that much of adaptation may be geographically local 
+even when selection pressures are wide-spread. 
+We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.
 
@@
   \caption{
-{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. Time
-runs down the page, and a single spatial dimension runs across the
+{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
+Time runs down the page, and a single spatial dimension runs across the
 page. The environment switches at $t=0$, and following that adaptive
@@
 \geq 0.05$),
-in order to have resulted in the equilibrium frequency seen today.
+in order to have resulted in the equilibrium frequency seen today in
+areas with endemic malaria.
 
@@
 plotted as a single black dotted line. 
+\plr{$y$ axis label: kilometers}
 } \label{Fig-G6PD-charlength}
@@
 and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
+\plr{$y$ axis label: kilometers}
 } \label{Fig-G6PD-charlength-sb}

### 2014-09-29 standing-variation-paper.tex
small coop changes
peter edited through

@@
 each point in space will be first reached by the wave whose origination point lies closest to it.
-Such a random tessellation of space is known as a Poisson-Voronoi tessellation \citep{Moller:94}
+This random tessellation of space is known as a Poisson-Voronoi tessellation \citep{Moller:94}
 (i.e.\ the cells formed by assigning regions of space to the nearest point in a Poisson process).
@@
   \begin{center}
-    \includegraphics{spreading_alleles_trimmed}
+    \includegraphics[width=\textwidth]{spreading_alleles_trimmed}
   \end{center}
@@
 {\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
-Time runs down the page, and a single spatial dimension runs across the
-page. The environment switches at $t=0$, and following that adaptive
-alleles that are successful, at escaping loss by drift, spread from
-standing variants (lightning bolts) and new mutations (stars). 
+Time runs down the page, and a single spatial dimension runs across the page. 
+The environment switches at $t=0$, 
+following which adaptive alleles that successfully escape loss by drift
+spread locally from both standing variants (lightning bolts) and new mutations (stars). 
 The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
 and the mean density in space $\times$ time of successful new mutations is $\lambda$.
-Our successful alleles spread spatially outwards at speed $v$ and hence carve
-space-time cones in previously unadapted areas. When our adaptive
-alleles meet their rapid advance is halted due to them selectively
-excluding one another, and so their boundaries form straight lines on
-short time-scales (overly longer time-scales they will mix into each
-other). The grey dashed line shows a cone radiating backward in space-time,
-originating from a point $(x,t)$, with slope $v$, its radius at $t=0$
-is $v t$. Any successful new allele that arose within this backwards
-pointing cone, and any successful standing variant present on the base of this cone, would have spread to $x$ by time $t$.
+These successful alleles spread spatially outwards at speed $v$, 
+so the space-time profile of their spread forms a cone.
+When these adaptive alleles meet, 
+their rapid advance is halted because they no longer have an advantage,
+so their boundaries form straight lines on short time-scales 
+(over longer time-scales, they will mix into each other). 
+The grey dashed line shows a cone radiating backward in space-time,
+originating from a point $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
+Point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
 }
@@
 Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to a variety of compounds, notably
-those present in fava beans (although, It is unclear if this alone is enough of a
-selection pressure to maintain the polymorphism in the face of these alleles
-considerable malaria resistance advantage). 
+leading to hemolytic anemia when exposed to a variety of compounds, notably those present in fava beans 
+(although it is unclear if this selective pressure alone suffices to maintain the polymorphism 
+in the face of the considerable malaria resistance advantage). 
 The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
@@
 Assuming that the underlying causes and strength of this drop in fitness have not changed, 
-we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
-\geq 0.05$),
-in order to have resulted in the equilibrium frequency seen today in
-areas with endemic malaria.
+we estimate that $s_d$ is upward of $0.05$ (if $sb \geq 0.05$)
+in order to have resulted in the equilibrium frequency seen today in areas with endemic malaria.
 
@@
  (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
-showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1--4}$.) 
+showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
@@
 \caption{ %
-{\bf Characteristic length, as a function of selective disadvantage}, 
+{\bf Characteristic length, in kilometers, as a function of selective disadvantage}, 
 compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
@@
 plotted as a single black dotted line. 
-\plr{$y$ axis label: kilometers}
 } \label{Fig-G6PD-charlength}
@@
 \caption{ %
-{\bf Characteristic length, as a function of selective advantage},
+{\bf Characteristic length, in kilometers, as a function of selective advantage},
 holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
@@
 and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
-\plr{$y$ axis label: kilometers}
 } \label{Fig-G6PD-charlength-sb}
@@
 geographic space covered by standing variants for parameters chosen to match our G6PD example.
-Even for relatively large deleterious costs prior to the environmental
-switch standing variants still makes up quite large proportion of the
-adaptive alleles, and an even larger proportion of the range (as they
-get to occupy a larger area than new mutations, as they start earlier).
+Even for relatively large deleterious costs prior to the environmental switch, 
+standing variants still makes up quite large proportion of the adaptive alleles, 
+and an even larger proportion of the range 
+(they get to occupy a larger area than new mutations, since they start earlier).
 
@@
 
-\gc{As discussed above, we could incorporate differences in the beneficial selection
-coefficients of the different classes of alleles, to allow for
-negative pleiotropic effects acting to surpress the advantage of an
-allele once the environment switches (effect one above). One simple way to do
-this is to simply replace $s_b$ with $s_{b,j}$ resulting in a class-specific new allele establishment rates ($\lambda_{j}$) and rates of spread ($v_j$).
-These then could be used in equation
-\eqref{eqn-prop-space-allele-j}. We could imagine our two class model, outlined
-above, with class one having the additional advantage of
-$s_{1,b}>s_{2,b}$. This would further increase the contribution of the rarer class,
-because our rare class would overcome drift more often and spread more
-rapidly. However, a straight-forward application of the logic of used
-to construct equation \eqref{eqn-prop-space-allele-j},
-ignores the fact that classes of alleles with higher $s_b$ is not
-completely selected excluded from spreading into geographic areas occupied by alleles with lower
-$s_b$. Given enough time the most advantageous type (type 1, in this case) would arise and
-spread everywhere, and so if we used different $s_b$s in equation
-\eqref{eqn-prop-space-allele-j}, it would only be a short-term approximation to
-a longer term dynamic. Even after the initial tessellation has formed the allele 1s would have an selective advantage $\delta s
-=s_{b,1}-s_{b,2}$ in areas where allele 2 have already spread, and so
-would arise at rate $2\rho \mu_1 \delta s $ in these areas and spread through them at
-speed $\sigma \sqrt{\delta s}$. An extension of our Poisson process model could incorporate these
- effects, by thinning the Poisson process of establishing mutations by
- correctly. 
-Whether allele $2$ persists would
-depend on the linkage arrangements between loci. If the loci of allele
-1 and 2 are unlinked, then allele 1 can spread without disrupting
-allele 2. However, if they are linked the spread of allele $1$ may
-push allele $2$ out of the population. More complicated dynamics,
-including spatial Dobzhansky-Muller incompatibilities \citep{,Kondrashov:03,ralphcoop2010}, could ensue if
-there are epistatic interactions between the alleles. }
+It is natural to also incorporate differences in the beneficial selection coefficients of the different classes of alleles, 
+to allow for negative pleiotropic effects acting to suppress the advantage of an allele once the environment switches.
+One simple way to do this is to simply replace $s_b$ with $s_{b,j}$ 
+resulting in a class-specific new allele establishment rates ($\lambda_{j}$) and rates of spread ($v_j$).
+These then could be used in equation \eqref{eqn-prop-space-allele-j}. 
+For instance, we could extend the two class model above 
+so that class one has the additional advantage of $s_{1,b}>s_{2,b}$. 
+This would further increase the contribution of the rarer class,
+because this class would both overcome drift more often and spread more rapidly. 
+However, a straightforward application of the logic of used to construct equation \eqref{eqn-prop-space-allele-j}
+fails once the different allelic types meet,
+since the assumption of selective exclusion no longer holds:
+alleles with higher $s_b$ will spread, at a lower speed, into regions occupied by alleles with lower $s_b$.
+Given enough time, the most advantageous type (type 1, in this case) would spread everywhere, 
+and so substituting multiple values for $s_b$ in equation \eqref{eqn-prop-space-allele-j}
+would only provide a short-term approximation to a longer term dynamic. 
+Even if the initial tessellation has formed with purely class 2 alleles, 
+the first allele would have an selective advantage $\delta s =s_{b,1}-s_{b,2}$,
+and so would arise at rate $2\rho \mu_1 \delta s $ and would spread at speed $\sigma \sqrt{\delta s}$. 
+An extension of our Poisson process model could incorporate these effects, 
+by thinning the Poisson process of establishing mutations by correctly,
+but is considerably less tractable.
+Whether allele $2$ persists would depend on the linkage arrangements between loci. 
+If the loci underlying allele 1 and 2 are unlinked, then allele 1 can spread without disrupting allele 2. 
+However, if they are linked, the spread of allele $1$ may push allele $2$ out of the population. 
+More complicated dynamics,
+including spatial Dobzhansky-Muller incompatibilities \citep{Kondrashov:03,ralphcoop2010} 
+could ensue if there are epistatic interactions between the alleles.
 
@@
 From this we know that $\tau \approx (1/s_b) (\log (\sigma^2 \rho) - \log W)$;
-although more detailed information is available (e.g.\ a CLT for $\tau$, \citet{nagaev1971limit}),
+although more detailed information is available (e.g.\ a central limit theorem for $\tau$, \citet{nagaev1971limit}),
 we will stick to the loose interpretation.
@@
 New mutations will appear and become established during this time if
-\gc{$4 \rho \sigma^2 \mu s_b \ge 1/T$,
-i.e.\ if $4 \rho \sigma^2 \mu \ge 1/\log (\sigma^2 \rho)$}.
+$4 \rho \sigma^2 \mu s_b \ge 1/T$,
+i.e.\ if $4 \rho \sigma^2 \mu \ge 1/\log (\sigma^2 \rho)$.
 Our model will still be a good approximation, however, 
@@
 
-\gc{The results of \citet{softsweepsI} and \citet{softsweepsII} suggest that in a
+The results of \citet{softsweepsI} and \citet{softsweepsII} suggest that in a
 panmictic population the number of independent alleles (and their frequencies), in a sample,
-is relatively independent of $s_b$ and $s_d$  \citep[although this
+is nearly independent of $s_b$ and $s_d$  \citep[although this
 breaks down with fluctuating population size,
@@
 increases the rate at which independent mutations become established,
-it also accelerates the frequency gain of established alleles decreasing the time period in which new alleles can arise and
-hope to be at significant frequency in the population. These two
-effects cancel each other out leading to no strong effect of $s_b$ on
-the number of independent alleles. 
-Decreasing $s_d$ increases the number of standing variants within a
-population, increasing the number of alleles that manage to establish
-and spread from standing variation
-\citep{softsweepsI,Orr:01}. However, having more established standing alleles
-acts to exclude the spread of new alleles that arise once the environment
-switches. These two opposing effects again cancel out leading to little overall
+it also accelerates the frequency gain of established alleles, 
+hence decreasing the time period in which new alleles can arise and hope to be at significant frequency in the population. 
+These two effects cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
+Decreasing $s_d$ increases the number of standing variants within a population, 
+increasing the number of alleles that manage to establish and spread from standing variation \citep{softsweepsI,Orr:01}. 
+However, having more established standing alleles
+acts to exclude the spread of new alleles that arise once the environment switches. 
+These two opposing effects again cancel out leading to little overall
 effect of $s_d$ on the number of independent alleles.
-In contrast our results show that the
-characteristic length (closely related to the density of independent
-alleles) depends on both $s_b$ and $s_d$ in a geographical spread population. 
-Like the panmictic model our alleles also act to exclude one and
-other, however, the geographic spread of an allele is slow (area grows
-as $s_b$) compared to the initial exponential growth of an allele in a
-panmictic population. That means that the role of selection in helping
-alleles become established can dominate leading to more independent origins, both by
+In contrast, our results show that the characteristic length 
+(closely related to the density of independent alleles) 
+depends on both $s_b$ and $s_d$ in a geographical spread population. 
+Like the panmictic model, in our model alleles also act to exclude each other; 
+however, the geographic spread of an allele is slow compared to the initial exponential growth of an allele in a panmictic population. 
+That means that the role of selection in helping
+alleles become established can dominate, leading to more independent origins, both by
 being  weaker before, and stronger after, the environmental switch.
-
-}
 
@@
 convergence among populations.
-While at face value this increase seems unsurprising, as explained above, this
-differs markedly from the case of multiple competing alleles in a
-panmictic population \citep{softsweepsI,softsweepsII}. 
+While at face value this increase seems unsurprising,
+this relationship differs markedly from the case of multiple competing alleles in a panmictic population \citep{softsweepsI,softsweepsII}. 
 Importantly, allowing standing variation may greatly lower the time until the species becomes adapted 
@@
 has led to the suggestion that these individuals had dark
-pigmentation \citep{olalde2014derived,lazaridis2013ancient, Wilde2014}. However, given our results and the partially convergent
+pigmentation \citep{olalde2014derived,lazaridis2013ancient, Wilde2014}. 
+However, given our results and the partially convergent
 basis of skin pigmentation between Europeans and East Asians
 \citep{norton-human-pigmentation,edwards2010pigmentation} it seems just as plausible that these ancient individuals
-simply had adapted to high latitudes via a different complement of ``light-skin'' pigmentation
-alleles. Such convergence may considerably complicate the 
+adapted to high latitudes via a different complement of ``light-skin'' pigmentation alleles;
+we have no evidence either way.
+Such convergence may considerably complicate the 
 exploration of phenotypes and adaptation among populations using variants mapped in a
@@
 the alleles concerned have no bearing on the ecological differences
-across these breaks \citep[see][ for a wide-ranging discussion of how
-allelic differentiation may build along particular zones]{bierne2011coupling}. 
-\gc{We are rarely as fortunate, to know as much about the 
- genetics, phenotypic distributions, 
-and potential selection agents as we do for malaria resistance. And so we must be wary of mistaking the
-strange spatial distributions of particular alleles for 
-adaptation to some very specific selection pressure (e.g. the distribution of the Mahidol allele in Figure
-\ref{fig-G6PD-map}), when they are simply elements of a larger geographic mosaic of
-alleles responding to a broadly shared selection pressure. }
+across these breaks \citep[see][for a wide-ranging discussion of how allelic differentiation may build along particular zones]{bierne2011coupling}. 
+We are rarely so fortunate as to know as much about the genetics, phenotypic distributions, 
+and potential selection agents as we do for malaria resistance in humans. 
+Therefore, we must be wary of mistaking the strange spatial distributions of particular alleles 
+for adaptation to some very specific selection pressure (e.g.\ the distribution of the Mahidol allele in Figure \ref{fig-G6PD-map}), 
+when they are simply elements of a larger geographic mosaic of alleles responding to a broadly shared selection pressure.
 
@@
 If many loci contribute to variation in a trait, then
-selection on any one allele may be weak, which might adaptation to use
-the same alleles in different populations.
-However, populations separated by a reasonable amount of drift may still
+selection on any one allele may be weak, 
+which might lead adaptation to use the same alleles in different populations.
+However, sufficiently differentiated populations may still
 adapt via different genetic routes, as the constellation of alleles
@@
 % i.e.\ those at intermediate frequency, will be somewhat different. 
-So it may be the case that polygenic traits are even less susceptible to a shared
+Therefore, it may be the case that polygenic traits are even less susceptible to a shared
 genetic basis to adaptation across populations than simple traits.   
@@
 Mayr and many others have argued that species are coherent
-evolutionary units because they are united by shared gene flow \citep[pages 521--522][]{Mayr:SpeciesEvol}. 
+evolutionary units because they are united by shared gene flow \citep[pages 521--522 in][]{Mayr:SpeciesEvol}. 
 However, this argument has been questioned by a number of authors 
 based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
-Even if gene flow is not high enough to prevent neutral
-differentiation or local adaptation, a number of authors have argued
-that that species are cohesive as gene flow were still high enough for
-globally selected alleles, and hitchhiking haplotypes, to
-spread across entire species \citep[see also][]{Rieseberg2001,MorjanRieseberg:04,Ellstrand2014}.
+Even if gene flow is not high enough to prevent neutral differentiation or local adaptation, 
+a number of authors have argued that species are cohesive 
+if gene flow is high enough for globally selected alleles 
+(and their hitchhiking haplotypes)
+to spread across entire species 
+\citep[see also][]{Rieseberg2001,MorjanRieseberg:04,Ellstrand2014}.
 At present, large scale genotyping and sequencing projects, 
@@
 \citep[see also][]{ralphcoop2010,RalphCoop:14} 
-bring the idea that this gene flow will lead to the rapid species-wide spread of
-selected alleles partially into question. 
-Our results suggest that species should often adapt to
+suggests that species should often adapt to
 wide-spread selection pressures through convergent evolution rather
@@
 
-In support of this idea, putative, recent selective sweeps seem to often be 
-geographically restricted \citep{Pickrell:09,Coop:09,Granka:14},
-rather than range-wide \citep[but see][ for a potential example]{Clark:07,Long:13}. 
+In support of this idea, putative recent selective sweeps seem to often be geographically restricted \citep{Pickrell:09,Coop:09,Granka:14},
+rather than species-wide \citep[but see][ for a potential example]{Clark:07,Long:13}. 
 This is likely in part due to the relatively low incidence of wide-spread selection pressures,
 but as noted above even when we know of wide-spread selection
-pressures (e.g.\ malaria) the response is usually convergent, not
-shared, across large spatial scales. 
+pressures (e.g.\ malaria) the response is usually convergent, not shared, across large spatial scales. 
 On the other hand, 
@@
 Currently we can only scan genomes for species-wide sweeps
-in those few organisms with population-scale sequence data, and so we
-are very limited in our view .
+in those few organisms with population-scale sequence data, 
+and so we do not know if these observations generalize to most species.
 This is rapidly changing,
@@
 induced by extirpation of the species over parts of its range
-\citep[see ][and reference therein for how such a model could be constructed]{barton2013modelling}. 
+\citep[see ][and references therein for how such a model could be constructed]{barton2013modelling}. 
 Such drops in range size may fix, or radically change the species-wide frequency, 
@@
 \citep[see ][for a recent exchange on this]{Pennell:14,Venditti:14,Pennell:14B}. 
-We first note that neutral substitutions are unaffected by this process, 
+Neutral substitutions are unaffected by this process, 
 because they accumulate in a clocklike manner along lineages, 
@@
 Turning to the accumulation of adaptive substitutions, 
-it likely that splits in phylogenies are likely only a tiny proportion of all incipient
-speciation events because extinction rates may be high
-\citep{Rosenblum:12}, such that every lineage has likely passed through
-many ``speciation events'' in addition to the observed ones.
-Furthermore, under this view, 
-spatial polymorphisms may accumulate gradually in geographically restricted populations
-between large-scale biogeographic events leading to their fixation or loss. 
-Such that the origination time of alleles and when they fix could be 
-reasonably uncoupled by the geographic structure \citep[in a manner,
-similar to that argued by ][]{Gillespie:94}.
+it is likely that splits in phylogenies are only a tiny proportion of all incipient speciation events, 
+because extinction rates may be high \citep{Rosenblum:12}, 
+and so every lineage has likely passed through many ``speciation events'' in addition to the observed ones.
+Under these assumptions,
+spatial polymorphisms could accumulate gradually in geographically restricted populations
+between the large-scale biogeographic events that cause their fixation or loss. 
+This effectively decorrelates the time at which new alleles arise and when they fix in the species,
+an effect similar to that pointed out by \citet{Gillespie:94}.
 
@@
 whereby genetic drift allows
-populations to cross fitness valleys and substitute novel epistatic
-combinations. While geographic lineage sorting via speciation and
-extinction can be thought of as very large-scale genetic drift events,
-the initial spread of alleles is due to selection they are only later sorted 
-into substitutions via these large scale drift events \citep[see also ][for discussion]{Futuyma:89}.
+populations to cross fitness valleys and substitute novel epistatic combinations. 
+However, although geographic lineage sorting via speciation and extinction 
+can be thought of as very large-scale genetic drift events,
+in the models we study here the initial spread of alleles is due to selection, not drift
+\citep[see also ][for discussion]{Futuyma:89}.
 
@@
 
-There is genome-wide evidence that a reasonable fraction of 
-substitutions are fixed by positive selection in a number of species
-\citep[most notably Drosophila, ][]{Sella:09}. Under this geographic view of
-fixation, selection has played a strong
-role in the establishment of these alleles locally. As we get more broadly geographic population
-genomics sampling for a range of species there is an opportunity to
-study whether the class of alleles that contribute 
-to local differentiation are the similar as those underlying species
-divergence, and the extent to which the answer to this depends on the
+There is evidence that a reasonable fraction of genome-wide substitutions are fixed by positive selection in a number of species
+\citep[most notably Drosophila, ][]{Sella:09}. 
+Under the geographic view of fixation, 
+selection has played a strong role in the establishment of these alleles locally. 
+As we get more broadly geographic population genomics sampling for a range of species 
+we will have the opportunity to study whether the class of alleles that contribute to local differentiation 
+are similar to those underlying species divergence, 
+and the extent to which the answer to this depends on the
 age and type of population structure within species.
 
-Finally, we close by noting that range expansion, and speciation, are obviously not separate from
-adaptive differentiation. The invasion of novel geographic spaces may
-lead to burst of adaptive differentiation, at least at a
-subset of genes, and speciation may be associated with rapidly
-adaption to novel environments. Conversely, the relatively slow geographic spread of adaptive
-alleles within ranges, e.g. if they only offer a local advantage or
-if they are selective excluded, may allow Dobzhansky-Muller incompatibilities to arise
-within species effectively offering a mechanism for 
-hybrid incompatibilities to evolve in parapatry effectively fracturing
-the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. Indeed the
-alleles that act as components in many of the Dobzhansky-Muller incompatibilities studied to
-date are geographically restricted \citep[see][]{Cutter:12}. As such
-it seems possible that populations within species may often be on the
-verge of speciation, and that as outlined here that this may drive
-some proportion of molecular divergence.
+Finally, we close by noting that range expansion and speciation 
+are obviously not separate from adaptive differentiation. 
+The invasion of new geographic areas may lead to a burst of adaptive differentiation, 
+at least in a subset of genes, and speciation may be associated with rapidly adaption to novel environments. 
+Conversely, if the geographic spread of adaptive alleles within ranges is slow, 
+e.g.\ if they only offer a local advantage or if they are selective excluded, 
+may allow Dobzhansky-Muller incompatibilities to arise within species, 
+effectively offering a mechanism for hybrid incompatibilities to evolve in parapatry,
+and fracturing the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. 
+The alleles that act as components in many of the Dobzhansky-Muller incompatibilities studied to date 
+are geographically restricted \citep[see][]{Cutter:12}. 
+Therefore, it seems possible that populations within species may be tending towards speciation, 
+and that as outlined here that this may drive some proportion of molecular divergence.
 

### 2014-09-29 g6pd_figures.R
small coop changes

@@ -18,7 +18,7 @@ layout(t(1:2))
 par(mar=c(4,4.1,2,0.5))
 for(j in 1:2){
 	rho=rhos[j]
-	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(chi),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
+	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(chi, km),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
 	if(j == 2) plot(sbvals,sbvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.4,main=expression(rho==2))
 	for(i in 1:3){
 	sigma=sigmas[i]; charlength.sb<-sapply(sbvals, function (sb) { charLength(mu,rho,sb,sd=sd,sigma)$value } )
@@ -196,7 +196,7 @@ dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_standing_v
  }
 # legend(x="topright",legend=rep("blah",4),lty=1:4)
 
-other.params<-as.expression( apply(cbind(c(2,2,0.2,0.2),c(10,100,10,100) ),1, function(x) {substitute( rho== myrho~ sigma==mysigma, list(myrho=x[1],mysigma=x[2]))})) 
+other.params<-as.expression( apply(cbind(c(2,2,0.2,0.2),c(10,100,10,100) ),1, function(x) {substitute(list( rho== myrho, sigma==mysigma), list(myrho=x[1],mysigma=x[2]))})) 
 mut.rates<-as.expression( lapply(c(-7,-6,-5) , function(x) {substitute( mu[2]== 10^mymu, list(mymu=x))}))
 my.exp<- as.expression(c(other.params,"Standing Var. Only",mut.rates))
  

### 2014-09-29 standing-variation-paper.tex

@@
 Assuming that the underlying causes and strength of this drop in fitness have not changed, 
-we estimate that $s_d$ is upward of $0.05$ (if $sb \geq 0.05$)
-in order to have resulted in the equilibrium frequency seen today in areas with endemic malaria.
+we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
+\geq 0.05$),
+in order to have resulted in the equilibrium frequency seen today in
+areas with endemic malaria \citep[see also][]{Ruwende-g6pd}.
 

### 2014-09-29 standing-variation-paper.tex
hopefully, close to final changes

@@
 
+Note that the conflicting roles of $\rho$ and $\sigma$ mean thar even in
+species where levels of neutral differentiation are low, geographic
+convergent adaptation may be common. This is because low levels of
+neutral genetic differentiation can be due to high population
+densities rather than high dispersal distances, and high population
+densities would allow convergent adaptation. As such geographic convergent
+evolution may be common even in species with little neutral population structure.
+
 \begin{figure}[ht]
@@
 Importantly, allowing standing variation may greatly lower the time until the species becomes adapted 
-across the geographic range of the selection pressure. 
-Adaptation through standing variation also biases the type of variation towards those alleles 
+across the geographic range of the selection pressure. Adaptation through standing variation also biases the type of variation towards those alleles 
 with fewer pleiotropic effects, since these are more common as standing variation before the environmental shift. 
@@
 adapted to high latitudes via a different complement of ``light-skin'' pigmentation alleles;
-we have no evidence either way.
+to our knowledge, we have no strong evidence either way.
 Such convergence may considerably complicate the 
@@
 than waiting for a single allele to migrate across the range.
+
 % over length-scales determined by our characteristic length.
@@
 Such ideas are not completely new and represent a perhaps logical consequence of
-an allopatric, or at least parapatric, view of the geography of speciation.
+an allopatric or parapatric view of the biogeography of speciation.
 However, it is worth revisiting this idea as geographically broad population genomic
@@
 adaptation is lost to local extinction and the collapse of population structure
-\citep[see also][]{lieberman:96,Eldredge:05}. 
+\citep[see also][]{lieberman:96,Eldredge:05, Futuyma:10}. 
 Furthermore, he suggests that speciation
@@
 have argued that the rate of speciation may well be quite high,
-but that the majority of incipient species do not persist long due to reabsorption or extinction. 
+but that the majority of incipient species do not persist long due to
+reabsorption or extinction. 
+Changes in range size, due to local extinction, can also be very rapid 
+on the time-scales over which alleles may spread on the landscape \citep{gaston2003structure,HEWITT1996}.
 Repeated bouts of extinction and speciation will send waves of alleles
@@
 are geographically restricted \citep[see][]{Cutter:12}. 
-Therefore, it seems possible that populations within species may be tending towards speciation, 
+Therefore, it seems possible that populations within species may often be tending towards speciation, 
 and that as outlined here that this may drive some proportion of molecular divergence.
@@
 \subsubsection*{Acknowledgements}
-We thank Michael Turelli, Jon Seger and the Coop lab for helpful conversations.
+We thank Michael Turelli, Jon Seger, and the Coop lab for helpful conversations.
 

### 2014-09-29 g6pd_figures.R

@@ -15,10 +15,10 @@ my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
 sigmas=c(100,50,10)
 
 layout(t(1:2))
-par(mar=c(4,4.1,2,0.5))
+par(mar=c(4,4.3,2,0.5))
 for(j in 1:2){
 	rho=rhos[j]
-	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(chi, km),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
+	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(list(chi, km)),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
 	if(j == 2) plot(sbvals,sbvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.4,main=expression(rho==2))
 	for(i in 1:3){
 	sigma=sigmas[i]; charlength.sb<-sapply(sbvals, function (sb) { charLength(mu,rho,sb,sd=sd,sigma)$value } )
@@ -44,11 +44,11 @@ dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlength
 sigmas=c(100,50,10)
 
 layout(t(1:2))
-par(mar=c(4,4.1,2,0.5))
+par(mar=c(4,4.3,2,0.5))
 
 for(j in 1:2){
 	rho=rhos[j]
-	if(j == 1) plot(sdvals,sdvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab=expression(chi),cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
+	if(j == 1) plot(sdvals,sdvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab=expression(list(chi,km)),cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
[…]

### 2015-02-15 standing-variation-paper.tex
made changes in response to reviews and comments by Molly

@@
 \newcommand{\gc}[1]{{\it\color{green}(#1)} }
+\newcommand{\gc2}[1]{{\it\color{purple}(#1)} }
+
 \newcommand{\plr}[1]{{\it\color{blue}(#1)}}
-
+\usepackage{lineno}  %%%% FOR AMNAT
 \newcommand{\var}{\mathop{\mbox{Var}}}
@@
 \newcommand{\R}{\mathbb{R}}
+
+%\usepackage{setspace}
+%\doublespacing
 
@@
 % selective sweeps and species coherence, gene flow
+
+\linenumbers
+\modulolinenumbers[2]
 
@@
 can bias the subset of alleles that get to contribute to a species adaptive response. 
-We apply the results to the many geographically localized G6PD deficiency alleles thought to confer resistance to malaria,
-whose large mutational target size and deleterious effects make them likely candidates to have been present as deleterious standing variation.
+We apply the results to the many geographically localized G6PD
+deficiency alleles thought to confer resistance to malaria, \gc2{where the large
+mutational target size of G6PD makes likely candidate for adaptation
+from standing variation, despite the selective cost of G6PD
+deficiency.}
 We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
-which suggest that these arose both from standing variation and new mutations since the advent of malaria.
+\gc2{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
 Our results suggest that much of adaptation may be geographically local 
-even when selection pressures are wide-spread. 
+even when selection pressures are widespread. \gc2{Therefore, we argue
+that caution must be exercised when arguing that strongly
+geographically restricted alleles are necessarily the outcome of local adaptation.}
 We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.
@@
 adapted to similar environments by means of independent genetic changes. 
-In some cases this convergent evolution has produced different phenotypes
-despite shared selection pressures; 
+In some cases this convergent evolution \gc2{is the result of quite
+distinct genetic changes, involving very different genes and pathways} despite shared selection pressures; 
 in other cases independent adaptations are identical down the same
-nucleotide change  \citep{Jeong20141,stern2013genetic,MartinOrgogozo:13}. 
+nucleotide change
+\citep{Jeong20141,stern2013genetic,MartinOrgogozo:13, Conte2012}. 
 Such convergent evolution within populations has been seen for many
@@
 whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
-and to what degree should species be regarded to cohesive units. 
+and to \gc2{what degree species should be regarded} as cohesive units. 
 Convergent evolution also affects our ability to detect adaptation from population genomic data,
@@
 For example, 
-patchy geographically selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
-since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
-In \citet{ralphcoop2010} we formulated a simple model of parallel adaptation in a geographic setting
-where the selection pressure was constant across the entire environment, 
-and a single mutational change was sufficient to adapt after the change in environment.
-We assumed that there was no standing variation for the adaptive allele, 
-so that parallel mutation must be due to multiple new mutations occurring after the environmental shift. 
-In that setting, 
-we found the characteristic geographic scale over which 
+\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
+since alleles are unable to spread through intervening populations
+\citep{RalphCoop:14}. 
+In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
+ a spatially spread population, where dispersal is local,
+that is exposed to some novel, spatially homogeneous selection pressure.
+We assumed that a single mutational change was sufficient to adapt
+the population after the change in environment. Under this assumption,
+selected alleles arise and spread locally. However, if the geographic
+area is large enough multiple selected alleles can
+arise independently and spread before any one has spread across all of
+space ( Figure \ref{fig:sim-spread}).  one allele's rapid spatial spread can be
+halted by running into convergent allele. See Figure
+\ref{fig:sim-spread}. A somewhat analogous situation also arises in spatial models of clonal
+inference in asexuals \citep{martens_interfering_2011}.}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
+  \end{center}
+  \caption{
+{\bf Simulation of the geographic spread of new mutations}.
+Simulation on a grid of $101 \times 101$ populations with 100 individuals in each, showing spread of a beneficial allele with advantage $s=0.1$.
+Times are in numbers of generations, with probability of mutation per generation 0.3.
+}
+  \label{fig:sim-spread}
+\end{figure}
+
+\gc{Under this model we previously derived} the characteristic geographic scale over which 
 multiple instances of the adaptive allele are expected to arise in parallel. 
 This characteristic length could be expressed in terms of a simple compound
-parameter determined by our parameters of interest. 
+parameter \gc2{comprised of} our parameters of interest. \gc{In
+\citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
+allele (e.g. because the allele was very strongly deleterious before
+the environmental change)}, 
+so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.
+
 In this paper, we extend this spatial model to include standing variation 
 present at mutation-selection balance before the selection pressures switch.
-
 Below,
-we show that convergent adaptation within a wide-spread species is likely to be
-common, as has already been seen for a number of traits. 
-On the basis of this we argue 
-that the genetics of adaptation may often be geographically local, 
+we show that convergent adaptation within a widespread species is likely to be
+common \gc{when ranges, population sizes, or mutational targets are large}, as has already been seen for a number of traits. 
+On this basis owe argue 
+that the genetics of adaptation may often be geographically local \gc{even
+when selection pressures are geographically broad}, 
 % and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
-and that most widespread selective sweeps should occur
+and that widespread selective sweeps \gc2{should tend to} occur
 when adaptation is highly constrained (e.g.\ by small mutation rate or
 the need for a linked combination of alleles).
-We discuss some implications, and history, of this view for the evolutionary
+We discuss the \gc{history, and implications,} of this view for the evolutionary
 coherence of species and molecular evolution.
@@
 is a large, homogeneous, one-- or two--dimensional region. 
-There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
+There are two selective classes -- the \gc{ {\em ancestral} } type,
+and the {\em mutated} type \gc{(which will offer a fitness benefit after the
+environmental switch)}.
 We assume that separately arising mutations are distinguishable --
-either as selectively equivalent mutations, or by linked variation.
+either as selectively equivalent mutations, or by linked \gc{neutral} variation.
+\gc{We assume that all alleles of the mutated type are selectively
+  equivalent (both in terms how deleterious they were before, and how
+  advantagous they are after, the
+  environmental shift). Note that these mutations do not have to arise at the same locus,
+  just that all of the mutated type of selective equivalent. As such,
+  these could be mutations arising at the same base pair or knockout
+  mutations of any one of a number of genes in a pathway as long as carrying at least one of these alleles
+  is sufficient to adapt an individual to the new environment.}
+
 We also suppose that 
@@
 important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
-homozygote for the mutant alleles.
-The number of offspring has finite variance, 
-which would divide the probability of establishment of advantageous alleles below,
-but to keep the formulas simpler we assume to be equal to one.
-(We leave aside the case where mutations are recessive, however.)
+homozygote for the mutant alleles. \gc2{Note that this means that we
+  are ignoring the case of totally recessive alleles.}
+\gc{We assume that the variance in offspring number is one, see
+  \cite{ralphcoop2010} for how this assumption can be relaxed.}
+%The number of offspring has finite variance, 
+%which would divide the probability of establishment of advantageous alleles below,
+%but to keep the formulas simpler we assume to be equal to one.
 As for the other parameters,
 suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
-and that the mean squared geographic distance between parent and child is $\sigma^2$.
+and that the mean squared geographic distance between parent and child
+is $\sigma^2$. \gc2{Note that we assume that the dispersal kernel is
+  not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
+  discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
+  on rapid spreading due to very long distance dispersal.}
 The species occupies an area with mean density
-$\rho$ of haploid individuals (or chromosomes) per unit area.
+$\rho$ of haploid individuals \gc{(twice the number of diploids)} per unit area.
 
@@
 is well--approximated by a Poisson process in space and time.
+\gc{By this we mean that there is a constant rate across space and
+  time at which new
+  mutations arise, and the occurance of new mutations in any region of
+  space-time is independent of that in non-overlapping intervals.}
 The rate of this Poisson process,
@@
 appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
-is approximately $\lambda = 2 \mu \rho s_b$.
+is approximately 
+\begin{equation}
+\lambda = 2 s_b \mu \rho 
+\end{equation}
+\gc{In geographic areas already
+ occupied by the mutant type (i.e. in regions it has spread to already), new mutations are selectively excluded
+ (i.e. are effectively neutral), and so unlikely to establish over the
+ short-time scales considered here.  }
 
@@
 The genetic descendants of each new mutation are (with high probability) doomed to extinction,
-but may persist for some time (note that we have assumed that $s_d$ is not too small).
-Since the times and locations of mutants before $t=0$ also
+but may persist for \gc2{a short} time (note that we have assumed that $s_d$ is not too small).
+\gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
 well--approximated by a 
-Poisson process with rate $\mu \rho$,
+Poisson process with rate $\mu \rho$ \gc{(if $\mu \rho \ll 1$) as
+mutations are rare events. Therefore,} 
 the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
-by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
+by the Poisson Mapping Theorem, a Poisson process with rate we define
+to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
+\gc2{Our $\lambda_0$ is the mean density
+of standing variants that are present and escape loss when the
+environment shifts.}
 % (In principle, more than one member of a mutant family may fix,
@@
 we can then treat each such family as equivalent to a single new mutation,
-but with somewhat larger probability of local establishment.
+but with somewhat larger probability of local establishment,
+\gc{following the environmental shift}.
 (This approximation will be good 
@@
 and the spatial distribution of each is small relative to the spread between the families.)
-To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
+To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let \gc{$Z_T$ be the number of its descendants at time $0$}.
 At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
@@
     \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
-        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
+        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
+        & =  \frac{\lambda }{ (-log(1-s_d)) }.\\
 \end{align}
 Note that this is closely related to the equilibrium frequency of an
-allele at mutation selection balance $\mu/s_d$ \citep{Haldane:27,Haldane:37}.
+allele at mutation selection balance $\mu/s_d$
+\citep{Haldane:27,Haldane:37, Orr:01}.
 
@@
 
-As we outlined in \citet{ralphcoop2010} this model of
+As we outlined in \citet{ralphcoop2010}, this model of
 geographic convergent evolution, when $\lambda_0=0$, is analogous to a model of crystallization due to \citet{kolmogorov-crystallization}.
@@
 the cells formed by a Voronoi tessellation have straight sides,
-but the introduction of new mutations cause these to curve.
+but the introduction of new mutations cause these to curve \citep[see
+Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.
 
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{spreading_alleles_trimmed}
+    \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
   \end{center}
@@
 following which adaptive alleles that successfully escape loss by drift
-spread locally from both standing variants (lightning bolts) and new mutations (stars). 
+spread locally from both standing variants (lightning bolts) and new
+mutations (stars). \gc{(Note that the standing variants arose at times
+prior to $t=0$.)}
 The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
@@
 These successful alleles spread spatially outwards at speed $v$, 
-so the space-time profile of their spread forms a cone.
+so the space-time profile of their spread forms a cone. Any mutation
+that arises in an area already within a cone to will not have a selective advantage.
 When these adaptive alleles meet, 
@@
 (over longer time-scales, they will mix into each other). 
-The grey dashed line shows a cone radiating backward in space-time,
-originating from a point $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
+\gc{One of the ways we derive various results in the text is by taking a
+space-time point $(x,t)$ and asking the probability that no successful
+alleles have spread there yet.} The grey dashed line shows a cone radiating backward in space-time,
+originating $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
 Point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
@@
 Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
-and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
+and spread through human populations in areas where malaria \gc2{has been} endemic \citep{Kwiatkowski:05}. 
 A number of these alleles appear to be examples of convergent adaptation, 
@@
 \citep{Flint:98,ralphcoop2010}.
-Another particularly impressive case of convergent evolution in response to
-selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
+Another particularly impressive case of convergent evolution \gc2{is} the numerous changes throughout the X-linked G6PD gene, 
 with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
@@
 and the Mahidol allele found Myanmar and Thailand.
-The $A-$ and Mahidol alleles have been shown
-to be protective against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
-hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
+\gc2{ 
+The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
+across study populations and the form of malaria considered
+\citep[see ][ for recent discussion]{manjurano_2015, malaria_network_2014}.  
+The $A-$ and Mahidol alleles are thought to offer some protective effects against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
+heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
+  Louicharoen-g6pd, manjurano_2015}. }
 Haplotype-based analysis of genetic diversity surrounding
@@
 
-Given such a strong pressure on these alleles they should have risen
+%although it is unclear if this selective pressure alone suffices to maintain the polymorphism 
+%in the face of the considerable malaria resistance advantage)
+
+%leading to hemolytic anemia when exposed to a variety of compounds, notably those present in fava beans
+
+Given \gc2{such strong selection the alleles} should have risen
 quickly to fixation, so their presence at intermediate frequency,
@@
 Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
-leading to hemolytic anemia when exposed to a variety of compounds, notably those present in fava beans 
-(although it is unclear if this selective pressure alone suffices to maintain the polymorphism 
-in the face of the considerable malaria resistance advantage). 
+and homozygote females may also not be protected against malaria \citep{manjurano_2015, malaria_network_2014}. 
 The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
@@
 in order to have resulted in the equilibrium frequency seen today in
-areas with endemic malaria \citep[see also][]{Ruwende-g6pd}.
+areas with endemic malaria \citep[based on heterozygote advantage
+calculations for the X chromosome, results not shown, see also][]{Ruwende-g6pd}.
 
@@
 %(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)
-
+\gc{CHECK FACTOR of $1/3$ in cone volume.}
 Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
 If it does not encounter other locally fixed, beneficial mutants,
-it will cover a distance $L$ in time $L/v$.
-The number of other mutations appearing in the circle it has covered up until this time is Poisson with mean
+it will cover a distance $L$ in time $L/v$. \gc{In doing so (in two-dimensions) it will have
+started from a point in space and spread out to cover an geographic
+circular area of size $\pi
+L^2$ and carved out a cone in space-time with this as a base and
+height $L/v$ (with a volume of $\pi L^2 \times (L/v)/3$)
+Therefore, the number of other successfully established mutations that
+would have appeared} in the circle it has covered up until this time is Poisson with mean
 $\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
@@
 Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
-\begin{equation}
+\begin{equation} \label{eqn:chi_def_1}
     \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
@@
 before it is expected that one other successful mutation would have arisen in the area covered so far.
-The explicit formula for $\chi$ in two dimensions is cumbersome; here we omit it.
-(In one dimension the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.)
+The explicit formula for $\chi$ in two dimensions \gc{can be found by
+rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome;} here we omit
+it. \gc{In one dimension things are a little prettier} and the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.
 Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
-we can rewrite this as
+\gc{in two-dimensions} we can rewrite eqn. \eqref{eqn:chi_def_1} as
 \begin{equation} \label{eqn:defines_chi}
@@
 From this we see that $\chi$ decreases with $\rho$ and $\mu$.
-Furthermore, for large $\sigma$, the characteristic length is close to the value obtained just from standing variation:
+Furthermore, for large $\sigma$, the characteristic length \gc{approaches}
+the value we would just from standing variation:
 \begin{equation} \label{eqn:chi_standing}
@@
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
-then the characteristic length is close to the value from \citet{ralphcoop2010}:
+\gc{then standing variation is} unimportant the characteristic length is \gc{approaches} the value from \citet{ralphcoop2010}:
 \begin{equation} \label{eqn:chi_new}
@@
 
-In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb} we show the range of characteristic lengths as a function 
+In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb}, we show the range of characteristic lengths as a function 
 of various parameters chosen to match the evolution of malaria
 resistance at G6PD. 
-These curves match our intuition that population densities result in
+These curves match our intuition that \gc{higher} population densities result in
 smaller characteristic lengths (as would higher mutation rates). 
-Allowing standing variation increases the mutational
-input and so decreases the characteristic length below that predicted
+Allowing standing variation increases \gc{the 
+input of new alleles} and so decreases the characteristic length below that predicted
 by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
@@
 Higher dispersal distances lead to larger characteristic lengths, 
-since more rapid geographic spread blocks others from arising. 
+since more rapid geographic spread block other mutations from establishing. 
 A larger selective advantage ($s_b$) acts in two conflicting
 ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
-escape drift and become established. 
-The effect of helping establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
+\gc2{to} escape drift and become established. 
+The effect of helping\gc{locally} establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
  (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
@@
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter. 
-The dependence of the characteristic length on $s_b$ when only new mutations contribute is much weaker ($s_b^{-1/6}$, equation \eqref{eqn:chi_new})
+as in that case the speed of spread does not matter \gc2{only the
+  initial density of established alleles after the environmental shift}. 
+The dependence of the characteristic length on $s_b$ when only new
+mutations contribute is much weaker (\gc{of order} $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
 Overall, the range of characteristic lengths observed are reasonably
@@
 800km, especially for the lower population density, as long as the
-fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was not too low.
-
-Note that the conflicting roles of $\rho$ and $\sigma$ mean thar even in
+fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \gc2{high}.
+
+Note that the conflicting roles of $\rho$ and $\sigma$ mean that even in
 species where levels of neutral differentiation are low, geographic
@@
 It is also straightforward to compute the mean time until adaptation.
-Imagine ourselves at some geographic location, and let $\tau \ge 0$ be the time at which we are first reached by some advantageous mutation.
+\gc{Imagine a geographic location, and let $\tau \ge 0$ be the time at
+which this location} is first reached by some advantageous mutation.
 Then, as can be seen from the
@@
 \caption{ %
-{\bf Mean adaptation times}, as a function of selective disadvantage
-and , compared to the corresponding quantity without standing variation, and to the quantity only considering standing variants.
+{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage 
+as well as the population density.} We compare these to the corresponding quantities without standing variation and only considering standing variants.
 } \label{G6PD_chartimes}
@@
 \end{equation}
-Therefore, the mean proportion of patches that come from standing variation
+Therefore, the mean proportion of \gc{adapted} patches that come from standing variation
 is 
@@
 There are $\lambda_0 + \nu_+$ patches per unit area, so
-the typical patch (i.e.\ with distribution given by the Palm measure) occupies area $1/(\lambda_0 + \nu_+)$.
+the typical patch (i.e.\ with distribution given by the Palm measure)
+occupies area $1/(\lambda_0 + \nu_+)$. \gc{NEED REF FOR PALM}
 
@@
 mutation with probability given by equation \eqref{eqn-no-mut-t}.
-The probability, given that it has not been reached by $t$, that it will be reached by time $t+dt$ 
+\gc2{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
 by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
@@
 expanded circle at $t=0$, which has gone from a radius $vt$ to
-$v(t+dt)$. The corresponding probability the point is reached by a new
+$v(t+dt)$. The corresponding probability that the point is reached by a new
 variant is 
@@
 sliver of  extra volume in our space-time cone at time $t+dt$ compared to that at time $t$.
-Therefore, as is standard for competing exponentials,
-the probability a given location is reached first by a standing variant,
-and therefore the mean proportion of space covered  by standing variants,
+Therefore, the mean proportion of space covered by standing variants,
 is
 \begin{equation} \label{prop_space_standing}
-    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt .
+    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( -
+      \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt,
 \end{equation}
+\gc2{this is the probability a given location is reached first by a
+standing variant (which follows from competing our two exponential
+waiting times).}
 To evaluate this integral, again see Appendix \ref{apx:integrals}.
@@
 Even for relatively large deleterious costs prior to the environmental switch, 
-standing variants still makes up quite large proportion of the adaptive alleles, 
+standing variants still makes up quite a large proportion of the adaptive alleles, 
 and an even larger proportion of the range 
-(they get to occupy a larger area than new mutations, since they start earlier).
+(they \gc2{occupy a larger area than new mutations, since they get a
+headstart}).
 
@@
 \subsection{Multiple variant types}
-Another problem that we can seek to address with this work is
+Another problem that we can address with this work is
 the extent to which pleiotropy biases 
@@
 This first effect has been well studied by a number of authors
-\citep{Orr:00,Otto:04,WelchWaxman:03,Chevin:10} and its role in precise genetic
+\citep{Orr:00,Otto:04,WelchWaxman:03,Chevin:10} and its role in genetic
 convergence examined \citep{Orr:05,Chevin:10,Unckless:09}. 
@@
 Let's imagine for the moment that all classes of beneficial allele 
-have the same beneficial selection coefficient. 
-Or, at least that beneficial selection coefficients are similar enough that our selective exclusion
+have the same beneficial selection coefficient, \gc{or} at least that beneficial selection coefficients are similar enough that our selective exclusion
 approximation holds over the time scale on which we examine the process.
@@
 To illustrate some of the properties of this model, 
-let's imagine the somewhat extreme scenario that there is a single base pair 
+let's imagine the somewhat extreme scenario in which there is a single base pair 
 at which a possible mutation is relatively free of negative pleiotropy (call this class $1$);
@@
 the rarer, class $1$ mutations in Figure \ref{fig:pleiotropy_calc}.
-Intuitively, the contribution of the rarer mutation decreases as the
+\gc2{As expected intuitively,} the contribution of the rarer mutation decreases as the
 mutational target of the second class becomes larger, and as the
@@
 Even if the initial tessellation has formed with purely class 2 alleles, 
-the first allele would have an selective advantage $\delta s =s_{b,1}-s_{b,2}$,
+the first allele would have a selective advantage $\delta s =s_{b,1}-s_{b,2}$,
 and so would arise at rate $2\rho \mu_1 \delta s $ and would spread at speed $\sigma \sqrt{\delta s}$. 
@@
 
-In the above (and in \citet{ralphcoop2010}) we have assumed that once the mutation appears,
+In the above (and in \citet{ralphcoop2010}), we have assumed that once the mutation appears,
 conditional on eventual fixation, it begins to spread spatially at speed $v$ instantly,
@@
 We are assuming that shortly after a new mutation appears, 
-it can be approximated by a branching process growing at rate $s$
+it can be approximated by a branching process growing at rate \gc{$s_b$}
 until the point that it grows large enough to ``feel'' spatial structure,
@@
 Although we are not aware of good analysis of this transition, 
-the relevant size when spatial structure becomes important
+the relevant size \gc{of the branching process} when spatial structure becomes important
 should be something close to $\sigma^2 \rho$ 
@@
 New mutations will appear and become established during this time if
-$4 \rho \sigma^2 \mu s_b \ge 1/T$,
-i.e.\ if $4 \rho \sigma^2 \mu \ge 1/\log (\sigma^2 \rho)$.
+$2 \rho \sigma^2 \mu s_b \ge 1/T$,
+i.e.\ if $2 \rho \sigma^2 \mu \ge 1/\log (\sigma^2 \rho)$.
 Our model will still be a good approximation, however, 
@@
 This can be worked out,
-but it is simpler to note that if the converse is true, 
+but it is simpler to note that if the converse is true \gc2{(i.e. that
+reach local fixation is slow compared to the spread of the wave)}, 
 the process is largely unaffected by spatial structure,
 and so the panmictic model is a good approximation for the true process.
-Since both our spatial model and the panmictic model underestimate the true degree of parallel adaptation,
-it suffices to compare results from the two models to avoid applying the wrong model.
+%Since both our spatial model and the panmictic model underestimate the true degree of parallel adaptation,
+%it suffices to compare results from the two models to avoid applying the wrong model.
 
@@
 The results of \citet{softsweepsI} and \citet{softsweepsII} suggest that in a
-panmictic population the number of independent alleles (and their frequencies), in a sample,
+panmictic population the number of independent alleles (and their frequencies) in a sample
 is nearly independent of $s_b$ and $s_d$  \citep[although this
@@
 hence decreasing the time period in which new alleles can arise and hope to be at significant frequency in the population. 
-These two effects cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
+These two effects \gc2{approximately} cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
 Decreasing $s_d$ increases the number of standing variants within a population, 
@@
 acts to exclude the spread of new alleles that arise once the environment switches. 
-These two opposing effects again cancel out leading to little overall
+These two opposing effects again cancel out, leading to little overall
 effect of $s_d$ on the number of independent alleles.
@@
 depends on both $s_b$ and $s_d$ in a geographical spread population. 
-Like the panmictic model, in our model alleles also act to exclude each other; 
+Like the panmictic model, in our model, alleles also act to exclude each other; 
 however, the geographic spread of an allele is slow compared to the initial exponential growth of an allele in a panmictic population. 
@@
 alleles become established can dominate, leading to more independent origins, both by
-being  weaker before, and stronger after, the environmental switch.
+being  weaker before and stronger after the environmental switch.
 
@@
 
-Our results further suggest that convergent evolution among populations may be quite common, 
-especially when standing variation is not so deleterious that it is present before environmental shifts. 
 When the geographical area where a species experiences a 
@@
 length we should expect multiple independent alleles to arise and
-spread in that area. Allowing standing variation in our model  
-decreases the characteristic length, i.e. it increases the extent of
-convergence among populations.
+spread in that area. Our results suggest that convergent evolution
+among populations may be quite common, \gc{at least when population
+  densities and mutational target sizes are large, and dispersal is
+  limited across the range.  Including standing variation is present
+before environmental shifts can greatly increase the probability of convergence. }
 While at face value this increase seems unsurprising,
-this relationship differs markedly from the case of multiple competing alleles in a panmictic population \citep{softsweepsI,softsweepsII}. 
+this relationship differs markedly from the case of multiple competing
+alleles in a panmictic population \citep[see discussion above and in ][]{softsweepsI,softsweepsII}. 
 Importantly, allowing standing variation may greatly lower the time until the species becomes adapted 
-across the geographic range of the selection pressure. Adaptation through standing variation also biases the type of variation towards those alleles 
+across the geographic range of the selection pressure. We have also
+shown that adaptation through standing variation biases the type of variation towards those alleles 
 with fewer pleiotropic effects, since these are more common as standing variation before the environmental shift. 
@@
 
-More generally, if geographic convergence is common we should often expect to see
-selected alleles which are strongly geographically restricted as they
+More generally, if geographic convergence is common, we should often expect to see
+selected alleles that are strongly geographically restricted as they
 have simply not had time for neutral mixing to spread them across the landscape.
@@
 migration and drift, 
-potentially leading to a signal of a sweep from multiple independent
+potentially leading to a \gc2{within population} signal of a sweep from multiple independent
 mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
@@
 initially rare within populations, i.e.\ they are reasonably
-deleterious before the environment switches. In contrast, even distant populations could have
-a shared basis of adaptation if they adapt via common variation,
-e.g.\ previously neutral (or nearly neutral) variation, shared among populations. 
-Our ability to detect such shared convergent events is at the moment quite limited. 
+deleterious before the environment switches. \gc2{ In contrast, if they adapt via common variation even distant populations could have
+a shared basis of adaptation, e.g.\ previously neutral (or nearly neutral) variation, shared among populations. 
 If many loci contribute to variation in a trait, then
@@
 adapt via different genetic routes, as the constellation of alleles
-best able to respond to selection will be somewhat different \citep{Barton:89}.
+that respond to selection quickest will be somewhat different due to
+drift among populations} \citep{Barton:89}.
 % best positioned to initially respond to positive selection, 
@@
 suggests that species should often adapt to
-wide-spread selection pressures through convergent evolution rather
+widespread selection pressures through convergent evolution rather
 than waiting for a single allele to migrate across the range.
@@
 rather than species-wide \citep[but see][ for a potential example]{Clark:07,Long:13}. 
-This is likely in part due to the relatively low incidence of wide-spread selection pressures,
-but as noted above even when we know of wide-spread selection
+This is likely in part due to the relatively low incidence of widespread selection pressures,
+but as noted above even when we know of widespread selection
 pressures (e.g.\ malaria) the response is usually convergent, not shared, across large spatial scales. 
@@
 However, at least some of these cases 
-are likely caused by introgression of haplotype complexes
+may be caused by introgression of haplotype complexes
 consisting of many, tightly linked, beneficial alleles (that perhaps
@@
 divergence among species. 
+\gc{Slight selective differences in the pleiotropic effects (or linked
+  background) among convergent alleles could allow one allele to
+  press into areas occupied by other alleles.}
 Furthermore, repeated bouts of adaptation in particular genomic regions may act 
 to push a subset of previously selected alleles to fixation across the
-species range, through the spread of the genetic backgrounds on which they arise.  
+species range, through the spread of the genetic backgrounds on which
+they arise.  
 However, it seems likely that this is a slow process compared to the
@@
 At face value the rate of species formation seems too low to contribute to this process. 
-However, \citep{Rosenblum:12}, and many others, 
+However, \citet{Rosenblum:12}, and many others, 
 have argued that the rate of speciation may well be quite high,

### 2015-02-15 g6pd_figures.R

@@ -6,6 +6,7 @@ library("wesanderson")
 my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
 
 
+######  Figure 1
 ###################################################
 ###################Char length as function of sb
 ###################################################
@@ -36,6 +37,7 @@ for(j in 1:2){
 
 dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlengths_sb.pdf")
 
+######  Figure 2
 ###################################################
 ###################Char length as function of sd
 ###################################################
@@ -67,47 +69,20 @@ for(j in 1:2){
  	
 dev.copy2pdf(file="~/Dropbox/postdocs/Peter/standing_parallelism/G6PD_charlengths.pdf")
 
-###############################################################
-####Mean time standing as func. of sd
-###############################################################
- layout(t(1:2))
-par(mar=c(4,4.1,2,0.3))
-for(j in 1:2){
-	rho=rhos[j]
-	if(j == 1) plot(sdvals,sdvals,type="n",ylim=c(0,250),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab="Generations",cex.lab=1.5,cex.axis=1.5,main=expression(rho==0.2))
-	if(j == 2) plot(sdvals,sdvals,type="n",ylim=c(0,250),lwd=2,col=my.col[1],xlab=expression(s[d]),ylab="",cex.lab=1.5,cex.axis=1.5,main=expression(rho==2))
-	
-	for(i in 1:3){
-	
-	sigma=sigmas[i]; 
-	charlength.sd<-sapply(sdvals, function (sd) {  meanTime(mu, rho, sb, sd, sigma)$value } )
-	 lines(sdvals,charlength.sd,lwd=2,col=my.col[i])
-	 abline(h= meanTime(mu, rho, sb, sd=1, sigma)$value,col=my.col[i],lwd=2,lty=2)  ##new mut. only
- 	 charlength.sd<-sapply(sdvals, function (sd) {  meanTime(mu, rho, sb, sd, sigma,include.new=FALSE)$value } )
-	 lines(sdvals,charlength.sd,lwd=2,lty=3,col=my.col[i])
-
-	}
-	
-
[…]

### 2015-02-15 standing-variation-paper.tex
making changes in response to reviewer and molly comments

@@
 \usepackage{color}
-\newcommand{\gc}[1]{{\it\color{green}(#1)} }
-\newcommand{\gc2}[1]{{\it\color{purple}(#1)} }
-
-\newcommand{\plr}[1]{{\it\color{blue}(#1)}}
+\newcommand{\gc}[1]{{\it\color{blue} #1} }
+\newcommand{\mfp}[1]{{\it\color{red} #1} }
+
+\newcommand{\plr}[1]{{\it\color{green}(#1)}}
 \usepackage{lineno}  %%%% FOR AMNAT
@@
 We apply the results to the many geographically localized G6PD
-deficiency alleles thought to confer resistance to malaria, \gc2{where the large
-mutational target size of G6PD makes likely candidate for adaptation
+deficiency alleles thought to confer resistance to malaria, \mfp{where the large
+mutational target size makes it a likely candidate for adaptation
 from standing variation, despite the selective cost of G6PD
-deficiency.}
+deficiency alleles in the absence of malaria.}
 We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
-\gc2{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
+\mfp{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
 Our results suggest that much of adaptation may be geographically local 
-even when selection pressures are widespread. \gc2{Therefore, we argue
+even when selection pressures are widespread. \mfp{Therefore, we argue
 that caution must be exercised when arguing that strongly
@@
 adapted to similar environments by means of independent genetic changes. 
-In some cases this convergent evolution \gc2{is the result of quite
+In some cases this convergent evolution \mfp{is the result of quite
 distinct genetic changes, involving very different genes and pathways} despite shared selection pressures; 
@@
 whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
-and to \gc2{what degree species should be regarded} as cohesive units. 
+and to \mfp{what degree species should be regarded} as cohesive units. 
 Convergent evolution also affects our ability to detect adaptation from population genomic data,
@@
 \ref{fig:sim-spread}. A somewhat analogous situation also arises in spatial models of clonal
-inference in asexuals \citep{martens_interfering_2011}.}
+inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}
 
@@
   \caption{
-{\bf Simulation of the geographic spread of new mutations}.
-Simulation on a grid of $101 \times 101$ populations with 100 individuals in each, showing spread of a beneficial allele with advantage $s=0.1$.
-Times are in numbers of generations, with probability of mutation per generation 0.3.
+{\bf Simulation of the geographic spread of new mutations}. Simulation
+on a grid of $101 \times 101$ populations with 100 haploid individuals
+in each, showing spread of a beneficial allele with advantage $s=0.1$.
+Times are in numbers of generations, with probability of mutation per
+generation 0.3. Different colours show the different independent
+origins and spreads of selected alleles. Alleles quick establish and
+start to spread, but because they are selective equivalent once they
+spread into each other they only mix slowly (note the lack of change
+from generation 600 onward). More details of the
+simulation are given in \citet{ralphcoop2010}. 
 }
@@
 This characteristic length could be expressed in terms of a simple compound
-parameter \gc2{comprised of} our parameters of interest. \gc{In
+parameter \mfp{comprised of} our parameters of interest. \gc{In
 \citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
@@
 % and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
-and that widespread selective sweeps \gc2{should tend to} occur
+and that widespread selective sweeps \mfp{should tend to} occur
 when adaptation is highly constrained (e.g.\ by small mutation rate or
@@
 important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
-homozygote for the mutant alleles. \gc2{Note that this means that we
+homozygote for the mutant alleles. \mfp{Note that this means that we
   are ignoring the case of totally recessive alleles.}
@@
 and that the mean squared geographic distance between parent and child
-is $\sigma^2$. \gc2{Note that we assume that the dispersal kernel is
+is $\sigma^2$. \mfp{Note that we assume that the dispersal kernel is
   not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
@@
 The genetic descendants of each new mutation are (with high probability) doomed to extinction,
-but may persist for \gc2{a short} time (note that we have assumed that $s_d$ is not too small).
+but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
 \gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
@@
 to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
-\gc2{Our $\lambda_0$ is the mean density
+\mfp{Our $\lambda_0$ is the mean density
 of standing variants that are present and escape loss when the
@@
 Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
-and spread through human populations in areas where malaria \gc2{has been} endemic \citep{Kwiatkowski:05}. 
+and spread through human populations in areas where malaria \mfp{has been} endemic \citep{Kwiatkowski:05}. 
 A number of these alleles appear to be examples of convergent adaptation, 
@@
 \citep{Flint:98,ralphcoop2010}.
-Another particularly impressive case of convergent evolution \gc2{is} the numerous changes throughout the X-linked G6PD gene, 
+Another particularly impressive case of convergent evolution \mfp{is} the numerous changes throughout the X-linked G6PD gene, 
 with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
@@
 and the Mahidol allele found Myanmar and Thailand.
-\gc2{ 
+\mfp{ 
 The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
@@
 
-Given \gc2{such strong selection the alleles} should have risen
+Given \mfp{such strong selection the alleles} should have risen
 quickly to fixation, so their presence at intermediate frequency,
@@
 ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
-\gc2{to} escape drift and become established. 
+\mfp{to} escape drift and become established. 
 The effect of helping\gc{locally} establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
@@
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter \gc2{only the
+as in that case the speed of spread does not matter \mfp{only the
   initial density of established alleles after the environmental shift}. 
@@
 800km, especially for the lower population density, as long as the
-fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \gc2{high}.
+fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \mfp{high}.
 
@@
 mutation with probability given by equation \eqref{eqn-no-mut-t}.
-\gc2{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
+\mfp{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
 by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
@@
 \end{equation}
-\gc2{this is the probability a given location is reached first by a
+\mfp{this is the probability a given location is reached first by a
 standing variant (which follows from competing our two exponential
@@
 and an even larger proportion of the range 
-(they \gc2{occupy a larger area than new mutations, since they get a
+(they \mfp{occupy a larger area than new mutations, since they get a
 headstart}).
@@
 the rarer, class $1$ mutations in Figure \ref{fig:pleiotropy_calc}.
-\gc2{As expected intuitively,} the contribution of the rarer mutation decreases as the
+\mfp{As expected intuitively,} the contribution of the rarer mutation decreases as the
 mutational target of the second class becomes larger, and as the
@@
 This can be worked out,
-but it is simpler to note that if the converse is true \gc2{(i.e. that
+but it is simpler to note that if the converse is true \mfp{(i.e. that
 reach local fixation is slow compared to the spread of the wave)}, 
@@
 hence decreasing the time period in which new alleles can arise and hope to be at significant frequency in the population. 
-These two effects \gc2{approximately} cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
+These two effects \mfp{approximately} cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
 Decreasing $s_d$ increases the number of standing variants within a population, 
@@
 migration and drift, 
-potentially leading to a \gc2{within population} signal of a sweep from multiple independent
+potentially leading to a \mfp{within population} signal of a sweep from multiple independent
 mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
@@
 initially rare within populations, i.e.\ they are reasonably
-deleterious before the environment switches. \gc2{ In contrast, if they adapt via common variation even distant populations could have
+deleterious before the environment switches. \mfp{ In contrast, if they adapt via common variation even distant populations could have
 a shared basis of adaptation, e.g.\ previously neutral (or nearly neutral) variation, shared among populations. 
@@
 populations to cross fitness valleys and substitute novel epistatic combinations. 
-However, although geographic lineage sorting via speciation and extinction 
+Although geographic lineage sorting via speciation and extinction 
 can be thought of as very large-scale genetic drift events,
@@
 at least in a subset of genes, and speciation may be associated with rapidly adaption to novel environments. 
-Conversely, if the geographic spread of adaptive alleles within ranges is slow, 
-e.g.\ if they only offer a local advantage or if they are selective excluded, 
-may allow Dobzhansky-Muller incompatibilities to arise within species, 
+Conversely, if the geographic spread of adaptive alleles within ranges is slow 
+\mfp{(e.g.\ if they only offer a local advantage or if they are selectively excluded)} 
+this may allow Dobzhansky-Muller incompatibilities to arise within species, 
 effectively offering a mechanism for hybrid incompatibilities to evolve in parapatry,

### 2015-02-17 standing-variation-paper.tex
changes in response to reviews

@@
         &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
-        & =  \frac{\lambda }{ (-log(1-s_d)) }.\\
+        & =  \frac{\lambda }{ (-log(1-s_d)) }.
 \end{align}
-Note that this is closely related to the equilibrium frequency of an
-allele at mutation selection balance $\mu/s_d$
-\citep{Haldane:27,Haldane:37, Orr:01}.
+Note that for small $s_d$ this can also be found by taking the expected frequency under mutation-selection
+balance $\mu/s_d $ \citep{Haldane:27,Haldane:37} multiplying it by the
+population density to obtain the expected number of chromosomes per unit
+carrying the deleterious allele, with each of these having a
+probability $2s_b$ of escaping loss \citep[an analogous approach to
+that taken by ][]{Orr:01}
+.
 
@@
 $\nu_+ = \lambda \E[\tau] $.
-Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
-\begin{equation}
-    \nu_+ = (\mbox{mean density of new patches}) = \frac{1}{1-\log(1-s_d) \E[\tau]} .
-\end{equation}
+%Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
+%\begin{equation}
+ %   \nu_+ = (\mbox{mean density of new patches}) = \frac{1}{1-\log(1-s_d) \E[\tau]} .
+%\end{equation}
 Therefore, the mean proportion of \gc{adapted} patches that come from standing variation
@@
 \begin{equation} \label{prop_patches_standing}
-\lambda_0 / (\lambda_0 + \nu_+) .
+\lambda_0 / (\lambda_0 + \nu_+)  = \frac{1}{1-\log(1/(1-s_d)) \E[\tau]} .
 \end{equation}
@@
 There are $\lambda_0 + \nu_+$ patches per unit area, so
-the typical patch (i.e.\ with distribution given by the Palm measure)
-occupies area $1/(\lambda_0 + \nu_+)$. \gc{NEED REF FOR PALM}
+the typical patch (informally, the patch around a randomly chosen
+successful mutation; formally, with the Palm measure
+\citep{cox1980point}) occupies area $1/(\lambda_0 + \nu_+)$.
 
@@
 \subsubsection*{Acknowledgements}
-We thank Michael Turelli, Jon Seger, and the Coop lab for helpful conversations.
+We thank Michael Turelli, Jon Seger, and the Coop lab for helpful
+conversations. We thank Florence D\'ebarre, Douglas Futuyma, Michael   %Débarre
+Whitlock, Molly Przeworski, and  Sam Yeaman for helpful comments on an
+earlier version of the paper. All figures and calculations were
+performed in \citet{R_stats}. The color paletes were generated using
+Karthik Ram's Wes Anderson R color palette \url{https://github.com/karthik/wesanderson}.
 

### 2015-02-17 g6pd_figures.R

@@ -1,5 +1,4 @@
-
-source('~/Dropbox/postdocs/Peter/Spatial_adaptation/Spatial_adaptation/standing-variation-fns.R')
+source("~/Dropbox/postdocs/Peter/standing_parallelism/standing-variation-fns.R")
 #install.packages("wesanderson")
 library("wesanderson")
 
@@ -31,7 +30,7 @@ for(j in 1:2){
 	 ##contribution of standing var only, i.e. old variation
 	 charlength.standing.only<-sapply(sbvals, function (sb) {  oldCharLength(mu, rho, sb, sd=sd, sigma)$value})
 	 lines(sbvals,charlength.standing.only ,lwd=2,col="black",lty=3)
-	  if(j == 2) legend("topright",col=c("grey","grey","black",my.col[1:3]),lwd=2,lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
+	  if(j == 2) legend("topright",col=c("grey","grey","grey",my.col[1:3]),lwd=2,lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
 	 	lapply(sigmas , function(x){  {substitute( sigma==myS, list(myS=x) )} }))))
  	}
 
@@ -63,7 +62,7 @@ for(j in 1:2){
 	 ##contribution of standing var only, i.e. old variation
 	 charlength.standing.only<-sapply(sdvals, function (sd) {  oldCharLength(mu, rho, sb, sd, sigma)$value})
 	 lines(sdvals,charlength.standing.only ,lwd=2,col="black",lty=3)
-	  if(j == 2) legend("topright",col=c("grey","grey","black",my.col[1:3]),lwd=2,lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
+	  if(j == 2) legend("topright",col=c("grey","grey","grey",my.col[1:3]),lwd=2,lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
[…]

### 2015-02-18 standing-variation-paper.tex
response

@@
 \end{align}
-For means of evaluating this integral, see Appendix \ref{apx:integrals}.
+\gc{For applications we evaluate this integral numerically. }  %For means of evaluating this integral, see Appendix \ref{apx:integrals}.
 
@@
 waiting times).}
-To evaluate this integral, again see Appendix \ref{apx:integrals}.
+\gc{For applications we evaluate this integral numerically.}
+%To evaluate this integral, again see Appendix \ref{apx:integrals}.
 %\marginnote{put in this value in terms of Gaussian CDF?}
@@
 earlier version of the paper. All figures and calculations were
-performed in \citet{R_stats}. The color paletes were generated using
+performed in \citet{R_stats}. The figure color paletes were generated using
 Karthik Ram's Wes Anderson R color palette \url{https://github.com/karthik/wesanderson}.
@@
 
-\appendix
-
-%%%%%%%%%%%%
-\section{Integrals appearing in the text}
-    \label{apx:integrals}
-
-In the text at several points appear integrals of the form
-\begin{align}
-  \int_0^\infty t^c \exp \left( - \alpha t^d - \beta t^{d+1} \right) dt 
-\end{align}
-where $c$ is a positive integer, $d$ is the dimension, and $\alpha$ and $\beta$ are positive real numbers.
-This could be evaluated through standard numerical methods; below we describe a power series expansion.
-Changing variables to $u = \beta t^{d+1}$, this becomes
-\begin{align}
-    \left( (d+1)^{-1} \beta^{ (1-c)/(d+1) } \right) \int_0^\infty u^{(c+1-d)/(d+1)} \exp\left( - \alpha \beta^{-d/(d+1)} u^{d/(d+1)} - u \right) du ,
-\end{align}
-so it suffices to evaluate the function
-\begin{equation}
-    G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt ,
-\end{equation}
-in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic parameters.
-Since $a$ and $b$ only depend on the dimension and the quantity being computed,
-we are interested in $G$ as a function of $x$.
-At least in the case $b=d/(d+1)$ it is possible to express $G$ as a finite sum of gamma functions,
-but we proceed with a simpler method.
-Note that $\partial_x G(a,b,x) = -G(a+b,b,x)$,
-and that at $x=0$, the function $G$ is the gamma function $G(a,b,0) = \Gamma(a+1)$.
-Therefore, a Taylor series for $G$ would be
-\[
-    G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
-\]
-It is easy to check using Stirling's formula that $\limsup_{n \to \infty} ( x^n \Gamma(a+nb+1)/n! )^{1/n} = 0$
-if $b<1$, so the sum converges.
+%\appendix
+
+% %%%%%%%%%%%%
+% \section{Integrals appearing in the text}
+%     \label{apx:integrals}
+
+% In the text at several points appear integrals of the form
+% \begin{align}
+%   \int_0^\infty t^c \exp \left( - \alpha t^d - \beta t^{d+1} \right) dt 
+% \end{align}
+% where $c$ is a positive integer, $d$ is the dimension, and $\alpha$ and $\beta$ are positive real numbers.
+% This could be evaluated through standard numerical methods; below we describe a power series expansion.
+% Changing variables to $u = \beta t^{d+1}$, this becomes
+% \begin{align}
+%     \left( (d+1)^{-1} \beta^{ (1-c)/(d+1) } \right) \int_0^\infty u^{(c+1-d)/(d+1)} \exp\left( - \alpha \beta^{-d/(d+1)} u^{d/(d+1)} - u \right) du ,
+% \end{align}
+% so it suffices to evaluate the function
+% \begin{equation}
+%     G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt ,
+% \end{equation}
+% in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic parameters.
+% Since $a$ and $b$ only depend on the dimension and the quantity being computed,
+% we are interested in $G$ as a function of $x$.
+% At least in the case $b=d/(d+1)$ it is possible to express $G$ as a finite sum of gamma functions,
+% but we proceed with a simpler method.
+% Note that $\partial_x G(a,b,x) = -G(a+b,b,x)$,
+% and that at $x=0$, the function $G$ is the gamma function $G(a,b,0) = \Gamma(a+1)$.
+% Therefore, a Taylor series for $G$ would be
+% \[
+%     G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
+% \]
+% It is easy to check using Stirling's formula that $\limsup_{n \to \infty} ( x^n \Gamma(a+nb+1)/n! )^{1/n} = 0$
+% if $b<1$, so the sum converges.
 

### 2015-02-21 standing-variation-paper.tex
finishing up changes in response to review

@@
 The species occupies an area with mean density
-$\rho$ of haploid individuals \gc{(twice the number of diploids)} per unit area.
+$\rho$ of haploid individuals \gc{(twice the number of diploid individuals)} per unit area.
 
@@
 treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
-as a branching process.
+as a branching process \gc{in continuous time, measured in generations
+(average inter-birth intervals)}. 
 After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
@@
 Assuming a mutation rate of $\approx 10^{-8}$ per base pair per generation, 
-we can take as an order-of-magnitude estimate $\mu \approx 1 \times 10^{-6}$ per generation. 
+we can take as an order-of-magnitude estimate $\mu \approx 10^{-6}$ per generation. 
 The dispersal and demographic parameters of humans in the past few thousand years is unclear,
@@
 %as well as a similar compound parameter describing the relative importance of standing variation to the process of adaptation.
-%(Note that a similar approach, summarizing with a {\em characteristic length} was taken by \citep{slatkin1973geneflow} for a different problem.)
-\gc{CHECK FACTOR of $1/3$ in cone volume.}
+%(Note that a similar approach, summarizing with a {\em characteristic
+%length} was taken by \citep{slatkin1973geneflow} for a different
+%problem.)
+
 Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
@@
 circular area of size $\pi
-L^2$ and carved out a cone in space-time with this as a base and
-height $L/v$ (with a volume of $\pi L^2 \times (L/v)/3$)
+L^2$ and carved out a cylinder in space-time with this circle as a base and
+height $L/v$ (the volume of this cylinder is $\pi L^2 \times (L/v)$)
 Therefore, the number of other successfully established mutations that
@@
 From this we see that $\chi$ decreases with $\rho$ and $\mu$.
-Furthermore, for large $\sigma$, the characteristic length \gc{approaches}
-the value we would just from standing variation:
+Furthermore, for large $\sigma$, the characteristic length \gc{approaches
+the value we would obtain just} from standing variation:
 \begin{equation} \label{eqn:chi_standing}
@@
 The other parameters are chosen to match those of G6PD,
-$\mu = 10^{-5}$ and $s_b = 0.05$. As the characteristic length with
+\gc{$\mu = 10^{-6}$} and $s_b = 0.05$. As the characteristic length with
 only standing variation is independent of dispersal distance it is
@@
 {\bf Characteristic length, in kilometers, as a function of selective advantage},
-holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
+holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at \gc{$\mu=10^{-6}$}.  
 We compare this to the corresponding quantity without standing
@@
 $\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$
-extending back to $t=0$ (light grey dashed lines) is empty of successful mutations.
-Since we assume these are a Poisson process, 
+extending back to $t=0$ (light grey dashed lines) is empty of
+successful mutations. Since we assume that \mfp{our successful alleles} arise as a Poisson process, 
 in two dimensions
@@
 environment shifts.
-We will define $\nu_+$ to be the mean density of newly arising alleles
+We will define $\gamma$ to be the mean density of newly arising alleles
 that spread having arisen in an area free of other adaptive alleles.
@@
 is $\P\{ \tau > t \}$,
-we can see  $\nu_+ = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
-$\nu_+ = \lambda \E[\tau] $.
+\gc{we can see  $\gamma = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
+$\gamma = \lambda \E[\tau] $.}
 %Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
@@
 \begin{equation} \label{prop_patches_standing}
-\lambda_0 / (\lambda_0 + \nu_+)  = \frac{1}{1-\log(1/(1-s_d)) \E[\tau]} .
+\lambda_0 / (\lambda_0 + \gamma)  = \frac{1}{1-\log(1/(1-s_d)) \E[\tau]} ,
 \end{equation}
+using the fact that $\lambda_0 = \lambda / \log(1/(1-s_d))$.
 % this gives us the exact quantity we approximated by $\pi_0$ in \eqref{eqn:pizero}.
-There are $\lambda_0 + \nu_+$ patches per unit area, so
+There are $\lambda_0 + \gamma$ patches per unit area, so
 the typical patch (informally, the patch around a randomly chosen
 successful mutation; formally, with the Palm measure
-\citep{cox1980point}) occupies area $1/(\lambda_0 + \nu_+)$.
+\citep{cox1980point}) occupies area $1/(\lambda_0 + \gamma)$.
 
@@
 among populations may be quite common, \gc{at least when population
-  densities and mutational target sizes are large, and dispersal is
-  limited across the range.  Including standing variation is present
-before environmental shifts can greatly increase the probability of convergence. }
+  densities and mutational target sizes are not too small, and dispersal is
+  limited across the range.  While considering smaller mutational
+target sizes (e.g. less than the 100bp in our example) or lower population densities would lower the probability of convergence
+within a given geographical area, it would also act to considerably lengthen
+the time to adaptation. In such cases adaptation may simply fail to
+occur at all or populations may adapt via some other means (i.e. by using a
+broader mutational target). 
+}
+
+\gc{Our inclusion of standing variation 
+before environmental change greatly increase the probability of
+convergence with a species. }
 While at face value this increase seems unsurprising,
@@
 selected alleles that are strongly geographically restricted as they
-have simply not had time for neutral mixing to spread them across the landscape.
+have simply not had time for neutral mixing to spread them across the
+landscape. \gc{Such convergent alleles will be $F_{ST}$-outliers compared to
+  older neutral variation, untill there is sufficient time for
+  migration to smooth out them out.}
 This pattern may be very hard to distinguish from local adaptation using population genomic approaches
@@
 in which local sweeps occur independently across the range.
-Lineages that are close to the sweep ($\sim v/\chi$ Morgans) will be moved
+Lineages that are \gc{closely linked to the sweeping allele} ($\sim v/\chi$ Morgans) will be moved
 towards the center of the sweep (a displacement $O(\chi)$), 

### 2015-03-19 standing-variation-paper.tex
almost final changes

@@
 In some cases this convergent evolution \mfp{is the result of quite
-distinct genetic changes, involving very different genes and pathways} despite shared selection pressures; 
+distinct genetic changes, involving very different genes and pathways,} despite shared selection pressures; 
 in other cases independent adaptations are identical down the same
@@
 or due to multiple adaptive alleles that arise after selection pressures switch \citep{softsweepsII}. 
-The previous work has shown that 
+Previous work has shown that 
 a primary determinant of the probability that multiple alleles contribute to adaptation
@@
 the population after the change in environment. Under this assumption,
-selected alleles arise and spread locally. However, if the geographic
+selected alleles arise and spread locally.  See Figure
+\ref{fig:sim-spread}. However, if the geographic
 area is large enough multiple selected alleles can
 arise independently and spread before any one has spread across all of
-space ( Figure \ref{fig:sim-spread}).  one allele's rapid spatial spread can be
-halted by running into convergent allele. See Figure
-\ref{fig:sim-spread}. A somewhat analogous situation also arises in spatial models of clonal
-inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}
+space. Therefore, one allele's rapid spatial spread can be halted by
+running into convergent allele.A somewhat analogous situation also
+arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}
 
@@
 common \gc{when ranges, population sizes, or mutational targets are large}, as has already been seen for a number of traits. 
-On this basis owe argue 
+On this basis we argue 
 that the genetics of adaptation may often be geographically local \gc{even
@@
 % and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
-and that widespread selective sweeps \mfp{should tend to} occur
+and that widespread selective sweeps \mfp{should tend to occur only}
 when adaptation is highly constrained (e.g.\ by small mutation rate or
@@
   environmental shift). Note that these mutations do not have to arise at the same locus,
-  just that all of the mutated type of selective equivalent. As such,
+  just that all of the mutated type are selective equivalent. As such,
   these could be mutations arising at the same base pair or knockout
@@
 and that the mean squared geographic distance between parent and child
-is $\sigma^2$. \mfp{Note that we assume that the dispersal kernel is
+is $\sigma^2$. \mfp{We assume that the dispersal kernel is
   not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
@@
 The species occupies an area with mean density
-$\rho$ of haploid individuals \gc{(twice the number of diploid individuals)} per unit area.
+$\rho$ of alleles per unit area \gc{(twice the number of diploid
+  individuals per unit area)}.
 
@@
 \gc{In geographic areas already
- occupied by the mutant type (i.e. in regions it has spread to already), new mutations are selectively excluded
- (i.e. are effectively neutral), and so unlikely to establish over the
+ occupied by the mutant type (i.e. in regions were the beneficial
+ allele has spread to), new mutations are selectively excluded,
+ i.e. are effectively neutral, and so unlikely to establish over the
  short-time scales considered here.  }
@@
 Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
-(See \citet{ralphcoop2010} for a more thorough review of these travelling waves.)
+See \citet{ralphcoop2010} for a more thorough review of these travelling waves.
 Note that the speed of the wave will vary
@@
 originating $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
-Point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
+Our point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
 }
@@
 before environmental change greatly increase the probability of
-convergence with a species. }
+convergence within a species. }
 While at face value this increase seems unsurprising,
@@
 selected alleles that are strongly geographically restricted as they
-have simply not had time for neutral mixing to spread them across the
+have simply not had time for neutral gene flow to spread them across the
 landscape. \gc{Such convergent alleles will be $F_{ST}$-outliers compared to
   older neutral variation, untill there is sufficient time for
-  migration to smooth out them out.}
+  migration to smooth out them out across the landscape.}
 This pattern may be very hard to distinguish from local adaptation using population genomic approaches

### 2015-04-02 standing-variation-paper.tex
peter edited through

@@
 \usepackage{color}
-\newcommand{\gc}[1]{{\it\color{blue} #1} }
-\newcommand{\mfp}[1]{{\it\color{red} #1} }
+\usepackage[utf8]{inputenc}  % for special characters in authors' names
+
+\newcommand{\gc}[1]{{\it\color{blue} #1}}
+\newcommand{\mfp}[1]{{\it\color{red} #1}}
 
@@
 In this paper, 
-we explore how standing, deleterious genetic variation contributes to
+we explore how standing, genetic variation contributes to
 convergent genetic responses in a geographically spread population,
@@
 Finally, we explore how negative pleiotropic effects of alleles before an environment change 
-can bias the subset of alleles that get to contribute to a species adaptive response. 
+can bias the subset of alleles that contribute to the species' adaptive response. 
 We apply the results to the many geographically localized G6PD
@@
 We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
-\mfp{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
+\mfp{consistent with the view that they arose from a combination of} 
+standing variation and new mutations since the advent of malaria.
 Our results suggest that much of adaptation may be geographically local 
-even when selection pressures are widespread. \mfp{Therefore, we argue
-that caution must be exercised when arguing that strongly
+even when selection pressures are homogeneous. 
+\mfp{Therefore, we argue that caution must be exercised when arguing that strongly
 geographically restricted alleles are necessarily the outcome of local adaptation.}
-We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.
+We close by discussing the implications of these results for 
+ideas of species coherence and the nature of divergence between species.
 
@@
 In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
- a spatially spread population, where dispersal is local,
+a spatially spread population with local dispersal
 that is exposed to some novel, spatially homogeneous selection pressure.
@@
 the population after the change in environment. Under this assumption,
-selected alleles arise and spread locally.  See Figure
-\ref{fig:sim-spread}. However, if the geographic
-area is large enough multiple selected alleles can
-arise independently and spread before any one has spread across all of
-space. Therefore, one allele's rapid spatial spread can be halted by
-running into convergent allele.A somewhat analogous situation also
+selected alleles arise and spread locally, as shown in Figure \ref{fig:sim-spread}. 
+If the geographic area is large enough multiple selected alleles can
+arise independently and spread before any one has spread across all of space. 
+A somewhat analogous situation also
 arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}
@@
 \gc{Under this model we previously derived} the characteristic geographic scale over which 
-multiple instances of the adaptive allele are expected to arise in parallel. 
-This characteristic length could be expressed in terms of a simple compound
-parameter \mfp{comprised of} our parameters of interest. \gc{In
-\citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
-allele (e.g. because the allele was very strongly deleterious before
-the environmental change)}, 
+multiple instances of the adaptive allele are expected to arise in parallel, 
+a characteristic length \mfp{expressed in terms the parameters of interest.} 
+\gc{In \citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
+allele (e.g., because the allele was very strongly deleterious before the environmental change)}, 
 so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.
@@
 is a large, homogeneous, one-- or two--dimensional region. 
-There are two selective classes -- the \gc{ {\em ancestral} } type,
+There are two selective classes -- the \gc{\textit{ancestral}} type,
 and the {\em mutated} type \gc{(which will offer a fitness benefit after the
@@
 equivalent (both in terms how deleterious they were before, and how
-  advantagous they are after, the
-  environmental shift). Note that these mutations do not have to arise at the same locus,
-  just that all of the mutated type are selective equivalent. As such,
-  these could be mutations arising at the same base pair or knockout
-  mutations of any one of a number of genes in a pathway as long as carrying at least one of these alleles
+advantagous they are after, the environmental shift). 
+Note that these mutations do not have to arise at the same locus,
+just that they are selectively equivalent:
+these could be mutations arising at the same base pair, 
+or knockout mutations of any one of a number of genes in a pathway,
+as long as carrying at least one of these alleles
 is sufficient to adapt an individual to the new environment.}
@@
 is $\sigma^2$. \mfp{We assume that the dispersal kernel is
-  not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
+  not heavy tailed, i.e., falling into the Gaussian domain of attraction, see \citet{ralphcoop2010} for
   discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
@@
 as a branching process \gc{in continuous time, measured in generations
-(average inter-birth intervals)}. 
+(average parental age at birth)}. 
 After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
@@
   mutations arise, and the occurance of new mutations in any region of
-  space-time is independent of that in non-overlapping intervals.}
+  space-time is independent of that in non-overlapping intervals
+  in which the mutated type has not already appeared.}
 The rate of this Poisson process,
@@
 appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
-is approximately 
+\plr{is the product of the haploid population density, denoted $\rho$,
+the mutation rate ($\mu$),
+and the probability of fixation $p_s$, or}
+approximately 
 \begin{equation}
@@
 \end{equation}
-\gc{In geographic areas already
- occupied by the mutant type (i.e. in regions were the beneficial
- allele has spread to), new mutations are selectively excluded,
- i.e. are effectively neutral, and so unlikely to establish over the
- short-time scales considered here.  }
+\gc{In geographic areas already occupied by the mutant type, 
+ new mutations are effectively neutral, and so unlikely to establish over the
+ short-time scales considered here, and are hence excluded.  }
 
@@
 but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
-\gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
+\gc{The} times and locations of new mutations before $t=0$ also \gc{will} be 
 well--approximated by a 
-Poisson process with rate $\mu \rho$ \gc{(if $\mu \rho \ll 1$) as
-mutations are rare events. Therefore,} 
-the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
-by the Poisson Mapping Theorem, a Poisson process with rate we define
-to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
-\mfp{Our $\lambda_0$ is the mean density
+Poisson process with rate $\mu \rho$ % PLR: this is not required, and depends on the units you measure area in: (if $\mu \rho \ll 1$) 
+\gc{since mutations are rare events. Therefore,} 
+the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is 
+a thinning of the original process, 
+and so, by the Poisson Mapping Theorem, is also a Poisson process.
+\mfp{We define $\lambda_0$ to be the mean density of this process, 
+i.e., the geographic density
 of standing variants that are present and escape loss when the
@@
 and the spatial distribution of each is small relative to the spread between the families.)
-To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let \gc{$Z_T$ be the number of its descendants at time $0$}.
+To find $\lambda_0$, consider a mutation that arose \plr{$T$ generations ago}, 
+and let \gc{$Z_T$ be the number of its descendants at time $0$}.
 At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
@@
         &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
-        & =  \frac{\lambda }{ (-log(1-s_d)) }.
+        & =  \frac{\lambda }{ (-\log(1-s_d)) }.
 \end{align}
@@
 balance $\mu/s_d $ \citep{Haldane:27,Haldane:37} multiplying it by the
-population density to obtain the expected number of chromosomes per unit
+population density to obtain the expected number of chromosomes per unit area
 carrying the deleterious allele, with each of these having a
@@
 the cells formed by a Voronoi tessellation have straight sides,
-but the introduction of new mutations cause these to curve \citep[see
-Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.
+but the introduction of new mutations cause these to curve.
+% \citep[see Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.
 
@@
 The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
-and the mean density in space $\times$ time of successful new mutations is $\lambda$.
+and the mean density in (space $\times$ time) of successful new mutations is $\lambda$.
 These successful alleles spread spatially outwards at speed $v$, 
 so the space-time profile of their spread forms a cone. Any mutation
-that arises in an area already within a cone to will not have a selective advantage.
+that arises in an area already within a cone will not have a selective advantage.
 When these adaptive alleles meet, 
@@
 alleles have spread there yet.} The grey dashed line shows a cone radiating backward in space-time,
-originating $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
-Our point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
+originating at $(x,t)$, with slope $v$; its radius at $t=0$ is $v t$. 
+This shows the location $x$ is not yet adapted at time $t$ because no successful alleles appear within this grey cone.
 }
@@
 \citep{Flint:98,ralphcoop2010}.
-Another particularly impressive case of convergent evolution \mfp{is} the numerous changes throughout the X-linked G6PD gene, 
+Another particularly impressive case of convergent evolution \mfp{is presented by} the numerous changes throughout the X-linked G6PD gene, 
 with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
@@
 Assuming that the underlying causes and strength of this drop in fitness have not changed, 
-we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
-\geq 0.05$),
+we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $s_b \geq 0.05$),
 in order to have resulted in the equilibrium frequency seen today in
@@
 on the order of ten million square kilometers.
-In that area there are at least $15$ common variants \citep[see Figure \ref{fig-G6PD-map}][]{Howes-g6pd-variants}. 
+In that area there are at least $15$ common, \plr{clinically relevant} variants \citep[see Figure \ref{fig-G6PD-map}, from][]{Howes-g6pd-variants}. 
+\plr{(These are type 2 variants that express at ${}<50\%$ enzyme activity, 
+predispose individuals to haemolytic anaemia, 
+and are found in at least 10 localities;
+see \citet{Howes-g6pd-variants} for more details.)}
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
@@
 it will cover a distance $L$ in time $L/v$. \gc{In doing so (in two-dimensions) it will have
-started from a point in space and spread out to cover an geographic
-circular area of size $\pi
-L^2$ and carved out a cylinder in space-time with this circle as a base and
-height $L/v$ (the volume of this cylinder is $\pi L^2 \times (L/v)$)
+started from a point in space and spread out to cover a circular geographic region of area $\pi L^2$.
+The cylinder in space-time with this circle as a base and
+height $L/v$ has volume of this cylinder is $\pi L^2 \times (L/v)$.
 Therefore, the number of other successfully established mutations that
@@
 $\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
-(and $2 \lambda_0 L + 2 \lambda \pi L^2 /v$ in one dimension).
+(and $2 \lambda_0 L + 2 \lambda L^2 /v$ in one dimension).
 % $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
@@
 \caption{ %
-{\bf Characteristic length, in kilometers, as a function of selective advantage},
+{\bf Characteristic length}, in kilometers, as a function of selective advantage, at two population densities,
 holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at \gc{$\mu=10^{-6}$}.  
@@
 extending back to $t=0$ (light grey dashed lines) is empty of
-successful mutations. Since we assume that \mfp{our successful alleles} arise as a Poisson process, 
+successful mutations. Since we assume that \mfp{successful alleles} arise as a Poisson process, 
 in two dimensions
@@
 \end{equation}
-i.e. the combined probability that area of a circle of radius $vt$ surrounding our
+i.e., the combined probability that area of a circle of radius $vt$ surrounding our
 point at $t=0$ was free of successful standing variants, and no
@@
 \caption{ %
-{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage 
-as well as the population density.} We compare these to the corresponding quantities without standing variation and only considering standing variants.
+{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage.}
+We compare these to the corresponding quantities without standing variation and only considering standing variants.
 } \label{G6PD_chartimes}
@@
 \subsection{The contribution of standing variation.}
-We can use our framework to address what proportion of new adaptive variants arise from standing variation.
+We can also ask in our framework to address what proportion of new adaptive variants arise from standing variation.
 We have defined $\lambda_0$ to be the mean density
@@
 the typical patch (informally, the patch around a randomly chosen
-successful mutation; formally, with the Palm measure
+successful mutation; formally, drawn from the Palm measure
 \citep{cox1980point}) occupies area $1/(\lambda_0 + \gamma)$.
@@
 Even for relatively large deleterious costs prior to the environmental switch, 
-standing variants still makes up quite a large proportion of the adaptive alleles, 
+standing variants still make up quite a large proportion of the adaptive alleles, 
 and an even larger proportion of the range 
-(they \mfp{occupy a larger area than new mutations, since they get a
-headstart}).
+(they \mfp{occupy a larger area than new mutations, since they get a head start}).
 
@@
 \caption{ %
-{\bf Mean proportion of patches arising from standing variation.} The
-parameters given are for the G6PD example.
-The left and right panels show the expected proportion of patches that
-arise from standing variation and the expected proportion of
+{\bf Mean proportion of patches arising from standing variation,} 
+\plr{as a function of selective disadvantage, at two population densities.}
+The parameters given are for the G6PD example.
+Both panels show the expected proportion of patches that
+arise from standing variation (dotted lines, equation \eqref{prop_patches_standing}),
+and the expected proportion of
 geographic space that is covered by adaptation from standing variation
-(equation \eqref{prop_patches_standing} and \eqref{prop_space_standing} respectively).
+(solid lines, equation \eqref{prop_space_standing}).
 We set $\mu=10^{-5}$ and $s_b=0.05$ to match our G6PD example.} \label{G6PD_standing_var_proportion}
@@
 assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
-and think of the second class of alleles as arising at one of ten, a hundred, or one thousand base pairs. 
+and think of the second class of alleles as arising at one of ten, one hundred, or one thousand base pairs. 
 We show the expected proportion of space covered by
@@
 reasonable rule of thumb, especially for parameter combinations, 
-like higher population densities and high dispersal distances, 
+such as high population densities and high dispersal distances, 
 that increase the contribution of standing variation (and similarly for high $s_b$).
@@
 
-In the above (and in \citet{ralphcoop2010}), we have assumed that once the mutation appears,
+In the above, we have assumed that once the mutation appears,
 conditional on eventual fixation, it begins to spread spatially at speed $v$ instantly,
@@
 This can be worked out,
-but it is simpler to note that if the converse is true \mfp{(i.e. that
-reach local fixation is slow compared to the spread of the wave)}, 
+but it is simpler to note that if the converse is true 
+\mfp{(i.e., that reaching local fixation is slow compared to the spread of the wave)}, 
 the process is largely unaffected by spatial structure,
@@
 (closely related to the density of independent alleles) 
-depends on both $s_b$ and $s_d$ in a geographical spread population. 
+depends on both $s_b$ and $s_d$ in a geographically spread population. 
 Like the panmictic model, in our model, alleles also act to exclude each other; 
@@
 selection pressure is greater than our characteristic
-length we should expect multiple independent alleles to arise and
+length we expect multiple independent alleles to arise and
 spread in that area. Our results suggest that convergent evolution
@@
 limited across the range.  While considering smaller mutational
-target sizes (e.g. less than the 100bp in our example) or lower population densities would lower the probability of convergence
+target sizes (e.g., less than the 100bp in our example) or lower population densities would lower the probability of convergence
 within a given geographical area, it would also act to considerably lengthen
 the time to adaptation. In such cases adaptation may simply fail to
-occur at all or populations may adapt via some other means (i.e. by using a
+occur at all or populations may adapt via some other means (e.g., by using a
 broader mutational target). 
@@
 before environmental change greatly increase the probability of
-convergence within a species. }
+convergence within a species.}
 While at face value this increase seems unsurprising,
@@
 of adaptation. For example, the absence of European skin pigmentation alleles in
-ancient DNA from European recovered from several thousand years ago
+ancient DNA from Europeans who lived several thousand years ago
 has led to the suggestion that these individuals had dark
@@
 earlier version of the paper. All figures and calculations were
-performed in \citet{R_stats}. The figure color paletes were generated using
-Karthik Ram's Wes Anderson R color palette \url{https://github.com/karthik/wesanderson}.
+performed in R \citep{R_stats}. The figure color paletes were generated using
+Karthik Ram's ``Wes Anderson'' R color palette \url{https://github.com/karthik/wesanderson}.
 

### 2015-04-02 response-amnat-reviews.tex
added response

\documentclass[12pt,letterpaper]{article}
\bibliographystyle{evolution}
\usepackage{epsfig}                 
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{mathabx}
\renewcommand{\baselinestretch}{1.6}
\large
\pagenumbering{arabic}
\usepackage[usenames,dvipsnames]{color}
\usepackage{fullpage}
\usepackage{multirow}
\newcommand{\ye}{\hat{y}}
\newcommand{\xe}{\hat{x}}
\usepackage{color}
\usepackage[normalem]{ulem}  
	\newcommand{\gc}[1]{{ \color{red} #1}}
	\newcommand{\yb}[1]{{ \color{blue} #1}}

	\newcommand{\response}[1]{\emph{ \color{blue} #1}}

\usepackage{lineno}
%\linenumbers 
\begin{document}

Dear Michael,\\
Thanks to Flo, Sam, and you for all of the useful comments, they were very helpful. We have revised the manuscript in light of those comments, and additional comments that we received from others. We have responded point by point to the reviews below (our responses are in blue). In the main text changes made in response to the reviewers are in blue, changes made by us in response to other comments are marked in red.

Best wishes\\
Peter Ralph and Graham Coop\\
3/19/14
\newpage

%Dear Graham and Peter-

%Thanks for your nice paper on the geographic patterns of convergent evolution. I really like the ideas in this paper, and it is going to be a nice match to the special issue.

%I have two reviews that also both like the paper (see below). Also I have read it carefully myself, and I’ll add my comments below. I think the general consensus is that the paper covers very interesting and novel territory, but perhaps it could be easier to read with some revision.

%Please use the comments to make the paper better in the ways that you see best. I think that you will be able to use some of these comments in ways that you will agree are improvements. I look forward to the revision. This already is a nice paper.

%Thanks for contributing to this issue. I think it is going to turn out to be pretty nice.

%All my best,
%Mike

{\bf Mike’s comments:}

I’ll add my comments in order that they appear in the presentation of the paper, rather than trying to rank them by importance.

64 –ish: Be clear about whether you are assuming that this variation is allelic

\response{Good point. We have extended our discussion there. }

68: Especially as this is in an issue about local adaption, you probably want to be very clear there that the alleles have the same fitness everywhere.

\response{We have expanded on this.}

77: “below”? \& 77: “we assume to be equal to one” – re-write please, it isn't clear at first reading what you are assuming to be one.

\response{We have edited this in response to reviewers' comments.}

After 90: “somewhat larger probability of local establishment” -- it is not at all clear to me what local establishment means in the case of a deleterious allele. Can you unpack this please?

\response{We meant establishment following the environmental shift, we have altered the text to reflect that.}

Equation 2: If the mutation selection balance is $\mu/s_d$, then the number of copies of the allele on average per unit area is $\rho mu/ s_d$. The probability that this leads to a successfully establishment is $2 s_b$ per copy approximately, which gives $2 s_b \mu \rho / s_d$ for $\lambda_0$. This seems like an easier derivation to me. (For large $s_d$, this diverges from your result -- which is more correct?)

\response{We have extended the description here to more clearly give the ``classic'' mutation-selection balance interpretation. We think it is important to work through this in the simple spatial case (via branching processes) rather than just move straight from the result people are familiar with deriving under the panmictic model. This allows folks to see that the result is valid. }

\gc{Peter is the difference coming about in part because the deteministic result is an approximation? Or because we are working in continuous time (or another reason?)?}

P3, somewhere. It would be worth explicitly giving the formula for lambda somewhere.

\response{Done.}

Figure 1: The grey cone for (x,t) is too faint to see easily. I didn't notice it until trying to find what the caption was referring to. And in the caption, it would be worth explaining why you are looking at this cone before describing it.

\response{Good points. We have edited the caption and darkened the cone. }

192: You need to be more specific about where these two formulas for the means come from. It is not clear at all.

\response{We have expanded on this section to explain how these calculations are performed.}

210: It may be confusing to refer to standing variation as mutational input. If they already exist they are not mutations anymore, but alleles.

\response{Changed}

225-229 – Cool!

\response{Thanks! We like this point.}

231-232 – I thought the first person here “ourselves” was a little jarring.

\response{Changed.}

246: Can you use something other than nu? It looks too much like the v that you have already used for something else.
\response{Changed to $\gamma$}

Figure 5: The caption was missing words or symbols in my copy. Also remove the last comma.

\response{Fixed.}

250: Perhaps this would be better as “the mean proportion of adapted patches”

\response{Done.}
251: Please give a reference for the Palm measure.

\response{Added more description and a reference.}

Figure 6, and many other places, especially in the discussion. You assume a quite large mutation rate here. This may be appropriate for a loss of function mutation ( is this the case for this example?), but it is very high for many types of mutations per locus. Many of your conclusions would be very different with lower mutation rates. It would be nice to know ( at least qualitatively ) what would happen if the mutation rates were lower. You say that convergent evolution may be quite high (e.g. start of discussion), but this is not true if the mutation rate were an order of magnitude or two lower than you assume here, I think.

\response{We have expanded our discussion at the start of the discussion to make this clearer. The G6PD alleles are not knockouts but do correspond to reduced activity. While a 100bp target may be high for some traits, it perhaps is not when you consider all of the possible genes in which a change could occcur. Also if you move to considering smaller target sizes the time to adaptation goes up considerably, both as there is a longer waiting time for new mutations to arise and standing variants are much rarer. }

284 plus a couple of lines: The sentence fragment beginning with “Or,…” is a sentence fragment.

\response{Modified.}

Discussion, somewhere: Can you say anything about the pattern of differentiation of these convergent selected genes relative to neutral differentiation?

\response{Added a sentence to the ``The confusing signal of geographic convergent evolution'' section pointing out that they will stick out against neutral variation. In Ralph and Coop 2010 we gace a more complete version of this argument. However, in practice it will depend on the true per generation rate of migration among demes, so it's hard to say anything to definite when that is rarely known.}

\newpage

{\bf Reviewer 1 (Flo)}\\

%  Extending their previous work on the topic, Ralph \& Coop investigate the role of standing variation in geographic convergent adaptation, or, in other words, how already existing mutations contribute to convergent adaptation (the independent evolution and coexistence of multiple beneficial alleles) in a geographically spread population subjected to a switch in selection pressures. The authors identify a characteristic length $\chi$, which gives the spatial scale at which different adapted mutants will establish. If the scale of the environment is larger than $\chi$, we can expect geographic convergent evolution to occur or have occurred. Importantly, and particularly relevant for the themed issue, the authors warn that the resulting pattern of geographic convergent adaptation can be mistaken for a pattern local adaptation.

The manuscript is well written, interesting, insightful, and dense (in a good way!). It is also sometimes a bit hard to read. The main issue is that it is rather likely that the average AmNat reader will not have read Ralph \& Coop 2010, or, even if they have, they will not remember some specific details. I had to read Ralph \& Coop 2010 (hereafter RC10) to better understand this manuscript, so I will try to highlight the points that could be clarified, in the hope that this will help make the manuscript more accessible to less mathematically-inclined readers.

== Comments on the model and the analysis (clarifications needed) ==

- The introduction is written as if it were obvious that there is a switch in selection pressures (e.g. line 41), while other scenarios are conceivable (e.g. a species encountering a novel habitat type while expanding its range; or, other types of temporal variation (not just a switch)). Maybe clarify near the beginning of the introduction the type of environmental variation you are considering.

\response{Good Points. We had originally wanted to avoid too much overlap with the previous paper, but agree that makes it hard to read as a stand-alone paper. We have extended our introduction to more fully describe our previous work on the topic, the homogeneous nature of the selection pressure, and adding a simulation figure to illustrate the previous intuition. }

- line 64: it is mainly about vocabulary here: the word "class" is used to refer to the "types", while in section 1.6 "class" is used to refer to the diversity of mutations (i.e., within type variability).

\response{We think that this is okay as is, but are happy to change this if folks disagree.}

- line 65: add "neutral" before "variation" (if appropriate)

\response{added neutral, thanks.}

- somewhere in the model description, it would be good to mention that time is measured in generations (and clarify whether it is continuous/discrete).

\response{Good point, added a setence to ``Rates of origination of standing and new mutations''.}

- line 76-77: this sentence was opaque to me until I read RC10 and the formula $2s/\xi^2$.

\response{In hindsight that is pretty cryptic. We changed this to ``We assume that the variance in offspring number is one, see
  Ralphc \& Coop 2010 for how this assumption can be relaxed.''}

- line 86: the way mutation works is not clear yet. If I understand correctly, a neutral type's offspring is mutated with probability $\mu$, but there are no further mutations among the mutated types (i.e. there cannot be a new cone growing within a cone). So "each new offspring" is the offspring of neutral type individuals, right? It could be also useful to recap somewhere the main assumptions of the analysis, e.g. the decomposition of time scales between local establishment and then spatial spread (which is discussed in 1.7), or why there are not further mutants emerging within mutant cones.

\response{We have now added ``In geographic areas already
 occupied by the mutant type (i.e. in regions it has spread to already), new mutations are selectively excluded
 (are effectively neutral), and so unlikely to establish over the
 short-time scales considered here.''. Hopefully that clarifes that we are not banning subsequent mutations, only assuming that they are not selectively favoured.}

- line 88: It would be great to add a few lines to explain what it is a Poisson process in space and time (the explanation in RC10, p651, is very clear).

\response{Added text to the section ``Rates of origination of standing and new mutations'' to help with this.}

- [unnumbered line that would otherwise be 94] "*Since* the times and locations…" has it been proven? \& [unnumbered line that would otherwise be 96] "Poisson Mapping Theorem" maybe add reference?

\response{We have edited this part for clarity.}

- [unnumbered line that would otherwise be 103] Using the letter $s$ to refer to a time is a bit confusing!!
(same line 334, where $s$ is now a rate)

\response{Good point, changed this around.}

\gc{I think I'm right in saying that $s$ at line 334 is actually suppose to be $s_b$, I've changed this.}

- [unnumbered line that would otherwise be 106] Do you actually need to use $\zeta$? Could not the equation that follows simply be written with $E[(1-p_s)^{Z_T}$]? Also, isn't $\zeta(u,T)$ the generating function of $Z_T$?

\gc{need to respond.}

- equation (2), since this will be used later on, maybe add a line $\lambda_0 = \lambda / (-log(1-s_d))$.

\response{Added.}

- Figure 1: I think that it would really help the reader if the figure were in 3D (space x space x time) instead of 2D (space x time), especially to visualize the "cones" carved out in space-time. Also, the lightening bolts give the false impression that the mutants appear right at $t=0$.

\response{We have kept this figure relatively unchanged, although we agree that it's not perfect. We are not convinced that showing the picture in 3D would help, as other details would have to be omitted. The lightening bolts do show the standing variants at time $t=0$, rather than the original mutations, so we have modifed the caption to indicate this.}

- line 208: higher population densities (adjective missing)

\response{Added higher.}

- equation (9) I do not understand where it comes from! (i.e. how using the proportionality between $\lambda_0$ and $\lambda$ give this result). Do we actually need (9) anyway to obtain expression (10)?

\response{Thanks for spotting this slip, we had accidently merged together two different results and a typo. We have corrected this now.}

- line 251: I do not know what is the "Palm measure", so a reference would be welcome here (or consider removing?)

\response{Added more description and a reference.}

- line 343: I do not understand why there is a factor 4 here (and not 2).

\response{Sorry, we had accidently reverted to $\rho$ being the density of diploids, we've changed these to a factor or 2 instead of 4. Thanks! }

- equation (16) I am not sure about the exponents: I find that the exponent for $\beta$ is $-(1+c)/(d+1)$ and that the exponent for u (outside of exp) is $(c-d)/(d+1)$, but I may be wrong?

\response{In the end we evaluate all of the integrals numerically, without the aid of this trick, so we have now dropped the entire appendix section.}

== Minor remarks ==

- line 58: "the genetics of adaptation may often be geographically local" the meaning of this phrases was not obvious at first.

\response{edited this sentence.}

- line 61: A more natural order seems to be first history, then implications.

\response{done}

- line 80-81: If there are more than 1 chromosome per haploid individual, this is not exactly the same?

\response{We meant that diploids carried twice the number of alleles as haploids. Edited in response.}

- line 91: that's probably personal, but I usually prefer when the terms are grouped depending on where they come from, and so I would find it clearer to write $2 s_b \mu \rho$.

\response{changed.}

[- bottom of page 4, the lines are unnumbered, but you can avoid this bug of the lineno package by leaving an empty space before begin equation]

- line 126: I do not understand which the sides curve, but this is probably not a crucial point!

\response{When there is just standing variation, we simple need to bissect up the area into regions closer to one origin or another, that leads to polygons. However, when the mutations are arising through time we have to consider whether a point lands in the cone of one mutation or another, this leads to curved edges. We've added a reference to Ralph and Coop 2010, which graphically depicted this point more clearly.}

- line 200 "close to the value obtained just from standing variation" obtained by whom/how? \&  line 202: "close to the value from" RC10: how different are the values and why?

\response{Sorry, we meant obtain by us, and by ``close to'' we meant that they approach these limits at these extremes. Edited accordingly.}

- line 216: add "locally" in front of "established", if appropriate.

\response{added.}

- line 219: add "of order" or an equivalent in front of $s_b^{1/2}$, if appropriate.

\response{added.}

- Figure 2: Not sure why "Both" is in black in the legend with the two other line types are in gre/ay.

\response{Changed this over to being grey for all three.}

- line 258: the expression for the "sliver" is actually not used in (11), but will be useful in (13).

\response{Agreed, but we think it is useful to introduce all of the components first.}

- the paragraph lines 304--323 seems to rather belong to the Discussion section.

\response{We agree that they are discussion flavo(u)red but they naturally go here, having to move this into the discussion would complicate the (already long) discussion more. So we think this is better left here.}

- line 336: "size" could refer to population size or the area of a patch, maybe add "population" to clarify?

\response{We meant the size (i.e. count) of the branching process. Added to text.}

\gc{right?}

- line 354: "total amount of area" isn't the environment effectively infinite? Or should one rather understand "sampled area"?

\response{In practice we mean the area where the selection pressure is present.}

- line 419--420: I am not sure I understand here; do you mean that the distant populations could share the same allele? But with the work you have done, including standing variance, shouldn't we expect them to be of independent origin, even if they are neutral?

\response{We think is correct as written. We are referring to a model where a common allele (of single origin) is wide-spread before selection pressures switch. Let us know if we need to clarify further.}

- line 459: distance between the lineages is in time$^{-1}$, but the displacement is in distance?

\response{Clarified this setence to make clear that $ v/\chi$ Morgans refers to how closely linked lineages must be to hitchike (spatially) with a selected allele. }

- line 467: another answer is that it is unlikely that all the alleles will strictly have the same beneficial effect $s_b$!

\response{We have extended the text to reflect this.}

== Typos that will probably be anyway corrected when the article is typeset ==

- wide-spread or widespread (lines 57, 59)

\response{corrected}

- line 225: that (thar)
\response{Fixed.}

- line 495: citet instead of citep

\response{Fixed.}
- the abstract on the title page contains a few typos.

\response{Thanks for all your useful comments Flo!}

\newpage 
{\bf Reviewer 2 (Sam)}

The authors have presented an intriguing approach to the problem of convergent genetic evolution, showing the relative importance of standing variation vs. new mutations and how spatial structure interacts with this. Generally, the manuscript is clearly written and covers a lot of new and interesting ground using very nice analytical models. I very much appreciate the presentation and discussion of the results and the big-picture issues covered at the end on speciation and extinction. As I am not primarily a mathematician, there are certain places where I struggled, and I have identified a few spots that could be clarified to make it easier to understand for a general audience. The derivations appear to be correct and to have made reasonable assumptions, but I did not re-derive them exhaustively.

I have a few comments about the discussion and the general implications of this paper, that might be worth considering. I would like to see more discussion of when this kind of evolution will matter, and if/when we can "safely ignore" it. The examples used in the paper are human-focused, but many species have much smaller total population sizes and ranges, and might never experience this kind of evolution. Is there some way we can get a general prediction about when this might matter?

Consider the following interpretation of Eq. 4: if we want to know whether convergent genetic evolution will occur at all in a given species, then the scale we are interested in is the entire species range, and rho = N. If convergent genetic evolution is unimportant, then the characteristic length has to be larger than the total species range size, and we can find out where this happens by setting chi = 1 and asking what parameters result in $\chi < 1$. This simplifies things quite a bit, as simplifies the $\chi^2$ and $\chi^3$ terms, so that we can rearrange the equation to solve for rho (which is now N):

$$Ncrit == (\sigma * \log[1-s]) / (\sqrt{2}  \mu  \pi * (\sqrt{s} \log[1-s] - \sqrt{2}\sigma) )$$

(see attached mathematica file for a more clear version). If N is less than this value, then the characteristic length would be greater than 1, and convergent genetic evolution would not be expected. I'm sure there are some nicer ways that this could be expressed to show various features of importance when certain parameters are small (for example, when sigma is small, it drops out of the denominator, and then Ncrit is a simple function of sigma in the numerator). The scale of sigma is now expressed as a proportion of the total species range, so sigma can realistically be quite small for some species. Of course, this doesn't help us know how fast the characteristic length decreases once N > Ncrit, but Eq. 4 does that. I'm not sure if I've made an obvious blunder in here somewhere, but perhaps this is a useful approach? In any case, it seems worth acknowledging in the discussion that some species are quite unlikely to have convergent genetic evolution, due to their small range size, total population size, or high dispersal distance. The discussion currently says that convergent genetic evolution may be quite common (and line 466 even suggests it may be rare for gene flow to spread a selected allele across the range of a species). Delineating when this is important is especially interesting because the signal of convergent evolution can resemble local adaptation (e.g., line 400-408). If this approach makes sense, it could be discussed along with the section from lines 225-229, which presents a very important conclusion arising from this work.

\gc{Not sure this helps much. Thoughts Peter? By equating $\rho$ to N is he implicitly measuring $\sigma$ in units of range radius?}

Another general consideration is that the trait needs to be non-genetically-redundant for the kind of genetic convergent evolution discussed here, in that there needs to be few genes that could evolve to give the phenotype of interest (ideally, just one, given the model formulation). In both skin pigmentation and malaria resistance, there is only one (or a few) gene(s) that can be mutated to give the phenotype, but many quantitative traits are much more genetically redundant. In a highly genetically redundant case, a very different but interesting prediction can be made: if the characteristic length is small, then we might expect different genetic solutions to the same evolutionary problem in populations separated by distance > chi (i.e., genetically non-convergent, but phenotypically convergent evolution). This is an interesting corollary of the work that could be explored in the future. Then, the ratio of {mutational target size within a single gene} vs. {mutation target size in all other genes that affect the trait with similar s} becomes important for whether evolution is genetically convergent or not. While redundant mutations in quantitative traits would likely have small s, they would also have a higher total target mutation rate across many genes, so species with relatively smaller total size might have small chi and high probability of evolving different genetic solutions for the same phenotype. The section on lines 418-426 touches on this a bit. This suggests that the model could be generalized by considering three classes of mutation with the same (redundant) value of s: 1) same nucleotide, 2) same gene, 3) all other genes. The characteristic length just gives the likelihood of there being multiple independent evolutionary lineages that arise and fix locally, making a tessellation (I think?), and the probability of the alleles being from class 1, 2, or 3 is then just determined by their relative mutational target sizes/rates. The type of convergence/parallelism is then defined by whether it is at the level of QTN (class 1), QTL (class 2), or phenotypic with a different genetic basis (class 3). If the approach I suggested above is sound, then this type of evolution would be common for most quantitative traits in most species (because the high mutation rate at the whole-trait level would compensate for the smaller total population size).

\response{In general we think that highly polygenic traits may often result in very different sets of alleles/loci being used  in different populations. }

Perhaps I am carrying this too far, but it suggests part of the reason why we might sometimes see so many outliers when we use approaches like Bayenv to look at putatively locally adapted quantitative traits. If different populations are evolving in response to a spatially homogenous temporal change in selection pressure (e.g., climate change) via highly redundant traits, then as long as the characteristic length is small, they would evolve different alleles at different loci in different populations. Because climatic variables are correlated with space, it is natural that they would often match up with the tesselations, resulting in associations between snps and local environment. As long as the environment changes faster than the period required for drift to homogenize these responses and wipe out the structure, this signature would be prevalent, but spurious. With suitably rapid changes in environment, alleles could have quite different values of s without enough time to homogenize them across the range of the species (e.g., as discussed around line 316). Obviously, I like how the paper discusses how this kind of evolution would generate signatures resembling local adaptation! I guess I'm arguing that this problem could be much more general, and extend to non-genetically convergent evolution.

\response{Agreed. A somewhat related argument is being made in the Bierne et al. 2011 we cite in the discussion.}

Some more specific comments:

- It might be helpful to discuss convergence in terms of "identical by descent" versus "identical in state", at least for the case where the same nucleotide is mutating. The phrase "different or shared genetic routes" (line 31) is a bit ambiguous in this regard, because it's not clear whether "genetic routes" is referring to the evolutionary lineage history of the gene or the actual genetic changes being compared. This becomes clear when you read it a few times. There is a similar problem with evolution producing "different phenotypes" (line 25) because this seems contrary to the definition of convergent evolution (evolution produces similar phenotypes, independently).

- I couldn't reproduce the exact values for the lines in figures 3 \& 4. I've attached the approach I used in the same mathematica file as above; the main text says $\mu = 10^{-6}$, while the legend says $\mu=10^{-5}$, which seems to match better, but Eq. 5 still doesn't seem to match that way. I assume I did something wrong there but I wanted to mention it just in case.

\response{Sorry there was a typo in the caption, we are using a mutation rate of $10^{-5}$ to match the 100 different changes seen at G6PD. We think our calculation is correct, the R code is available (function charLength in standing-var-fns.R). It also gives the correct answers in the limiting cases (no standing variation and only standing variation), so we think it is correctly implemented.}

- line 124-126: I don't understand why introducing new mutations results in curvature of the tesellations. Wouldn't it just result in a bunch of smaller-sized cells being intermingled with the larger cells that arose from standing variation, without changing the tesselation shape? I guess I can't intuitively see why a wave front arising from new mutation would behave differently than a wave front arising from standing variation, but maybe I missed the point here?

\response{see response to reviewer 1 above, we now give a reference to a figure in Ralph and Coop 2010 that makes this point clearer.}

- Conte et al. (2012); Proceedings of the Royal Society B) reviewed cases of gene reuse and parallel and convergent evolution, and found a high probability of gene reuse, further suggesting the importance of this kind of evolution.

\response{Added this reference to the introduction, thanks.}

- line 46: patchy geographically -> geographically patchy

\response{Fixed.}
- bracket at the bottom of page 4, "this approximation will…": this statement seemed unsubstantiated. Why the log of family size? Perhaps add more justification or move it to somewhere where it fits better?

\response{The log follows from the fact that it takes $\sim log(k)/s$ generations for a selected mutation with selection coefficient $s$ to have $k$ descendents.}

\gc{MORE NEEDED?}

- line 168: I'm not clear why sd = 0.05, just because sb >= 0.05? Perhaps more description here of why this is the case.

\response{This is based on heterozygote advantage calculations for the X chromosome given the observed frequency and $s_b$. We have now explained that more fully. However, we have not shown the calculations as similar ones were presented in Ruwende et al..}

- line 190-196: the alternation between the one-dimensional and two-dimensional equations here is a little confusing and could be explained more clearly. When you say you omit the explicit formula, it takes some looking back and forth to figure out that the explicit formula for 2D is still there in Eq. 4, it's just not rearranged so that there is a single chi term on the left hand side.

\response{We have worked to unpack this a little more for clarity. Thanks for pointing this out.}

- line 201: could add "and standing variation is unimportant" after "is highly deleterious" to clarify.

\response{Done.}

- line 316: have an -> have a

\response{fixed.}

- Figure 5: legend is missing a word ("benefit"?)

\response{Fixed the caption.}

- line 333-335: If it's "feeling" spatial structure, doesn't that mean that it's interfering with neighbours? Doesn't it grow deterministically until it "feels" structure?

\response{The initial growth is generated by a branching process, where each lineage branch independently at a rate $1+s_b$, so its initial growth is not deterministic.}

\end{document}

### 2015-04-02 response-amnat-reviews.tex
peter edited reviews

@@
 	\newcommand{\gc}[1]{{ \color{red} #1}}
+	\newcommand{\plr}[1]{{ \color{green} #1}}
 	\newcommand{\yb}[1]{{ \color{blue} #1}}
@@
 Peter Ralph and Graham Coop\\
-3/19/14
+3/19/15
 \newpage
@@
 %I have two reviews that also both like the paper (see below).
-Also I have read it carefully myself, and I’ll add my comments below.
+Also I have read it carefully myself, and I'll add my comments below.
 I think the general consensus is that the paper covers very interesting and novel territory, but perhaps it could be easier to read with some revision.
@@
 
-{\bf Mike’s comments:}
+{\bf Mike's comments:}
 
-I’ll add my comments in order that they appear in the presentation of the paper, rather than trying to rank them by importance.
+I'll add my comments in order that they appear in the presentation of the paper, rather than trying to rank them by importance.
 
-64 –ish: Be clear about whether you are assuming that this variation is allelic
+64 --ish: Be clear about whether you are assuming that this variation is allelic
 
@@
 
-77: “below”? \& 77: “we assume to be equal to one” – re-write please, it isn't clear at first reading what you are assuming to be one.
+77: ``below''? \& 77: ``we assume to be equal to one'' -- re-write please, it isn't clear at first reading what you are assuming to be one.
 
@@
 
-After 90: “somewhat larger probability of local establishment” -- it is not at all clear to me what local establishment means in the case of a deleterious allele. Can you unpack this please?
+After 90: ``somewhat larger probability of local establishment'' -- it is not at all clear to me what local establishment means in the case of a deleterious allele. Can you unpack this please?
 
@@
 
-\gc{Peter is the difference coming about in part because the deteministic result is an approximation? Or because we are working in continuous time (or another reason?)?}
+\gc{Peter is the difference coming about in part because the deterministic result is an approximation? Or because we are working in continuous time (or another reason?)?}
+\plr{Continuous time: in a discrete-time model, the mean density of individuals resulting from a rate-1 Poisson rain of branching processes would be $1/s_d$; in continuous time, it would be $1/\log(1/(1-s_d))$.
+So, I wouldn't say that one is more valid that the other.
+Hmph, confusing.}
 
@@
 
-225-229 – Cool!
+225-229 -- Cool!
 
@@
 
-231-232 – I thought the first person here “ourselves” was a little jarring.
+231-232 -- I thought the first person here ``ourselves'' was a little jarring.
 
@@
 
-246: Can you use something other than nu? It looks too much like the v that you have already used for something else.
+246: Can you use something other than nu? It looks too much like the $v$ that you have already used for something else.
 \response{Changed to $\gamma$}
@@
 
-250: Perhaps this would be better as “the mean proportion of adapted patches”
+250: Perhaps this would be better as ``the mean proportion of adapted patches''
 
@@
 The G6PD alleles are not knockouts but do correspond to reduced activity.
-While a 100bp target may be high for some traits, it perhaps is not when you consider all of the possible genes in which a change could occcur.
+While a 100bp target may be high for some traits, it perhaps is not when you consider all of the possible genes in which a change could occur.
 Also if you move to considering smaller target sizes the time to adaptation goes up considerably, both as there is a longer waiting time for new mutations to arise and standing variants are much rarer.
@@
 
-284 plus a couple of lines: The sentence fragment beginning with “Or,…” is a sentence fragment.
+284 plus a couple of lines: The sentence fragment beginning with ``Or, \ldots'' is a sentence fragment.
 
@@
 \response{Added a sentence to the ``The confusing signal of geographic convergent evolution'' section pointing out that they will stick out against neutral variation.
-In Ralph and Coop 2010 we gace a more complete version of this argument.
+In Ralph and Coop 2010 we gave a more complete version of this argument.
 However, in practice it will depend on the true per generation rate of migration among demes, so it's hard to say anything to definite when that is rarely known.}
@@
 
-== Comments on the model and the analysis (clarifications needed) ==
+\textbf{Comments on the model and the analysis (clarifications needed)}
 
@@
 
-\response{We think that this is okay as is, but are happy to change this if folks disagree.}
+\response{Good point, although it is consistent in calling a ``class'' a category of mutations with distinct properties? We've modified section 1.6 a bit to set the usage up like this.}
 
@@
 
-\response{Good point, added a setence to ``Rates of origination of standing and new mutations''.}
+\response{Good point, added a sentence to ``Rates of origination of standing and new mutations''.}
 
@@
 \response{In hindsight that is pretty cryptic. We changed this to ``We assume that the variance in offspring number is one, see
-  Ralphc \& Coop 2010 for how this assumption can be relaxed.''}
+  Ralph \& Coop 2010 for how this assumption can be relaxed.''}
 
@@
  (are effectively neutral), and so unlikely to establish over the
- short-time scales considered here.''. Hopefully that clarifes that we are not banning subsequent mutations, only assuming that they are not selectively favoured.}
+ short-time scales considered here.''. Hopefully that clarifies that we are not banning subsequent mutations, only assuming that they are not selectively favoured.}
 
@@
 
-- [unnumbered line that would otherwise be 94] "*Since* the times and locations…" has it been proven? \& [unnumbered line that would otherwise be 96] "Poisson Mapping Theorem" maybe add reference?
+- [unnumbered line that would otherwise be 94] "*Since* the times and locations \ldots" has it been proven?
+\& [unnumbered line that would otherwise be 96] "Poisson Mapping Theorem" maybe add reference?
 
@@
 
-\gc{I think I'm right in saying that $s$ at line 334 is actually suppose to be $s_b$, I've changed this.}
-
 - [unnumbered line that would otherwise be 106] Do you actually need to use $\zeta$?
@@
 
-\gc{need to respond.}
+\response{Sure, good suggestion.  We've removed $\zeta$.}
 
-- equation (2), since this will be used later on, maybe add a line $\lambda_0 = \lambda / (-log(1-s_d))$.
+- equation (2), since this will be used later on, maybe add a line $\lambda_0 = \lambda / (-\log(1-s_d))$.
 
@@
 We are not convinced that showing the picture in 3D would help, as other details would have to be omitted.
-The lightening bolts do show the standing variants at time $t=0$, rather than the original mutations, so we have modifed the caption to indicate this.}
+The lightening bolts do show the standing variants at time $t=0$, rather than the original mutations, so we have modified the caption to indicate this.}
 
@@
 
-\response{Thanks for spotting this slip, we had accidently merged together two different results and a typo. We have corrected this now.}
+\response{Thanks for spotting this slip, we had accidentally merged together two different results and a typo. We have corrected this now.}
 
@@
 
-\response{Sorry, we had accidently reverted to $\rho$ being the density of diploids, we've changed these to a factor or 2 instead of 4. Thanks! }
+\response{Sorry, we had accidentally reverted to $\rho$ being the density of diploids, we've changed these to a factor or 2 instead of 4. Thanks! }
 
@@
 
-== Minor remarks ==
+\textbf{Minor remarks}
 
@@
 
-\response{When there is just standing variation, we simple need to bissect up the area into regions closer to one origin or another, that leads to polygons.
+\response{When there is just standing variation, we simple need to bisect up the area into regions closer to one origin or another, that leads to polygons.
 However, when the mutations are arising through time we have to consider whether a point lands in the cone of one mutation or another, this leads to curved edges.
-We've added a reference to Ralph and Coop 2010, which graphically depicted this point more clearly.}
+We've added a reference to Ralph and Coop 2010, which graphically depicted this point more clearly, and the teaser ``because the radii of the colliding circles differ'' (as it is the special case of equal radii circles that leads to the special property of Voronoi cells).
+A longer explanation: when two expanding circles meet, the direction their intersections move in is the average of their tangents; for non-equally sized circles (as with different ages), this direction changes as they expand.}
 
@@
 
-\gc{right?}
-
 - line 354: "total amount of area" isn't the environment effectively infinite? Or should one rather understand "sampled area"?
@@
 
-\response{Clarified this setence to make clear that $ v/\chi$ Morgans refers to how closely linked lineages must be to hitchike (spatially) with a selected allele. }
+\response{Clarified this sentence to make clear that $ v/\chi$ Morgans refers to how closely linked lineages must be to hitchhike (spatially) with a selected allele. }
 
@@
 
-== Typos that will probably be anyway corrected when the article is typeset ==
+\textbf{Typos that will probably be anyway corrected when the article is typeset}
 
@@
 \response{Fixed.}
+
 - the abstract on the title page contains a few typos.
@@
 The scale of sigma is now expressed as a proportion of the total species range, so sigma can realistically be quite small for some species.
-Of course, this doesn't help us know how fast the characteristic length decreases once N > Ncrit, but Eq.
+Of course, this doesn't help us know how fast the characteristic length decreases once $N > Ncrit$, but Eq.
 4 does that.
@@
 \gc{Not sure this helps much. Thoughts Peter? By equating $\rho$ to N is he implicitly measuring $\sigma$ in units of range radius?}
+\plr{I like it.  (and, he mentions that $\sigma$ is as a fraction of the range, er, should be range diameter, I think.)
+What this comes down to is $N > \frac{ \sigma }{ \sqrt{2 s_b} \mu \pi ( 1 + \sigma \sqrt{s_b}/\log(1/(1-s_d)))}$,
+which since $\pi \sqrt{2s_b}$ will be something of order 1, is roughly just $N > \sigma/\mu$, or $N \mu > \sigma$.
+This is a bit surprising, but very concrete.  What do you think?
+}
 
@@
 In both skin pigmentation and malaria resistance, there is only one (or a few) gene(s) that can be mutated to give the phenotype, but many quantitative traits are much more genetically redundant.
-In a highly genetically redundant case, a very different but interesting prediction can be made: if the characteristic length is small, then we might expect different genetic solutions to the same evolutionary problem in populations separated by distance > chi (i.e., genetically non-convergent, but phenotypically convergent evolution).
+In a highly genetically redundant case, a very different but interesting prediction can be made: if the characteristic length is small, then we might expect different genetic solutions to the same evolutionary problem in populations separated by distance $> \chi$ i.e., genetically non-convergent, but phenotypically convergent evolution).
 This is an interesting corollary of the work that could be explored in the future.
@@
 \response{In general we think that highly polygenic traits may often result in very different sets of alleles/loci being used  in different populations. }
+\plr{er, should say a bit more here, but unsure how to say something that's not ``yes, we know.'' Isn't there more discussion of this in RC10?}
 
@@
 
+\response{We've actually left this ambiguous on purpose, as the results don't depend on whether the mutations are actually the same,
+at the same genes, or in totally different places (as you note).  We've added clarification of this point in the first paragraph of ``Model description''.}
+
 - I couldn't reproduce the exact values for the lines in figures 3 \& 4.
@@
 It also gives the correct answers in the limiting cases (no standing variation and only standing variation), so we think it is correctly implemented.}
+\plr{Ooops, typo again: shouldn't it be $10^{-6}$?}
 
-- line 124-126: I don't understand why introducing new mutations results in curvature of the tesellations.
+- line 124-126: I don't understand why introducing new mutations results in curvature of the tessellations.
 Wouldn't it just result in a bunch of smaller-sized cells being intermingled with the larger cells that arose from standing variation, without changing the tesselation shape?
@@
 
-\response{see response to reviewer 1 above, we now give a reference to a figure in Ralph and Coop 2010 that makes this point clearer.}
+\response{See response to reviewer 1 above, we now give a reference to a figure in Ralph and Coop 2010 that makes this point clearer.}
 
@@
 
-- line 46: patchy geographically -> geographically patchy
+- line 46: patchy geographically $\rightarrow$ geographically patchy
 
 \response{Fixed.}
-- bracket at the bottom of page 4, "this approximation will…": this statement seemed unsubstantiated.
+
+- bracket at the bottom of page 4, "this approximation will \ldots": this statement seemed unsubstantiated.
 Why the log of family size?
@@
 
-\response{The log follows from the fact that it takes $\sim log(k)/s$ generations for a selected mutation with selection coefficient $s$ to have $k$ descendents.}
+\response{The log follows from the fact that it takes $\sim log(k)/s$ generations for a selected mutation with selection coefficient $s$ to have $k$ descendents; we've added a note explaining this further.}
 
-\gc{MORE NEEDED?}
-
-- line 168: I'm not clear why sd = 0.05, just because sb >= 0.05? Perhaps more description here of why this is the case.
+- line 168: I'm not clear why $s_d = 0.05$, just because $s_b \ge 0.05$? Perhaps more description here of why this is the case.
 
@@
 
-- line 316: have an -> have a
+- line 316: have an $\rightarrow$ have a
 

### 2015-04-02 standing-variation-paper.tex
peter edited reviews

@@
 
-\newcommand{\gc}[1]{{\it\color{blue} #1}}
-\newcommand{\mfp}[1]{{\it\color{red} #1}}
-
-\newcommand{\plr}[1]{{\it\color{green}(#1)}}
+\newcommand{\gc}[1]{{\it\color{blue}#1}}
+\newcommand{\mfp}[1]{{\it\color{red}#1}}
+
+\newcommand{\plr}[1]{{\it\color{green}#1}}
 \usepackage{lineno}  %%%% FOR AMNAT
@@
 if the logarithm of the size of each extant family is small relative to the establishment time,
+\plr{since then new families quickly ``catch up'' to the size of already-extant families,}
 and the spatial distribution of each is small relative to the spread between the families.)
@@
 Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
-so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
+% so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
 the mean number of clusters of standing variants destined to fix locally, per unit area, is
 \begin{align*}
-    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
+    % \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
+    \mu \rho \int_0^\infty \E\left[1- (1-p_s)^{Z_T}\right] dT .
 \end{align*}
 For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
-we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
+\plr{we know that $\E[(1-2s_b)^{Z_T}] \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,}
 resulting in the approximation
@@
 the cells formed by a Voronoi tessellation have straight sides,
-but the introduction of new mutations cause these to curve.
-% \citep[see Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.
+but the introduction of new mutations cause these to curve
+\citep[because the radii of the colliding circles differ; see Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.
 
@@
 convergent adaptation may be common. This is because low levels of
-neutral genetic differentiation can be due to high population
+neutral genetic differentiation between geographic regions can be due to high population
 densities rather than high dispersal distances, and high population
@@
 Here we focus primarily on the second effect. 
-Let's imagine for the moment that all classes of beneficial allele 
-have the same beneficial selection coefficient, \gc{or} at least that beneficial selection coefficients are similar enough that our selective exclusion
-approximation holds over the time scale on which we examine the process.
+Let's imagine for the moment \plr{that there are several classes of beneficial allele,}
+all having the same beneficial selection coefficient 
+(\gc{or} at least that beneficial selection coefficients are similar enough that our selective exclusion
+approximation holds over the time scale on which we examine the process).
 Each class of mutations $j$ has its own mutation rate $\mu_j$ and selective disadvantage $s_{d,j}$ prior to the environmental switch. 

### 2015-04-15 standing-variation-paper.tex
added Ne bit in discussion

@@
 \usepackage{color}
+\usepackage[dvipsnames]{xcolor}
 \usepackage[utf8]{inputenc}  % for special characters in authors' names
@@
 \newcommand{\mfp}[1]{{\it\color{red}#1}}
-
-\newcommand{\plr}[1]{{\it\color{green}#1}}
+\newcommand{\plr}[1]{{\it\color{Fuchsia}#1}}
+
 \usepackage{lineno}  %%%% FOR AMNAT
@@
 divergence among species. 
+\plr{Of course, effective population sizes estimated from genetic diversity
+    are nearly always small enough that fixation of neutral alleles by drift
+    is expected to take less time than the divergence time between species.
+    However, this observation is somewhat circular,
+    and as reviewed by \citet{leffler2012revisiting},
+    we do not understand why the implied ``effective population sizes'' differ so much
+    from census population sizes;
+    selective processes (as described here) are likely part of the explanation \citep{corbettdetig2015natural}.
+}
 \gc{Slight selective differences in the pleiotropic effects (or linked

### 2015-04-20 response-amnat-reviews.tex
Graham's final changes

@@
 Peter Ralph and Graham Coop\\
-3/19/15
+4/7/15
 \newpage
@@
 
-\response{We have extended the description here to more clearly give the ``classic'' mutation-selection balance interpretation.
+\response{This difference arises from taking a continuous time approach.
+We have extended the description here to more clearly give the ``classic'' mutation-selection balance interpretation.
 We think it is important to work through this in the simple spatial case (via branching processes) rather than just move straight from the result people are familiar with deriving under the panmictic model.
-This allows folks to see that the result is valid.
+This allows folks to see that the result is valid in our spatial setting.
 }
 
-\gc{Peter is the difference coming about in part because the deterministic result is an approximation? Or because we are working in continuous time (or another reason?)?}
-\plr{Continuous time: in a discrete-time model, the mean density of individuals resulting from a rate-1 Poisson rain of branching processes would be $1/s_d$; in continuous time, it would be $1/\log(1/(1-s_d))$.
+%\gc{Peter is the difference coming about in part because the deterministic result is an approximation? Or because we are working in continuous time (or another reason?)?}
+%\plr{Continuous time: in a discrete-time model, the mean density of individuals resulting from a rate-1 Poisson rain of branching processes would be $1/s_d$; in continuous time, it would be $1/\log(1/(1-s_d))$.
 So, I wouldn't say that one is more valid that the other.
@@
 
-\gc{Not sure this helps much. Thoughts Peter? By equating $\rho$ to N is he implicitly measuring $\sigma$ in units of range radius?}
-\plr{I like it.  (and, he mentions that $\sigma$ is as a fraction of the range, er, should be range diameter, I think.)
-What this comes down to is $N > \frac{ \sigma }{ \sqrt{2 s_b} \mu \pi ( 1 + \sigma \sqrt{s_b}/\log(1/(1-s_d)))}$,
-which since $\pi \sqrt{2s_b}$ will be something of order 1, is roughly just $N > \sigma/\mu$, or $N \mu > \sigma$.
-This is a bit surprising, but very concrete.  What do you think?
-}
+\response{Thanks, this is a nice way of interpreting the result! We have added a paragraph giving this intepretation in roughly the locaiton in the text suggested.}
+
+%\gc{
+
+%Not sure this helps much. Thoughts Peter? By equating $\rho$ to N is he implicitly measuring $\sigma$ in units of range radius?}
+%\plr{I like it.  (and, he mentions that $\sigma$ is as a fraction of the range, er, should be range diameter, I think.)
+%What this comes down to is $N > \frac{ \sigma }{ \sqrt{2 s_b} \mu \pi ( 1 + \sigma \sqrt{s_b}/\log(1/(1-s_d)))}$,
+%which since $\pi \sqrt{2s_b}$ will be something of order 1, is roughly just $N > \sigma/\mu$, or $N \mu > \sigma$.
+%This is a bit surprising, but very concrete.  What do you think?
+%}
 
@@
 
-\response{In general we think that highly polygenic traits may often result in very different sets of alleles/loci being used  in different populations. }
-\plr{er, should say a bit more here, but unsure how to say something that's not ``yes, we know.'' Isn't there more discussion of this in RC10?}
+\response{We are interested in this area, and agree with the reviewer, but we don't want to expand on it given that the paper is already heavy on discussion.
+In general we think that highly polygenic traits may often result in very different sets of alleles/loci being used in different populations.
+We are currently thinking on expanding on this point to form another paper.
+}
+%\plr{er, should say a bit more here, but unsure how to say something that's not ``yes, we know.'' Isn't there more discussion of this in RC10?}
 

### 2015-04-20 standing-variation-paper.tex
tidied up .bib
Graham's final changes

@@
 \usepackage{fullpage}
-\usepackage{amsmath,amssymb}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage[authoryear]{natbib}
-\usepackage{marginnote}
 \usepackage{graphicx}
@@
 \newcommand{\mfp}[1]{{\it\color{red}#1}}
-\newcommand{\plr}[1]{{\it\color{Fuchsia}#1}}
+\newcommand{\plr}[1]{{\it\color{blue}#1}}
 
@@
 \newcommand{\E}{\mathbb{E}}
-\newcommand{\R}{\mathbb{R}}
 
@@
 %\doublespacing
-
-%\iflatexml  % no margin notes in latexml
-% \newcommand{\marginnote}[1]{}
-%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
-%\fi
 
@@
 
+\gc{Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
+the characteristic length, is not
+particularly intuitive we can use it to ask when we should expect
+multiple adaptative alleles within a large geographic region  where our selective pressure
+is present (thanks to Sam Yeaman for proposing this
+interpretation). Consider the case where this geographic area is a fairly regular
+shape of area $G$ and diameter $\sqrt{G}$. Denote the total effective population size
+over this area by $N = \rho G$, and the standard deviation of dispersal as a
+fraction of the diameter of this area by $\sigma_G  = \sigma/
+\sqrt{G}$. Measuring distance in units of the diameter of $\sqrt{G}$,
+we expect multiple mutations when $\chi<1$ a
+condition which will be met when 
+\begin{equation}
+N > \frac{\sigma_G}{ \mu \pi \sqrt{2s_b} \left(1 -  \sigma_G \sqrt{2s_b} /\log (1-s_d) \right)},
+\end{equation}
+this is found by rearranging eqn. \eqref{eqn:defines_chi} having set
+$\chi=1$. This nicely shows that $\sigma_G/\mu$ is ae primary
+determinant of the critical population density necessary for
+convergent evolution.}
+
 Note that the conflicting roles of $\rho$ and $\sigma$ mean that even in
@@
 %To evaluate this integral, again see Appendix \ref{apx:integrals}.
-%\marginnote{put in this value in terms of Gaussian CDF?}
+%\plr{put in this value in terms of Gaussian CDF?}
 

### 2015-04-20 response-amnat-reviews.tex
Graham's final changes

@@
 
-\response{Sorry there was a typo in the caption, we are using a mutation rate of $10^{-5}$ to match the 100 different changes seen at G6PD.
+\response{Sorry there was a typo in the caption, we are using a mutation rate of $10^{-6}$ to match the 100 different changes seen at G6PD.
 We think our calculation is correct, the R code is available (function charLength in standing-var-fns.R).
 It also gives the correct answers in the limiting cases (no standing variation and only standing variation), so we think it is correctly implemented.}
-\plr{Ooops, typo again: shouldn't it be $10^{-6}$?}
 

### 2015-04-22 standing-variation-paper.tex
final changes

@@
 % G6PD a combination of new and standing
-% interpolates betw panmixia and 2010
+% interpolates   panmixia and 2010
 % pleiotropy pre-shift biases adaptation
@@
 \linenumbers
-\modulolinenumbers[2]
+%\modulolinenumbers[2]
 
@@
 equivalent (both in terms how deleterious they were before, and how
-advantagous they are after, the environmental shift). 
+advantageous they are after, the environmental shift). 
 Note that these mutations do not have to arise at the same locus,
@@
   time at which new
-  mutations arise, and the occurance of new mutations in any region of
+  mutations arise, and the occurrence of new mutations in any region of
   space-time is independent of that in non-overlapping intervals
@@
 this is found by rearranging eqn. \eqref{eqn:defines_chi} having set
-$\chi=1$. This nicely shows that $\sigma_G/\mu$ is ae primary
+$\chi=1$. This nicely shows that $\sigma_G/\mu$ is a primary
 determinant of the critical population density necessary for
-convergent evolution.}
+convergent evolution, when $\sigma_G$ is small. As $\sigma_G
+s_b/s_d$ becomes large, e.g. because the allele is not too deleterious
+before the environmental shift, we get our standing variation only
+case where the dependence of  $\sigma_G$
+drops out leaving the critical population size as $1/(2 \mu \pi
+s_b/s_d)$}
 
@@
 landscape. \gc{Such convergent alleles will be $F_{ST}$-outliers compared to
-  older neutral variation, untill there is sufficient time for
+  older neutral variation, until there is sufficient time for
   migration to smooth out them out across the landscape.}
@@
 divergence among species. 
-\plr{Of course, effective population sizes estimated from genetic diversity
-    are nearly always small enough that fixation of neutral alleles by drift
-    is expected to take less time than the divergence time between species.
-    However, this observation is somewhat circular,
-    and as reviewed by \citet{leffler2012revisiting},
-    we do not understand why the implied ``effective population sizes'' differ so much
-    from census population sizes;
-    selective processes (as described here) are likely part of the explanation \citep{corbettdetig2015natural}.
-}
+%\plr{Of course, effective population sizes estimated from genetic diversity
+ %   are nearly always small enough that fixation of neutral alleles by drift
+  %  is expected to take less time than the divergence time between species.
+   % However, this observation is somewhat circular,
+   % and as reviewed by \citet{leffler2012revisiting},
+   % we do not understand why the implied ``effective population sizes'' differ so much
+   % from census population sizes;
+   % selective processes (as described here) are likely part of the explanation \citep{corbettdetig2015natural}.
+%}
 \gc{Slight selective differences in the pleiotropic effects (or linked
@@
 Karthik Ram's ``Wes Anderson'' R color palette \url{https://github.com/karthik/wesanderson}.
+This work was supported by grants from the
+National Science Foundation under Grant No. 1262645 to PLR and GC and by the
+National Institute of General Medical Sciences of the National
+Institutes of Health under award numbers NIH RO1GM83098 and
+RO1GM107374 to GC.
 

### 2015-06-15 standing-variation-paper.tex
changes

@@
 
-\usepackage{lineno}  %%%% FOR AMNAT
+%\usepackage{lineno}  %%%% FOR AMNAT
 \newcommand{\var}{\mathop{\mbox{Var}}}
@@
 
-\linenumbers
+%\linenumbers
 %\modulolinenumbers[2]
@@
 \section*{Abstract}
-The extent to which populations experiencing shared selective pressures
-adapt through a shared genetic response
-is relevant to many questions in evolutionary biology. 
-In a number of well studied traits and species, 
-it appears that convergent evolution within species is common. 
-In this paper, 
-we explore how standing, genetic variation contributes to
-convergent genetic responses in a geographically spread population,
-extending our previous work on the topic.
-Geographically limited dispersal slows the spread of each selected allele,
-hence allowing other alleles 
--- newly arisen mutants or present as standing variation -- 
-to spread before any one comes to dominate the population. 
-When such alleles meet, their progress is substantially slowed 
--- if the alleles are selectively equivalent, 
-they mix slowly, dividing the species range into a random tessellation,
-which can be well understood by analogy to a Poisson process model of crystallization. 
-In this framework, 
-we derive the geographic scale over which a typical allele is expected to dominate, 
-the time it takes the species to adapt as a whole,
-and the proportion of adaptive alleles that arise from standing variation.
-Finally, we explore how negative pleiotropic effects of alleles before an environment change 
-can bias the subset of alleles that contribute to the species' adaptive response. 
-We apply the results to the many geographically localized G6PD
-deficiency alleles thought to confer resistance to malaria, \mfp{where the large
-mutational target size makes it a likely candidate for adaptation
-from standing variation, despite the selective cost of G6PD
-deficiency alleles in the absence of malaria.}
-We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
-\mfp{consistent with the view that they arose from a combination of} 
-standing variation and new mutations since the advent of malaria.
-Our results suggest that much of adaptation may be geographically local 
-even when selection pressures are homogeneous. 
-\mfp{Therefore, we argue that caution must be exercised when arguing that strongly
-geographically restricted alleles are necessarily the outcome of local adaptation.}
-We close by discussing the implications of these results for 
-ideas of species coherence and the nature of divergence between species.
+
+%SUBMITTED ABSTRACT
+The extent to which populations experiencing shared selective pressures adapt through a shared genetic response is relevant to many questions in evolutionary biology.
+In this paper, we explore how standing, genetic variation contributes to convergent genetic responses in a geographically spread population.
+Geographically limited dispersal slows the spread of each selected allele, hence allowing other alleles to spread before any one comes to dominate the population.
+When selectively equivalent alleles meet, their progress is substantially slowed, dividing the species range into a random tessellation, which can be well understood by analogy to a Poisson process model of crystallization.
+In this framework, we derive the geographic scale over which an allele dominates and the proportion of adaptive alleles that arise from standing variation.
+Finally, we explore how negative pleiotropic effects of alleles can bias the subset of alleles that contribute to the species' adaptive response.
+We apply the results to the malaria resistance G6PD deficiency alleles, where the large mutational target size makes it a likely candidate for adaptation from deleterious standing variation.
+Our results suggest that convergent adaptation may be common.
+Therefore, caution must be exercised when arguing that strongly geographically restricted alleles are the outcome of local adaptation.
+We close by discussing the implications of these results for ideas of species coherence and the nature of divergence between species.
+
+%OLD
+%The extent to which populations experiencing shared selective pressures adapt through a shared genetic response is relevant to many questions in evolutionary biology.
+In a number of well studied traits and species, it appears that convergent evolution within species is common.
+In this paper, we explore how standing, genetic variation contributes to convergent genetic responses in a geographically spread population, extending our previous work on the topic.
+Geographically limited dispersal slows the spread of each selected allele, hence allowing other alleles -- newly arisen mutants or present as standing variation -- to spread before any one comes to dominate the population.
+When such alleles meet, their progress is substantially slowed -- if the alleles are selectively equivalent, they mix slowly, dividing the species range into a random tessellation, which can be well understood by analogy to a Poisson process model of crystallization.
+In this framework, we derive the geographic scale over which a typical allele is expected to dominate, the time it takes the species to adapt as a whole, and the proportion of adaptive alleles that arise from standing variation.
+Finally, we explore how negative pleiotropic effects of alleles before an environment change can bias the subset of alleles that contribute to the species' adaptive response.
+We apply the results to the many geographically localized G6PD deficiency alleles thought to confer resistance to malaria, where the large mutational target size makes it a likely candidate for adaptation from standing variation, despite the selective cost of G6PD deficiency alleles in the absence of malaria.
+We find the numbers and geographic spread of these alleles matches our predictions reasonably well, consistent with the view that they arose from a combination of standing variation and new mutations since the advent of malaria.
+Our results suggest that much of adaptation may be geographically local even when selection pressures are homogeneous.
+Therefore, we argue that caution must be exercised when arguing that strongly geographically restricted alleles are necessarily the outcome of local adaptation.
+We close by discussing the implications of these results for ideas of species coherence and the nature of divergence between species.
 
@@
 adapted to similar environments by means of independent genetic changes. 
-In some cases this convergent evolution \mfp{is the result of quite
-distinct genetic changes, involving very different genes and pathways,} despite shared selection pressures; 
+In some cases this convergent evolution is the result of quite
+distinct genetic changes, involving very different genes and pathways, despite shared selection pressures; 
 in other cases independent adaptations are identical down the same
@@
 whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
-and to \mfp{what degree species should be regarded} as cohesive units. 
+and to what degree species should be regarded as cohesive units. 
 Convergent evolution also affects our ability to detect adaptation from population genomic data,
@@
 For example, 
-\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
+geographically patchy selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
 since alleles are unable to spread through intervening populations
 \citep{RalphCoop:14}. 
-In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
+In \citet{ralphcoop2010} we formulated a simple model of convergent adaptation in 
 a spatially spread population with local dispersal
@@
 A somewhat analogous situation also
-arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
-  \end{center}
-  \caption{
-{\bf Simulation of the geographic spread of new mutations}. Simulation
-on a grid of $101 \times 101$ populations with 100 haploid individuals
-in each, showing spread of a beneficial allele with advantage $s=0.1$.
-Times are in numbers of generations, with probability of mutation per
-generation 0.3. Different colours show the different independent
-origins and spreads of selected alleles. Alleles quick establish and
-start to spread, but because they are selective equivalent once they
-spread into each other they only mix slowly (note the lack of change
-from generation 600 onward). More details of the
-simulation are given in \citet{ralphcoop2010}. 
-}
-  \label{fig:sim-spread}
-\end{figure}
-
-\gc{Under this model we previously derived} the characteristic geographic scale over which 
+arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.
+
+Under this model we previously derived the characteristic geographic scale over which 
 multiple instances of the adaptive allele are expected to arise in parallel, 
-a characteristic length \mfp{expressed in terms the parameters of interest.} 
-\gc{In \citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
-allele (e.g., because the allele was very strongly deleterious before the environmental change)}, 
+a characteristic length expressed in terms the parameters of interest. 
+In \citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
+allele (e.g., because the allele was very strongly deleterious before the environmental change), 
 so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.
@@
 we show that convergent adaptation within a widespread species is likely to be
-common \gc{when ranges, population sizes, or mutational targets are large}, as has already been seen for a number of traits. 
+common when ranges, population sizes, or mutational targets are large, as has already been seen for a number of traits. 
 On this basis we argue 
-that the genetics of adaptation may often be geographically local \gc{even
-when selection pressures are geographically broad}, 
+that the genetics of adaptation may often be geographically local even
+when selection pressures are geographically broad, 
 % and that we should expect to see relatively few wide-spread sweeps even when selection pressures are wide-spread. 
-and that widespread selective sweeps \mfp{should tend to occur only}
+and that widespread selective sweeps should tend to occur only
 when adaptation is highly constrained (e.g.\ by small mutation rate or
 the need for a linked combination of alleles).
-We discuss the \gc{history, and implications,} of this view for the evolutionary
+We discuss the history, and implications, of this view for the evolutionary
 coherence of species and molecular evolution.
@@
 is a large, homogeneous, one-- or two--dimensional region. 
-There are two selective classes -- the \gc{\textit{ancestral}} type,
-and the {\em mutated} type \gc{(which will offer a fitness benefit after the
-environmental switch)}.
+There are two selective classes -- the \textit{ancestral} type,
+and the {\em mutated} type which will offer a fitness benefit after the
+environmental switch).
 We assume that separately arising mutations are distinguishable --
-either as selectively equivalent mutations, or by linked \gc{neutral} variation.
-\gc{We assume that all alleles of the mutated type are selectively
+either as selectively equivalent mutations, or by linked neutral variation.
+We assume that all alleles of the mutated type are selectively
 equivalent (both in terms how deleterious they were before, and how
@@
 as long as carrying at least one of these alleles
-is sufficient to adapt an individual to the new environment.}
+is sufficient to adapt an individual to the new environment.
 
@@
 important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
-homozygote for the mutant alleles. \mfp{Note that this means that we
-  are ignoring the case of totally recessive alleles.}
-\gc{We assume that the variance in offspring number is one, see
-  \cite{ralphcoop2010} for how this assumption can be relaxed.}
+homozygote for the mutant alleles. Note that this means that we
+  are ignoring the case of totally recessive alleles.
+We assume that the variance in offspring number is one, see
+  \cite{ralphcoop2010} for how this assumption can be relaxed.
 %The number of offspring has finite variance, 
@@
 and that the mean squared geographic distance between parent and child
-is $\sigma^2$. \mfp{We assume that the dispersal kernel is
+is $\sigma^2$. We assume that the dispersal kernel is
   not heavy tailed, i.e., falling into the Gaussian domain of attraction, see \citet{ralphcoop2010} for
   discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
-  on rapid spreading due to very long distance dispersal.}
+  on rapid spreading due to very long distance dispersal.
 The species occupies an area with mean density
-$\rho$ of alleles per unit area \gc{(twice the number of diploid
-  individuals per unit area)}.
+$\rho$ of alleles per unit area (twice the number of diploid
+  individuals per unit area).
 
@@
 treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
-as a branching process \gc{in continuous time, measured in generations
-(average parental age at birth)}. 
+as a branching process in continuous time, measured in generations
+(average parental age at birth). 
 After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
@@
 is well--approximated by a Poisson process in space and time.
-\gc{By this we mean that there is a constant rate across space and
+By this we mean that there is a constant rate across space and
   time at which new
@@
   space-time is independent of that in non-overlapping intervals
-  in which the mutated type has not already appeared.}
+  in which the mutated type has not already appeared.
 The rate of this Poisson process,
@@
 appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
-\plr{is the product of the haploid population density, denoted $\rho$,
+is the product of the haploid population density, denoted $\rho$,
 the mutation rate ($\mu$),
-and the probability of fixation $p_s$, or}
+and the probability of fixation $p_s$, or
 approximately 
@@
 \end{equation}
-\gc{In geographic areas already occupied by the mutant type, 
+In geographic areas already occupied by the mutant type, 
  new mutations are effectively neutral, and so unlikely to establish over the
- short-time scales considered here, and are hence excluded.  }
+ short-time scales considered here, and are hence excluded.
 
@@
 The genetic descendants of each new mutation are (with high probability) doomed to extinction,
-but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
-\gc{The} times and locations of new mutations before $t=0$ also \gc{will} be 
+but may persist for a short time (note that we have assumed that $s_d$ is not too small).
+The times and locations of new mutations before $t=0$ also will be 
 well--approximated by a 
 Poisson process with rate $\mu \rho$ % PLR: this is not required, and depends on the units you measure area in: (if $\mu \rho \ll 1$) 
-\gc{since mutations are rare events. Therefore,} 
+since mutations are rare events. Therefore,
 the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is 
@@
 and so, by the Poisson Mapping Theorem, is also a Poisson process.
-\mfp{We define $\lambda_0$ to be the mean density of this process, 
+We define $\lambda_0$ to be the mean density of this process, 
 i.e., the geographic density
 of standing variants that are present and escape loss when the
-environment shifts.}
+environment shifts.
 % (In principle, more than one member of a mutant family may fix,
@@
 but with somewhat larger probability of local establishment,
-\gc{following the environmental shift}.
+following the environmental shift.
 (This approximation will be good 
 if the logarithm of the size of each extant family is small relative to the establishment time,
-\plr{since then new families quickly ``catch up'' to the size of already-extant families,}
+since then new families quickly ``catch up'' to the size of already-extant families,
 and the spatial distribution of each is small relative to the spread between the families.)
-To find $\lambda_0$, consider a mutation that arose \plr{$T$ generations ago}, 
-and let \gc{$Z_T$ be the number of its descendants at time $0$}.
+To find $\lambda_0$, consider a mutation that arose $T$ generations ago, 
+and let $Z_T$ be the number of its descendants at time $0$.
 At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
@@
 For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
-\plr{we know that $\E[(1-2s_b)^{Z_T}] \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,}
+we know that $\E[(1-2s_b)^{Z_T}] \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
 resulting in the approximation
@@
 
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
-  \end{center}
-  \caption{
-{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
-Time runs down the page, and a single spatial dimension runs across the page. 
-The environment switches at $t=0$, 
-following which adaptive alleles that successfully escape loss by drift
-spread locally from both standing variants (lightning bolts) and new
-mutations (stars). \gc{(Note that the standing variants arose at times
-prior to $t=0$.)}
-The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
-and the mean density in (space $\times$ time) of successful new mutations is $\lambda$.
-These successful alleles spread spatially outwards at speed $v$, 
-so the space-time profile of their spread forms a cone. Any mutation
-that arises in an area already within a cone will not have a selective advantage.
-When these adaptive alleles meet, 
-their rapid advance is halted because they no longer have an advantage,
-so their boundaries form straight lines on short time-scales 
-(over longer time-scales, they will mix into each other). 
-\gc{One of the ways we derive various results in the text is by taking a
-space-time point $(x,t)$ and asking the probability that no successful
-alleles have spread there yet.} The grey dashed line shows a cone radiating backward in space-time,
-originating at $(x,t)$, with slope $v$; its radius at $t=0$ is $v t$. 
-This shows the location $x$ is not yet adapted at time $t$ because no successful alleles appear within this grey cone.
-}
-  \label{fig:cartoon}
-\end{figure}
-
 %%%%%%%%% %%%%%%%%%%%
@@
 Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
-and spread through human populations in areas where malaria \mfp{has been} endemic \citep{Kwiatkowski:05}. 
+and spread through human populations in areas where malaria has been endemic \citep{Kwiatkowski:05}. 
 A number of these alleles appear to be examples of convergent adaptation, 
@@
 \citep{Flint:98,ralphcoop2010}.
-Another particularly impressive case of convergent evolution \mfp{is presented by} the numerous changes throughout the X-linked G6PD gene, 
+Another particularly impressive case of convergent evolution is presented by the numerous changes throughout the X-linked G6PD gene, 
 with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
@@
 and the Mahidol allele found Myanmar and Thailand.
-\mfp{ 
 The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
@@
 heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
-  Louicharoen-g6pd, manjurano_2015}. }
+  Louicharoen-g6pd, manjurano_2015}. 
 Haplotype-based analysis of genetic diversity surrounding
@@
 
-Given \mfp{such strong selection the alleles} should have risen
+Given such strong selection the alleles should have risen
 quickly to fixation, so their presence at intermediate frequency,
@@
 on the order of ten million square kilometers.
-In that area there are at least $15$ common, \plr{clinically relevant} variants \citep[see Figure \ref{fig-G6PD-map}, from][]{Howes-g6pd-variants}. 
-\plr{(These are type 2 variants that express at ${}<50\%$ enzyme activity, 
+In that area there are at least $15$ common, clinically relevant variants \citep[see Figure \ref{fig-G6PD-map}, from][]{Howes-g6pd-variants}.
+(These are type 2 variants that express at ${}<50\%$ enzyme activity,
 predispose individuals to haemolytic anaemia, 
 and are found in at least 10 localities;
-see \citet{Howes-g6pd-variants} for more details.)}
+see \citet{Howes-g6pd-variants} for more details.)
 Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
@@
 
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
-\caption{ 
-{\bf Map of G6PD-deficiency allele frequencies across Asia.} 
-The pie chart shows the frequency of G6PD-deficiency alleles. 
-The size of the pie chart indicates the number of G6PD-deficient individuals sampled.
-Countries with endemic malaria are colored yellow. 
-Figure taken from \citet{Howes-g6pd-variants}
-\url{http://www.malariajournal.com/content/12/1/418}. 
-} \label{fig-G6PD-map}
-\end{center}
-\end{figure}
-
 % \begin{figure}[ht]
@@
 If it does not encounter other locally fixed, beneficial mutants,
-it will cover a distance $L$ in time $L/v$. \gc{In doing so (in two-dimensions) it will have
+it will cover a distance $L$ in time $L/v$. In doing so (in two-dimensions) it will have
 started from a point in space and spread out to cover a circular geographic region of area $\pi L^2$.
@@
 Therefore, the number of other successfully established mutations that
-would have appeared} in the circle it has covered up until this time is Poisson with mean
+would have appeared in the circle it has covered up until this time is Poisson with mean
 $\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
@@
 before it is expected that one other successful mutation would have arisen in the area covered so far.
-The explicit formula for $\chi$ in two dimensions \gc{can be found by
-rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome;} here we omit
-it. \gc{In one dimension things are a little prettier} and the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.
+The explicit formula for $\chi$ in two dimensions can be found by
+rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome; here we omit
+it. In one dimension things are a little prettier and the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.
 Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
-\gc{in two-dimensions} we can rewrite eqn. \eqref{eqn:chi_def_1} as
+in two-dimensions we can rewrite eqn. \eqref{eqn:chi_def_1} as
 \begin{equation} \label{eqn:defines_chi}
@@
 From this we see that $\chi$ decreases with $\rho$ and $\mu$.
-Furthermore, for large $\sigma$, the characteristic length \gc{approaches
-the value we would obtain just} from standing variation:
+Furthermore, for large $\sigma$, the characteristic length approaches
+the value we would obtain just from standing variation:
 \begin{equation} \label{eqn:chi_standing}
@@
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
-\gc{then standing variation is} unimportant the characteristic length is \gc{approaches} the value from \citet{ralphcoop2010}:
+then standing variation is unimportant the characteristic length is approaches the value from \citet{ralphcoop2010}:
 \begin{equation} \label{eqn:chi_new}
@@
 resistance at G6PD. 
-These curves match our intuition that \gc{higher} population densities result in
+These curves match our intuition that higher population densities result in
 smaller characteristic lengths (as would higher mutation rates). 
-Allowing standing variation increases \gc{the 
-input of new alleles} and so decreases the characteristic length below that predicted
+Allowing standing variation increases the 
+input of new alleles and so decreases the characteristic length below that predicted
 by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
@@
 ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
-\mfp{to} escape drift and become established. 
-The effect of helping\gc{locally} establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
+to escape drift and become established. 
+The effect of helping locally establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
  (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
@@
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter \mfp{only the
-  initial density of established alleles after the environmental shift}. 
+as in that case the speed of spread does not matter only the
+  initial density of established alleles after the environmental shift. 
 The dependence of the characteristic length on $s_b$ when only new
-mutations contribute is much weaker (\gc{of order} $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
+mutations contribute is much weaker (of order $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
 Overall, the range of characteristic lengths observed are reasonably
@@
 800km, especially for the lower population density, as long as the
-fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was \mfp{high}.
-
-\gc{Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
+fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was high.
+
+Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
 the characteristic length, is not
@@
 drops out leaving the critical population size as $1/(2 \mu \pi
-s_b/s_d)$}
+s_b/s_d)$
 
@@
 evolution may be common even in species with little neutral population structure.
-
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
-\caption{ %
-{\bf Characteristic length, in kilometers, as a function of selective disadvantage}, 
-compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
-and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
-The other parameters are chosen to match those of G6PD,
-\gc{$\mu = 10^{-6}$} and $s_b = 0.05$. As the characteristic length with
-only standing variation is independent of dispersal distance it is
-plotted as a single black dotted line. 
-} \label{Fig-G6PD-charlength}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
-\caption{ %
-{\bf Characteristic length}, in kilometers, as a function of selective advantage, at two population densities,
-holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at \gc{$\mu=10^{-6}$}.  
-We compare this to the corresponding quantity without standing
-variation (equation \eqref{eqn:chi_new}) 
-and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
-} \label{Fig-G6PD-charlength-sb}
-\end{center}
-\end{figure}
 
@@
 It is also straightforward to compute the mean time until adaptation.
-\gc{Imagine a geographic location, and let $\tau \ge 0$ be the time at
-which this location} is first reached by some advantageous mutation.
+Imagine a geographic location, and let $\tau \ge 0$ be the time at
+which this location is first reached by some advantageous mutation.
 Then, as can be seen from the
@@
 extending back to $t=0$ (light grey dashed lines) is empty of
-successful mutations. Since we assume that \mfp{successful alleles} arise as a Poisson process, 
+successful mutations. Since we assume that successful alleles arise as a Poisson process, 
 in two dimensions
@@
 \end{align}
-\gc{For applications we evaluate this integral numerically. }  %For means of evaluating this integral, see Appendix \ref{apx:integrals}.
+For applications we evaluate this integral numerically.   %For means of evaluating this integral, see Appendix \ref{apx:integrals}.
 
@@
 
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
-\caption{ %
-{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage.}
-We compare these to the corresponding quantities without standing variation and only considering standing variants.
-} \label{G6PD_chartimes}
-\end{center}
-\end{figure}
-
 % proportion of haplotypes (and of space) that are from standing variation
@@
 is $\P\{ \tau > t \}$,
-\gc{we can see  $\gamma = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
-$\gamma = \lambda \E[\tau] $.}
+we can see  $\gamma = \int_0^\infty \lambda \P\{\tau>t\} dt$, and hence
+$\gamma = \lambda \E[\tau] $.
 %Using that $\lambda_0 = \lambda / \log(1/(1-s_d))$, this gives us that
@@
 %\end{equation}
-Therefore, the mean proportion of \gc{adapted} patches that come from standing variation
+Therefore, the mean proportion of adapted patches that come from standing variation
 is 
@@
 mutation with probability given by equation \eqref{eqn-no-mut-t}.
-\mfp{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
+Given that it has not been reached by $t$, the probability that it will be reached by time $t+dt$ 
 by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
@@
 \end{equation}
-\mfp{this is the probability a given location is reached first by a
+this is the probability a given location is reached first by a
 standing variant (which follows from competing our two exponential
-waiting times).}
-\gc{For applications we evaluate this integral numerically.}
+waiting times).
+For applications we evaluate this integral numerically.
 %To evaluate this integral, again see Appendix \ref{apx:integrals}.
@@
 and an even larger proportion of the range 
-(they \mfp{occupy a larger area than new mutations, since they get a head start}).
-
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
-\caption{ %
-{\bf Mean proportion of patches arising from standing variation,} 
-\plr{as a function of selective disadvantage, at two population densities.}
-The parameters given are for the G6PD example.
-Both panels show the expected proportion of patches that
-arise from standing variation (dotted lines, equation \eqref{prop_patches_standing}),
-and the expected proportion of
-geographic space that is covered by adaptation from standing variation
-(solid lines, equation \eqref{prop_space_standing}).
-We set $\mu=10^{-5}$ and $s_b=0.05$ to match our G6PD example.} \label{G6PD_standing_var_proportion}
-\end{center}
-\end{figure}
+(they occupy a larger area than new mutations, since they get a head start).
 
@@
 Here we focus primarily on the second effect. 
-Let's imagine for the moment \plr{that there are several classes of beneficial allele,}
+Let's imagine for the moment that there are several classes of beneficial allele,
 all having the same beneficial selection coefficient 
-(\gc{or} at least that beneficial selection coefficients are similar enough that our selective exclusion
+(or at least that beneficial selection coefficients are similar enough that our selective exclusion
 approximation holds over the time scale on which we examine the process).
@@
 the rarer, class $1$ mutations in Figure \ref{fig:pleiotropy_calc}.
-\mfp{As expected intuitively,} the contribution of the rarer mutation decreases as the
+As expected intuitively, the contribution of the rarer mutation decreases as the
 mutational target of the second class becomes larger, and as the
@@
 
-\begin{figure}[ht]
-\begin{center}
-  \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
-\caption{ %
-{\bf Proportion of space covered by rarer but less negatively pleiotropic
-  mutation}. Empty circles give the result for standing variation only,
-  equation \eqref{eqn-prop-space-allele-j-standing-room-only}. Lines give
-  the result allowing both standing and \textit{de novo} mutations (equation \eqref{eqn-prop-space-allele-j}).
-  Here we hold $s_b=s_{d,2}=0.05$ and $\mu_1=1 \times 10^{-8}$.
-  }   \label{fig:pleiotropy_calc}
-\end{center}
-\end{figure}
-
 \subsection{Local establishment and comparison to panmixia}
@@
 We are assuming that shortly after a new mutation appears, 
-it can be approximated by a branching process growing at rate \gc{$s_b$}
+it can be approximated by a branching process growing at rate $s_b$
 until the point that it grows large enough to ``feel'' spatial structure,
@@
 Although we are not aware of good analysis of this transition, 
-the relevant size \gc{of the branching process} when spatial structure becomes important
+the relevant size of the branching process when spatial structure becomes important
 should be something close to $\sigma^2 \rho$ 
@@
 but it is simpler to note that if the converse is true 
-\mfp{(i.e., that reaching local fixation is slow compared to the spread of the wave)}, 
+(i.e., that reaching local fixation is slow compared to the spread of the wave), 
 the process is largely unaffected by spatial structure,
@@
 hence decreasing the time period in which new alleles can arise and hope to be at significant frequency in the population. 
-These two effects \mfp{approximately} cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
+These two effects approximately cancel each other out leading to no strong effect of $s_b$ on the number of independent alleles. 
 Decreasing $s_d$ increases the number of standing variants within a population, 
@@
 spread in that area. Our results suggest that convergent evolution
-among populations may be quite common, \gc{at least when population
+among populations may be quite common, at least when population
 densities and mutational target sizes are not too small, and dispersal is
@@
 broader mutational target).
-}
-
-\gc{Our inclusion of standing variation 
+
+Our inclusion of standing variation 
 before environmental change greatly increase the probability of
-convergence within a species.}
+convergence within a species.
 While at face value this increase seems unsurprising,
@@
 have simply not had time for neutral gene flow to spread them across the
-landscape. \gc{Such convergent alleles will be $F_{ST}$-outliers compared to
+landscape. Such convergent alleles will be $F_{ST}$-outliers compared to
   older neutral variation, until there is sufficient time for
-  migration to smooth out them out across the landscape.}
+  migration to smooth out them out across the landscape.
 This pattern may be very hard to distinguish from local adaptation using population genomic approaches
@@
 migration and drift, 
-potentially leading to a \mfp{within population} signal of a sweep from multiple independent
+potentially leading to a within population signal of a sweep from multiple independent
 mutations if our selected allele occurs at the same locus \citep{softsweepsIII}, 
@@
 initially rare within populations, i.e.\ they are reasonably
-deleterious before the environment switches. \mfp{ In contrast, if they adapt via common variation even distant populations could have
+deleterious before the environment switches. In contrast, if they adapt via common variation even distant populations could have
 a shared basis of adaptation, e.g.\ previously neutral (or nearly neutral) variation, shared among populations. 
@@
 that respond to selection quickest will be somewhat different due to
-drift among populations} \citep{Barton:89}.
+drift among populations \citep{Barton:89}.
 % best positioned to initially respond to positive selection, 
@@
 in which local sweeps occur independently across the range.
-Lineages that are \gc{closely linked to the sweeping allele} ($\sim v/\chi$ Morgans) will be moved
+Lineages that are closely linked to the sweeping allele ($\sim v/\chi$ Morgans) will be moved
 towards the center of the sweep (a displacement $O(\chi)$), 
@@
 %}
-\gc{Slight selective differences in the pleiotropic effects (or linked
+Slight selective differences in the pleiotropic effects (or linked
   background) among convergent alleles could allow one allele to
-  press into areas occupied by other alleles.}
+  press into areas occupied by other alleles.
 Furthermore, repeated bouts of adaptation in particular genomic regions may act 
@@
 Conversely, if the geographic spread of adaptive alleles within ranges is slow 
-\mfp{(e.g.\ if they only offer a local advantage or if they are selectively excluded)} 
+(e.g.\ if they only offer a local advantage or if they are selectively excluded) 
 this may allow Dobzhansky-Muller incompatibilities to arise within species, 
@@
 performed in R \citep{R_stats}. The figure color paletes were generated using
-Karthik Ram's ``Wes Anderson'' R color palette \url{https://github.com/karthik/wesanderson}.
+Karthik Ram's ``Wes Anderson'' R color palette
+\url{https://github.com/karthik/wesanderson}. Scripts to create all of
+the figures for the paper are included as supplementary files.
 This work was supported by grants from the
-National Science Foundation under Grant No. 1262645 to PLR and GC and by the
+National Science Foundation under Grant No. 1262645 to PLR and GC and
+Grant No. 1353380 to GC by the
 National Institute of General Medical Sciences of the National
@@
 
-%\appendix
-
-% %%%%%%%%%%%%
-% \section{Integrals appearing in the text}
-%     \label{apx:integrals}
-
-% In the text at several points appear integrals of the form
-% \begin{align}
-%   \int_0^\infty t^c \exp \left( - \alpha t^d - \beta t^{d+1} \right) dt 
-% \end{align}
-% where $c$ is a positive integer, $d$ is the dimension, and $\alpha$ and $\beta$ are positive real numbers.
-% This could be evaluated through standard numerical methods; below we describe a power series expansion.
-% Changing variables to $u = \beta t^{d+1}$, this becomes
-% \begin{align}
-%     \left( (d+1)^{-1} \beta^{ (1-c)/(d+1) } \right) \int_0^\infty u^{(c+1-d)/(d+1)} \exp\left( - \alpha \beta^{-d/(d+1)} u^{d/(d+1)} - u \right) du ,
-% \end{align}
-% so it suffices to evaluate the function
-% \begin{equation}
-%     G(a,b,x) := \int_0^\infty  t^a \exp\left( -x t^b - t \right) dt ,
-% \end{equation}
-% in the case that $a=(c-d+1)/(d+1)$, $b=d/(d+1)$, and $x$ is a function of the demographic parameters.
-% Since $a$ and $b$ only depend on the dimension and the quantity being computed,
-% we are interested in $G$ as a function of $x$.
-% At least in the case $b=d/(d+1)$ it is possible to express $G$ as a finite sum of gamma functions,
-% but we proceed with a simpler method.
-% Note that $\partial_x G(a,b,x) = -G(a+b,b,x)$,
-% and that at $x=0$, the function $G$ is the gamma function $G(a,b,0) = \Gamma(a+1)$.
-% Therefore, a Taylor series for $G$ would be
-% \[
-%     G(a,b,x) = \sum_{n \ge 0} \frac{(-x)^n}{n!} \Gamma(a+nb+1) .
-% \]
-% It is easy to check using Stirling's formula that $\limsup_{n \to \infty} ( x^n \Gamma(a+nb+1)/n! )^{1/n} = 0$
-% if $b<1$, so the sum converges.
+%%%Figure 1
+\begin{figure}[ht]
+  \begin{center}
+%    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
+  \end{center}
+  \caption{
+{\bf Simulation of the geographic spread of new mutations}. Simulation
+on a grid of $101 \times 101$ populations with 100 haploid individuals
+in each, showing spread of a beneficial allele with advantage $s=0.1$.
+Times are in numbers of generations, with probability of mutation per
+generation 0.3. Different colours show the different independent
+origins and spreads of selected alleles. Alleles quick establish and
+start to spread, but because they are selective equivalent once they
+spread into each other they only mix slowly (note the lack of change
+from generation 600 onward). More details of the
+simulation are given in \citet{ralphcoop2010}. 
+}
+  \label{fig:sim-spread}
+\end{figure}
+
+%%Figure 2
+\begin{figure}[ht]
+  \begin{center}
+  %  \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
+  \end{center}
+  \caption{
+{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
+Time runs down the page, and a single spatial dimension runs across the page. 
+The environment switches at $t=0$, 
+following which adaptive alleles that successfully escape loss by drift
+spread locally from both standing variants (lightning bolts) and new
+mutations (stars). (Note that the standing variants arose at times
+prior to $t=0$.)
+The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
+and the mean density in (space $\times$ time) of successful new mutations is $\lambda$.
+These successful alleles spread spatially outwards at speed $v$, 
+so the space-time profile of their spread forms a cone. Any mutation
+that arises in an area already within a cone will not have a selective advantage.
+When these adaptive alleles meet, 
+their rapid advance is halted because they no longer have an advantage,
+so their boundaries form straight lines on short time-scales 
+(over longer time-scales, they will mix into each other). 
+One of the ways we derive various results in the text is by taking a
+space-time point $(x,t)$ and asking the probability that no successful
+alleles have spread there yet. The grey dashed line shows a cone radiating backward in space-time,
+originating at $(x,t)$, with slope $v$; its radius at $t=0$ is $v t$. 
+This shows the location $x$ is not yet adapted at time $t$ because no successful alleles appear within this grey cone.
+}
+  \label{fig:cartoon}
+\end{figure}
+
+%%%Figure 3
+\begin{figure}[ht]
+\begin{center}
+%  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
+\caption{ 
+{\bf Map of G6PD-deficiency allele frequencies across Asia.} 
+The pie chart shows the frequency of G6PD-deficiency alleles. 
+The size of the pie chart indicates the number of G6PD-deficient individuals sampled.
+Countries with endemic malaria are colored yellow. 
+Figure taken from \citet{Howes-g6pd-variants}
+\url{http://www.malariajournal.com/content/12/1/418}. 
+} \label{fig-G6PD-map}
+\end{center}
+\end{figure}
+
+%%Figure 4
+\begin{figure}[ht]
+\begin{center}
+ % \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
+\caption{ %
+{\bf Characteristic length, in kilometers, as a function of selective disadvantage}, 
+compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
+and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
+The other parameters are chosen to match those of G6PD,
+$\mu = 10^{-6}$ and $s_b = 0.05$. As the characteristic length with
+only standing variation is independent of dispersal distance it is
+plotted as a single black dotted line. 
+} \label{Fig-G6PD-charlength}
+\end{center}
+\end{figure}
+
+%%Figure 5
+\begin{figure}[ht]
+\begin{center}
+  %\includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
+\caption{ %
+{\bf Characteristic length, in kilometers, as a function of selective advantage}, at two population densities,
+holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-6}$.  
+We compare this to the corresponding quantity without standing
+variation (equation \eqref{eqn:chi_new}) 
+and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
+} \label{Fig-G6PD-charlength-sb}
+\end{center}
+\end{figure}
+
+%%%Figure 6
+\begin{figure}[ht]
+\begin{center}
+ % \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
+\caption{ %
+{\bf Mean adaptation times} as a function of selective advantage and disadvantage.
+We compare these to the corresponding quantities without standing variation and only considering standing variants.
+} \label{G6PD_chartimes}
+\end{center}
+\end{figure}
+
+%%Figure 7
+\begin{figure}[ht]
+\begin{center}
+  %\includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
+\caption{ %
+{\bf Mean proportion of patches arising from standing variation,} 
+as a function of selective disadvantage, at two population densities.
+The parameters given are for the G6PD example.
+Both panels show the expected proportion of patches that
+arise from standing variation (dotted lines, equation \eqref{prop_patches_standing}),
+and the expected proportion of
+geographic space that is covered by adaptation from standing variation
+(solid lines, equation \eqref{prop_space_standing}).
+We set $\mu=10^{-5}$ and $s_b=0.05$ to match our G6PD example.} \label{G6PD_standing_var_proportion}
+\end{center}
+\end{figure}
+
+%%Figure 8.
+\begin{figure}[ht]
+\begin{center}
+ % \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
+\caption{ %
+{\bf Proportion of space covered by rarer but less negatively pleiotropic
+  mutation}. Empty circles give the result for standing variation only,
+  equation \eqref{eqn-prop-space-allele-j-standing-room-only}. Lines give
+  the result allowing both standing and \textit{de novo} mutations (equation \eqref{eqn-prop-space-allele-j}).
+  Here we hold $s_b=s_{d,2}=0.05$ and $\mu_1=1 \times 10^{-8}$.
+  }   \label{fig:pleiotropy_calc}
+\end{center}
+\end{figure}
 

### 2015-06-15 g6pd_figures.R

@@ -2,41 +2,9 @@ source("~/Dropbox/postdocs/Peter/standing_parallelism/standing-variation-fns.R")
 #install.packages("wesanderson")
 library("wesanderson")
 
-my.col<-wes.palette(5, "Darjeeling")  #wes.palette(2, "Royal1")
+my.col<-wes_palette(name="Darjeeling",n=5)
 
-
-######  Figure 1
-###################################################
-###################Char length as function of sb
-###################################################
-
- mu=1e-6; rhos=c(0.2,2); sd=0.05; sbvals=seq(0.00001,0.3,length=1600); 
-
-sigmas=c(100,50,10)
-
-layout(t(1:2))
-par(mar=c(4,4.3,2,0.5))
-for(j in 1:2){
-	rho=rhos[j]
-	if(j == 1) plot(sbvals,sbvals,type="n",ylim=c(100,1500),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab=expression(list(chi, km)),cex.lab=1.5,cex.axis=1.4,main=expression(rho==0.2))
-	if(j == 2) plot(sbvals,sbvals,type="n",ylim=c(100,800),lwd=2,col=my.col[1],xlab=expression(s[b]),ylab="",cex.lab=1.5,cex.axis=1.4,main=expression(rho==2))
-	for(i in 1:3){
-	sigma=sigmas[i]; charlength.sb<-sapply(sbvals, function (sb) { charLength(mu,rho,sb,sd=sd,sigma)$value } )
-	 lines(sbvals,charlength.sb,lwd=2,col=my.col[i]); 
-	 
-	 charlength.new.only<-sapply(sbvals, function (sb) {  newCharLength(mu, rho, sb, sd=NA, sigma)$value})
-	 lines(sbvals,charlength.new.only,col=my.col[i],lwd=2,lty=2)  ##new mut. only
-	}
-	 ##contribution of standing var only, i.e. old variation
-	 charlength.standing.only<-sapply(sbvals, function (sb) {  oldCharLength(mu, rho, sb, sd=sd, sigma)$value})
-	 lines(sbvals,charlength.standing.only ,lwd=2,col="black",lty=3)
-	  if(j == 2) legend("topright",col=c("grey","grey","grey",my.col[1:3]),lwd=2,lty=c(2,3,1,rep(NA,3)),pch=c(NA,NA,NA,rep(19,3)),legend= as.expression( c("No Standing var.","No new mut.","Both",
[…]

### 2015-06-15 standing-variation-paper.tex
changed all inline sqrt to power 1/2s for AmNat

@@
 \citet{fisher1937wave} and \citet*{KPP1937}. 
-Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
+Under reasonably general conditions, the speed of advance of this wave is $v = \sigma (2s)^{1/2}$. 
 See \citet{ralphcoop2010} for a more thorough review of these travelling waves.
@@
 see \citet{Howes-g6pd-variants} for more details.)
-Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
+Therefore, the average width of an area occupied by an allele is $(10^7/15)^{1/2} \approx  800$km. 
 The coding region of G6PD is 515 codons long, 
@@
 rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome; here we omit
-it. In one dimension things are a little prettier and the characteristic length is $\chi = ( \sqrt{\lambda_0^2 + 2 \lambda/v} - \lambda_0 )/( 2 \lambda / v )$.
+it. In one dimension things are a little prettier and the
+characteristic length is $\chi = ( \left(\lambda_0^2 + 2 \lambda/v \right) - \lambda_0 )^{1/2}/( 2 \lambda / v )$.
 Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
@@
 The effect of helping locally establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
- (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=\sqrt{s_b}$,
+ (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=s_b^{1/2}$,
 showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
@@
 interpretation). Consider the case where this geographic area is a fairly regular
-shape of area $G$ and diameter $\sqrt{G}$. Denote the total effective population size
+shape of area $G$ and diameter $G^{1/2}$. Denote the total effective population size
 over this area by $N = \rho G$, and the standard deviation of dispersal as a
 fraction of the diameter of this area by $\sigma_G  = \sigma/
-\sqrt{G}$. Measuring distance in units of the diameter of $\sqrt{G}$,
+G^{1/2}$. Measuring distance in units of the diameter of $G^{1/2}$,
 we expect multiple mutations when $\chi<1$ a
@@
 the first allele would have a selective advantage $\delta s =s_{b,1}-s_{b,2}$,
-and so would arise at rate $2\rho \mu_1 \delta s $ and would spread at speed $\sigma \sqrt{\delta s}$. 
+and so would arise at rate $2\rho \mu_1 \delta s $ and would spread at
+speed $\sigma \delta s^{1/2}$. 
 An extension of our Poisson process model could incorporate these effects, 

### 2015-06-15 standing-variation-paper.tex
added switch to turn on figs

@@
 \bibliographystyle{amnat}
+
+% turn on/off figures
+\newif\iffigures
+\figuresfalse
 
@@
   \begin{center}
-%    \includegraphics[width=\textwidth]{sims/Peter_sims.png}
+      \iffigures
+      \includegraphics[width=\textwidth]{sims/Peter_sims}
+      \fi
   \end{center}
@@
   \begin{center}
-  %  \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
+      \iffigures
+        \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
+      \fi
   \end{center}
@@
 \begin{center}
-%  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
+    \fi
 \caption{ 
@@
 \begin{center}
- % \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
+    \fi
 \caption{ %
@@
 \begin{center}
-  %\includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
+    \fi
 \caption{ %
@@
 \begin{center}
- % \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
+    \fi
 \caption{ %
@@
 \begin{center}
-  %\includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
+    \fi
 \caption{ %
@@
 \begin{center}
- % \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
+    \iffigures
+      \includegraphics[width=1.0\textwidth]{pleiotropy_calc}
+    \fi
 \caption{ %

### 2015-06-15 standing-variation-paper.tex
removed ugly green boxes from links
minor fix

@@
 \usepackage{graphicx}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 %\usepackage{latexml} % for \iflatexml
@@
 determinant of the critical population density necessary for
-convergent evolution, when $\sigma_G$ is small. As $\sigma_G
-s_b/s_d$ becomes large, e.g. because the allele is not too deleterious
-before the environmental shift, we get our standing variation only
-case where the dependence of  $\sigma_G$
-drops out leaving the critical population size as $1/(2 \mu \pi
-s_b/s_d)$
+convergent evolution, when $\sigma_G$ is small. 
+As $\sigma_G s_b/s_d$ becomes large, 
+e.g., because the allele is not too deleterious before the environmental shift, 
+this reduces to the situation with only standing variation,
+in which the dependence of  $\sigma_G$
+drops out, leaving the critical population size as $1/(2 \mu \pi s_b/s_d)$.
 

### 2015-08-20 standing-variation-paper.tex
minor grammar fixes from amnat

@@
 The cylinder in space-time with this circle as a base and
-height $L/v$ has volume of this cylinder is $\pi L^2 \times (L/v)$.
+height $L/v$ has a volume of $\pi L^2 \times (L/v)$.
 Therefore, the number of other successfully established mutations that
@@
 On the other hand, if the mutant allele is highly deleterious before $t=0$,
-then standing variation is unimportant the characteristic length is approaches the value from \citet{ralphcoop2010}:
+then standing variation is unimportant and the characteristic length is approaches the value from \citet{ralphcoop2010}:
 \begin{equation} \label{eqn:chi_new}
@@
 This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
-as in that case the speed of spread does not matter only the
-  initial density of established alleles after the environmental shift. 
+as in that case only the initial density of established alleles after the environmental shift matters,
+not the speed of spread.
 The dependence of the characteristic length on $s_b$ when only new
@@
 G^{1/2}$. Measuring distance in units of the diameter of $G^{1/2}$,
-we expect multiple mutations when $\chi<1$ a
-condition which will be met when 
+we expect multiple mutations when $\chi<1$, 
+a condition which will be met when 
 \begin{equation}
